### 2019-09-25 SPregress.tex
Initial checkin

\newcommand{\NWtarget}[2]{#2}
\newcommand{\NWlink}[2]{#2}
\newcommand{\NWtxtMacroDefBy}{Fragment defined by}
\newcommand{\NWtxtMacroRefIn}{Fragment referenced in}
\newcommand{\NWtxtMacroNoRef}{Fragment never referenced}
\newcommand{\NWtxtDefBy}{Defined by}
\newcommand{\NWtxtRefIn}{Referenced in}
\newcommand{\NWtxtNoRef}{Not referenced}
\newcommand{\NWtxtFileDefBy}{File defined by}
\newcommand{\NWtxtIdentsUsed}{Uses:}
\newcommand{\NWtxtIdentsNotUsed}{Never used}
\newcommand{\NWtxtIdentsDefed}{Defines:}
\newcommand{\NWsep}{${\diamond}$}
\newcommand{\NWnotglobal}{(not defined globally)}
\newcommand{\NWuseHyperlinks}{}
\documentclass[reqno]{amsart}
\usepackage[margin=1in]{geometry}
\usepackage[colorlinks=true,linkcolor=blue]{hyperref}
\renewcommand{\NWtarget}[2]{\hypertarget{#1}{#2}}
\renewcommand{\NWlink}[2]{\hyperlink{#1}{#2}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bq}{\mathbf{q}}
\newcommand{\bpi}{\text{\boldmath $\pi$}}
\newcommand{\leqst}{\mathrel{\preceq^{st}}}
\newcommand{\geqst}{\mathrel{\succeq^{st}}}

\title{Semi-parametric relative risk model for correlated binary data}
\author{Aniko Szabo}
\date{\today}

\begin{document}
\maketitle

Based on the exchangeability and marginal compatibility assumptions, we propose the following semi-parametric model to describe  $\lambda_k(z), k=0,\ldots, N$ as a function of covariates $z$:
\begin{equation}\label{E:semipara}
	\lambda_k(z) = {\mu_k} \times {\theta(\beta' z)^k}, k = 0, \dots, N,
\end{equation}
where $\theta(\beta' z)$ is a known function with range $[0,1]$; and $\mu_k$ is a non-parametric baseline vector of joint probabilities.

Since $\lambda_1$ equals the marginal probability of a response $\pi$, model \eqref{E:semipara} can be interpreted as a multiplicative model for it:
\begin{equation*}
  \pi(z) = \mu_1 \times \theta(\beta'z).
\end{equation*}
Here $\mu_1$ is the maximal marginal probability that can be achieved by the model, since $\theta(.) \leq 1$. It can also be interpreted as the probability of response for $z=z_1$ for which $\theta(\beta'z_1) = 1$ (potentially in limit).

We will allow two options: having $\mu_1$ as a fixed pre-defined constant, or estimating it from the data.

\section{Initial setup}
\begin{flushleft} \small\label{scrap1}\raggedright\small
\NWtarget{nuweb1a}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {1a}}$\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@require(CorrBin)@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtFileDefBy\ \NWlink{nuweb1a}{1a}\NWlink{nuweb1b}{b}\NWlink{nuweb2}{, 2}\NWlink{nuweb3b}{, 3b}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\subsection{Exchangeable model}

\begin{equation}\label{E:1to1}
	p_{r,n}= \binom{n}{r} \sum_{j=0}^{n-r} {(-1)^j} \binom{n-r}{j} \lambda_{r+j} = \binom{n}{r} \sum_{s=r}^{n} {(-1)^(s-r)} \binom{n-r}{s-r} \lambda_{s},
\end{equation}

\begin{equation}\label{E:also1to1}
	\lambda_k = \sum_{j=0}^{n-k} \frac{\binom{n-k}{j} {p_{n-j,n}}}{\binom{n}{n-j}} = \sum_{r=k}^{n} \frac{\binom{n-k}{n-r} {p_{r,n}}}{\binom{n}{r}},
\end{equation}

\begin{flushleft} \small\label{scrap2}\raggedright\small
\NWtarget{nuweb1b}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {1b}}$\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@# lambda-to-p weight matrix, rows:r, cols:s@\\
\mbox{}\verb@# (-1)^(s-r)*choose(n,r)*choose(n-r,s-r)@\\
\mbox{}\verb@@\\
\mbox{}\verb@weight_mat <- function(n){@\\
\mbox{}\verb@  s <- r <- 0:n@\\
\mbox{}\verb@  res <- outer(r, s, function(x,y)(-1)^(y-x)*choose(n,x)*choose(n-x,y-x))@\\
\mbox{}\verb@  res@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@\\
\mbox{}\verb@p_from_lambda <- function(lambda.vec, n){@\\
\mbox{}\verb@  H <- weight_mat(n)@\\
\mbox{}\verb@  c(H %*% lambda.vec[1:(n+1)])@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@\\
\mbox{}\verb@# p-to-lambda weight matrix, rows:k, cols:r@\\
\mbox{}\verb@# (choose(n-k, n-r)/choose(n,r)@\\
\mbox{}\verb@@\\
\mbox{}\verb@weight_mat2 <- function(n){@\\
\mbox{}\verb@  k <- r <- 0:n@\\
\mbox{}\verb@  res <- outer(k, r, function(x,y)choose(n-x, n-y)/choose(n,y))@\\
\mbox{}\verb@  res@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@\\
\mbox{}\verb@lambda_from_p <- function(p.vec, n=length(p.vec)-1){@\\
\mbox{}\verb@  H <- weight_mat2(n)@\\
\mbox{}\verb@  c(H %*% p.vec)@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtFileDefBy\ \NWlink{nuweb1a}{1a}\NWlink{nuweb1b}{b}\NWlink{nuweb2}{, 2}\NWlink{nuweb3b}{, 3b}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\section{Defining the model}

Parameter estimation is based on the maximization likelihood estimator with respect to likelihood function of observed data. Under the marginal compatibility assumption, the parameters ${\lambda}$ are independent of cluster sizes. The likelihood function based on the observed data is written as follows:
\begin{equation}\label{D:imcompleteL}
	L = {\prod_{i=1}^{I}} f_i\log p_{r_i,n_i}(z_i),
\end{equation}
where $p(r,n)(z)$ is the probability of observing $r$ responses in a cluster of size $n$ given covariates $z$, that can be calculated from \eqref{E:semipara} using the connection between $\lambda$s and probabilities in \eqref{E:1to1}, and $f_i$ are observation weights.

\begin{flushleft} \small\label{scrap3}\raggedright\small
\NWtarget{nuweb2}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {2}}$\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@#'@{\tt @}\verb@rdname sprr@\\
\mbox{}\verb@#'@{\tt @}\verb@param formula a one-sided formula of the form \code{cbind(r, s) ~ predictors} where \code{r} and \code{s} give the number of responses and non-responses within each cluster, respectively (so the cluster size is \code{r+s}), and \code{predictors} describes the covariates.@\\
\mbox{}\verb@#'@{\tt @}\verb@param data  an optional matrix or data frame containing the variables in the formula \code{formula}. By default the variables are taken from \code{environment(formula).}@\\
\mbox{}\verb@#'@{\tt @}\verb@param subset  an optional vector specifying a subset of observations to be used.@\\
\mbox{}\verb@#'@{\tt @}\verb@param weight  an optional vector specifying observation weights.@\\
\mbox{}\verb@#'@{\tt @}\verb@param link      a link function for the binomial distribution, the inverse of which models the covariate effects.@\\
\mbox{}\verb@#'@{\tt @}\verb@param mu1       an optional value between 0 and 1 giving the maximal predicted marginal probability. If set to NULL (default), the algorithm will try to estimate it from the data.@\\
\mbox{}\verb@#'@{\tt @}\verb@param start     an optional list with elements named \code{beta}, \code{q}, and/or \code{mu1} giving starting values for estimation. If none or only some are specified, starting values are selected automatically.@\\
\mbox{}\verb@#'@{\tt @}\verb@param control a list with parameters controlling the algorithm.@\\
\mbox{}\verb@#'@{\tt @}\verb@return an object of class \code{sprr} with the fitted model.@\\
\mbox{}\verb@#'@{\tt @}\verb@export@\\
\mbox{}\verb@#' @{\tt @}\verb@importFrom stats terms model.matrix@\\
\mbox{}\verb@@\\
\mbox{}\verb@#' Semi-parametric relative risk model@\\
\mbox{}\verb@sprr <- function(formula, data, subset, weights, link="cloglog", mu1=NULL, start=NULL, control=list(eps=0.001, maxit=100), ...){@\\
\mbox{}\verb@    fam <- binomial(link=link)@\\
\mbox{}\verb@    model_fun <- fam$linkinv@\\
\mbox{}\verb@@\\
\mbox{}\verb@    @\hbox{$\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize \NWlink{nuweb3a}{3a}}$\,\rangle$}\verb@@\\
\mbox{}\verb@    @\hbox{$\langle\,${\itshape Fit model}\nobreak\ {\footnotesize \NWlink{nuweb6}{6}}$\,\rangle$}\verb@@\\
\mbox{}\verb@@\\
\mbox{}\verb@    mt <- attr(mf, "terms")@\\
\mbox{}\verb@    res <- list(coefficients = beta_new, q=q_new, niter = iter, loglik=logl,@\\
\mbox{}\verb@                link = link, call = mc, terms = mt,@\\
\mbox{}\verb@                xlevels = .getXlevels(mt, mf),@\\
\mbox{}\verb@                data_object=list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights),@\\
\mbox{}\verb@                model_fun=model_fun)@\\
\mbox{}\verb@    class(res) <- "sprr"@\\
\mbox{}\verb@    res@\\
\mbox{}\verb@@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtFileDefBy\ \NWlink{nuweb1a}{1a}\NWlink{nuweb1b}{b}\NWlink{nuweb2}{, 2}\NWlink{nuweb3b}{, 3b}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\begin{flushleft} \small\label{scrap4}\raggedright\small
\NWtarget{nuweb3a}{} $\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize {3a}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@   if (missing(formula) || (length(formula) != 3L))@\\
\mbox{}\verb@        stop("'formula' missing or incorrect")@\\
\mbox{}\verb@   if (missing(data))@\\
\mbox{}\verb@        data <- environment(formula)@\\
\mbox{}\verb@    mc <- match.call(expand.dots = FALSE)@\\
\mbox{}\verb@    m <- match(c("formula", "data", "subset", "weights"), names(mc), 0L)@\\
\mbox{}\verb@    m <- mc[c(1L,m)]@\\
\mbox{}\verb@    if (is.matrix(eval(m$data, parent.frame())))@\\
\mbox{}\verb@        m$data <- as.data.frame(data)@\\
\mbox{}\verb@    m[[1L]] <- quote(stats::model.frame)@\\
\mbox{}\verb@    mf <- eval(m, parent.frame())@\\
\mbox{}\verb@@\\
\mbox{}\verb@    # extract and check response variable@\\
\mbox{}\verb@    Y <- model.response(mf)@\\
\mbox{}\verb@    if (ncol(Y) != 2) stop("The response variable should be a 2-column matrix")@\\
\mbox{}\verb@@\\
\mbox{}\verb@    # create model matrix@\\
\mbox{}\verb@    mm <- model.matrix(formula, data=mf)@\\
\mbox{}\verb@@\\
\mbox{}\verb@    # extract weights@\\
\mbox{}\verb@    weights <- as.vector(model.weights(mf))@\\
\mbox{}\verb@    if (!is.null(weights) && !is.numeric(weights))@\\
\mbox{}\verb@        stop("'weights' must be a numeric vector")@\\
\mbox{}\verb@    if (!is.null(weights) && any(weights < 0))@\\
\mbox{}\verb@        stop("negative weights not allowed")@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb2}{2}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
Using the model \eqref{E:semipara}, and the 1-to-1 relationship equation \eqref{E:1to1}, for cluster size $N$
\begin{equation}\label{E:Qresp}
\begin{split}
p_{r,N}(z) &= \binom{N}{r} {\sum_{j=0}^{N-r}} {(-1)^j} \binom{N-r}{j} {\mu_{r+j}} {\theta(\beta'z)^{r+j}} \\
      &= \sum_{y=r}^N \binom{y}{r} {\theta(\beta'z)^r} {(1-\theta(\beta'z))}^{y-r} {q_y}.
\end{split}
\end{equation}

Then for other cluster sizes, using marginal compatibility,
\begin{equation*}
p_{r,n}(z) = \sum_{s=0}^{N} h(r,s,n,N)p_{s,N}(z).
\end{equation*}

\subsection{Model predictions and likelihood}
We define internal functions that calculate the model-based predicted values for a set of parameters, and the log-likelihood.

\begin{flushleft} \small\label{scrap5}\raggedright\small
\NWtarget{nuweb3b}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {3b}}$\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@pred_lp <- function(beta, data_object){@\\
\mbox{}\verb@    lp <- data_object$model_matrix %*% beta@\\
\mbox{}\verb@    c(lp)@\\
\mbox{}\verb@}@\\
\mbox{}\verb@pred_theta <- function(beta, data_object, model_fun){@\\
\mbox{}\verb@    lp <- pred_lp(beta, data_object)@\\
\mbox{}\verb@    theta <- model_fun(lp)@\\
\mbox{}\verb@    theta@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@\\
\mbox{}\verb@pred_lambda <- function(beta, q, data_object, model_fun){@\\
\mbox{}\verb@    lp <- data_object$model_matrix %*% beta@\\
\mbox{}\verb@    theta <- model_fun(lp)@\\
\mbox{}\verb@    N <- length(q)-1@\\
\mbox{}\verb@    lambda_N <- lambda_from_p(q)@\\
\mbox{}\verb@    th <- sapply(0:N, function(k)theta^k)@\\
\mbox{}\verb@    lambda <- apply(th, 1, function(t)t * lambda_N)@\\
\mbox{}\verb@    rownames(lambda) <- 0:N@\\
\mbox{}\verb@    t(lambda)@\\
\mbox{}\verb@}@\\
\mbox{}\verb@pred_pvec <- function(beta, q, data_object, model_fun){@\\
\mbox{}\verb@   ll <- pred_lambda(beta, q, data_object, model_fun)@\\
\mbox{}\verb@   cs <- data_object$n@\\
\mbox{}\verb@   pp <- lapply(seq_along(cs), @\\
\mbox{}\verb@                function(i){res <- p_from_lambda(ll[i,], n=cs[i]);@\\
\mbox{}\verb@                            names(res) <- 0:cs[i];@\\
\mbox{}\verb@                            res})@\\
\mbox{}\verb@   pp@\\
\mbox{}\verb@}@\\
\mbox{}\verb@pred_p <- function(beta, q, data_object, model_fun){@\\
\mbox{}\verb@   pvec <- pred_pvec(beta, q, data_object, model_fun)@\\
\mbox{}\verb@   rvec <- data_object$resp[,1]@\\
\mbox{}\verb@   pp <- sapply(seq_along(rvec), function(i)pvec[[i]][rvec[i]+1])@\\
\mbox{}\verb@   pp@\\
\mbox{}\verb@}@\\
\mbox{}\verb@loglik <- function(beta, q, data_object, model_fun){@\\
\mbox{}\verb@    p <- pred_p(beta=beta, q=q, data_object = data_object, model_fun=model_fun)@\\
\mbox{}\verb@    w <- data_object$weights@\\
\mbox{}\verb@    if (is.null(w)) ll <- sum(log(p))@\\
\mbox{}\verb@    else ll <- sum(ifelse(w==0, 0, w*log(p)))@\\
\mbox{}\verb@    ll@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtFileDefBy\ \NWlink{nuweb1a}{1a}\NWlink{nuweb1b}{b}\NWlink{nuweb2}{, 2}\NWlink{nuweb3b}{, 3b}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\subsection{Methods for the \texttt{sprr} class}

First, define a printing method which does not show the saved data and model matrix

\begin{flushleft} \small\label{scrap6}\raggedright\small
\NWtarget{nuweb4}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {4}}$\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@# based on print.lm@\\
\mbox{}\verb@print.sprr <- function(x, digits = max(3L, getOption("digits") - 3L),...){@\\
\mbox{}\verb@  cat("\nA semi-parametric relative risk regression model fit\n")@\\
\mbox{}\verb@  cat("\nCall:\n", paste(deparse(x$call), sep = "\n", collapse = "\n"),@\\
\mbox{}\verb@        "\n\n", sep = "")@\\
\mbox{}\verb@  if (length(coef(x))) {@\\
\mbox{}\verb@    cat("Coefficients:\n")@\\
\mbox{}\verb@    print.default(format(x$coefficients, digits = digits),@\\
\mbox{}\verb@            print.gap = 2, quote = FALSE)@\\
\mbox{}\verb@    }@\\
\mbox{}\verb@    else cat("No coefficients\n")@\\
\mbox{}\verb@@\\
\mbox{}\verb@    if (length(x$q)){@\\
\mbox{}\verb@        cat("Baseline joint probabilities (mu):\n")     @\\
\mbox{}\verb@        N <- length(x$q)-1@\\
\mbox{}\verb@        ll <- c(lambda_from_p(x$q))@\\
\mbox{}\verb@        names(ll) <- 0:N@\\
\mbox{}\verb@        print.default(format(ll, digits = digits),@\\
\mbox{}\verb@            print.gap = 2, quote = FALSE)@\\
\mbox{}\verb@    }@\\
\mbox{}\verb@    else cat("No baseline joint probabilities\n")@\\
\mbox{}\verb@@\\
\mbox{}\verb@   cat("Log-likelihood: ", format(x$loglik, digits=digits), "\n")@\\
\mbox{}\verb@   invisible(x)@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtFileDefBy\ \NWlink{nuweb1a}{1a}\NWlink{nuweb1b}{b}\NWlink{nuweb2}{, 2}\NWlink{nuweb3b}{, 3b}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
The prediction method will predict for a variety of scenarios:
\begin{itemize}
\item Input data set:
  \begin{itemize}
    \item the data used in the fitting;
    \item new data.
  \end{itemize}
\item Results:
  \begin{itemize}
    \item $p_{r,n}(z)$: the probability of observing the given $r$ responses with cluster size $n$, given $z$;
    \item $\{p_{\cdot,n}(z)\}$: the entire vector of response probabilities for cluster size $n$, given $z$ (will be a list due to varying lengths);
    \item $\lambda_{1,n}(z)$: the marginal probability of response for cluster size $n$, given $z$;
    \item $\{\lambda_{\cdot,n}(z)\}$: the entire vector of joint probabilities $\lambda$ for cluster size $n$, given $z$;
    \item $\beta`z$: the linear predictor value $z$;
    \item $\theta(\beta'z)$: the relative risk at predictor value $z$.
  \end{itemize}
\end{itemize}

\begin{flushleft} \small\label{scrap7}\raggedright\small
\NWtarget{nuweb5}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {5}}$\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@predict.sprr <- function(object, newdata=NULL,@\\
\mbox{}\verb@                              type=c("mean", "relrisk", "likelihood", "probvec", "lvec", "lp"),@\\
\mbox{}\verb@                              newn=NULL, ...){@\\
\mbox{}\verb@  type <- match.arg(type)@\\
\mbox{}\verb@  tt <- terms(object)@\\
\mbox{}\verb@  if (!missing(newdata)){    @\\
\mbox{}\verb@    Terms <- delete.response(tt)@\\
\mbox{}\verb@    m <- model.frame(Terms, newdata, xlev = object$xlevels)@\\
\mbox{}\verb@    if (!is.null(cl <- attr(Terms, "dataClasses")))@\\
\mbox{}\verb@            .checkMFClasses(cl, m)@\\
\mbox{}\verb@    mm <- model.matrix(Terms, m)@\\
\mbox{}\verb@    data_object <- list(model_matrix=mm)@\\
\mbox{}\verb@    if (!missing(newn)){@\\
\mbox{}\verb@       if (!(length(newn) == 1L || length(newn) == nrow(newdata)))@\\
\mbox{}\verb@          stop("'newn' should have length 1 or equal to the number of rows of 'newdata'")@\\
\mbox{}\verb@        data_object$n <- rep(newn, length=nrow(newdata))@\\
\mbox{}\verb@       }@\\
\mbox{}\verb@  } else {@\\
\mbox{}\verb@    data_object <- object$data_object@\\
\mbox{}\verb@  }@\\
\mbox{}\verb@  @\\
\mbox{}\verb@@\\
\mbox{}\verb@  if (type=="likelihood"){@\\
\mbox{}\verb@     if (!missing(newdata))@\\
\mbox{}\verb@        stop("Type = 'likelihood' is not available for new data. Consider using type='probvec' to get vector of likelihood values.")@\\
\mbox{}\verb@     pred <- pred_p(beta=object$coefficients, q=object$q,@\\
\mbox{}\verb@                    data_object=data_object,@\\
\mbox{}\verb@                    model_fun=object$model_fun)@\\
\mbox{}\verb@  } else@\\
\mbox{}\verb@  if (type %in% c("mean", "lvec")){@\\
\mbox{}\verb@     ll <- pred_lambda(beta=object$coefficients, q=object$q,@\\
\mbox{}\verb@                    data_object=data_object,@\\
\mbox{}\verb@                    model_fun=object$model_fun)@\\
\mbox{}\verb@     pred <- if (type=="mean") ll[,2] else ll           @\\
\mbox{}\verb@  } else@\\
\mbox{}\verb@  if (type == "lp"){@\\
\mbox{}\verb@     pred <- pred_lp(beta=object$coefficients, data_object=data_object)@\\
\mbox{}\verb@  } else@\\
\mbox{}\verb@  if (type == "relrisk"){@\\
\mbox{}\verb@     pred <- pred_theta(beta=object$coefficients, data_object=data_object,@\\
\mbox{}\verb@                        model_fun=object$model_fun)@\\
\mbox{}\verb@  }@\\
\mbox{}\verb@  if (type == "probvec"){@\\
\mbox{}\verb@    if (!missing(newdata) && missing(newn))@\\
\mbox{}\verb@       stop("For prediction of probability vectors with new data, cluster sizes should be specified in 'newn'")@\\
\mbox{}\verb@    pred <- pred_pvec(beta=object$coefficients, q=object$q, data_object=data_object,@\\
\mbox{}\verb@                        model_fun=object$model_fun) @\\
\mbox{}\verb@  }@\\
\mbox{}\verb@  return(pred)    @\\
\mbox{}\verb@}@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtFileDefBy\ \NWlink{nuweb1a}{1a}\NWlink{nuweb1b}{b}\NWlink{nuweb2}{, 2}\NWlink{nuweb3b}{, 3b}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\section{Fitting the model via an EM-MM algorithm}

We implement an algorithm called the Expectation Maximization Minorize-Maximize, abbreviated as EM MM, to estimate parameters.  Catalina Stefanescu and Bruce W. Turnbull have proved that the marginal compatibility assumption is equivalent to assuming that clusters are from a sample of clusters sharing the same cluster size (the maximum cluster size $N$ is a good choice and is used in our study), but some observations are completely missing at random, abbreviated as MCAR \cite{stefanescu2003likelihood}. The expectation step in the EM MM algorithm can be performed based on the MCAR assumption. The estimation of the non-parametric backbone $\mu_k$ will be done through the corresponding pdf $q_t, t=0,\ldots,N$, because enforcing the complete monotonicity of $\mu$ is much more demanding than enforcing $\sum q_t=1$. Thus the set of parameters to be estimated is $\phi = (q, \beta)$, under the restrictions $\sum_{y=0}^N q_y=1$ and $\frac 1N \sum_{y=0}^{N}yq_y=\mu_1$.

In the EM setup, the missing data for each cluster $i$ comes from its representation as a MCAR sample from a cluster of size $N$ with $s_i$ responses. So the complete data are $\mathcal{D}=\{r_i, n_i, z_i; s_i)\}_{i=1}^I$ and the observed data are $\mathcal{D}=\{r_i, n_i, z_i; s_i)\}_{i=1}^I$. The complete data log-likelihood is
\begin{equation*}
  \log L_c (\phi \mid \mathcal{D}_c) = \sum_{i=1}^I f_i \log p_{s_i}(z_i) = \sum_{i=1}^I \sum_{s=0}^N f_i I(s_i=s) \log \Big[ \sum_{y=s}^N \binom{y}{s} {\theta(\beta'z_i)^s} {(1-\theta(\beta'z_i))}^{y-s} {q_y}\Big].
\end{equation*}
Its expectation given the observed data and a current set of parameter estimates $\phi^{(k)}$ is
\begin{equation*}
  E[\log L_c \mid \phi^{(k)}, \mathcal{D}]  = \sum_{i=1}^I \sum_{s=0}^N f_i P(S_i=s \mid \phi^{(k)}, \mathcal{D}) \log  \Big[\sum_{y=s}^N \binom{y}{s} {\theta(\beta'z_i)^s} {(1-\theta(\beta'z_i))}^{y-s} {q_y}\Big].
\end{equation*}

\begin{flushleft} \small\label{scrap8}\raggedright\small
\NWtarget{nuweb6}{} $\langle\,${\itshape Fit model}\nobreak\ {\footnotesize {6}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@    @\hbox{$\langle\,${\itshape Define internal functions}\nobreak\ {\footnotesize \NWlink{nuweb9a}{9a}}$\,\rangle$}\verb@@\\
\mbox{}\verb@    @\hbox{$\langle\,${\itshape Set initial values}\nobreak\ {\footnotesize \NWlink{nuweb9c}{9c}}$\,\rangle$}\verb@@\\
\mbox{}\verb@@\\
\mbox{}\verb@    @\hbox{$\langle\,${\itshape Setup for E-step}\nobreak\ {\footnotesize \NWlink{nuweb8a}{8a}}$\,\rangle$}\verb@@\\
\mbox{}\verb@    iter <- 0@\\
\mbox{}\verb@    diff <- 100@\\
\mbox{}\verb@    while ((diff > control$eps) & (iter < control$maxit)){@\\
\mbox{}\verb@        iter <- iter + 1@\\
\mbox{}\verb@        beta_old <- beta_new@\\
\mbox{}\verb@        q_old <- q_new@\\
\mbox{}\verb@@\\
\mbox{}\verb@        @\hbox{$\langle\,${\itshape E-step}\nobreak\ {\footnotesize \NWlink{nuweb8b}{8b}}$\,\rangle$}\verb@@\\
\mbox{}\verb@        @\hbox{$\langle\,${\itshape M-step for beta}\nobreak\ {\footnotesize \NWlink{nuweb8c}{8c}}$\,\rangle$}\verb@@\\
\mbox{}\verb@        @\hbox{$\langle\,${\itshape M-step for q}\nobreak\ {\footnotesize \NWlink{nuweb9b}{9b}}$\,\rangle$}\verb@@\\
\mbox{}\verb@@\\
\mbox{}\verb@        diff <- sum(abs(beta_old - beta_new)) + sum(abs(q_old - q_new))@\\
\mbox{}\verb@    }@\\
\mbox{}\verb@    logl <- loglik(beta_new, q_new, list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights), model_fun=model_fun)@\\
\mbox{}\verb@    names(beta_new) <- colnames(mm)@\\
\mbox{}\verb@@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb2}{2}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\subsection{E-step}
Using the Bayes theorem,
\begin{equation*}
  e_{is}^{(k)} = P(s_i=s \mid \phi^{(k)}, \mathcal{D}) \propto h(r_i, s, n_i, N) P(S_i=s \mid \phi^{(k)}, \mathcal{D}),
\end{equation*}
where the coefficient of proportionality is chosen so that these values add up to 1 over $s=0,\ldots, N$, and
\begin{equation*}
   P(S_i=s \mid \phi^{(k)}, \mathcal{D}) = \sum_{y=s}^N \binom{y}{s} \theta(\beta^{(k)'}z_i)^s (1-\theta(\beta^{(k)'}z_i))^{y-s} q^{(k)}_y.
\end{equation*}

\subsection{M-step}

The maximization step uses Minorize-Maximize to update the parameters $\phi$. We apply Jensen's inequality to bound from below the sum within the logarithm by assigning element-wise weights to each $q_{y}$, and transforming the log sum expression to obtain an update for parameters $\phi$.

\begin{multline*}
\log \Big[\sum_{y=s}^N \binom{y}{s} {\theta(\beta'z_i)^s} {(1-\theta(\beta'z_i))}^{y-s} {q_y}\Big] =
\log \Big[\sum_{y=s}^N w_{iys}^{(k)} \frac{\binom{y}{s} {\theta(\beta'z_i)^s} {(1-\theta(\beta'z_i))}^{y-s} {q_y}}{w_{iys}^{(k)}}\Big] \geq \\
\sum_{y=s}^N w_{iys}^{(k)} \log\Big[\frac{\binom{y}{s} {\theta(\beta'z_i)^s} {(1-\theta(\beta'z_i))}^{y-s} {q_y}}{w_{iys}^{(k)}}\Big] = \\
\sum_{y=s}^N w_{iys}^{(k)} \log\Big[\binom{y}{s} {\theta(\beta'z_i)^s} {(1-\theta(\beta'z_i))}^{y-s} {q_y}\Big] -\sum_{y=s}^N w_{iys}^{(k)} \log w_{iys}^{(k)}.
\end{multline*}

Therefore, the lower bound of the expected complete data log-likelihood function is:
\begin{multline}\label{E:LowerBound}
	 E[\log L_c(\phi) \mid \phi^{(k)}, \mathcal{D}] \geq  \\
 \sum_{i=1}^I \sum_{s=0}^N \sum_{y=s}^N f_i e_{is}^{(k)} w_{iys}^{(k)} \log\Big[\binom{y}{s} {\theta(\beta'z_i)^s} {(1-\theta(\beta'z_i))}^{y-s} \Big] +
  \sum_{i=1}^I \sum_{s=0}^N \sum_{y=s}^N f_i e_{is}^{(k)} w_{iys}^{(k)} \log{q_y} -\\
 \sum_{i=1}^I \sum_{s=0}^N \sum_{y=s}^N f_i e_{is}^{(k)} w_{iys}^{(k)}\log w_{iys}^{(k)} ,
\end{multline}
where the last term does not actually depend on the parameters, and can be ignored. Also note that the terms containing $\beta$ and $q$ are separated, and can be maximized individually.

The weights at the $k^{th}$ iteration are selected so that they add up to 1 and in the above equation equality holds for $\phi=\phi^{(k)}$,
\begin{equation}\label{D:weights}
	w_{iys}^{(k)} = \frac{\binom{y}{s} {\theta(\beta^{(k)'}z_i)^s} {(1-\theta(\beta^{(k)'}z_i))}^{y-s} {q^{(k)}_y} }%
                          {\sum_{\gamma=s}^N \binom{\gamma}{s} {\theta(\beta^{(k)'}z_i)^s} {(1-\theta(\beta^{(k)'}z_i))}^{\gamma-s} {q^{(k)}_\gamma}}.
\end{equation}

\subsection{Implementation of E-step}

In \eqref{E:LowerBound} we only need
$$e_{is}^{(k)}w_{iys}^{(k)} = \frac{ f_i h(r_i, s, n_i, N) a_{iys}^{(k)}}{\sum_{t=0}^{N}\sum_{\gamma=t}^{N} h(r_i, t, n_i, N) a_{i\gamma t}^{(k)}},$$
where $a_{iys}^{(k)} = \binom{y}{s} {\theta(\beta^{(k)'}z_i)^s} {(1-\theta(\beta^{(k)'}z_i))}^{y-s} {q^{(k)}_y}$ and the denominator is just a normalizing constant.

\begin{flushleft} \small\label{scrap9}\raggedright\small
\NWtarget{nuweb8a}{} $\langle\,${\itshape Setup for E-step}\nobreak\ {\footnotesize {8a}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@    # replace each cluster with N(N-1)/2 clusters of size N@\\
\mbox{}\verb@    new <- cbind(s=rep(0:N, each=N+1), y=rep(0:N, times=N+1))@\\
\mbox{}\verb@    new <- new[new[,"s"] <= new[,"y"],]@\\
\mbox{}\verb@@\\
\mbox{}\verb@    rep_idx <- rep(1:nrow(Y), each=nrow(new))@\\
\mbox{}\verb@    Y2 <- cbind(1:nrow(Y), Y)[rep_idx,]@\\
\mbox{}\verb@    colnames(Y2) <- c("i", "resp","nonresp")@\\
\mbox{}\verb@    mm2 <- mm[rep_idx,,drop=FALSE]@\\
\mbox{}\verb@@\\
\mbox{}\verb@    rep_idx2 <- rep(1:nrow(new), times=nrow(Y))@\\
\mbox{}\verb@    Ycomb <- cbind(Y2, new[rep_idx2,])@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb6}{6}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\begin{flushleft} \small\label{scrap10}\raggedright\small
\NWtarget{nuweb8b}{} $\langle\,${\itshape E-step}\nobreak\ {\footnotesize {8b}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@    # calculate a_{iys}^k@\\
\mbox{}\verb@    lp <- mm2 %*% beta_old@\\
\mbox{}\verb@    theta <- model_fun(lp)@\\
\mbox{}\verb@    a <- dbinom(x=Ycomb[,"s"], size=Ycomb[,"y"], prob=theta) * q_old[Ycomb[,"y"]+1]@\\
\mbox{}\verb@@\\
\mbox{}\verb@    # calculate e_{is}^k w_{iys}^k@\\
\mbox{}\verb@    ew_num <- dhyper(x=Ycomb[,"resp"], m=Ycomb[,"s"], n=N - Ycomb[,"s"],@\\
\mbox{}\verb@                     k=Ycomb[,"resp"] + Ycomb[,"nonresp"]) * a@\\
\mbox{}\verb@    ew_denom <- tapply(ew_num, list(i=Ycomb[,"i"]), sum, simplify=TRUE)@\\
\mbox{}\verb@    ew <-  ew_num / ew_denom[Ycomb[,"i"]]@\\
\mbox{}\verb@    if (!is.null(weights)) ew <- weights * ew@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb6}{6}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\subsection{Implementation of the M-step}

The first term in \eqref{E:LowerBound} corresponds to a weighted binomial likelihood with $y$ as the cluster size, $s$ as the number of successes, a link function $\theta^{-1}$, and weights $f_i e_{is}^{(k)}w_{iys}^{(k)}$, so $\beta$ can be updated using logistic regression.

\begin{flushleft} \small\label{scrap11}\raggedright\small
\NWtarget{nuweb8c}{} $\langle\,${\itshape M-step for beta}\nobreak\ {\footnotesize {8c}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@    mod <- glm(cbind(Ycomb[,"s"], Ycomb[,"y"]-Ycomb[,"s"]) ~ mm2+0, family=fam, weights=ew)@\\
\mbox{}\verb@    beta_new <- coef(mod)@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb6}{6}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
The second term is a multinomial likelihood for $q$ with the usual restriction  $\sum_{y=0}^N q_y=1$, and an additional restriction for the mean $\sum_{y=0}^{N}yq_y=N \mu_1$. Rewriting the second term of \eqref{E:LowerBound} with $c^{(k)}_y = \sum_{i=1}^I \sum_{s=0}^N f_i e_{is}^{(k)} w_{iys}^{(k)}$ and including the equality constraints via a Lagrangian, we need to maximize
\begin{equation*}
  F(q,\alpha_1,\alpha_2) = \sum_{y=0}^{N}c^{(k)}_y\log q_y - \alpha_1 (\sum_{y=0}^{N}q_y - 1) - \alpha_2 (\sum_{y=0}^{N}yq_y - N\mu_1).
\end{equation*}
Taking partial derivatives, and setting them to 0, we can show that the solution is
\begin{equation}\label{E:q_update}
  q^{(k)}_y = \frac{c^{(k)}_y}{1+r y} \Big/ \Big[ \sum_{x=0}^{N}\frac{c^{(k)}_x}{1+r x}\Big],
\end{equation}
where $r=\hat\alpha_2 / \hat\alpha_2$ is the solution of the equation
\begin{equation}\label{E:r_equation}
   f(r) = \sum_{x=0}^{N}\frac{c^{(k)}_x (x - N\mu_1)}{1+r x} = 0.
\end{equation}
To ensure $q_y\geq 0$ for all $y=0,\ldots,N$, we need $r> -\frac 1N$. We reparameterize it as $r=-\frac 1N+\exp(\rho)$ to enforce the constraint.

\begin{flushleft} \small\label{scrap12}\raggedright\small
\NWtarget{nuweb9a}{} $\langle\,${\itshape Define internal functions}\nobreak\ {\footnotesize {9a}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@mean_constrained_probs <- function(cc, m=NULL){@\\
\mbox{}\verb@    if (is.null(m)) return(cc/sum(cc))          @\\
\mbox{}\verb@    N <- length(cc) - 1@\\
\mbox{}\verb@    r_eq <- Vectorize(function(rho){i <- 0:N; sum(cc * (i - m)/(1+(exp(rho)-1/N)*i))})@\\
\mbox{}\verb@    rho <- uniroot(r_eq, interval=c(-10, 10), extendInt="yes")@\\
\mbox{}\verb@    r <- exp(rho$root) - 1/N@\\
\mbox{}\verb@    q <- cc / (1 + r * (0:N))@\\
\mbox{}\verb@    q <- q/sum(q)@\\
\mbox{}\verb@    q@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb6}{6}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\begin{flushleft} \small\label{scrap13}\raggedright\small
\NWtarget{nuweb9b}{} $\langle\,${\itshape M-step for q}\nobreak\ {\footnotesize {9b}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@    c_vec <- tapply(ew, list(y=Ycomb[,"y"]), sum, simplify=TRUE)@\\
\mbox{}\verb@    if (is.null(mu1))@\\
\mbox{}\verb@       q_new <- mean_constrained_probs(c_vec)@\\
\mbox{}\verb@    else@\\
\mbox{}\verb@       q_new <- mean_constrained_probs(c_vec, N*mu1)@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb6}{6}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\subsection{Initial values}

The initial values for $\beta$ can be selected using linear regression on transformed estimates of the marginal probabilities:
\begin{equation*}
  \theta^{-1}\Big(\frac{\pi(z)}{\mu_1}\Big) = \beta'z
\end{equation*}
%
%For $q$, we need to ensure that it gives a marginal probability of $\mu_1$. We will use a mixture of the discrete uniform distribution on $0,\ldots,N$ to ensure that no probabilities %are zero, and a point mass at either $0$ or $N$, depending on whether $\mu_1$ is below or over 0.5.

For $q$, we will get the marginally compatible estimate for the pooled dataset, ensure a minimal probability of 0.01 at each value, then shift it to have mean $\mu_1$.

\begin{flushleft} \small\label{scrap14}\raggedright\small
\NWtarget{nuweb9c}{} $\langle\,${\itshape Set initial values}\nobreak\ {\footnotesize {9c}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@ N <- max(rowSums(Y))@\\
\mbox{}\verb@@\\
\mbox{}\verb@ if (is.null(start$beta)){   @\\
\mbox{}\verb@    p0 <- (Y[,1] + 0.5)/(Y[,1] + Y[,2]+ 1)@\\
\mbox{}\verb@    if (is.null(mu1) && is.null(start$mu1))@\\
\mbox{}\verb@       lp0 <- fam$linkfun(pmin(1-.Machine$double.eps, p0))@\\
\mbox{}\verb@    else if (!is.null(mu1))@\\
\mbox{}\verb@       lp0 <- fam$linkfun(pmin(1-.Machine$double.eps, p0/mu1))@\\
\mbox{}\verb@    else if (!is.null(start$mu1))@\\
\mbox{}\verb@       lp0 <- fam$linkfun(pmin(1-.Machine$double.eps, p0/start$mu1))@\\
\mbox{}\verb@    if (is.null(weights))       @\\
\mbox{}\verb@        lm0 <- lm.fit(x=mm, y=lp0)@\\
\mbox{}\verb@    else         @\\
\mbox{}\verb@        lm0 <- lm.wfit(x=mm, y=lp0, w=weights)@\\
\mbox{}\verb@    beta_new <- coef(lm0)@\\
\mbox{}\verb@ } else {@\\
\mbox{}\verb@    beta_new <- start$beta@\\
\mbox{}\verb@ }@\\
\mbox{}\verb@ @\\
\mbox{}\verb@ if (is.null(start$q)){@\\
\mbox{}\verb@    pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y)), trt="Trt",@\\
\mbox{}\verb@                     clustersize="ClusterSize", nresp="NResp")@\\
\mbox{}\verb@    est <- mc.est(pooled)@\\
\mbox{}\verb@    q0 <- est$Prob[est$ClusterSize == N]@\\
\mbox{}\verb@    if (is.null(mu1) && is.null(start$mu1))@\\
\mbox{}\verb@        q_new <- mean_constrained_probs(pmax(q0, 0.01))@\\
\mbox{}\verb@     else if (!is.null(mu1))@\\
\mbox{}\verb@        q_new <- mean_constrained_probs(pmax(q0, 0.01), N*mu1)@\\
\mbox{}\verb@     else if (!is.null(start$mu1))@\\
\mbox{}\verb@        q_new <- mean_constrained_probs(pmax(q0, 0.01), N*start$mu1)@\\
\mbox{}\verb@ } else {@\\
\mbox{}\verb@   q_new <- start$q@\\
\mbox{}\verb@   if (!is.null(start$mu1))@\\
\mbox{}\verb@      q_new <- mean_constrained_probs(q_new, N*start$mu1)@\\
\mbox{}\verb@  }@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb6}{6}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\end{document}

### 2019-09-25 R/SPreg.R

@@ -1,0 +1,309 @@
+
+require(CorrBin)
+
+# lambda-to-p weight matrix, rows:r, cols:s
+# (-1)^(s-r)*choose(n,r)*choose(n-r,s-r)
+
+weight_mat <- function(n){
+  s <- r <- 0:n
+  res <- outer(r, s, function(x,y)(-1)^(y-x)*choose(n,x)*choose(n-x,y-x))
+  res
+}
+
+p_from_lambda <- function(lambda.vec, n){
+  H <- weight_mat(n)
+  c(H %*% lambda.vec[1:(n+1)])
+}
+
+# p-to-lambda weight matrix, rows:k, cols:r
+# (choose(n-k, n-r)/choose(n,r)
+
+weight_mat2 <- function(n){
+  k <- r <- 0:n
+  res <- outer(k, r, function(x,y)choose(n-x, n-y)/choose(n,y))
+  res
+}
+
+lambda_from_p <- function(p.vec, n=length(p.vec)-1){
+  H <- weight_mat2(n)
+  c(H %*% p.vec)
+}
+
+#'@rdname sprr
+#'@param formula a one-sided formula of the form \code{cbind(r, s) ~ predictors} where \code{r} and \code{s} give the number of responses and non-responses within each cluster, respectively (so the cluster size is \code{r+s}), and \code{predictors} describes the covariates.
+#'@param data  an optional matrix or data frame containing the variables in the formula \code{formula}. By default the variables are taken from \code{environment(formula).}
+#'@param subset an optional vector specifying a subset of observations to be used.
+#'@param weight an optional vector specifying observation weights.
+#'@param link     a link function for the binomial distribution, the inverse of which models the covariate effects.
+#'@param mu1      an optional value between 0 and 1 giving the maximal predicted marginal probability. If set to NULL (default), the algorithm will try to estimate it from the data.
+#'@param start    an optional list with elements named \code{beta}, \code{q}, and/or \code{mu1} giving starting values for estimation. If none or only some are specified, starting values are selected automatically.
+#'@param control        a list with parameters controlling the algorithm.
+#'@return an object of class \code{sprr} with the fitted model.
+#'@export
+#' @importFrom stats terms model.matrix
+
+#' Semi-parametric relative risk model
[…]

### 2019-09-25 SPreg_develop.R

@@ -1,0 +1,49 @@
+library(devtools)
+source('z:/RForge/Nuweb.R')
+source('/home/aszabo/RForge/Nuweb_linux.R')
+
+run_nuweb(file="SPregress.w", path=getwd())
+
+source('R/SPreg.R', echo=FALSE)
+
+data(shelltox)
+a <- sprr(cbind(NResp, ClusterSize-NResp) ~ Trt, data=shelltox, weights=shelltox$Freq,
+                control=list(eps=0.01,maxit=100))
+
+nd <- data.frame(Trt = unique(shelltox$Trt))
+p <- predict(a, newdata=nd)
+
+ab <- sprr(cbind(NResp, ClusterSize-NResp) ~ Trt, data=shelltox, weights=shelltox$Freq,
+           control=list(eps=0.01,maxit=100), start=list(beta=c(0,-1,1,1)))
+
+aa <- sprr(cbind(NResp, ClusterSize-NResp) ~ Trt, data=shelltox, 
+          control=list(eps=0.01,maxit=100))
+a0 <- sprr(cbind(NResp, ClusterSize-NResp) ~ 1, 
+                 data=shelltox, weights=shelltox$Freq,
+                 control=list(eps=0.01,maxit=100))
+a1 <- sprr(cbind(NResp, ClusterSize-NResp) ~ as.numeric(Trt), 
+                 data=shelltox, weights=shelltox$Freq,
+                 control=list(eps=0.01,maxit=100))
+
+data(egde)
+b <- sprr(cbind(NResp, ClusterSize-NResp) ~ Trt+log(ClusterSize),
+                data=egde, weights=egde$Freq,
+                control=list(eps=0.01,maxit=100))
+
+
+
+#########
+
+
+# create native routine registration file
+tools::package_native_routine_registration_skeleton("z://RForge/corrbin/pkg/CorrBin",
+                                                    con="z://RForge/corrbin/pkg/CorrBin/src/init.c")
+
+document(cb)
+run_examples(cb)  # or dev_example("ran.CMData")
+load_all(cb)
+
+build(cb)
+check(cb,  check_dir = "z:/RForge", check_version = TRUE, cran = TRUE)
+
+

### 2019-09-25 SPregress.tex
Experiment with fitting starting values

@@
 \begin{flushleft} \small\label{scrap1}\raggedright\small
-\NWtarget{nuweb1a}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {1a}}$\equiv$
+\NWtarget{nuweb?}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {?}}$\equiv$
 \vspace{-1ex}
@@
 \begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
-\item \NWtxtFileDefBy\ \NWlink{nuweb1a}{1a}\NWlink{nuweb1b}{b}\NWlink{nuweb2}{, 2}\NWlink{nuweb3b}{, 3b}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}.
+\item \NWtxtFileDefBy\ \NWlink{nuweb?}{?}\NWlink{nuweb?}{, ?}\NWlink{nuweb?}{, ?}\NWlink{nuweb?}{, ?}\NWlink{nuweb?}{, ?}\NWlink{nuweb?}{, ?}.
 
@@
 \begin{flushleft} \small\label{scrap2}\raggedright\small
-\NWtarget{nuweb1b}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {1b}}$\equiv$
+\NWtarget{nuweb?}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {?}}$\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap3}\raggedright\small
-\NWtarget{nuweb2}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {2}}$\equiv$
+\NWtarget{nuweb?}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {?}}$\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize \NWlink{nuweb3a}{3a}}$\,\rangle$}\verb@@\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape Fit model}\nobreak\ {\footnotesize \NWlink{nuweb6}{6}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize \NWlink{nuweb?}{?}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape Fit model}\nobreak\ {\footnotesize \NWlink{nuweb?}{?}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@@\\
@@
 \begin{flushleft} \small\label{scrap4}\raggedright\small
-\NWtarget{nuweb3a}{} $\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize {3a}}$\,\rangle\equiv$
+\NWtarget{nuweb?}{} $\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize {?}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
-\item \NWtxtMacroRefIn\ \NWlink{nuweb2}{2}.
+\item \NWtxtMacroRefIn\ \NWlink{nuweb?}{?}.
 
@@
 \begin{flushleft} \small\label{scrap5}\raggedright\small
-\NWtarget{nuweb3b}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {3b}}$\equiv$
+\NWtarget{nuweb?}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {?}}$\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap6}\raggedright\small
-\NWtarget{nuweb4}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {4}}$\equiv$
+\NWtarget{nuweb?}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {?}}$\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap7}\raggedright\small
-\NWtarget{nuweb5}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {5}}$\equiv$
+\NWtarget{nuweb?}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {?}}$\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap8}\raggedright\small
-\NWtarget{nuweb6}{} $\langle\,${\itshape Fit model}\nobreak\ {\footnotesize {6}}$\,\rangle\equiv$
+\NWtarget{nuweb?}{} $\langle\,${\itshape Fit model}\nobreak\ {\footnotesize {?}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape Define internal functions}\nobreak\ {\footnotesize \NWlink{nuweb9a}{9a}}$\,\rangle$}\verb@@\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape Set initial values}\nobreak\ {\footnotesize \NWlink{nuweb9c}{9c}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape Define internal functions}\nobreak\ {\footnotesize \NWlink{nuweb?}{?}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape Set initial values}\nobreak\ {\footnotesize \NWlink{nuweb?}{?}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@@\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape Setup for E-step}\nobreak\ {\footnotesize \NWlink{nuweb8a}{8a}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape Setup for E-step}\nobreak\ {\footnotesize \NWlink{nuweb?}{?}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@    iter <- 0@\\
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@        @\hbox{$\langle\,${\itshape E-step}\nobreak\ {\footnotesize \NWlink{nuweb8b}{8b}}$\,\rangle$}\verb@@\\
-\mbox{}\verb@        @\hbox{$\langle\,${\itshape M-step for beta}\nobreak\ {\footnotesize \NWlink{nuweb8c}{8c}}$\,\rangle$}\verb@@\\
-\mbox{}\verb@        @\hbox{$\langle\,${\itshape M-step for q}\nobreak\ {\footnotesize \NWlink{nuweb9b}{9b}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@        @\hbox{$\langle\,${\itshape E-step}\nobreak\ {\footnotesize \NWlink{nuweb?}{?}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@        @\hbox{$\langle\,${\itshape M-step for beta}\nobreak\ {\footnotesize \NWlink{nuweb?}{?}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@        @\hbox{$\langle\,${\itshape M-step for q}\nobreak\ {\footnotesize \NWlink{nuweb?}{?}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@@\\
@@
 \begin{flushleft} \small\label{scrap9}\raggedright\small
-\NWtarget{nuweb8a}{} $\langle\,${\itshape Setup for E-step}\nobreak\ {\footnotesize {8a}}$\,\rangle\equiv$
+\NWtarget{nuweb?}{} $\langle\,${\itshape Setup for E-step}\nobreak\ {\footnotesize {?}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
-\item \NWtxtMacroRefIn\ \NWlink{nuweb6}{6}.
+\item \NWtxtMacroRefIn\ \NWlink{nuweb?}{?}.
 
@@
 \begin{flushleft} \small\label{scrap10}\raggedright\small
-\NWtarget{nuweb8b}{} $\langle\,${\itshape E-step}\nobreak\ {\footnotesize {8b}}$\,\rangle\equiv$
+\NWtarget{nuweb?}{} $\langle\,${\itshape E-step}\nobreak\ {\footnotesize {?}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap11}\raggedright\small
-\NWtarget{nuweb8c}{} $\langle\,${\itshape M-step for beta}\nobreak\ {\footnotesize {8c}}$\,\rangle\equiv$
+\NWtarget{nuweb?}{} $\langle\,${\itshape M-step for beta}\nobreak\ {\footnotesize {?}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap12}\raggedright\small
-\NWtarget{nuweb9a}{} $\langle\,${\itshape Define internal functions}\nobreak\ {\footnotesize {9a}}$\,\rangle\equiv$
+\NWtarget{nuweb?}{} $\langle\,${\itshape Define internal functions}\nobreak\ {\footnotesize {?}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap13}\raggedright\small
-\NWtarget{nuweb9b}{} $\langle\,${\itshape M-step for q}\nobreak\ {\footnotesize {9b}}$\,\rangle\equiv$
+\NWtarget{nuweb?}{} $\langle\,${\itshape M-step for q}\nobreak\ {\footnotesize {?}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap14}\raggedright\small
-\NWtarget{nuweb9c}{} $\langle\,${\itshape Set initial values}\nobreak\ {\footnotesize {9c}}$\,\rangle\equiv$
+\NWtarget{nuweb?}{} $\langle\,${\itshape Set initial values}\nobreak\ {\footnotesize {?}}$\,\rangle\equiv$
 \vspace{-1ex}

### 2019-09-25 SPreg_develop.R

@@ -1,6 +1,6 @@
 library(devtools)
 source('z:/RForge/Nuweb.R')
-source('/home/aszabo/RForge/Nuweb_linux.R')
+#source('/home/aszabo/RForge/Nuweb_linux.R')
 
 run_nuweb(file="SPregress.w", path=getwd())
 
@@ -8,10 +8,18 @@ source('R/SPreg.R', echo=FALSE)
 
 data(shelltox)
 a <- sprr(cbind(NResp, ClusterSize-NResp) ~ Trt, data=shelltox, weights=shelltox$Freq,
-                control=list(eps=0.01,maxit=100))
+                control=list(eps=0.001, maxit=1000), 
+          start = list(mu1=0.4))
 
 nd <- data.frame(Trt = unique(shelltox$Trt))
-p <- predict(a, newdata=nd)
+nd$p <- predict(a, newdata=nd)
+nd
+
+a2 <- sprr(cbind(NResp, ClusterSize-NResp) ~ Trt + log(ClusterSize), 
+           data=shelltox, weights=shelltox$Freq,
+           control=list(eps=0.001,maxit=100),
+           start = list(beta=c(coef(a),0), q=a$q))
+
 
 ab <- sprr(cbind(NResp, ClusterSize-NResp) ~ Trt, data=shelltox, weights=shelltox$Freq,
            control=list(eps=0.01,maxit=100), start=list(beta=c(0,-1,1,1)))

### 2019-09-30 SPsim.Rmd
Simulate one dataset

---
title: "Simulation study"
output: html_notebook
---

```{r setup}
source('R/SPreg.R', echo=FALSE)

```

Goal:

 * simulate data with a q-power backbone distribution that follows the relative-risk regression model
 * fit the data using our semi-parametric regression model, beta-binomial regression (misspecified correlation structure), and GEE (marginal only)

$$
	\lambda_k(z) = {\mu_k} \times {\theta(\beta' z)^k}, k = 0, \dots, K,
$$

## Simulating data
### Cluster-size distributions

 1. All small (3); 
 2. All large (10); 
 3. varying (0.5*(1+binom(9, 0.5)) + 0.5*Unif(2,10))

```{r Clustersizes}
cs.small <- function(n) rep(3, n)
cs.large <- function(n) rep(10, n)
cs.varied <- function(n) {
  p <- rbinom(n, size=1, prob=0.5)
  x <- ifelse(p, 1+rbinom(n, size=9, prob=0.5), sample(2:10, size=n, replace=TRUE))
  x
}
```

### Simulate one dataset

The inputs will be

 * `n` - number of clusters to simulate data
 * `clustersize.gen` - function to generate the clustersizes
 * `backbone.gen` - function to generate the PDF of the backbone
 * `covariate.gen` - function to generate the covariates
 * `beta` - coefficients of the linear predictor
 * `link` - the link-function for the semi-parametric model; the inverse link is the $\theta$ function in the equation

```{r SimSetup}
sim.sp <- function(n, clustersize.gen, backbone.gen, 
                   covariate.gen, beta, link="cloglog"){

   model_fun <- binomial(link=link)$linkinv

   cs <- clustersize.gen(n)
   K <- max(cs)
   cov.mat <- covariate.gen(n)

   lp <- cov.mat %*% beta
   theta <- model_fun(lp)

   q0 <- backbone.gen(K)
   lambda0 <- lambda_from_p(q0)

   th <- sapply(0:R, function(k)theta^k)
   lambda <- apply(th, 1, function(t)t * lambda0)
   rownames(lambda) <- 0:R

   q <- lapply(1:n,
               function(idx)p_from_lambda(lambda[,idx], n=cs[idx]))
   nresp <- lapply(q, function(qq)sample(seq_along(qq)-1, 
                                         size=1, prob=qq))
   res <- data.frame(ID=1:n,
                     ClusterSize = cs,
                     NResp = nresp)
   res <- cbind(res, cov.mat)
   res
}

```

### 2019-09-30 SPsim.Rmd
Simulate one example dataset

@@
 source('R/SPreg.R', echo=FALSE)
-
+library(ggplot2)
+theme_set(theme_classic())
 ```
@@
 
+### Covariate-generating functions
+
+Returns model matrix with randomly generated covariates
+
+```{r Covgen}
+cov.trt <- function(n, ntrts=4){
+  obs <- rep(LETTERS[1:ntrts], length = n)
+  dat <- data.frame(Trt = obs)
+  dat
+}
+
+cov.bin_norm <- function(n, p=0.5){
+  bin <- rbinom(n, size=1, prob=p)
+  norm <- rnorm(n)
+  data.frame(X1 = bin, X2 = norm)
+}
+```
+
 ### Simulate one dataset
@@
 sim.sp <- function(n, clustersize.gen, backbone.gen, 
-                   covariate.gen, beta, link="cloglog"){
+                   covariate.gen, formula, beta, link="log"){
 
@@
    K <- max(cs)
-   cov.mat <- covariate.gen(n)
+   covs <- covariate.gen(n)
+   cov.mat <- model.matrix(formula, data=covs)
 
@@
 
-   th <- sapply(0:R, function(k)theta^k)
+   th <- sapply(0:K, function(k)theta^k)
    lambda <- apply(th, 1, function(t)t * lambda0)
-   rownames(lambda) <- 0:R
+   rownames(lambda) <- 0:K
 
@@
                function(idx)p_from_lambda(lambda[,idx], n=cs[idx]))
-   nresp <- lapply(q, function(qq)sample(seq_along(qq)-1, 
+   nresp <- sapply(q, function(qq)sample(seq_along(qq)-1, 
                                          size=1, prob=qq))
@@
                      NResp = nresp)
-   res <- cbind(res, cov.mat)
+   res <- cbind(res, covs)
    res
@@
 ```
+
+### Example of simulated data
+
+```{r SimEx}
+simd <- sim.sp(n = 100,
+               clustersize.gen = cs.varied,
+               backbone.gen = function(n)qpower.pdf(p=0.8, rho=0.5, n=n),
+               covariate.gen = cov.trt,
+               formula = ~Trt,
+               beta = c(0, -0.5, -1, -2),
+               link = "log")
+
+ggplot(simd, aes(x=Trt, y=NResp/ClusterSize, size=ClusterSize)) +
+  geom_jitter(width=0.1, height=0, alpha=0.5) +
+  scale_size_area(breaks=c(1,5,10))
+```
+
+### Export multiple simulated sets
+
+The GEE and beta-binomial models with log-links will be fitted in SAS.

### 2019-10-01 R/SPreg.R
Add starting parameter values for binomial glm fit

@@ -159,7 +159,8 @@ sprr <- function(formula, data, subset, weights, link="cloglog", mu1=NULL, start
                 if (!is.null(weights)) ew <- weights * ew
             
             
-                mod <- glm(cbind(Ycomb[,"s"], Ycomb[,"y"]-Ycomb[,"s"]) ~ mm2+0, family=fam, weights=ew)
+                mod <- glm(cbind(Ycomb[,"s"], Ycomb[,"y"]-Ycomb[,"s"]) ~ mm2+0, 
+                           family=fam, weights=ew, start=beta_old)
                 beta_new <- coef(mod)
             
             

### 2019-10-04 Talk/Exchreg_talk.tex
ICOSDA talk initila version from Xinran's proposal

\documentclass[hyperref={bookmarks=false},aspectratio=169]{beamer}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{silence,lmodern}
\usepackage[style=verbose,backend=biber]{biblatex}
%\addbibresource{./figures/vv.bib}
% Filter warnings issued by package biblatex starting with "Patching footnotes failed"
\WarningFilter{biblatex}{Patching footnotes failed}

% ---------------  Define theme and color scheme  -----------------
\usetheme[sidebarleft]{Caltech}  % 3 options: minimal, sidebarleft, sidebarright

% ------------  Information on the title page  --------------------
\title[Semi-parametric model]
{\bfseries{Semi-parametric Model for Clustered Binary Outcome}}

\author[Aniko Szabo, Xinran Qi]
{Aniko Szabo\& Xinran Qi}

\institute[MCW]
{Medical College of Wisconsin}

\date[October 11, 2019]
{\footnotesize ICOSDA 2019}
%------------------------------------------------------------

%------------------------------------------------------------
%The next block of commands puts the table of contents at the
%beginning of each section and highlights the current section:

\AtBeginSection[]
{
  \begin{frame}
    \frametitle{Table of Contents}
    \tableofcontents[currentsection]
  \end{frame}
}
%------------------------------------------------------------

\begin{document}

\frame{\titlepage}  % Creates title page

%---------   table of contents after title page  ------------
\begin{frame}
\frametitle{Table of Contents}
\tableofcontents
\end{frame}
%---------------------------------------------------------

\section{Introduction}

%---------------------------------------------------------
%Two columns
\begin{frame}
\frametitle{Developmental toxicity of boron acid (BA) \footcite{heindel1994developmental}}

\begin{columns}

\column{0.6\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/Rat.png}
    \caption{Developmental toxicity experiment of boron acid}
    \label{fig:ratExperiment}
\end{figure}

\column{0.4\textwidth}
\begin{itemize}
    \item National Toxicology Program provided pregnant mice with feed containing BA at different dose levels.
    \item Number of fetuses in litter NOT affected by dose exposure.
    \item Number of resorptions, deaths, or other abnormalities.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Two columns
\begin{frame}
\frametitle{Developmental toxicity of boron acid (BA) Cont.}

\begin{columns}

\column{0.5\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/RatBarplot.png}
    \caption{Bar plot for developmental toxicity experiment of boron acid}
    \label{fig:ratExperimentBarplot}
\end{figure}

\column{0.5\textwidth}
\begin{itemize}
    \item Primary parameter of interest: entire distribution for combined endpoints from varying cluster sizes {\&} under different dose levels exposure.
    \item Correlated fetuses: gestate in same litter {\&} exposed to same/different concentrations of BA.
    \item Model within-cluster effects {\&} exogenous risks: second/higher order correlation coefficients as parameterization.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Two columns
\begin{frame}
\frametitle{Developmental toxicity of boron acid (BA) Cont.}

\begin{columns}

\column{0.5\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/RatBarplotRatAdded.png}
    \caption{Bar plot for developmental toxicity experiment of boron acid}
    \label{fig:ratExperimentBarplotRatAdded}
\end{figure}

\column{0.5\textwidth}
\begin{itemize}
    \item Primary parameter of interest: entire distribution for combined endpoints from varying cluster sizes {\&} under different dose levels exposure.
    \item Correlated fetuses: gestate in same litter {\&} exposed to same/different concentrations of BA.
    \item Model within-cluster effects {\&} exogenous risks: second/higher order correlation coefficients as parameterization.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

\section{Clustered Binary Outcomes}

%---------------------------------------------------------
%Two columns
\begin{frame}
\frametitle{Notations for Clustered Binary Outcomes}

\begin{columns}

\column{0.38\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/notations.png}
    \caption{Illustration of notations for clustered binary outcomes exposed to single dose level}
    \label{fig:notations}
\end{figure}

\column{0.62\textwidth}
\begin{itemize}
    \item A sample of clusters $(i = 1, \dots, I)$ with cluster sizes $(N_1, \dots, N_i, \dots, N_I)$.
    \item $\left\lbrace X_{i1}, \dots, X_{ij}, \dots, X_{iN_i} \right\rbrace$, $j \in \left\lbrace 1, \dots, N_i \right\rbrace$: observations in $i^{th}$ cluster of size $N_i$.
    \item $X_{ij}$: binary outcome, 1 indicating success {\&} 0 indicating failure.
    \item $Y_i = \sum_{j=1}^{N_i} X_{ij}$: total number of successes from cluster of size $N_i$.
    \item $q_{y,N} = \textbf{Pr}_N(\sum_{j=1}^{N} X_j = y)$: marginal probability of achieving $y$ successes from cluster of size $N$.
    \item $\mathcal{N} \geq \max{(N_{i})}$, where $1 \leq i \leq I$: maximal possible cluster size.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Modeling Exchangeable Clustered Binary Outcomes}

\begin{block}{Assumption: Exchangeability}
A sequence of binary random variables $\left\lbrace X_1, X_2, \dots, X_N \right \rbrace$ being exchangeable if for any $N$, any vector $(x_1, x_2, \dots, x_N)'$ and any permutation of indexes $\left\lbrace 1, 2, \dots, N \right\rbrace$:

$\textbf{Pr}(X_{\varphi(1)} = x_1, \dots, X_{\varphi(N)} = x_N) = \textbf{Pr}(X_1=x_1, \dots, X_N=x_N)$.
\end{block}

\textbf{Fixed or random effect models}

\begin{itemize}
    \item Fixed effect: Tallis [1962] added additional parameter $\rho$ to probability generating functions {\&} derived joint probability distribution of correlated observations within same cluster \footcite{tallis1962use}.
    \item Random effect: Williams [1975] proposed beta-binomial model assuming binary responses subject to Bernoulli whose probability varied among clusters according to Beta distribution \footcite{WilliamsDA1975}.
\end{itemize}

\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Modeling Exchangeable Clustered Binary Outcomes Cont.}

\textbf{Generalized estimating equation (GEE) models}

\begin{itemize}
    \item Expectation {\&} covariance matrix obtained using properties of generalized exponential family/direct derivation from given distributions {\&} definitions.
    \item Robust estimates of regression coefficients under misspecified working correlation matrix.
    \item Research objectives: marginal mean {\&} pairwise correlation \footcite{PrenticeZhao1991}, odds ratio \footcite{lipsitz1991generalized}, mean response {\&} intra-cluster correlation \footcite{BowmanChen1995}.
\end{itemize}

\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Modeling Exchangeable Clustered Binary Outcomes Cont.}

\textbf{Higher order moments and correlations}

\begin{itemize}
    \item Bowman and George [1995] proposed to use joint probability $\left\lbrace \lambda_k, k = 1, \dots, N \right\rbrace$ to reparametrize likelihood function \footcite{bowman1995saturated}.
    \item $\lambda_k = \textbf{Pr}(X_{i_1} = X_{i_2} = \dots = X_{i_k} = 1)$, with $\lambda_0=1$, $\lambda_1 = \textbf{Pr} (X_1 = 1)$, $\lambda_2 = \textbf{Pr} (X_1 = X_2 = 1)$.
\end{itemize}

\begin{block}{1-1 Mapping: Joint probability $\left\lbrace \lambda_k \right\rbrace$ {\&} marginal probability $\left\lbrace q_y \right\rbrace$}
\begin{equation*}\label{E:1to1}
q_y = \binom{N}{y} \sum_{k=0}^{N-y} {(-1)^k} \binom{N-y}{k} \lambda_{y+k} ,
\end{equation*}

\begin{equation*}\label{E:also1to1}
\lambda_k = \sum_{y=0}^{N-k} \frac{\binom{N-k}{y} {q_{N-y}}}{\binom{N}{N-y}} , y = 0, 1, \dots, N.
\end{equation*}
\end{block}

\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Modeling Exchangeable Clustered Binary Outcomes Cont.}

\textbf{Higher order moments and correlations}

\begin{itemize}
    \item A sequence of $\lambda$'s define valid distribution $\Longleftrightarrow$ completely monotone \footcite{bowman1995saturated}.
\end{itemize}

\begin{block}{Definition: Forward differences in $\left\lbrace \lambda_k \right\rbrace$}
\begin{equation*}\label{D:CM}
\Delta^{N-k} (\lambda_k) \geq 0, k = 0, \dots, N,
\end{equation*}
where ${\Delta}(\lambda_k) = \lambda_k - \lambda_{k+1}, {\Delta^2}(\lambda_k) = {\Delta}(\lambda_k) - {\Delta}(\lambda_{k+1})$, etc.
\end{block}

\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Modeling Exchangeable Clustered Binary Outcomes Cont.}

\textbf{Higher order moments and correlations}

\begin{itemize}
    \item Stefanescu and Turnbull [2003] proposed additional assumption “marginal compatibility” to estimate  $\left\lbrace \lambda_k \right\rbrace$ under varying cluster sizes \footcite{stefanescu2003likelihood}.
    \item Pang and Kuk [2007] proposed penalized kernel smoothing method which performed smoothing in both covariate and response space \footcite{pang2007test}.
\end{itemize}

\begin{block}{Assumption: Marginal compatibility}
For a cluster of size $N_1$ $(< N_2)$,
\begin{equation*}\label{A:marginalCompatibility}
\begin{split}
\lambda_{k,N_1} &= \boldsymbol{\Pr} \left( X_{1,N_1} = \dots = X_{k,N_1} = 1 \right) \\
                &= \boldsymbol{\Pr} \left( X_{1,N_2} = \dots = X_{k,N_2} = 1\right) = \lambda_{k,N_2}.
\end{split}
\end{equation*}
\end{block}

\end{frame}
%---------------------------------------------------------

\section{Clustered Multinomial Outcomes}

%---------------------------------------------------------
%Two columns
\begin{frame}
\frametitle{Example of Clustered Multinomial Outcomes}

\begin{columns}

\column{0.35\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/DMF1.png}
    \caption{Illustration of DMF \newline \tiny{(Pictures downloaded from: https://www.vectorstock.com/royalty-free-vector)}}
    \label{fig:toothDMF}
\end{figure}

\column{0.65\textwidth}
\begin{itemize}
    \item NCHS conducted periodic survey, National Health and Nutrition Examination Survey (NHANES), based on complex, multi-stage sample design.
    \item Provide national estimates of health {\&} nutritional status of US civilian, non-institutionalized population aged two months and older.
    \item DMF Index: total number of teeth/surfaces decayed, missing, or filled in individual.
    \item DMFT index: applied to teeth, scores per individual range from 0 to 28/32.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Two columns
\begin{frame}
\frametitle{Example of Clustered Multinomial Outcomes}

\begin{columns}

\column{0.35\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/DMF3pics.png}
    \caption{Illustration of DMF \newline \tiny{(Pictures downloaded from: https://www.vectorstock.com/royalty-free-vector)}}
    \label{fig:toothDMF3Left}
\end{figure}

\column{0.65\textwidth}
\begin{itemize}
    \item NCHS conducted periodic survey, National Health and Nutrition Examination Survey (NHANES), based on complex, multi-stage sample design.
    \item Provide national estimates of health {\&} nutritional status of US civilian, non-institutionalized population aged two months and older.
    \item DMF Index: total number of teeth/surfaces decayed, missing, or filled in individual.
    \item DMFT index: applied to teeth, scores per individual range from 0 to 28/32.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Two columns
\begin{frame}
\frametitle{Example of Clustered Multinomial Outcomes}

\begin{columns}

\column{0.35\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/DMF2pics.png}
    \caption{Illustration of DMF \newline \tiny{(Pictures downloaded from: https://www.vectorstock.com/royalty-free-vector)}}
    \label{fig:toothDMF2Left}
\end{figure}

\column{0.65\textwidth}
\begin{itemize}
    \item NCHS conducted periodic survey, National Health and Nutrition Examination Survey (NHANES), based on complex, multi-stage sample design.
    \item Provide national estimates of health {\&} nutritional status of US civilian, non-institutionalized population aged two months and older.
    \item DMF Index: total number of teeth/surfaces decayed, missing, or filled in individual.
    \item DMFT index: applied to teeth, scores per individual range from 0 to 28/32.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Two columns
\begin{frame}
\frametitle{Example of Clustered Multinomial Outcomes}

\begin{columns}

\column{0.35\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/DMF1pics.png}
    \caption{Illustration of DMF \newline \tiny{(Pictures downloaded from: https://www.vectorstock.com/royalty-free-vector)}}
    \label{fig:toothDMF1Left}
\end{figure}

\column{0.65\textwidth}
\begin{itemize}
    \item NCHS conducted periodic survey, National Health and Nutrition Examination Survey (NHANES), based on complex, multi-stage sample design.
    \item Provide national estimates of health {\&} nutritional status of US civilian, non-institutionalized population aged two months and older.
    \item DMF Index: total number of teeth/surfaces decayed, missing, or filled in individual.
    \item DMFT index: applied to teeth, scores per individual range from 0 to 28/32.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

\section{Exponential Tilting}

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Definition of Exponential Tilting}

\begin{columns}

\column{0.4\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/p.png}
    \caption{Original probability distribution}
    \label{fig:OriginalPMF}
\end{figure}

\column{0.6\textwidth}
\begin{itemize}
    \item For random variable $Y$ with probability density distribution $\boldsymbol{p}$ {\&} mean at $\mu_0$.
    \item New distribution $\boldsymbol{f}$ with shortest Kullback - Leibler distance/divergence to distribution $\boldsymbol{p}$ {\&} mean at pre-specified value $\mu_1$.
    \item $f_y \propto p_y  \exp (\theta y)$: natural exponential family generated from distribution $\boldsymbol{p}$ via exponential tilting \footcite{EETCarloCamarda}.
    \item $\Longleftrightarrow$ distribution $\boldsymbol{p}$ with Lagrange multipliers for constraints define distribution $\boldsymbol{f}$.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Definition of Exponential Tilting}

\begin{columns}

\column{0.4\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/pf.png}
    \caption{Original and exponentially tilted probability distributions}
    \label{fig:OriginalETPMF1}
\end{figure}

\column{0.6\textwidth}
\begin{itemize}
    \item For random variable $Y$ with probability density distribution $\boldsymbol{p}$ {\&} mean at $\mu_0$.
    \item New distribution $\boldsymbol{f}$ with shortest Kullback - Leibler distance/divergence to distribution $\boldsymbol{p}$ {\&} mean at pre-specified value $\mu_1$.
    \item $f_y \propto p_y  \exp (\theta y)$: natural exponential family generated from distribution $\boldsymbol{p}$ via exponential tilting \footcite{EETCarloCamarda}.
    \item $\Longleftrightarrow$ distribution $\boldsymbol{p}$ with Lagrange multipliers for constraints define distribution $\boldsymbol{f}$.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Definition of Exponential Tilting}

\begin{columns}

\column{0.4\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/pff.png}
    \caption{Original and exponentially tilted probability distributions}
    \label{fig:OriginalETPMF2}
\end{figure}

\column{0.6\textwidth}
\begin{itemize}
    \item For random variable $Y$ with probability density distribution $\boldsymbol{p}$ {\&} mean at $\mu_0$.
    \item New distribution $\boldsymbol{f}$ with shortest Kullback - Leibler distance/divergence to distribution $\boldsymbol{p}$ {\&} mean at pre-specified value $\mu_1$.
    \item $f_y \propto p_y  \exp (\theta y)$: natural exponential family generated from distribution $\boldsymbol{p}$ via exponential tilting \footcite{EETCarloCamarda}.
    \item $\Longleftrightarrow$ distribution $\boldsymbol{p}$ with Lagrange multipliers for constraints define distribution $\boldsymbol{f}$.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Generalized Linear Models with Unspecified Reference Distributions for Continuous Outcomes \footcite{rathouz2008generalized}}

\begin{columns}

\column{0.4\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/orginalModel.png}
    \caption{Flowchart of Rathouz and Gao's semi-parametric generalized linear model}
    \label{fig:RathouzGaoSPGLM}
\end{figure}

\column{0.6\textwidth}
\begin{itemize}
    \item Rathouz and Gao [2008] proposed semi-parametric generalized linear model, specified with linear predictor {\&} link function for mean response $Y$.
    \item \alert{Conditional mean model}: $E\left( Y|\boldsymbol{Z};\boldsymbol{\beta} \right) =\mu\left( \boldsymbol{Z}, \boldsymbol{\beta} \right) \equiv \mu = h^{-1}\left( \eta\right)$, where $\eta = \boldsymbol{Z}' \boldsymbol{\beta}$.
    \item \alert{Probability density}: $f\left( y |\boldsymbol{Z};\boldsymbol{\beta}, f_0\right) \propto f_0\left( y\right) \exp(\theta y)$.
    \item Extend framework by estimating $f_0$ from data {\&} maintaining mean model at user-specified $h\left( \mu\right) =\eta$.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

\section{Research Aims for Dissertation}

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Aim 1: Develop A Semi-parametric Generalized Linear Model (SPGLM) for Clustered Binary Outcomes}

\begin{columns}

\column{0.4\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/modelAim1.png}
    \caption{Flowchart of extending SPGLM to model clustered binary outcomes with varying cluster sizes}
    \label{fig:Aim11}
\end{figure}

\column{0.6\textwidth}
\begin{itemize}
    \item Extend SPGLM to model clustered binary outcomes with varying cluster sizes.
    \item Response variable: number of successes $Y$ scaled by cluster size $N$.
    \item \alert{Assumption}: probability density of $Y_g/N$ exposed to $g^{th}$ ($g = 1, \dots, G$) dose level, $\left\lbrace q_{y,N}^{(g)} \right\rbrace$, is exponentially tilted version of reference dose level's density $\left\lbrace q_{y,N} \right\rbrace$.
    \item \alert{Conditional mean model}: \newline
    $E\left( \frac{Y_g}{N} | \boldsymbol{Z}_g, \boldsymbol{\beta}_g \right) = \mu(\boldsymbol{Z}_g, \boldsymbol{\beta}_g)  \equiv \mu_g = h^{-1} \left( \boldsymbol{Z}_g^{'} \boldsymbol{\beta}_g \right)$, where $\mu_g \in (0, 1)$.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Aim 1: Develop A Semi-parametric Generalized Linear Model (SPGLM) for Clustered Binary Outcomes Cont.}

\begin{columns}

\column{0.4\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/modelAim1.png}
    \caption{Flowchart of extending SPGLM to model clustered binary outcomes with varying cluster sizes}
    \label{fig:Aim12}
\end{figure}

\column{0.6\textwidth}
\begin{itemize}
    \item \alert{Probability density}: for a cluster of size $N$, $ q_{y,N}^{(g)} \left( \boldsymbol{Z}_g, \boldsymbol{\beta}_g \right) \propto q_{y,N} \times \exp{ \left[ - \omega_g(\boldsymbol{Z}_g, \boldsymbol{\beta}_g) y \right]} $.
    \item $\omega_g(\boldsymbol{Z}_g, \boldsymbol{\beta}_g)$: tilting parameter for $g^{th}$ dose level ($\omega_0 \equiv 0$ for reference group), $\Longrightarrow$ $\displaystyle \sum_{y=0}^N \dfrac{y q_{y,N}^{(g)}}{N} = \mu_g = h^{-1}\left( \boldsymbol{Z}_g^{'} \boldsymbol{\beta}_g \right)$.
    \item For identifiability, mean of $\left\lbrace q_{y,N}\right\rbrace$ equal to arbitrary fixed value.
    \item \alert{Extend models for varying cluster sizes by assuming marginal compatibility}.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Aim 1: Develop A Semi-parametric Generalized Linear Model (SPGLM) for Clustered Binary Outcomes Cont.}

\begin{columns}

\column{0.4\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/modelAim1.png}
    \caption{Flowchart of extending SPGLM to model clustered binary outcomes with varying cluster sizes}
    \label{fig:Aim13}
\end{figure}

\column{0.6\textwidth}
\begin{itemize}
    \item Similar Expectation Maximization Newton-Raphson algorithm proposed by Rathouz and Gao [2008] for parameter estimation.
    \item Increase {\&} decrease risk from reference probability density.
    \item Application: BA rat data.
    \item Simulation studies: finite-sample properties {\&} compare results with existing methods.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Aim 2: Generalize Concept of Marginal Compatibility for Clustered Binary Outcomes}

\textbf{Parametric extension}: include certain transformation of cluster size as covariate in linear predictor.

\begin{columns}

\column{0.4\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/modelAim2.png}
    \caption{Flowchart of generalizing marginal compatibility for clustered binary outcomes}
    \label{fig:Aim21}
\end{figure}

\column{0.6\textwidth}
\begin{itemize}
    \item \alert{Parametric conditional mean model}:
    \begin{equation*}\label{M:parametricSPGLM_GMCconditionalMean}
    \begin{split}
    E\left( \frac{Y_i}{N_i} | \boldsymbol{Z}_i, \boldsymbol{\beta} \right) &= \mu(\boldsymbol{Z}_i, \boldsymbol{\beta})  \equiv \mu_i  \\
    &= h^{-1} \left( \boldsymbol{Z}_i^{'} \boldsymbol{\beta} \right) \\
    &= h^{-1} \left( \beta_0 + \beta_1 \log(N_i) \right) . \\
    \end{split}
    \end{equation*}
    \item Cluster size included in $\boldsymbol{Z}$ as covariate, $\boldsymbol{Z}_i = \left( 1, \log(N_i) \right)^{'}$.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Aim 2: Generalize Concept of Marginal Compatibility for Clustered Binary Outcomes Cont.}

\textbf{Parametric extension}: include certain transformation of cluster size as covariate in linear predictor.

\begin{columns}

\column{0.4\textwidth}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{./figures/modelAim2.png}
    \caption{Flowchart of generalizing marginal compatibility for clustered binary outcomes}
    \label{fig:Aim22}
\end{figure}

\column{0.6\textwidth}
\begin{itemize}
    \item For a cluster of size $N_i$ ($\leq N$), \newline
    $q_{Y_i,N_i} =\binom{N_i}{Y_i} \displaystyle \sum_{t=Y_i}^{N-N_i+Y_i} \dfrac{\binom{N-N_i}{t-Y_i}} {\binom{N}{t}}$\alert{$q_{t,N|N_i}$}, $Y_i = 0, \dots, N_i$, $t=0, \dots, N$.
    \item \alert{Parametric generalized marginal compatibility model}: \newline
    $q_{t,N|N_i}\left( \boldsymbol{Z}_i, \boldsymbol{\beta} \right) \propto q_{t,N} \times \exp{ \left[ - \omega(\boldsymbol{Z}_i, \boldsymbol{\beta}) t \right]} $.
\end{itemize}

\end{columns}
\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Aim 2: Generalize Concept of Marginal Compatibility for Clustered Binary Outcomes Cont.}

\textbf{Non-parametric extension}: assign different tilting parameters to different cluster sizes.

\begin{itemize}
    \item \alert{Assumption}: for smaller cluster size $N'$ ($<N$), probability density $\left\lbrace q_{y,N'}, y = 0, \dots, N' \right\rbrace$, subject to \alert{marginal compatibility} {\&} derive from exponentially tilted version of maximum cluster size's density $\left \lbrace q_{t,N|N'}, t=0,\dots, N\right\rbrace$.
    \item $q_{y,N'} =\binom{N'}{y} \displaystyle \sum_{t=y}^{N-N'+y} \dfrac{\binom{N-N'}{t-y} q_{t,N|N'}}{\binom{N}{t}}$.
    \item \alert{Non-parametric conditional mean model}:
    \begin{equation*}\label{M:NONparametricSPGLM_GMCconditionalMean}
    E\left( \frac{Y}{N} | N', \boldsymbol{Z}, \boldsymbol{\beta} \right) = \mu(N', \boldsymbol{Z}, \boldsymbol{\beta})  \equiv \mu = h^{-1} \left( \boldsymbol{Z}^{'} \boldsymbol{\beta} \right) .
    \end{equation*}
\end{itemize}

\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Aim 2: Generalize Concept of Marginal Compatibility for Clustered Binary Outcomes Cont.}

\textbf{Non-parametric extension}: assign different tilting parameters to different cluster sizes.

\begin{itemize}
    \item \alert{Non-parametric generalized marginal compatibility model}:
    \begin{equation*}\label{M:NONparametricSPGLM_GMCexponentialTiltPDF}
    \begin{split}
    q_{y,N|N'} &= q_{y,N} \times \exp{ \left[ - \omega_{N'} y - b \left( \omega_{N'} \right) \right]} \\
    & \propto q_{y,N} \times \exp{ \left( - \omega_{N'} y \right)} .
    \end{split}
    \end{equation*}
    \item Similar Expectation Maximization Newton-Raphson algorithm \footcite{rathouz2008generalized} for parameter estimation.
    \item Additional penalty $\varrho \displaystyle\sum_{N'=1}^N (\omega_{N'} - \omega_{N'-1})^2$, $\omega_N=0$, assigned in Lagrangian.
\end{itemize}

\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Aim 3: Develop A Semi-parametric Model for Clustered Ordinal Outcomes}

\alert{Semi-parametric model}:
\begin{equation*}\label{M:SPforOrdinal1}
\begin{split}
    \boldsymbol{\Pr} \big( & X_1 = \dots = X_{C_1} = 1,  X_{C_1 + 1} = \dots = X_{C_1 + C_2} = 2, \cdots, \\
   & X_{C_1 + \dots + C_{a-1} + 1} = \dots = X_{C_1 + \cdots + C_{a-1} +  C_{a}} = a, \cdots, \\
   & X_{C_1 + \dots + C_{A-2} + 1} = \dots = X_{C_1 + \dots + C_{A-2} +  C_{A-1}} = A-1, \\
   & X_{C_1 + \dots + C_{A-1} + 1} = \dots = X_{C_1 + \dots + C_{A-1} +  C_{A}} = A | N \big) \\
   & {\color{white} = {\boldsymbol{\Pr}}^{(1)} \big( X_1 = \dots = X_{C_1} = 1 | N \big) \times {\boldsymbol{\Pr}}^{(2)} \big(X_{1} = \dots = X_{C_2} = 2 | N - C_1\big) \times \cdots }\\
   & {\color{white} \times {\boldsymbol{\Pr}}^{(a)} \big(X_{1} = \dots = X_{C_{a}} = a | N - C_1 - \cdots - C_{a-1} \big) \times \cdots }\\
   & {\color{white} \times {\boldsymbol{\Pr}}^{(A-1)} \big(X_{1} = \dots = X_{C_{A-1}} = A-1 | N - C_1 - \cdots - C_{A-2} \big) } \\
   & {\color{white} = \dfrac{q^{(1)}_{C_1,N}}{\binom{N}{C_1}} \times \dfrac{q^{(2)}_{C_2,N - C_1}}{\binom{N - C_1}{C_2}} \times \cdots \times \dfrac{q^{(a)}_{C_{a},N - C_1 - \cdots - C_{a-1}}}{\binom{N - C_1 - \cdots - C_{a-1}}{C_a}} \times \cdots \times \dfrac{q^{(A-1)}_{C_{A-1},N - C_1 - \cdots - C_{A-2}}}{\binom{N - C_1 - \cdots - C_{A-2}}{C_{A-1}}} }
\end{split}
\end{equation*}

\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Aim 3: Develop A Semi-parametric Model for Clustered Ordinal Outcomes}

\alert{Semi-parametric model}:
\begin{equation*}\label{M:SPforOrdinal2}
\begin{split}
    \boldsymbol{\Pr} \big( & X_1 = \dots = X_{C_1} = 1,  X_{C_1 + 1} = \dots = X_{C_1 + C_2} = 2, \cdots, \\
   & X_{C_1 + \dots + C_{a-1} + 1} = \dots = X_{C_1 + \cdots + C_{a-1} +  C_{a}} = a, \cdots, \\
   & X_{C_1 + \dots + C_{A-2} + 1} = \dots = X_{C_1 + \dots + C_{A-2} +  C_{A-1}} = A-1, \\
   & X_{C_1 + \dots + C_{A-1} + 1} = \dots = X_{C_1 + \dots + C_{A-1} +  C_{A}} = A | N \big) \\
   & = {\boldsymbol{\Pr}}^{(1)} \big( X_1 = \dots = X_{C_1} = 1 | N \big) \times {\boldsymbol{\Pr}}^{(2)} \big(X_{1} = \dots = X_{C_2} = 2 | N - C_1\big) \times \cdots \\
   & \times {\boldsymbol{\Pr}}^{(a)} \big(X_{1} = \dots = X_{C_{a}} = a | N - C_1 - \cdots - C_{a-1} \big) \times \cdots \\
   & \times {\boldsymbol{\Pr}}^{(A-1)} \big(X_{1} = \dots = X_{C_{A-1}} = A-1 | N - C_1 - \cdots - C_{A-2} \big) \\
   & {\color{white} = \dfrac{q^{(1)}_{C_1,N}}{\binom{N}{C_1}} \times \dfrac{q^{(2)}_{C_2,N - C_1}}{\binom{N - C_1}{C_2}} \times \cdots \times \dfrac{q^{(a)}_{C_{a},N - C_1 - \cdots - C_{a-1}}}{\binom{N - C_1 - \cdots - C_{a-1}}{C_a}} \times \cdots \times \dfrac{q^{(A-1)}_{C_{A-1},N - C_1 - \cdots - C_{A-2}}}{\binom{N - C_1 - \cdots - C_{A-2}}{C_{A-1}}} }
\end{split}
\end{equation*}

\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Aim 3: Develop A Semi-parametric Model for Clustered Ordinal Outcomes}

\alert{Semi-parametric model}:
\begin{equation*}\label{M:SPforOrdinal3}
\begin{split}
    \boldsymbol{\Pr} \big( & X_1 = \dots = X_{C_1} = 1,  X_{C_1 + 1} = \dots = X_{C_1 + C_2} = 2, \cdots, \\
   & X_{C_1 + \dots + C_{a-1} + 1} = \dots = X_{C_1 + \cdots + C_{a-1} +  C_{a}} = a, \cdots, \\
   & X_{C_1 + \dots + C_{A-2} + 1} = \dots = X_{C_1 + \dots + C_{A-2} +  C_{A-1}} = A-1, \\
   & X_{C_1 + \dots + C_{A-1} + 1} = \dots = X_{C_1 + \dots + C_{A-1} +  C_{A}} = A | N \big) \\
   & = {\boldsymbol{\Pr}}^{(1)} \big( X_1 = \dots = X_{C_1} = 1 | N \big) \times {\boldsymbol{\Pr}}^{(2)} \big(X_{1} = \dots = X_{C_2} = 2 | N - C_1\big) \times \cdots \\
   & \times {\boldsymbol{\Pr}}^{(a)} \big(X_{1} = \dots = X_{C_{a}} = a | N - C_1 - \cdots - C_{a-1} \big) \times \cdots \\
   & \times {\boldsymbol{\Pr}}^{(A-1)} \big(X_{1} = \dots = X_{C_{A-1}} = A-1 | N - C_1 - \cdots - C_{A-2} \big) \\
   & = \dfrac{q^{(1)}_{C_1,N}}{\binom{N}{C_1}} \times \dfrac{q^{(2)}_{C_2,N - C_1}}{\binom{N - C_1}{C_2}} \times \cdots \times \dfrac{q^{(a)}_{C_{a},N - C_1 - \cdots - C_{a-1}}}{\binom{N - C_1 - \cdots - C_{a-1}}{C_a}} \times \cdots \times \dfrac{q^{(A-1)}_{C_{A-1},N - C_1 - \cdots - C_{A-2}}}{\binom{N - C_1 - \cdots - C_{A-2}}{C_{A-1}}}
\end{split}
\end{equation*}

\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Aim 3: Develop A Semi-parametric Model for Clustered Ordinal Outcomes Cont.}

\begin{itemize}
    \item DMFT data from NHANES, patients records treated as random samples, which in fact based on complex, multi-stage design.
    \item \alert{Assumption}: probability density of smaller cluster of size $N'$ ($<N$) from exponentially tilted version of density for maximum cluster size $N$.
    \item $\left\lbrace q^{(a)}_{C_{a},N - C_1 - \cdots - C_{a-1}}, a=1, \dots, A-1 \right\rbrace$: marginal probability from smaller cluster of size $N' := {N - C_1 - \cdots - C_{a-1}}$.
    \item Subject to \alert{marginal compatibility} {\&} derive from exponentially tilted version of maximum cluster size's distribution $\left\lbrace q^{(a)}_{C_{a},N | N - C_1 - \cdots - C_{a-1}} \right\rbrace$.
    \item $q^{(a)}_{y,N'} =\binom{N'}{y} \displaystyle \sum_{t=y}^{N-N'+y} \dfrac{\binom{N-N'}{t-y} q^{(a)}_{t,N|N'}}{\binom{N}{t}}$.
\end{itemize}

\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
%Changing visivility of the text
\begin{frame}
\frametitle{Aim 3: Develop A Semi-parametric Model for Clustered Ordinal Outcomes Cont.}

\begin{itemize}
    \item \alert{Generalized marginal compatibility model for ordinal outcomes}:
    \begin{equation*}\label{M:GMCordinal_ExponentialTilting}
    \begin{split}
    q^{(a)}_{y,N|N'} &= q^{(a)}_{y,N} \times \exp{ \left[ - \omega^{(a)}_{N'} y - b \left( \omega^{(a)}_{N'} \right) \right]} \\
    & \propto q^{(a)}_{y,N} \times \exp{ \left( - \omega^{(a)}_{N'} y \right)} .
    \end{split}
    \end{equation*}
    \item $\omega^{(a)}_{N'}$: tilting parameter for cluster size $N'$ {\&} response category “$a$”.
    \item Similar Expectation Maximization Newton-Raphson algorithm \footcite{rathouz2008generalized} for parameter estimation.
    \item Additional penalty $\varrho_a \displaystyle\sum_{N'=1}^N (\omega_{N'} - \omega^{(a)}_{N'-1})^2$, $\omega^{(a)}_N=0$, assigned in Lagrangian.
    \item R package \textbf{CorrBin}.
\end{itemize}

\end{frame}
%---------------------------------------------------------

%---------------------------------------------------------
\begin{frame}[allowframebreaks]
     \printbibliography
 \end{frame}
%---------------------------------------------------------

\section{Acknowledgement}

%---------------------------------------------------------
\begin{frame}
     \begin{itemize}
     \item First of all, I would like to express my thanks to Dr. Szabo for her constant guidance throughout the proposal work and being my advisor. This proposal would not have been possible without her timely suggestion despite of her busy schedule. I also would like to express my deepest appreciation to my committee chair and advisor, Dr. Laud, for his continually and convincingly conveying a spirit of adventure in regard to research and scholarship, as well as an excitement in regard to teaching.
     \item  I will not forget to thank Dr. Wang, my RA advisor, whom for without his patience and guidance, I would not have made it so far, especially with my extended research in new developments on statistical machine learning.
     \item I also take this opportunity to specially thank Dr. Zhang, Dr. Banerjee and Dr. Okunseri for their valuable insights and kind help serving on my thesis committee. I would like extend my thanks to all the staff and faculty in the Division of Biostatistics.
     \item Finally, heartfelt gratitude to my friend, Ying, as well as family for the moral support and encouragement throughout graduation days.
     \end{itemize}
 \end{frame}
%---------------------------------------------------------
\end{document}

### 2019-11-22 SPsims/SPsim.Rmd
Move ICOSDA talk simulations to subfolder

---
title: "Simulation study"
output: html_notebook
---

```{r setup}
source('R/SPreg.R', echo=FALSE)
library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_classic())
```

Goal:

 * simulate data with a q-power backbone distribution that follows the relative-risk regression model
 * fit the data using our semi-parametric regression model, beta-binomial regression (misspecified correlation structure), and GEE (marginal only)

$$
	\lambda_k(z) = {\mu_k} \times {\theta(\beta' z)^k}, k = 0, \dots, K,
$$

## Simulating data
### Cluster-size distributions

 1. All small (3); 
 2. All large (10); 
 3. varying (0.5*(1+binom(9, 0.5)) + 0.5*Unif(2,10))

```{r Clustersizes}
cs.small <- function(n) rep(3, n)
cs.large <- function(n) rep(10, n)
cs.varied <- function(n) {
  p <- rbinom(n, size=1, prob=0.5)
  x <- ifelse(p, 1+rbinom(n, size=9, prob=0.5), sample(2:10, size=n, replace=TRUE))
  x
}
```

### Covariate-generating functions

Returns model matrix with randomly generated covariates

```{r Covgen}
cov.trt <- function(n, ntrts=4){
  obs <- rep(LETTERS[1:ntrts], length = n)
  dat <- data.frame(Trt = obs)
  dat
}

cov.bin_norm <- function(n, p=0.5){
  bin <- rbinom(n, size=1, prob=p)
  norm <- rnorm(n)
  data.frame(X1 = bin, X2 = norm)
}
```

### Simulate one dataset

The inputs will be

 * `n` - number of clusters to simulate data
 * `clustersize.gen` - function to generate the clustersizes
 * `backbone.gen` - function to generate the PDF of the backbone
 * `covariate.gen` - function to generate the covariates
 * `beta` - coefficients of the linear predictor
 * `link` - the link-function for the semi-parametric model; the inverse link is the $\theta$ function in the equation

```{r SimSetup}
sim.sp <- function(n, clustersize.gen, backbone.gen, 
                   covariate.gen, formula, beta, link="log"){

   model_fun <- binomial(link=link)$linkinv

   cs <- clustersize.gen(n)
   K <- max(cs)
   covs <- covariate.gen(n)
   cov.mat <- model.matrix(formula, data=covs)

   lp <- cov.mat %*% beta
   theta <- model_fun(lp)

   q0 <- backbone.gen(K)
   lambda0 <- lambda_from_p(q0)

   th <- sapply(0:K, function(k)theta^k)
   lambda <- apply(th, 1, function(t)t * lambda0)
   rownames(lambda) <- 0:K

   q <- lapply(1:n,
               function(idx)p_from_lambda(lambda[,idx], n=cs[idx]))
   nresp <- sapply(q, function(qq)sample(seq_along(qq)-1, 
                                         size=1, prob=qq))
   res <- data.frame(ID=1:n,
                     ClusterSize = cs,
                     NResp = nresp)
   res <- cbind(res, covs)
   list(data=res, probs=q)
}

```

### Example of simulated data

```{r SimEx}
simd <- sim.sp(n = 100,
               clustersize.gen = cs.varied,
               backbone.gen = function(n)qpower.pdf(p=0.8, rho=0.5, n=n),
               covariate.gen = cov.trt,
               formula = ~Trt,
               beta = c(0, -0.5, -1, -2),
               link = "log")$data

ggplot(simd, aes(x=Trt, y=NResp/ClusterSize, size=ClusterSize)) +
  geom_jitter(width=0.1, height=0, alpha=0.5) +
  scale_size_area(breaks=c(1,5,10))
```

### Simulate and export multiple simulated sets

The GEE and beta-binomial models with log-links will be fitted in SAS, so we will simulate all the datasets at once.

```{r ActualSim}
set.seed(10012019)
R <- 500
sim.list <- 
  lapply(1:R,
         function(idx){
           res <- sim.sp(n = 100,
               clustersize.gen = cs.varied,
               backbone.gen = function(n)qpower.pdf(p=0.5, rho=0.5, n=n),
               covariate.gen = cov.trt,
               formula = ~Trt,
               beta = c(0, -0.5, -1, -2),
               link = "log")$data
           res$Dataset <- idx
           res})
sim.all <- do.call(rbind, sim.list)
write.table(sim.all, file="SimData.txt", row.names=FALSE, quote=FALSE)
```

## Run analyses

```{r RunSAS}
if (.Platform$OS.type == "unix"){
  system("sas FitModels", wait=TRUE)
}
gee.res <- read.delim("GEEest.txt") %>%
  mutate(Coef = paste0(Parm, Level1)) %>%
  select(-Parm,-Level1) %>%
  group_by(Dataset) %>%
  spread(key=Coef, value=Estimate)

bb.res <- read.delim("BBest.txt")%>%
  mutate(Coef = paste0(Effect, Trt)) %>%
  select(-Effect,-Trt) %>%
  group_by(Dataset) %>%
  spread(key=Coef, value=Estimate)
```

```{r RunSP, warning=FALSE}
sp.res.list <- lapply(sim.list,
                 function(d){
                   tryCatch(fit <- sprr(cbind(NResp, ClusterSize-NResp) ~ Trt,
                               data=d, link="log",
                               control=list(maxit=100, eps=0.01)), silent=TRUE,
                            error = function(e)NULL)
                 })
sp.coef <- lapply(sp.res.list, 
                      function(z)if(is.null(z)) rep(NA,4) else coef(z))
sp.mu <- lapply(sp.res.list, 
                 function(z)if(is.null(z)) NA else lambda_from_p(z$q)[2])
sp.res <- as_tibble(do.call(rbind, sp.coef)) %>%
  bind_cols(mu = unlist(sp.mu)) %>%
  mutate(Dataset = 1:length(sp.coef)) %>%
  mutate(Intercept = `(Intercept)`+log(mu))

```

## Compare results

```{r Compare}
comb <- bind_rows(list(GEE=gee.res,
                       BB=bb.res,
                       SP=sp.res), .id="Method") %>%
  select(-TrtA, -`Scale Parameter`, -`(Intercept)`, -mu) %>%
  group_by(Dataset, Method) %>%
  gather(key="Parameter", value="Estimate", Intercept:TrtD)

correct <- tibble(Parameter=c("Intercept","TrtB","TrtC", "TrtD"),
                  Correct = c(log(0.5), -0.5, -1, -2))
comb <- left_join(comb, correct, by="Parameter")

scomb <- comb %>%
  group_by(Method, Parameter) %>%
  summarize(bias = mean(Estimate-Correct, na.rm=TRUE),
            se = sd(Estimate, na.rm=TRUE),
            rmse = sqrt(mean((Estimate-Correct)^2, na.rm=TRUE)))
kable(scomb, digits=2)
```

```{r ComparePlot}
ggplot(comb, aes(x=Method, y=Estimate, colour=Method)) +
  facet_wrap(~Parameter) +
  geom_hline(aes(yintercept=Correct)) +
  geom_jitter(alpha=0.3, height=0)
```

### Compare predicted probability of 0 events in a cluster of size 5

```{r ProbZero}
nd <- expand.grid(Trt=LETTERS[1:4])
sp.zero.list <- lapply(sp.res.list, 
                  function(fit){
                    pr <- predict(fit, newdata=nd, newn=5, type="probvec")
                    cbind(nd, p0=do.call(rbind, pr)[,1])})
sp.zero <- bind_rows(sp.zero.list, .id="Dataset") %>%
  mutate(Dataset = as.numeric(Dataset))

# rho =1/(1+phi)
bb.zero <- bb.res %>% 
  gather(key="TrtS", value="Estimate", TrtA:TrtD) %>%
  mutate(rho = 1/(1+`Scale Parameter`),
         mu = exp(Intercept + Estimate),
         Trt = substr(TrtS, 4, 4),
         p0 = dbetabinom(0, size=5, prob = mu, rho = rho)) %>%
  select(Dataset, Trt, p0)
```

Calculate correct values:
```{r CorrectProbZero}
p0c <- sim.sp(n = 4,
       clustersize.gen = function(n)rep(5,n),
       backbone.gen = function(n)qpower.pdf(p=0.5, rho=0.5, n=n),
       covariate.gen = cov.trt,
       formula = ~Trt,
       beta = c(0, -0.5, -1, -2),
       link = "log")$probs %>%
  sapply(., function(x)x[1])
correct.zero <- cbind(nd, Correct=p0c)
```

```{r ProbZeroComb}
comb.zero <- bind_rows(list(BB=bb.zero,
                            SP=sp.zero), .id="Method") %>%
  left_join(correct.zero, by="Trt")

scomb.zero <- comb.zero %>%
  group_by(Method, Trt) %>%
  summarize(m_p0 = mean(p0, na.rm = TRUE),
            bias = mean(p0 - Correct, na.rm = TRUE),
            se = sd(p0, na.rm=TRUE),
            rmse = sqrt(mean((p0-Correct)^2, na.rm=TRUE)))
scomb.zero
```

```{r CompareZeroPlot}
ggplot(comb.zero, aes(x=Method, y=p0, colour=Method)) +
  facet_wrap(~Trt) +
  geom_hline(aes(yintercept=Correct)) +
  geom_jitter(alpha=0.3, height=0)
```

### 2019-11-22 R/MixMod.R
Initial commit of mixture modeling code

@@ -1,0 +1,2 @@
+
+require(CorrBin)

### 2019-11-27 nuweb/Clustering.tex
Rearrange folder structure

\newcommand{\NWtarget}[2]{#2}
\newcommand{\NWlink}[2]{#2}
\newcommand{\NWtxtMacroDefBy}{Fragment defined by}
\newcommand{\NWtxtMacroRefIn}{Fragment referenced in}
\newcommand{\NWtxtMacroNoRef}{Fragment never referenced}
\newcommand{\NWtxtDefBy}{Defined by}
\newcommand{\NWtxtRefIn}{Referenced in}
\newcommand{\NWtxtNoRef}{Not referenced}
\newcommand{\NWtxtFileDefBy}{File defined by}
\newcommand{\NWtxtIdentsUsed}{Uses:}
\newcommand{\NWtxtIdentsNotUsed}{Never used}
\newcommand{\NWtxtIdentsDefed}{Defines:}
\newcommand{\NWsep}{${\diamond}$}
\newcommand{\NWnotglobal}{(not defined globally)}
\newcommand{\NWuseHyperlinks}{}
\documentclass[reqno]{amsart}
\usepackage[margin=1in]{geometry}
\usepackage[colorlinks=true,linkcolor=blue]{hyperref}
\renewcommand{\NWtarget}[2]{\hypertarget{#1}{#2}}
\renewcommand{\NWlink}[2]{\hyperlink{#1}{#2}}
\newcommand{\vv}[1]{\boldsymbol{#1}}
\newcommand{\ff}[1]{\dot{#1}}
\newcommand{\Z}[2][K]{\mathbb{Z}_{#1}^{#2}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\EM}{\mathcal{EM}}

\title{Gaussian mixture model for clustered observations}
\author{Aniko Szabo}
\date{\today}

\begin{document}
\maketitle

\section{Mixture model}
Let $Y_{ij}$, $i=1,\ldots,\I$, $j=1,\ldots,n_i$ denote a continuous random variable 
measured in subject $j$ of cluster $i$. We want to model the distribution of $Y$
as a mixture of $K$ latent normal distributions:
\begin{equation}
[Y_{ij} \mid \xi_{ij}=k] \sim N(\mu_k, \sigma_k),
\end{equation}
where $\xi_{ij}$ is the latent class indicator for subject $j$ of cluster $i$
taking values from 1 to $K$.

Due to the clustering we cannot assume that $\xi_{ij}$ are independent, however we
will assume that they follow and exchangeable multinomial distribution $\EM(K,\vv{q})$.

Denoting $\phi(y; \mu, \sigma)$  the normal pdf with mean $\mu$ and standard deviation 
$\sigma$, the likelihood is
\begin{align}
L = & \prod_{i=1}^\I \sum_{\vv{x}\in \Z{n_i}} Pr(\vv{\xi}_{i}=\vv{x})
\times \prod_{j=1}^{n_i} \phi(y_{ij}; \mu_{x_j}, \sigma_{x_j})\\
 = &  \prod_{i=1}^\I \sum_{\vv{x}\in \Z{n_i}} \frac{q_{\ff{\vv{x}}|n_i}}{\binom{n_i}{\ff{\vv{x}}}} 
\times \prod_{j=1}^{n_i} \phi(y_{ij}; \mu_{x_j}, \sigma_{x_j}),
\end{align}
where $\vv{x} = (x_1,\ldots,x_{n_i})$, $\vv{\xi}_{i}=(\xi_{i1},\ldots,\xi_{in_i})$, 
$\ff{\vv{x}} = (\sum_j I(x_j=1),\ldots,\sum_j I(x_j=K))$ is the vector of latent category 
frequencies, and
\begin{equation}
q_{\vv{r}|n_i} = \sum_{\vv{t}\in\mathbb{Z}_N: \vv{t}'\vv{1}=N} h(\vv{r}, \vv{t}, N, n_i) q_{\vv{t}|L}
\end{equation}
are the exchangeable and marginally reproducible outcome probabilities of the 
exchangeable multinomial distribution with maximal cluster size $N$ with 
$h(\vv{r}, \vv{t}, N, n_i)$ defined as multivariate hypergeometric probabilities.

\section{EM algorithm}

We can consider $\xi_{ij}$ missing data. Then the complete data likelihood is
\begin{equation}
L_c(\mu, \sigma, q | \xi_{ij}, y_{ij}) = 
 \sum_{i=1}^\I  \frac{q_{\ff{\vv{\xi}}|n_i}}{\binom{n_i}{\ff{\vv{\xi}}}}  \prod_{j=1}^{n_i} \log \phi(y_{ij}; \mu_{\xi_{ij}}, \sigma_{\xi_{ij}}) =
 \sum_{i=1}^\I \prod_{\vv{z}\in\Z{n_i}} \big[\frac{q_{\ff{\vv{z}}|n_i}}{\binom{n_i}{\ff{\vv{z}}}} 
    \prod_{j=1}^{n_i} \log \phi(y_{ij}; \mu_{z_j}, \sigma_{z_j})\big]^{I(\vv{\xi}_i = \vv{z})}.
\end{equation}

Its logarithm without the binomial term that does not contain unknown parameters is

\begin{equation}
\log L_c(\mu, \sigma, q | \xi_{ij}, y_{ij}) = 
  \sum_{i=1}^\I \sum_{\vv{z}\in\Z{n_i}} I(\vv{\xi}_i = \vv{z}) \big[\log q_{\ff{\vv{z}}|n_i}  +  
        \sum_{j=1}^{n_i} \log \phi(y_{ij}; \mu_{z_j}, \sigma_{z_j})\big]
\end{equation}

The expected complete data log-likelihood, given previous estimates $\mu^{(m)}, \sigma^{(m)}, q^{(m)}$ is
\begin{multline}
Q(\mu, \sigma, q) = E\big[\log L_c(\mu, \sigma, q \mid \mu^{(m)}, \sigma^{(m)}, q^{(m)}, y_{ij})\big] = \\
 \sum_{i=1}^\I \sum_{\vv{z}\in\Z{n_i}}  Pr(\vv{\xi}_{i}=\vv{z} \mid \mu^{(m)}, \sigma^{(m)}, q^{(m)}, y_{ij} ) 
 [\log q_{\ff{\vv{z}}|n_i}  +\sum_{j=1}^{n_i}\log \phi(y_{ij}; \mu_{x_j}, \sigma_{x_j})],
\end{multline}

\subsection{E-step}
Using the Bayes theorem
\begin{align}
 e_{i\vv{z}}^{(m)} = & Pr(\vv{\xi}_{i}=\vv{z} \mid \mu^{(m)}, \sigma^{(m)}, q^{(m)}, y_{ij} )=\\
  & \frac{Pr(\vv{y}_{i} \mid \vv{\xi}_i=\vv{z},  \mu^{(m)}, \sigma^{(m)}, q^{(m)}) Pr(\vv{\xi}_{i}=\vv{z} \mid \mu^{(m)}, \sigma^{(m)}, q^{(m)})}%
  {\sum_{\vv{w}} Pr(\vv{y}_{i} \mid \vv{\xi}_i=\vv{w}, \mu^{(m)}, \sigma^{(m)}, q^{(m)}) Pr(\vv{\xi}_{i}=\vv{w} \mid \mu^{(m)}, \sigma^{(m)}, q^{(m)}) } = 
  \frac{\prod_{j=1}^{n_i} \phi(y_{ij}; \mu_{z_j}^{(m)}, \sigma^{(m)}_{z_j})  q^{(m)}_{\ff{\vv{z}}|n_i}\big/ {\binom{n_i}{\ff{\vv{z}}}} }%
  {\sum_{\vv{w}}\prod_{j=1}^{n_i} \phi(y_{ij}; \mu_{w_j}^{(m)}, \sigma^{(m)}_{w_j})  q^{(m)}_{\ff{\vv{w}}|n_i} \big/{\binom{n_i}{\ff{\vv{w}}}}   }.  
\end{align}

\subsection{M-step}

\begin{align}
Q(\mu, \sigma, q) = & \sum_{i=1}^\I\sum_{\vv{z}\in\Z{n_i}} e_{i\vv{z}}^{(m)}   
 [\log q_{\ff{\vv{z}}|n_i}  +\sum_{j=1}^{n_i}\log \phi(y_{ij}; \mu_{z_j}, \sigma_{z_j})] =\\
 &  \sum_{i=1}^\I\sum_{\vv{z}\in\Z{n_i}} e_{i\vv{z}}^{(m)} \log q_{\ff{\vv{z}}|n_i} +
  \sum_{i=1}^\I\sum_{\vv{z}\in\Z{n_i}} e_{i\vv{z}}^{(m)}  \sum_{j=1}^{n_i}\log \phi(y_{ij}; \mu_{z_j}, \sigma_{z_j})  =\\
 &  \sum_{i=1}^\I\sum_{\vv{z}\in\Z{n_i}} e_{i\vv{z}}^{(m)} \log q_{\ff{\vv{z}}|n_i}+
  \sum_{k=1}^K \sum_{i=1}^\I \sum_{j=1}^{n_i}\sum_{\vv{z}: z_j=k} e_{i\vv{z}}^{(m)}  \log \phi(y_{ij}; \mu_k, \sigma_k),
\end{align}
where the two components can be maximized separately, and within the last component the terms corresponding to different $k$'s
can be maximized separately as well.

\subsubsection{Update for $q$}

In terms of $q$, the log-likelihood can be viewed as the $\EM$ log-likelihood for
an extended data set: for each cluster $i$ and for each possible frequency vector of responses $\vv{r}\in \mathbb{Z}_{n_i}$, $\vv{r}'\vv{1}=n_i$ 
we compute the `observed' frequency as $a_{i\vv{r}}^{(m)}   = \sum_{\vv{z}: \ff{\vv{z}}=\vv{r}}  e_{i\vv{z}}^{(m)}$. Then  
$q_{\vv{t}|N}^{(m+1)}$ can be obtained from the EM algorithm for fitting the $\EM$ model to these frequencies.

\subsubsection{Update for $\mu$ and $\sigma$}
In terms of the normal distribution parameters, we have a weighted normal log-likelihood for each $k$. 
The weight corresponding to observation $j$ in cluster $i$ is $b_{ijk}^{(m)} = \sum_{\vv{z}: z_j=k} e_{i\vv{z}}^{(m)}$.
Then 
\begin{align}
\mu_k^{(m+1)} = & \sum_{i=1}^\I \sum_{j=1}^{n_i} b_{ijk}^{(m)} y_{ij} \\
[\sigma_k^{(m+1)}]^2 = &  \sum_{i=1}^\I \sum_{j=1}^{n_i} b_{ijk}^{(m)} (y_{ij} - \mu_k^{(m+1)})^2
\end{align}

%For now, let's assume $K=2$ -- a two-component mixture.

\end{document}

\section{Initial setup}
\begin{flushleft} \small\label{scrap1}\raggedright\small
\NWtarget{nuweb?}{} \verb@"R/MixMod.R"@\nobreak\ {\footnotesize {?}}$\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@require(CorrBin)@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}

### 2019-11-27 nuweb/MixMod_develop.R

@@ -1,0 +1,12 @@
+library(devtools)
+source('z:/RForge/Nuweb.R')
+#source('/home/aszabo/RForge/Nuweb_linux.R')
+
+run_nuweb(file="Clustering.w", path=getwd())
+
+source('R/MixMod.R', echo=FALSE)
+
+
+### initial setup
+create_package("z:/exchreg")
+use_package("CorrBin")

### 2021-03-03 R/wgldrm.R
Fix minor bugs & typos in gldrm implementation; implement simple consistency check

@@ -1,32 +1,9 @@
-\documentclass[reqno]{amsart}
-\usepackage[margin=1in]{geometry}
-\usepackage[colorlinks=true,linkcolor=blue]{hyperref}
-\renewcommand{\NWtarget}[2]{\hypertarget{#1}{#2}}
-\renewcommand{\NWlink}[2]{\hyperlink{#1}{#2}}
-\newcommand{\bv}{\mathbf{v}}
-\newcommand{\bq}{\mathbf{q}}
-\newcommand{\bpi}{\text{\boldmath $\pi$}}
-\newcommand{\leqst}{\mathrel{\preceq^{st}}}
-\newcommand{\geqst}{\mathrel{\succeq^{st}}}
 
-\title{Generalized linear density ratio model (GLDRM) with frequency weights}
-\author{Aniko Szabo}
-\date{\today}
-
-
-The code is based on the \texttt{gldrm} package, with minor changes to incorporate frequency weights. The code is simplified a bit as well by removing some options:
-\begin{itemize}
-  \item sampling weights are removed
-  \item there is no variance/standard error calculation, and no inference against the null
-\end{itemize}
-
-
-@O R/wgldrm.R @{
 #' Main optimization function
 #'
 #' This function is called by the main \code{gldrm} function.
 #'
-#' @@keywords internal
+#' @keywords internal
 gldrm.control <- function(eps=1e-10, maxiter=100, returnfTiltMatrix=TRUE,
                           returnf0ScoreInfo=FALSE, print=FALSE,
                           betaStart=NULL, f0Start=NULL)
@@ -204,33 +181,30 @@ gldrmFit <- function(x, y, linkfun, linkinv, mu.eta, mu0=NULL, offset=NULL, weig
     class(fit) <- "gldrm"
     fit
 }
-@}
 
-
-@O R/wgldrm.R @{
 #' Beta optimization routing
 #'
-#' @@param x Covariate matrix.
-#' @@param y Response vector.
-#' @@param spt Vector of unique observed support points in the response.
-#' @@param ySptIndex Index of each \code{y} value within the \code{spt} vector.
-#' @@param f0 Current values of f0.
-#' @@param linkinv Inverse link function.
-#' @@param mu.eta Deriviative of inverse link function.
-#' @@param offset Vector of known offset values to be added to the linear
+#' @param x Covariate matrix.
+#' @param y Response vector.
+#' @param spt Vector of unique observed support points in the response.
[…]

### 2021-03-03 SPreg_develop.R

@@ -3,10 +3,88 @@ source('z:/RForge/Nuweb.R')
 #source('/home/aszabo/RForge/Nuweb_linux.R')
 
 run_nuweb(file="SPregress.w", path=getwd())
+run_nuweb(file="SPGLM.w", path=getwd())
+run_nuweb(file="wgldrm.w", path=getwd())
 
 source('R/SPreg.R', echo=FALSE)
+source('R/SPGLM.R', echo=FALSE)
+source('R/wgldrm.R', echo=FALSE)
 
 data(shelltox)
+######### Testing wGLDRM
+run_gldrm <- 
+function (formula, data = NULL, link = "identity", mu0 = NULL,  weights=NULL,
+          offset = NULL, gldrmControl = gldrm.control(), thetaControl = theta.control(), 
+          betaControl = beta.control(), f0Control = f0.control()) {
+  if (is.null(weights)) weights <- rep(1, nrow(data))
+  mf <- model.frame(formula, data)
+  x <- stats::model.matrix(attr(mf, "terms"), mf)
+  attributes(x)[c("assign", "contrasts")] <- NULL
+  y <- stats::model.response(mf, type = "numeric")
+  if (is.null(offset)) 
+    offset <- rep(0, nrow(x))
+  if (length(offset) != nrow(x)) 
+    stop("offset should be NULL or a vector with length equal to the number of observations.")
+  if (is.character(link)) {
+    link <- stats::make.link(link)
+  }
+  else if (!is.list(link) || !(all(c("linkfun", "linkinv", 
+                                     "mu.eta") %in% names(link)))) {
+    stop(paste0("link should be a character string or a list containing ", 
+                "functions named linkfun, linkinv, and mu.eta"))
+  }
+  yMin <- min(y)
+  yMax <- max(y)
+  yMed <- (yMin + yMax)/2
+  Z <- function(y) (y - yMed) * 2/(yMax - yMin)
+  Y <- function(z) z * (yMax - yMin)/2 + yMed
+  z <- Z(y)
+  linkfunZ <- function(muZ) link$linkfun(Y(muZ))
+  linkinvZ <- function(eta) Z(link$linkinv(eta))
+  mu.etaZ <- function(eta) 2/(yMax - yMin) * link$mu.eta(eta)
+  if (is.null(mu0)) {
[…]

### 2021-03-03 R/SPGLM.R
Add description of SPGLM fitting approach

@@ -1,0 +1,92 @@
+
+
+#'@rdname spglm
+#'@param formula a one-sided formula of the form \code{cbind(r, s) ~ predictors} where \code{r} and \code{s} give the number of responses and non-responses within each cluster, respectively (so the cluster size is \code{r+s}), and \code{predictors} describes the covariates.
+#'@param data  an optional matrix or data frame containing the variables in the formula \code{formula}. By default the variables are taken from \code{environment(formula).}
+#'@param subset an optional vector specifying a subset of observations to be used.
+#'@param weight an optional vector specifying observation weights.
+#'@param link     a link function for the mean.
+#'@param start    an optional list with elements named \code{beta}, \code{q}, and/or \code{mu1} giving starting values for estimation. If none or only some are specified, starting values are selected automatically.
+#'@param control        a list with parameters controlling the algorithm.
+#'@return an object of class \code{spglm} with the fitted model.
+#'@export
+#' @importFrom stats terms model.matrix
+
+#' Semi-parametric generalized linear model
+spglm <- function(formula, data, subset, weights, link="logit", start=NULL, control=list(eps=0.001, maxit=100), ...){
+    fam <- binomial(link=link)
+    
+    
+       if (missing(formula) || (length(formula) != 3L))
+            stop("'formula' missing or incorrect")
+       if (missing(data))
+            data <- environment(formula)
+        mc <- match.call(expand.dots = FALSE)
+        m <- match(c("formula", "data", "subset", "weights"), names(mc), 0L)
+        m <- mc[c(1L,m)]
+        if (is.matrix(eval(m$data, parent.frame())))
+            m$data <- as.data.frame(data)
+        m[[1L]] <- quote(stats::model.frame)
+        mf <- eval(m, parent.frame())
+
+        # extract and check response variable
+        Y <- model.response(mf)
+        if (ncol(Y) != 2) stop("The response variable should be a 2-column matrix")
+
+        # create model matrix
+        mm <- model.matrix(formula, data=mf)
+
+        # extract weights
+        weights <- as.vector(model.weights(mf))
+        if (!is.null(weights) && !is.numeric(weights))
+            stop("'weights' must be a numeric vector")
+        if (!is.null(weights) && any(weights < 0))
+            stop("negative weights not allowed")
+    
+    
+     while (iter<itermax & difference>tol) {
+        iter <- iter + 1
[…]

### 2021-03-03 SPreg_develop.R

@@ -2,6 +2,7 @@ library(devtools)
 source('z:/RForge/Nuweb.R')
 #source('/home/aszabo/RForge/Nuweb_linux.R')
 
+
 run_nuweb(file="SPregress.w", path=getwd())
 run_nuweb(file="SPGLM.w", path=getwd())
 run_nuweb(file="wgldrm.w", path=getwd())
@@ -9,6 +10,8 @@ run_nuweb(file="wgldrm.w", path=getwd())
 source('R/SPreg.R', echo=FALSE)
 source('R/SPGLM.R', echo=FALSE)
 source('R/wgldrm.R', echo=FALSE)
+
+shell("cd c:/exchreg/ && texify --pdf --quiet --run-viewer --clean SPGLM.tex")
 
 data(shelltox)
 ######### Testing wGLDRM
@@ -149,14 +152,14 @@ matplot(t(l), type="l")
 
 
 # create native routine registration file
-tools::package_native_routine_registration_skeleton("z://RForge/corrbin/pkg/CorrBin",
-                                                    con="z://RForge/corrbin/pkg/CorrBin/src/init.c")
+#tools::package_native_routine_registration_skeleton("z://RForge/corrbin/pkg/CorrBin",
+#                                                    con="z://RForge/corrbin/pkg/CorrBin/src/init.c")
 
-document(cb)
-run_examples(cb)  # or dev_example("ran.CMData")
-load_all(cb)
+#document(cb)
+#run_examples(cb)  # or dev_example("ran.CMData")
+#load_all(cb)
 
-build(cb)
-check(cb,  check_dir = "z:/RForge", check_version = TRUE, cran = TRUE)
+#build(cb)
+#check(cb,  check_dir = "z:/RForge", check_version = TRUE, cran = TRUE)
 
 

### 2021-03-04 nuweb/SPGLM.tex
Set up package infrastructure

\newcommand{\NWtarget}[2]{#2}
\newcommand{\NWlink}[2]{#2}
\newcommand{\NWtxtMacroDefBy}{Fragment defined by}
\newcommand{\NWtxtMacroRefIn}{Fragment referenced in}
\newcommand{\NWtxtMacroNoRef}{Fragment never referenced}
\newcommand{\NWtxtDefBy}{Defined by}
\newcommand{\NWtxtRefIn}{Referenced in}
\newcommand{\NWtxtNoRef}{Not referenced}
\newcommand{\NWtxtFileDefBy}{File defined by}
\newcommand{\NWtxtIdentsUsed}{Uses:}
\newcommand{\NWtxtIdentsNotUsed}{Never used}
\newcommand{\NWtxtIdentsDefed}{Defines:}
\newcommand{\NWsep}{${\diamond}$}
\newcommand{\NWnotglobal}{(not defined globally)}
\newcommand{\NWuseHyperlinks}{}
\documentclass[reqno]{amsart}
\usepackage[margin=1in]{geometry}
\usepackage[colorlinks=true,linkcolor=blue]{hyperref}
\renewcommand{\NWtarget}[2]{\hypertarget{#1}{#2}}
\renewcommand{\NWlink}[2]{\hyperlink{#1}{#2}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bq}{\mathbf{q}}
\newcommand{\bpi}{\text{\boldmath $\pi$}}
\newcommand{\leqst}{\mathrel{\preceq^{st}}}
\newcommand{\geqst}{\mathrel{\succeq^{st}}}
\newcommand{\qz}{\boldsymbol{q}^{(0)}}
\DeclareMathOperator{\E}{\mathbb{E}}

\usepackage{dsfont}

\title{Semi-parametric generalized linear model for correlated binary data}
\author{Aniko Szabo}
\date{\today}
\begin{document}
\maketitle

Note: this setup assumes that parts defined in \texttt{SPregress.w} are available.

\section{Introduction}

We extend the Rathouz-Gao semi-parametric generalized linear model to clustered binary outcomes with varying cluster sizes. Treating the number of events $Y_i$ from a cluster scaled by corresponding cluster size $n_i$ as the response variable, we use a parametric regression model for the marginal response probability $Y_/n_i$ and assume that for the maximum cluster size $N$ the probability density distribution of $Y_i$ is an exponentially tilted version of the reference density distribution.

The conditional mean model is as follows:
\begin{equation}\label{M:conditionalMeanSPGLM}
     E(\dfrac{Y_i}{n_i} | \boldsymbol{Z}_i; \boldsymbol{\beta})=\mu(\boldsymbol{Z}_i,\boldsymbol{\beta}) \equiv \mu_i = h^{-1}( \boldsymbol{Z}_i^{T} \boldsymbol{\beta})
\end{equation}
where $\boldsymbol{Z}_i$ is a matrix of cluster-level covariates, the function $h(\cdot)$ is a known strictly increasing link function.

The assumed probability density function for a cluster of maximal size $N$ is:
\begin{equation}\label{M:densitySPGLM}
    q_{y,N} (\boldsymbol{Z}, \boldsymbol{\beta}) \propto q_{y,N}^{(0)} \times \exp [ - \omega(\boldsymbol{Z}, \boldsymbol{\beta}) \times y]
\end{equation}
where $q_{y,N}^{(0)}$ is a non-parametric reference distribution, and $\omega(\boldsymbol{Z}, \boldsymbol{\beta})$ is the tilting parameter chosen so that the expectation of $Y/N$ equals to the value defined by the mean model \eqref{M:conditionalMeanSPGLM}, i.e.\  $\sum_{y=0}^{N} \dfrac{y}{N} q_{y,N}(\boldsymbol{Z}, \boldsymbol{\beta}) = \mu = h^{-1}( \boldsymbol{Z}^{T} \boldsymbol{\beta}) $. 
%We define the tilting parameter for the reference density to be $0$, i.e., $\omega_0 \equiv 0$.

For the purpose of identifiability, the mean of marginal probability $\{ q_{y,N}^{(0)} \}$ is set at an arbitrary fixed value, which can be the mean of the reference group, or the overall marginal mean of the data. We extend the conditional mean and probability density model for  cluster sizes smaller than the maximal size $N$ by assuming marginal compatibility.

\section{Main function}

\begin{flushleft} \small\label{scrap1}\raggedright\small
\NWtarget{nuweb1}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {1}}$\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@#' Fit semi-parametric GLM@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#'@{\tt @}\verb@rdname spglm@\\
\mbox{}\verb@#'@{\tt @}\verb@param formula a one-sided formula of the form \code{cbind(r, s) ~ predictors} where \code{r} and \code{s} give the number of responses and non-responses within each cluster, respectively (so the cluster size is \code{r+s}), and \code{predictors} describes the covariates.@\\
\mbox{}\verb@#'@{\tt @}\verb@param data  an optional matrix or data frame containing the variables in the formula \code{formula}. By default the variables are taken from \code{environment(formula).}@\\
\mbox{}\verb@#'@{\tt @}\verb@param subset  an optional vector specifying a subset of observations to be used.@\\
\mbox{}\verb@#'@{\tt @}\verb@param weight  an optional vector specifying observation weights.@\\
\mbox{}\verb@#'@{\tt @}\verb@param link      a link function for the mean.@\\
\mbox{}\verb@#'@{\tt @}\verb@param start     an optional list with elements named \code{beta}, \code{q}, and/or \code{mu1} giving starting values for estimation. If none or only some are specified, starting values are selected automatically.@\\
\mbox{}\verb@#'@{\tt @}\verb@param control a list with parameters controlling the algorithm.@\\
\mbox{}\verb@#'@{\tt @}\verb@return an object of class \code{spglm} with the fitted model.@\\
\mbox{}\verb@#'@{\tt @}\verb@export@\\
\mbox{}\verb@#' @{\tt @}\verb@importFrom stats terms model.matrix@\\
\mbox{}\verb@@\\
\mbox{}\verb@spglm <- function(formula, data, subset, weights, link="logit", start=NULL, control=list(eps=0.001, maxit=100), ...){@\\
\mbox{}\verb@    fam <- binomial(link=link)@\\
\mbox{}\verb@    @\\
\mbox{}\verb@    @\hbox{$\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize \NWlink{nuweb2}{2}}$\,\rangle$}\verb@@\\
\mbox{}\verb@    @\hbox{$\langle\,${\itshape Fit model}\nobreak\ {\footnotesize \NWlink{nuweb4}{4}}$\,\rangle$}\verb@@\\
\mbox{}\verb@@\\
\mbox{}\verb@    mt <- attr(mf, "terms")@\\
\mbox{}\verb@    res <- list(coefficients = beta_new, q=q_new, niter = iter, loglik=logl,@\\
\mbox{}\verb@                link = link, call = mc, terms = mt,@\\
\mbox{}\verb@                xlevels = .getXlevels(mt, mf),@\\
\mbox{}\verb@                data_object=list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights),@\\
\mbox{}\verb@                model_fun=model_fun)@\\
\mbox{}\verb@    class(res) <- "spglm"@\\
\mbox{}\verb@    res@\\
\mbox{}\verb@@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\begin{flushleft} \small\label{scrap2}\raggedright\small
\NWtarget{nuweb2}{} $\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize {2}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@   if (missing(formula) || (length(formula) != 3L))@\\
\mbox{}\verb@        stop("'formula' missing or incorrect")@\\
\mbox{}\verb@   if (missing(data))@\\
\mbox{}\verb@        data <- environment(formula)@\\
\mbox{}\verb@    mc <- match.call(expand.dots = FALSE)@\\
\mbox{}\verb@    m <- match(c("formula", "data", "subset", "weights"), names(mc), 0L)@\\
\mbox{}\verb@    m <- mc[c(1L,m)]@\\
\mbox{}\verb@    if (is.matrix(eval(m$data, parent.frame())))@\\
\mbox{}\verb@        m$data <- as.data.frame(data)@\\
\mbox{}\verb@    m[[1L]] <- quote(stats::model.frame)@\\
\mbox{}\verb@    mf <- eval(m, parent.frame())@\\
\mbox{}\verb@@\\
\mbox{}\verb@    # extract and check response variable@\\
\mbox{}\verb@    Y <- model.response(mf)@\\
\mbox{}\verb@    if (ncol(Y) != 2) stop("The response variable should be a 2-column matrix")@\\
\mbox{}\verb@@\\
\mbox{}\verb@    # create model matrix@\\
\mbox{}\verb@    mm <- model.matrix(formula, data=mf)@\\
\mbox{}\verb@@\\
\mbox{}\verb@    # extract weights@\\
\mbox{}\verb@    weights <- as.vector(model.weights(mf))@\\
\mbox{}\verb@    if (!is.null(weights) && !is.numeric(weights))@\\
\mbox{}\verb@        stop("'weights' must be a numeric vector")@\\
\mbox{}\verb@    if (!is.null(weights) && any(weights < 0))@\\
\mbox{}\verb@        stop("negative weights not allowed")@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb1}{1}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\section{EM algorithm for model fitting}
We will combine an Expectation Maximization with the Newton-Raphson algorithm proposed by Rathouz and Gao (2009) to estimate the model parameters. Recall that the observed data consists of the number of events, cluster size, and cluster-level covariates: $(y_i, N_i, \boldsymbol{Z}_i)$, $i=1,\cdots,I$.

\subsubsection{Observed data log-likelihood}
Under the marginal compatibility assumption, the observed log-likelihood function is:
\begin{equation}\label{E:loglikelihood}
\begin{split}
    \ell (q_{y,N}, \boldsymbol{\beta}) &= \sum_{i=1}^I \log \{ q_{y_i,N_i} (\boldsymbol{Z}_i, \boldsymbol{\beta}) \} \\
    &= \sum_{i=1}^I \log \bigg[ \binom{N_i}{y_i} \sum_{t=y}^{N-N_i+y_i} \dfrac{\binom{N-N_i}{t-y_i}}{\binom{N}{Y_i}} \times \dfrac{q_{t,N}^{(0)} \times \exp \{ \omega(\boldsymbol{Z}_i; \boldsymbol{\beta}) t \}} {\sum_{t'=0}^{N} q_{t',N}^{(0)} \times \exp \{ \omega(\boldsymbol{Z}_i; \boldsymbol{\beta}) t' \}}  \bigg].
\end{split}
\end{equation}

It is difficult to utilize the observed log-likelihood directly for parameter estimation. Its score function is complex because there is a logarithm of a sum in \eqref{E:loglikelihood}. Therefore, we implement an EM algorithm for parameter estimation.

\subsubsection{Missing data setup}
Stefanescu et al (2003) have proved that the marginal compatibility assumption is equivalent to assuming that each cluster of size $N'$ arises from a cluster of maximal cluster size $N$ with some binary observations missing completely missing at random (MCAR). The missing data setup and the expectation step from EM algorithm are based on this MCAR interpretation.

For the setup of the EM algorithm, the complete data are clusters of size $N$ with a corresponding number of events $s_i$, where $s_i \in \{y_i, \cdots, N \}$. The missing data correspond to the unobserved outcomes in of the $(N - N_i)$ cluster components. The missing data can be summarized as $(s_i - y_i)$ events out of $(N - N_i)$ elements.

The complete data log-likelihood is:
\begin{equation}\label{E:logLikelihoodCompleteMapFromObs}
\begin{split}
    \ell_{\text{complete}} (q_{y,N}, \boldsymbol{\beta} \mid N_i,s_i) &= \sum_{i=0}^I \log \{ q_{s_i,N}(\boldsymbol{Z}_i; \boldsymbol{\beta}) \} \\
    &= \sum_{i=0}^I \sum_{s=y_i}^N {\mathds{1}}(s_i=s) \times \log \{ q_{s,N}(\boldsymbol{Z}_i; \boldsymbol{\beta}) \}, 
\end{split}
\end{equation}
where $q_{s_i,N}$ is the probability of achieving $s_i$ events in a cluster of size $N$ in the complete data set. The $\mathds{1}$ denotes the indicator function.

\subsubsection{Expectation step}
%For complete data in which all clusters have $N$ observations, the marginal probability of achieving $y$ events out of $N$ observations within a cluster is denoted with parameter $q_{y,N}$. 
We take the expectation of complete data log-likelihood \eqref{E:logLikelihoodCompleteMapFromObs} given parameters from the previous $k^{th}$ iteration, which is shown as follows:
\begin{equation}\label{E:ExpectedlogLikelihoodCompleteMapFromObs}
\begin{split}
    Q \bigg( q_{y,N}^{(k)}, \boldsymbol{\beta}^{(k)} \bigg) &= 
     \E \bigg\{ \ell_{\text{complete}} (q_{y,N}, \boldsymbol{\beta} \mid N_i, s_i) \mid q_{y,N}^{(k)}, \boldsymbol{\beta}^{(k)}, N_i, y_i \bigg\} \\
    &= \E \Bigg[ \sum_{i=0}^I \sum_{s=0}^N {\mathds{1}}(s=s_i) \times \log \{ q_{s,N}(\boldsymbol{Z}_i; \boldsymbol{\beta}) \} \mid q_{y,N}^{(k)}, y_i,N_i \Bigg] \\
    &= \sum_{i=0}^I \sum_{s=0}^N \underbrace{ \E \big\{ {\mathds{1}}(s=s_i) \mid q_{y,N}^{(k)}, \boldsymbol{\beta}^{(k)}, y_i, N_i \big\}}_{p_{N_iy_is_i}^{(k)}} \times \log \{q_{s,N}(\boldsymbol{Z}_i; \boldsymbol{\beta}) \},
\end{split}
\end{equation}
where $p_{N'y'y} = \textbf{Pr}_N(Y=y \mid Y'=y',N')$ denotes the conditional probability of achieving $y$ events in the complete cluster of size $N$, given that $y'$ events have been observed in the original cluster of size $N'$. 
The expression for $p_{N'y'y}$ can be derived from $q_{y,N}$ using Bayes theorem and the MCAR interpretation as follows:
\begin{equation}\label{D:prst}
\begin{split}
p_{N'y'y} &= \textbf{Pr}_N(Y=y \mid Y'=y',N') \\ 
          &= \frac{\textbf{Pr}_{N'}(Y'=y' \mid Y=y) \textbf{Pr}_N(Y=y)}{\sum_{t=0}^{N} \textbf{Pr}_{y'}(Y'=y' \mid Y=t) \textbf{Pr}_N(Y=t)} \\ 
          &= \frac{\binom{y}{y'} \binom{N-y}{N'-y'} q_{y,N}}{\sum_{t=y'}^{N-N'+y'} \dbinom{t}{y'} \binom{N-t}{N'-y'} q_{t,N}},
\end{split}
\end{equation}
and $p_{N_iy_is_i}^{(k)}$ at the $k$th step of the iteration can be obtained by using $q_{y,N}^{(k)}$ on the right-hand side of the expression.

\subsubsection{Maximization step}

Equation \eqref{E:ExpectedlogLikelihoodCompleteMapFromObs} can be interpreted as the log-likelihood for the SPGLM with fixed cluster size $N$ for an expanded data set: each observation with cluster size $N_i$ and number of responses $y_i$ is expanded to $(N+1)$ replicates with $0, 1, \ldots, N$, responses, respectively. The probabilities $p_{N_iy_is_i}^{(k)}$ serve as cluster weights in the expanded data set. From here we can use a slight generalization of the Newton-Raphson algorithm developed by Wurm and Rathouz (2018) to estimate the baseline density distribution $ \boldsymbol{q}_N^{(0)} $ and the regression coefficients $\boldsymbol{\beta}$.

\begin{flushleft} \small\label{scrap3}\raggedright\small
\NWtarget{nuweb4}{} $\langle\,${\itshape Fit model}\nobreak\ {\footnotesize {4}}$\,\rangle\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@ while (iter < control$maxit & difference > control$eps) {@\\
\mbox{}\verb@    iter <- iter + 1@\\
\mbox{}\verb@    referencef0Pre <- referencef0@\\
\mbox{}\verb@    tiltingWsPre <- tiltingWs@\\
\mbox{}\verb@    betasPre <- betas@\\
\mbox{}\verb@    # passing parameter updates to the next iteration@\\
\mbox{}\verb@    if (iter == 1) {@\\
\mbox{}\verb@      startUpdates <- list(iter = iter,@\\
\mbox{}\verb@                           f0StartValue = NULL,@\\
\mbox{}\verb@                           thetaStartValue = NULL,@\\
\mbox{}\verb@                           betaStartValue = NULL)@\\
\mbox{}\verb@    } else {@\\
\mbox{}\verb@      startUpdates <- list(iter = iter,@\\
\mbox{}\verb@                           f0StartValue = referencef0Pre,@\\
\mbox{}\verb@                           thetaStartValue = tiltingWsPre,@\\
\mbox{}\verb@                           betaStartValue = betasPre)@\\
\mbox{}\verb@    }@\\
\mbox{}\verb@    # EM NR algorithm@\\
\mbox{}\verb@    res <- gldrMaximize(DesignMatrix.CONST, CBData, referencef0Pre, tiltingWsPre, startUpdates)@\\
\mbox{}\verb@    referencef0 <- res$density.ref @\\
\mbox{}\verb@    # notice that this tilting parameter estimates are w.r.t the intermediate dataset, original dataset extended by max(cluster size)+1 times@\\
\mbox{}\verb@    tiltingWs <- res$tiltParam[(0:(nrow(CBData)-1))*length(referencef0)+1]@\\
\mbox{}\verb@    betas <- res$regressionEst@\\
\mbox{}\verb@    # monitor parameter estimates for baseline distribution, tilting parameters, regression coefficients@\\
\mbox{}\verb@    difference <- sum(abs(referencef0Pre - referencef0)) + sum(abs(betasPre - betas))@\\
\mbox{}\verb@  }@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  list(referencef0 = referencef0,@\\
\mbox{}\verb@       tiltingWs = tiltingWs,@\\
\mbox{}\verb@       betas = betas,@\\
\mbox{}\verb@       EMgldrmFitRes = res)@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtMacroRefIn\ \NWlink{nuweb1}{1}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\end{document}

### 2021-03-04 nuweb/wgldrm.tex
Set up package infrastructure

\newcommand{\NWtarget}[2]{#2}
\newcommand{\NWlink}[2]{#2}
\newcommand{\NWtxtMacroDefBy}{Fragment defined by}
\newcommand{\NWtxtMacroRefIn}{Fragment referenced in}
\newcommand{\NWtxtMacroNoRef}{Fragment never referenced}
\newcommand{\NWtxtDefBy}{Defined by}
\newcommand{\NWtxtRefIn}{Referenced in}
\newcommand{\NWtxtNoRef}{Not referenced}
\newcommand{\NWtxtFileDefBy}{File defined by}
\newcommand{\NWtxtIdentsUsed}{Uses:}
\newcommand{\NWtxtIdentsNotUsed}{Never used}
\newcommand{\NWtxtIdentsDefed}{Defines:}
\newcommand{\NWsep}{${\diamond}$}
\newcommand{\NWnotglobal}{(not defined globally)}
\newcommand{\NWuseHyperlinks}{}
\documentclass[reqno]{amsart}
\usepackage[margin=1in]{geometry}
\usepackage[colorlinks=true,linkcolor=blue]{hyperref}
\renewcommand{\NWtarget}[2]{\hypertarget{#1}{#2}}
\renewcommand{\NWlink}[2]{\hyperlink{#1}{#2}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bq}{\mathbf{q}}
\newcommand{\bpi}{\text{\boldmath $\pi$}}
\newcommand{\leqst}{\mathrel{\preceq^{st}}}
\newcommand{\geqst}{\mathrel{\succeq^{st}}}

\title{Generalized linear density ratio model (GLDRM) with frequency weights}
\author{Aniko Szabo}
\date{\today}

\begin{document}
\maketitle

The code is based on the \texttt{gldrm} package, with minor changes to incorporate frequency weights. The code is simplified a bit as well by removing some options:
\begin{itemize}
  \item sampling weights are removed
  \item there is no variance/standard error calculation, and no inference against the null
\end{itemize}

\begin{flushleft} \small\label{scrap1}\raggedright\small
\NWtarget{nuweb1}{} \verb@"../R/wgldrm.R"@\nobreak\ {\footnotesize {1}}$\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@#' Main optimization function@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#' This function is called by the main \code{gldrm} function.@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#' @{\tt @}\verb@keywords internal@\\
\mbox{}\verb@gldrm.control <- function(eps=1e-10, maxiter=100, returnfTiltMatrix=TRUE,@\\
\mbox{}\verb@                          returnf0ScoreInfo=FALSE, print=FALSE,@\\
\mbox{}\verb@                          betaStart=NULL, f0Start=NULL)@\\
\mbox{}\verb@{@\\
\mbox{}\verb@    gldrmControl <- as.list(environment())@\\
\mbox{}\verb@    class(gldrmControl) <- "gldrmControl"@\\
\mbox{}\verb@    gldrmControl@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@\\
\mbox{}\verb@gldrmFit <- function(x, y, linkfun, linkinv, mu.eta, mu0=NULL, offset=NULL, weights=NULL,@\\
\mbox{}\verb@                     gldrmControl=gldrm.control(), thetaControl=theta.control(),@\\
\mbox{}\verb@                     betaControl=beta.control(), f0Control=f0.control())@\\
\mbox{}\verb@{@\\
\mbox{}\verb@    ## Extract control arguments@\\
\mbox{}\verb@    if (class(gldrmControl) != "gldrmControl")@\\
\mbox{}\verb@        stop("gldrmControl must be an object of class \'gldrmControl\' returned by@\\
\mbox{}\verb@              gldrmControl() function.")@\\
\mbox{}\verb@    eps <- gldrmControl$eps@\\
\mbox{}\verb@    maxiter <- gldrmControl$maxiter@\\
\mbox{}\verb@    returnHess <- gldrmControl$returnHess@\\
\mbox{}\verb@    returnfTiltMatrix <- gldrmControl$returnfTiltMatrix@\\
\mbox{}\verb@    returnf0ScoreInfo <- gldrmControl$returnf0ScoreInfo@\\
\mbox{}\verb@    print <- gldrmControl$print@\\
\mbox{}\verb@    betaStart <- gldrmControl$betaStart@\\
\mbox{}\verb@    f0Start <- gldrmControl$f0Start@\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Tabulation and summary of responses used in estimating f0@\\
\mbox{}\verb@    n <- length(y)@\\
\mbox{}\verb@    spt <- sort(unique(y))  # observed support@\\
\mbox{}\verb@    ySptIndex <- match(y, spt)  # index of each y value within support@\\
\mbox{}\verb@    # sptFreq <- table(ySptIndex)@\\
\mbox{}\verb@    # attributes(sptFreq) <- NULL@\\
\mbox{}\verb@@\\
\mbox{}\verb@  # create a weight matrix for score.logT1 calculation@\\
\mbox{}\verb@  weightsMatrix <- matrix(0, nrow=length(y), ncol=length(unique(y)), @\\
\mbox{}\verb@        dimnames=list(paste0("obs", 1:length(y)), paste0("response", sort(unique(y)))))@\\
\mbox{}\verb@  # fulfillment of weightMatrix@\\
\mbox{}\verb@  for (obs in 1:length(y)) {@\\
\mbox{}\verb@    weightsMatrix[obs, y[obs]==sort(unique(y))] <- weights[obs]@\\
\mbox{}\verb@  }@\\
\mbox{}\verb@  # weighted version of "sptFreq" @\\
\mbox{}\verb@  sptFreq.weighted <- colSums(weightsMatrix)@\\
\mbox{}\verb@@\\
\mbox{}\verb@@\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Initialize offset@\\
\mbox{}\verb@    if (is.null(offset))@\\
\mbox{}\verb@        offset <- rep(0, n)@\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Initialize mu0 if not provided by user@\\
\mbox{}\verb@    if (is.null(mu0)) {@\\
\mbox{}\verb@       mu0 <- weighted.mean(y, weights)@\\
\mbox{}\verb@    } else if (mu0<=min(spt) || mu0>=max(spt)) {@\\
\mbox{}\verb@        stop(paste0("mu0 must lie within the range of observed values. Choose a different ",@\\
\mbox{}\verb@                    "value or set mu0=NULL to use the default value, weighted.mean(y,weights)."))@\\
\mbox{}\verb@    }@\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Initialize f0@\\
\mbox{}\verb@    if (is.null(f0Start)) {@\\
\mbox{}\verb@      # weighted version of initial value for baseline distribution@\\
\mbox{}\verb@        f0 <- sptFreq.weighted / sum(sptFreq.weighted)@\\
\mbox{}\verb@        if (mu0 != weighted.mean(y,weights))@\\
\mbox{}\verb@            f0 <- getTheta(spt=spt, f0=f0, mu=mu0, weights=weights, ySptIndex=1, thetaStart=0,@\\
\mbox{}\verb@                           thetaControl=thetaControl)$fTilt[, 1]@\\
\mbox{}\verb@    } else {@\\
\mbox{}\verb@        if (length(f0Start) != length(spt))@\\
\mbox{}\verb@            stop("Length of f0Start should equal number of unique values in the response.")@\\
\mbox{}\verb@        if (any(f0Start <= 0))@\\
\mbox{}\verb@            stop("All values in f0Start should be strictly positive.")@\\
\mbox{}\verb@        f0 <- f0Start / sum(f0Start)@\\
\mbox{}\verb@        f0 <- getTheta(spt=spt, f0=f0, mu=mu0, weights=weights, ySptIndex=1, thetaStart=0,@\\
\mbox{}\verb@                       thetaControl=thetaControl)$fTilt[, 1]@\\
\mbox{}\verb@    }@\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Initialize beta@\\
\mbox{}\verb@    ## The starting values returned by lm.fit guarantee that all mu values are@\\
\mbox{}\verb@    ## within the support range, even if there is no intercept.@\\
\mbox{}\verb@    ## Offset could still create problems.@\\
\mbox{}\verb@    lmcoef <- stats::lm.wfit(x=x, y=linkfun(mu0) - offset, weights)$coef@\\
\mbox{}\verb@    if (is.null(betaStart)) {@\\
\mbox{}\verb@        beta <- lmcoef@\\
\mbox{}\verb@    } else {@\\
\mbox{}\verb@        if (length(betaStart) != ncol(x))@\\
\mbox{}\verb@            stop("Length of betaStart should equal the number of columns in the model matrix.")@\\
\mbox{}\verb@      beta <- betaStart@\\
\mbox{}\verb@    }@\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Drop coefficients if x is not full rank (add NA values back at the end)@\\
\mbox{}\verb@    naID <- is.na(lmcoef)@\\
\mbox{}\verb@    beta <- beta[!naID]@\\
\mbox{}\verb@    x <- x[, !naID, drop=FALSE]@\\
\mbox{}\verb@    eta <- c(x %*% beta + offset)@\\
\mbox{}\verb@    mu <- linkinv(eta)@\\
\mbox{}\verb@    if (ncol(x) >= n)@\\
\mbox{}\verb@    stop("gldrm requires n > p.")@\\
\mbox{}\verb@    if (any(mu<min(spt) | mu>max(spt)))@\\
\mbox{}\verb@    stop("Unable to find beta starting values that do not violate convex hull condition.")@\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Get initial theta and log likelihood@\\
\mbox{}\verb@    th <- getTheta(spt=spt, f0=f0, mu=mu, weights=weights, ySptIndex=ySptIndex,@\\
\mbox{}\verb@                   thetaStart=NULL, thetaControl=thetaControl)@\\
\mbox{}\verb@    llik <- th$llik@\\
\mbox{}\verb@@\\
\mbox{}\verb@    conv <- FALSE@\\
\mbox{}\verb@    iter <- 0@\\
\mbox{}\verb@    while (!conv && iter <= maxiter)@\\
\mbox{}\verb@    {@\\
\mbox{}\verb@        iter <- iter+1@\\
\mbox{}\verb@        betaold <- beta@\\
\mbox{}\verb@        f0old <- f0@\\
\mbox{}\verb@        llikold <- llik@\\
\mbox{}\verb@@\\
\mbox{}\verb@        ## update beta (mu) and theta, with fixed f0:@\\
\mbox{}\verb@        bb <- getBeta(x=x, y=y, spt=spt, ySptIndex=ySptIndex, f0=f0,@\\
\mbox{}\verb@                      linkinv=linkinv, mu.eta=mu.eta, offset=offset, weights=weights,@\\
\mbox{}\verb@                      betaStart=beta, thStart=th,@\\
\mbox{}\verb@                      thetaControl=thetaControl, betaControl=betaControl)@\\
\mbox{}\verb@        th <- bb$th@\\
\mbox{}\verb@        llik <- bb$llik@\\
\mbox{}\verb@        mu <- bb$mu@\\
\mbox{}\verb@        beta <- bb$beta@\\
\mbox{}\verb@@\\
\mbox{}\verb@        ## update f0 and theta, with fixed beta (mu)@\\
\mbox{}\verb@        ff <- getf0(y=y, spt=spt, ySptIndex=ySptIndex, sptFreq=sptFreq,@\\
\mbox{}\verb@                    weights=weights, sptFreq.weighted=sptFreq.weighted, mu=mu, mu0=mu0, f0Start=f0, thStart=th,@\\
\mbox{}\verb@                    thetaControl=thetaControl, f0Control=f0Control, trace=FALSE)@\\
\mbox{}\verb@        th <- ff$th@\\
\mbox{}\verb@        llik <- ff$llik@\\
\mbox{}\verb@        f0 <- ff$f0@\\
\mbox{}\verb@@\\
\mbox{}\verb@        ## Check convergence@\\
\mbox{}\verb@        del <- abs((llik - llikold) / llik)@\\
\mbox{}\verb@        if (llik == 0) del <- 0@\\
\mbox{}\verb@        conv <- del < eps@\\
\mbox{}\verb@@\\
\mbox{}\verb@        if (print) {@\\
\mbox{}\verb@            cat("iteration ", iter,@\\
\mbox{}\verb@                "\nrelative change in log-likelihood = ", del,@\\
\mbox{}\verb@                "  (eps = ", eps, ")\n")@\\
\mbox{}\verb@        }@\\
\mbox{}\verb@    }@\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Final values@\\
\mbox{}\verb@    eta <- linkfun(mu)@\\
\mbox{}\verb@    dmudeta <- mu.eta(eta)@\\
\mbox{}\verb@    llik <- ff$llik@\\
\mbox{}\verb@    theta <- th$theta@\\
\mbox{}\verb@    bPrime <- th$bPrime@\\
\mbox{}\verb@    bPrime2 <- th$bPrime2@\\
\mbox{}\verb@    fTilt <- th$fTilt[cbind(ySptIndex, seq_along(ySptIndex))]@\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Add NA values back into beta vector and varbeta if covariate matrix is not full rank@\\
\mbox{}\verb@    nBeta <- length(beta) + sum(naID)@\\
\mbox{}\verb@    betaTemp <- rep(NA, nBeta)@\\
\mbox{}\verb@    betaTemp[!naID] <- beta@\\
\mbox{}\verb@    beta <- betaTemp@\\
\mbox{}\verb@ @\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Return gldrm object@\\
\mbox{}\verb@    attributes(beta) <- NULL@\\
\mbox{}\verb@    attributes(f0) <- NULL@\\
\mbox{}\verb@@\\
\mbox{}\verb@    fit <- list(conv=conv, iter=iter, llik=llik,@\\
\mbox{}\verb@                beta=beta, mu=mu, eta=eta, f0=f0, spt=spt, mu0=mu0,@\\
\mbox{}\verb@                theta=theta, bPrime=bPrime, bPrime2=bPrime2, fTilt=fTilt, weights=weights)@\\
\mbox{}\verb@@\\
\mbox{}\verb@    if (returnfTiltMatrix)@\\
\mbox{}\verb@        fit$fTiltMatrix <- t(th$fTilt)@\\
\mbox{}\verb@@\\
\mbox{}\verb@    if (returnf0ScoreInfo) {@\\
\mbox{}\verb@        fit$score.logf0 <- ff$score.log@\\
\mbox{}\verb@        fit$info.logf0 <- ff$info.log@\\
\mbox{}\verb@    }@\\
\mbox{}\verb@@\\
\mbox{}\verb@    class(fit) <- "gldrm"@\\
\mbox{}\verb@    fit@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb4}{, 4}\NWlink{nuweb6}{, 6}\NWlink{nuweb10}{, 10}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\begin{flushleft} \small\label{scrap2}\raggedright\small
\NWtarget{nuweb4}{} \verb@"../R/wgldrm.R"@\nobreak\ {\footnotesize {4}}$\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@#' Beta optimization routing@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#' @{\tt @}\verb@param x Covariate matrix.@\\
\mbox{}\verb@#' @{\tt @}\verb@param y Response vector.@\\
\mbox{}\verb@#' @{\tt @}\verb@param spt Vector of unique observed support points in the response.@\\
\mbox{}\verb@#' @{\tt @}\verb@param ySptIndex Index of each \code{y} value within the \code{spt} vector.@\\
\mbox{}\verb@#' @{\tt @}\verb@param f0 Current values of f0.@\\
\mbox{}\verb@#' @{\tt @}\verb@param linkinv Inverse link function.@\\
\mbox{}\verb@#' @{\tt @}\verb@param mu.eta Deriviative of inverse link function.@\\
\mbox{}\verb@#' @{\tt @}\verb@param offset Vector of known offset values to be added to the linear@\\
\mbox{}\verb@#' combination (x' beta) for each observation. Mostly intended for likelihood ratio@\\
\mbox{}\verb@#' and score confidence intervals.@\\
\mbox{}\verb@#'@{\tt @}\verb@param sampprobs Optional matrix of sampling probabilities.@\\
\mbox{}\verb@#'@{\tt @}\verb@param betaStart Starting values for beta (typically the estimates from the@\\
\mbox{}\verb@#' previous iteration).@\\
\mbox{}\verb@#'@{\tt @}\verb@param thStart Starting theta values. Needs to be a list of values matching@\\
\mbox{}\verb@#' the output of the \code{getTheta} function.@\\
\mbox{}\verb@#'@{\tt @}\verb@param thetaConrol A "thetaControl" object returned from the \code{theta.control}@\\
\mbox{}\verb@#' function.@\\
\mbox{}\verb@#'@{\tt @}\verb@param betaControl A "betaControl" object returned from the \code{beta.control}@\\
\mbox{}\verb@#' function.@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#' @{\tt @}\verb@return A list containing the following:@\\
\mbox{}\verb@#' \itemize{@\\
\mbox{}\verb@#' \item \code{beta} Updated values.@\\
\mbox{}\verb@#' \item \code{mu} Updated mean for each observation.@\\
\mbox{}\verb@#' \item \code{th} Updated list returned from the \code{getTheta} function.@\\
\mbox{}\verb@#' \item \code{llik} Updated log-likelihood.@\\
\mbox{}\verb@#' \item \code{iter} Number of iterations until convergence. (Will always be@\\
\mbox{}\verb@#' one unless \code{maxiter} is increased to something greater than one using the@\\
\mbox{}\verb@#' \code{betaControl} argument.)@\\
\mbox{}\verb@#' \item \code{conv} Convergence indicator. (Will always be FALSE unless@\\
\mbox{}\verb@#' \code{maxiter} is increased to something greater than one using the@\\
\mbox{}\verb@#' \code{betaControl} argument.)@\\
\mbox{}\verb@#' }@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#' @{\tt @}\verb@keywords internal@\\
\mbox{}\verb@beta.control <- function (eps = 1e-10, maxiter = 1, maxhalf = 10) @\\
\mbox{}\verb@{@\\
\mbox{}\verb@    betaControl <- as.list(environment())@\\
\mbox{}\verb@    class(betaControl) <- "betaControl"@\\
\mbox{}\verb@    betaControl@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@\\
\mbox{}\verb@getBeta <- function(x, y, spt, ySptIndex, f0, linkinv, mu.eta, offset, weights,@\\
\mbox{}\verb@                    betaStart, thStart,@\\
\mbox{}\verb@                    thetaControl=theta.control(), betaControl=beta.control())@\\
\mbox{}\verb@{@\\
\mbox{}\verb@    ## Extract control arguments@\\
\mbox{}\verb@    if (class(betaControl) != "betaControl")@\\
\mbox{}\verb@      stop("betaControl must be an object of class betaControl returned by betaControl() function.")@\\
\mbox{}\verb@    eps <- betaControl$eps@\\
\mbox{}\verb@    maxiter <- betaControl$maxiter@\\
\mbox{}\verb@    maxhalf <- betaControl$maxhalf@\\
\mbox{}\verb@@\\
\mbox{}\verb@    sptMin <- min(spt)@\\
\mbox{}\verb@    sptMax <- max(spt)@\\
\mbox{}\verb@    beta <- betaStart@\\
\mbox{}\verb@    th <- thStart@\\
\mbox{}\verb@    llik <- th$llik@\\
\mbox{}\verb@@\\
\mbox{}\verb@    conv <- FALSE@\\
\mbox{}\verb@    maxhalfreached <- FALSE@\\
\mbox{}\verb@    iter <- 0@\\
\mbox{}\verb@    while (!conv && !maxhalfreached && iter < maxiter)@\\
\mbox{}\verb@    {@\\
\mbox{}\verb@        iter <- iter+1@\\
\mbox{}\verb@@\\
\mbox{}\verb@        ## Update mean vector and related quantities@\\
\mbox{}\verb@        eta <- c(x %*% beta + offset)@\\
\mbox{}\verb@        mu <- linkinv(eta)@\\
\mbox{}\verb@        dmudeta <- mu.eta(eta)@\\
\mbox{}\verb@        betaold <- beta@\\
\mbox{}\verb@        muold <- mu@\\
\mbox{}\verb@        thold <- th@\\
\mbox{}\verb@        llikold <- llik@\\
\mbox{}\verb@@\\
\mbox{}\verb@        ## Compute weighted least squares update@\\
\mbox{}\verb@        w <- weights * dmudeta^2 / th$bPrime2@\\
\mbox{}\verb@        ymm <- y - mu@\\
\mbox{}\verb@        r <-  ymm / dmudeta@\\
\mbox{}\verb@        @\\
\mbox{}\verb@        yeqmu <- which(abs(ymm) < 1e-15)@\\
\mbox{}\verb@        w[yeqmu] <- 0  # prevent 0/0@\\
\mbox{}\verb@        r[yeqmu] <- 0  # prevent 0/0@\\
\mbox{}\verb@        @\\
\mbox{}\verb@        if (any(w==Inf)) break@\\
\mbox{}\verb@        @\\
\mbox{}\verb@        betastep <- unname(coef(lm.wfit(x, r, w)))@\\
\mbox{}\verb@        betastep[is.na(betastep)] <- 0@\\
\mbox{}\verb@        @\\
\mbox{}\verb@        ## Let q = b''*(theta) / b''(theta)@\\
\mbox{}\verb@        ## W = diag{dmudeta^2 / b''(theta) * q}@\\
\mbox{}\verb@        ## r = (y - b'*(theta)) / (q * dmudeta)@\\
\mbox{}\verb@        ## We need to solve for beta such that I(betaHat) %*% beta = Score(betaHat),@\\
\mbox{}\verb@        ## or equivalently, X'WX = X'Wr, or equivalently W^{1/2}X = W^{1/2}r.@\\
\mbox{}\verb@        ## The linear system can be solved using qr.coef().@\\
\mbox{}\verb@@\\
\mbox{}\verb@        ### Update beta and take half steps if log-likelihood does not improve@\\
\mbox{}\verb@        beta <- beta + betastep@\\
\mbox{}\verb@        eta <- c(x %*% beta + offset)@\\
\mbox{}\verb@        mu <- linkinv(eta)@\\
\mbox{}\verb@        if (min(mu)<sptMin || max(mu)>sptMax) {@\\
\mbox{}\verb@            llik <- -Inf@\\
\mbox{}\verb@        } else {@\\
\mbox{}\verb@            th <- getTheta(spt=spt, f0=f0, mu=mu, weights=weights, ySptIndex=ySptIndex,@\\
\mbox{}\verb@                           thetaStart=thold$theta, thetaControl=thetaControl)@\\
\mbox{}\verb@            llik <- th$llik@\\
\mbox{}\verb@        }@\\
\mbox{}\verb@@\\
\mbox{}\verb@        nhalf <- 0@\\
\mbox{}\verb@        while ((llik<llikold) && (nhalf<maxhalf)) {@\\
\mbox{}\verb@            nhalf <- nhalf + 1@\\
\mbox{}\verb@            beta <- (beta + betaold) / 2@\\
\mbox{}\verb@            eta <- c(x %*% beta + offset)@\\
\mbox{}\verb@            mu <- linkinv(eta)@\\
\mbox{}\verb@            if (min(mu)<sptMin || max(mu)>sptMax) {@\\
\mbox{}\verb@                llik <- -Inf@\\
\mbox{}\verb@            } else {@\\
\mbox{}\verb@                th <- getTheta(spt=spt, f0=f0, mu=mu, weights=weights, ySptIndex=ySptIndex,@\\
\mbox{}\verb@                               thetaStart=thold$theta, thetaControl=thetaControl)@\\
\mbox{}\verb@                llik <- th$llik@\\
\mbox{}\verb@            }@\\
\mbox{}\verb@        }@\\
\mbox{}\verb@@\\
\mbox{}\verb@        if (llik < llikold) {@\\
\mbox{}\verb@            beta <- betaold@\\
\mbox{}\verb@            mu <- muold@\\
\mbox{}\verb@            th <- thold@\\
\mbox{}\verb@            llik <- llikold@\\
\mbox{}\verb@            conv <- FALSE@\\
\mbox{}\verb@            maxhalfreached <- TRUE@\\
\mbox{}\verb@        } else {@\\
\mbox{}\verb@            del <- (llik - llikold) / llik@\\
\mbox{}\verb@            if (llik == 0) del <- 0  # consider converged if model fit is perfect@\\
\mbox{}\verb@            conv <- del < eps@\\
\mbox{}\verb@        }@\\
\mbox{}\verb@    }@\\
\mbox{}\verb@@\\
\mbox{}\verb@    return(list(beta=beta, mu=mu, th=th, llik=llik, iter=iter, conv=conv))@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb4}{, 4}\NWlink{nuweb6}{, 6}\NWlink{nuweb10}{, 10}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\begin{flushleft} \small\label{scrap3}\raggedright\small
\NWtarget{nuweb6}{} \verb@"../R/wgldrm.R"@\nobreak\ {\footnotesize {6}}$\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@## Computes log(sum(exp(x))) with better precision@\\
\mbox{}\verb@logSumExp <- function(x)@\\
\mbox{}\verb@{@\\
\mbox{}\verb@    i <- which.max(x)@\\
\mbox{}\verb@    m <- x[i]@\\
\mbox{}\verb@    lse <- log1p(sum(exp(x[-i]-m))) + m@\\
\mbox{}\verb@    lse@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@\\
\mbox{}\verb@## g function (logit transformation from appendix)@\\
\mbox{}\verb@g <- function(mu, m, M) log(mu-m) - log(M-mu)@\\
\mbox{}\verb@@\\
\mbox{}\verb@#' Control arguments for \eqn{\theta} update algorithm@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#' This function returns control arguments for the \eqn{\theta} update algorithm.@\\
\mbox{}\verb@#' Each argument has a default value, which will be used unless a different@\\
\mbox{}\verb@#' value is provided by the user.@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#'@{\tt @}\verb@param eps Convergence threshold for theta updates. Convergence is@\\
\mbox{}\verb@#' evaluated separately for each observation. An observation has converged when@\\
\mbox{}\verb@#' the difference between \eqn{b'(\theta)} and \eqn{\mu} is less than \code{epsTheta}.@\\
\mbox{}\verb@#'@{\tt @}\verb@param maxiter Maximum number of iterations.@\\
\mbox{}\verb@#'@{\tt @}\verb@param maxhalf Maximum number of half steps allowed per iteration if the@\\
\mbox{}\verb@#' convergence criterion does not improve.@\\
\mbox{}\verb@#'@{\tt @}\verb@param maxtheta Absolute value of theta is not allowed to exceed \code{maxtheta}.@\\
\mbox{}\verb@#'@{\tt @}\verb@param logit Logical for whether logit transformation should be used. Use of@\\
\mbox{}\verb@#' this stabilizing transformation appears to be faster in general. Default is TRUE.@\\
\mbox{}\verb@#'@{\tt @}\verb@param logsumexp Logical argument for whether log-sum-exp trick should be used.@\\
\mbox{}\verb@#' This may improve numerical stability at the expense of computational time.@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#' @{\tt @}\verb@return Object of S3 class "thetaControl", which is a list of control arguments.@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#' @{\tt @}\verb@keywords internal@\\
\mbox{}\verb@theta.control <- function(eps=1e-10, maxiter=100, maxhalf=20, maxtheta=500,@\\
\mbox{}\verb@                          logit=TRUE, logsumexp=FALSE)@\\
\mbox{}\verb@{@\\
\mbox{}\verb@    thetaControl <- as.list(environment())@\\
\mbox{}\verb@    class(thetaControl) <- "thetaControl"@\\
\mbox{}\verb@    thetaControl@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@\\
\mbox{}\verb@#' getTheta@\\
\mbox{}\verb@#' Updates theta. Vectorized but only updates observations that have not converged.@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#'@{\tt @}\verb@param spt Support of the observed response variable. (This is the set of@\\
\mbox{}\verb@#' unique values observed, not the set of all possible values.)@\\
\mbox{}\verb@#'@{\tt @}\verb@param f0 Values of the baseline distribution corresponding to the values of spt@\\
\mbox{}\verb@#'@{\tt @}\verb@param mu The fitted mean for each observation. Note these values must lie@\\
\mbox{}\verb@#' strictly within the range of the support.@\\
\mbox{}\verb@#'@{\tt @}\verb@param sampprobs Matrix of sampling probabilities. The number of rows should@\\
\mbox{}\verb@#' equal the number of observations, and the number of columns should equal@\\
\mbox{}\verb@#' the number of unique observed support points.@\\
\mbox{}\verb@#'@{\tt @}\verb@param ySptIndex Vector containing index of each obervation's response value@\\
\mbox{}\verb@#' within the \code{spt} vector. This is only needed to calculate the log-likelihood@\\
\mbox{}\verb@#' after each update.@\\
\mbox{}\verb@#'@{\tt @}\verb@param thetaStart Vector of starting values. One value per observation. If@\\
\mbox{}\verb@#' \code{NULL}, zero is used as the starting value for each observation.@\\
\mbox{}\verb@#'@{\tt @}\verb@param thetaControl Object of class \code{thetaControl}, which is a list of@\\
\mbox{}\verb@#' control arguments returned by the \code{thetaControl} function.@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#' @{\tt @}\verb@return List containing the following:@\\
\mbox{}\verb@#' \itemize{@\\
\mbox{}\verb@#' \item \code{theta} Updated values.@\\
\mbox{}\verb@#' \item \code{fTilt} Matrix containing the exponentially tilted distribution for each@\\
\mbox{}\verb@#' observation, i.e. f(y|X=x). Each column corresponds to an observation and sums to one.@\\
\mbox{}\verb@#' \item \code{bPrime} Vector containing the mean of the exponentially tilted distribution@\\
\mbox{}\verb@#' for each observation. Should match \code{mu} argument very closely.@\\
\mbox{}\verb@#' \item \code{bPrime2} Vector containing the variance of the exponentially tilted@\\
\mbox{}\verb@#' distribution for each observation.@\\
\mbox{}\verb@#' \item \code{fTiltSW} Matrix containing the exponentially tilted distribution for each@\\
\mbox{}\verb@#' observation, conditional on that observation being sampled, i.e. f(y|X=x, S=1).@\\
\mbox{}\verb@#' If \code{sampprobs=NULL}, then \code{fTiltSW} matches \code{fTilt}.@\\
\mbox{}\verb@#' \item \code{bPrimeSW} Vector containing the mean for each observation, conditional@\\
\mbox{}\verb@#' on that observation being sampled. If \code{sampprobs=NULL}, then \code{bPrimeSW}@\\
\mbox{}\verb@#' matches \code{bPrime}.@\\
\mbox{}\verb@#' \item \code{bPrime2SW} Vector containing the variance for each observation, conditional@\\
\mbox{}\verb@#' on that observation being sampled. If \code{sampprobs=NULL}, then \code{bPrime2SW}@\\
\mbox{}\verb@#' matches \code{bPrime2}.@\\
\mbox{}\verb@#' \item \code{llik} Semiparametric log-likelihood, evaluated at the current beta@\\
\mbox{}\verb@#' and f0 values. If sampling weights are used, then the log-likelihood is conditional@\\
\mbox{}\verb@#' on each observation being sampled.@\\
\mbox{}\verb@#' \item \code{conv} Convergence indicator.@\\
\mbox{}\verb@#' \item \code{iter} Number of iterations until convergence was reached.@\\
\mbox{}\verb@#' }@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#' @{\tt @}\verb@keywords internal@\\
\mbox{}\verb@getTheta <- function(spt, f0, mu, weights, ySptIndex, thetaStart=NULL, thetaControl=theta.control())@\\
\mbox{}\verb@{@\\
\mbox{}\verb@    ## Extract control arguments@\\
\mbox{}\verb@    if (class(thetaControl) != "thetaControl")@\\
\mbox{}\verb@        stop("thetaControl must be an object of class \'thetaControl\' returned by@\\
\mbox{}\verb@             thetaControl() function.")@\\
\mbox{}\verb@    logit <- thetaControl$logit@\\
\mbox{}\verb@    eps <- thetaControl$eps@\\
\mbox{}\verb@    maxiter <- thetaControl$maxiter@\\
\mbox{}\verb@    maxhalf <- thetaControl$maxhalf@\\
\mbox{}\verb@    maxtheta <- thetaControl$maxtheta@\\
\mbox{}\verb@    logsumexp <- thetaControl$logsumexp@\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Define value from inputs@\\
\mbox{}\verb@    sptN <- length(spt)@\\
\mbox{}\verb@    m <- min(spt)@\\
\mbox{}\verb@    M <- max(spt)@\\
\mbox{}\verb@    n <- length(mu)@\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Format arguments@\\
\mbox{}\verb@    spt <- as.vector(spt)@\\
\mbox{}\verb@    f0 <- as.vector(f0)@\\
\mbox{}\verb@    mu <- as.vector(mu)@\\
\mbox{}\verb@    if (!is.null(thetaStart)) {@\\
\mbox{}\verb@        thetaStart <- as.vector(thetaStart)@\\
\mbox{}\verb@    } else {@\\
\mbox{}\verb@        thetaStart <- rep(0, n)@\\
\mbox{}\verb@    }@\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Argument checks@\\
\mbox{}\verb@    if (length(f0) != sptN)@\\
\mbox{}\verb@        stop("spt and f0 must be vectors of equal length.")@\\
\mbox{}\verb@    if (any(f0 < 0))@\\
\mbox{}\verb@        stop("f0 values cannot be negative.")@\\
\mbox{}\verb@    if (min(mu)<m || max(mu)>M)@\\
\mbox{}\verb@        stop("mu starting values must lie within the range of spt.")@\\
\mbox{}\verb@    if (length(thetaStart) != n)@\\
\mbox{}\verb@        stop("thetaStart must be a vector with length equal length(mu)")@\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Value does not change@\\
\mbox{}\verb@    gMu <- g(mu, m, M)@\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Initialize values@\\
\mbox{}\verb@    theta <- thetaStart  # initial values required@\\
\mbox{}\verb@    thetaOld <- bPrimeErrOld <- rep(NA, n)@\\
\mbox{}\verb@    conv <- rep(FALSE, n)@\\
\mbox{}\verb@    maxedOut <- rep(FALSE, n)@\\
\mbox{}\verb@@\\
\mbox{}\verb@    if (logsumexp) {@\\
\mbox{}\verb@        logf0 <- log(f0)@\\
\mbox{}\verb@        logfUnstd <- logf0 + tcrossprod(spt, theta)@\\
\mbox{}\verb@        logb <- apply(logfUnstd, 2, logSumExp)@\\
\mbox{}\verb@        fTilt <- exp(logfUnstd - rep(logb, each=sptN))@\\
\mbox{}\verb@        normfact <- colSums(fTilt)@\\
\mbox{}\verb@        normss <- which(normfact != 1)@\\
\mbox{}\verb@        fTilt[, normss] <- fTilt[, normss, drop=FALSE] / rep(normfact[normss], each=sptN)@\\
\mbox{}\verb@    } else {@\\
\mbox{}\verb@        fUnstd <- f0 * exp(tcrossprod(spt, theta))  # |spt| x n matrix of tilted f0 values@\\
\mbox{}\verb@        b <- colSums(fUnstd)@\\
\mbox{}\verb@        fTilt <- fUnstd / rep(b, each=sptN)  # normalized@\\
\mbox{}\verb@    }@\\
\mbox{}\verb@    bPrime <- colSums(spt*fTilt)  # mean as a function of theta@\\
\mbox{}\verb@    bPrime2 <- colSums(outer(spt, bPrime, "-")^2 * fTilt)  # variance as a function of theta@\\
\mbox{}\verb@    bPrimeErr <- bPrime - mu  # used to assess convergence@\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Update theta until convergence@\\
\mbox{}\verb@    conv <- (abs(bPrimeErr) < eps) |@\\
\mbox{}\verb@        (theta==maxtheta & bPrimeErr<0) |@\\
\mbox{}\verb@        (theta==-maxtheta & bPrimeErr>0)@\\
\mbox{}\verb@    s <- which(!conv)@\\
\mbox{}\verb@    iter <- 0@\\
\mbox{}\verb@    while(length(s)>0 && iter<maxiter) {@\\
\mbox{}\verb@        iter <- iter + 1@\\
\mbox{}\verb@        bPrimeErrOld[s] <- bPrimeErr[s]  # used to assess convergence@\\
\mbox{}\verb@@\\
\mbox{}\verb@        ## 1) Update theta@\\
\mbox{}\verb@        thetaOld[s] <- theta[s]@\\
\mbox{}\verb@        if (logit) {@\\
\mbox{}\verb@            tPrimeS <- (M-m) / ((bPrime[s]-m) * (M-bPrime[s])) * bPrime2[s]@\\
\mbox{}\verb@                # t'(theta) from paper: temporary variable@\\
\mbox{}\verb@                # only needed for the subset of observations that have not converged@\\
\mbox{}\verb@            tPrimeS[is.na(tPrimeS) | tPrimeS==Inf] <- 0@\\
\mbox{}\verb@                # If bPrime is on the boundary, then bPrime2 should be zero.@\\
\mbox{}\verb@                # Exceptions are due to rounding error.@\\
\mbox{}\verb@            thetaS <- theta[s] - (g(bPrime[s], m, M) - gMu[s]) / tPrimeS@\\
\mbox{}\verb@        } else {@\\
\mbox{}\verb@            thetaS <- theta[s] - bPrimeErr[s] / bPrime2[s]@\\
\mbox{}\verb@        }@\\
\mbox{}\verb@        thetaS[thetaS > maxtheta] <- maxtheta@\\
\mbox{}\verb@        thetaS[thetaS < -maxtheta] <- -maxtheta@\\
\mbox{}\verb@        theta[s] <- thetaS@\\
\mbox{}\verb@@\\
\mbox{}\verb@        ## 2) Update fTilt, bPrime, and bPrime2 and take half steps if bPrimeErr not improved@\\
\mbox{}\verb@        ss <- s@\\
\mbox{}\verb@        nhalf <- 0@\\
\mbox{}\verb@        while(length(ss)>0 && nhalf<maxhalf) {@\\
\mbox{}\verb@            ## 2a) Update fTilt, bPrime, and bPrime2@\\
\mbox{}\verb@            if (logsumexp) {@\\
\mbox{}\verb@                logfUnstd[, ss] <- logf0 + tcrossprod(spt, theta[ss])@\\
\mbox{}\verb@                logb[ss] <- apply(logfUnstd[, ss, drop=FALSE], 2, logSumExp)@\\
\mbox{}\verb@                fTilt[, ss] <- exp(logfUnstd[, ss, drop=FALSE] - rep(logb[ss], each=sptN))@\\
\mbox{}\verb@                normfact <- colSums(fTilt[, ss, drop=FALSE])@\\
\mbox{}\verb@                normss <- which(normfact != 1)@\\
\mbox{}\verb@                fTilt[, ss[normss]] <- fTilt[, ss[normss], drop=FALSE] / rep(normfact[normss], each=sptN)@\\
\mbox{}\verb@            } else {@\\
\mbox{}\verb@                fUnstd[, ss] <- f0*exp(tcrossprod(spt, theta[ss]))  # |spt| x n matrix of tilted f0 values@\\
\mbox{}\verb@                b[ss] <- colSums(fUnstd[, ss, drop=FALSE])@\\
\mbox{}\verb@                fTilt[, ss] <- fUnstd[, ss, drop=FALSE] / rep(b[ss], each=sptN)  # normalized@\\
\mbox{}\verb@            }@\\
\mbox{}\verb@            bPrime[ss] <- colSums(spt*fTilt[, ss, drop=FALSE])  # mean as a function of theta@\\
\mbox{}\verb@            bPrime2[ss] <- colSums(outer(spt, bPrime[ss], "-")^2 * fTilt[, ss, drop=FALSE])  # variance as a function of theta@\\
\mbox{}\verb@            bPrimeErr[ss] <- bPrime[ss] - mu[ss]  # used to assess convergence@\\
\mbox{}\verb@@\\
\mbox{}\verb@            ## 2b) Take half steps if necessary@\\
\mbox{}\verb@            ss <- ss[abs(bPrimeErr[ss]) > abs(bPrimeErrOld[ss])]@\\
\mbox{}\verb@            if (length(ss) > 0) nhalf <- nhalf + 1@\\
\mbox{}\verb@            theta[ss] <- (theta[ss] + thetaOld[ss]) / 2@\\
\mbox{}\verb@        }@\\
\mbox{}\verb@        ## If maximum half steps are exceeded, set theta to previous value@\\
\mbox{}\verb@        maxedOut[ss] <- TRUE@\\
\mbox{}\verb@        theta[ss] <- thetaOld[ss]@\\
\mbox{}\verb@@\\
\mbox{}\verb@        ## 3) Check convergence@\\
\mbox{}\verb@        conv[s] <- (abs(bPrimeErr[s]) < eps) |@\\
\mbox{}\verb@            (theta[s]==maxtheta & bPrimeErr[s]<0) |@\\
\mbox{}\verb@            (theta[s]==-maxtheta & bPrimeErr[s]>0)@\\
\mbox{}\verb@        s <- s[!conv[s] & !maxedOut[s]]@\\
\mbox{}\verb@    }@\\
\mbox{}\verb@@\\
\mbox{}\verb@@\\
\mbox{}\verb@    fTiltSW <- fTilt@\\
\mbox{}\verb@    bPrimeSW <- bPrime@\\
\mbox{}\verb@    bPrime2SW <- bPrime2@\\
\mbox{}\verb@  @\\
\mbox{}\verb@@\\
\mbox{}\verb@    ## Calculate log-likelihood@\\
\mbox{}\verb@   # llik <- sum(log(fTiltSW[cbind(ySptIndex, seq_along(ySptIndex))]))@\\
\mbox{}\verb@    fTiltSW.extracted <- fTiltSW[cbind(ySptIndex, seq_along(ySptIndex))]@\\
\mbox{}\verb@    llik <- sum(weights[fTiltSW.extracted>0] * log(fTiltSW.extracted[fTiltSW.extracted>0]))@\\
\mbox{}\verb@ @\\
\mbox{}\verb@    list(theta=theta, fTilt=fTilt, bPrime=bPrime, bPrime2=bPrime2,@\\
\mbox{}\verb@         fTiltSW=fTiltSW, bPrimeSW=bPrimeSW, bPrime2SW=bPrime2SW,@\\
\mbox{}\verb@         llik=llik, conv=conv, iter=iter)@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb4}{, 4}\NWlink{nuweb6}{, 6}\NWlink{nuweb10}{, 10}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\begin{flushleft} \small\label{scrap4}\raggedright\small
\NWtarget{nuweb10}{} \verb@"../R/wgldrm.R"@\nobreak\ {\footnotesize {10}}$\equiv$
\vspace{-1ex}
\begin{list}{}{} \item
\mbox{}\verb@@\\
\mbox{}\verb@@\\
\mbox{}\verb@#' Control arguments for f0 update algorithm@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#' This function returns control arguments for the \eqn{f_0} update algorithm.@\\
\mbox{}\verb@#' Each argument has a default value, which will be used unless a different@\\
\mbox{}\verb@#' value is provided by the user.@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#'@{\tt @}\verb@param eps Convergence threshold. The update has converged when the relative@\\
\mbox{}\verb@#' change in log-likelihood between iterations is less than \code{eps}.@\\
\mbox{}\verb@#' absolute change is less than \code{thesh}.@\\
\mbox{}\verb@#'@{\tt @}\verb@param maxiter Maximum number of iterations allowed.@\\
\mbox{}\verb@#'@{\tt @}\verb@param maxhalf Maximum number of half steps allowed per iteration if@\\
\mbox{}\verb@#' log-likelihood does not improve between iterations.@\\
\mbox{}\verb@#'@{\tt @}\verb@param maxlogstep Maximum optimization step size allowed on the@\\
\mbox{}\verb@#' \code{log(f0)} scale.@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#' @{\tt @}\verb@return Object of S3 class "f0Control", which is a list of control arguments.@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#' @{\tt @}\verb@keywords internal@\\
\mbox{}\verb@f0.control <- function(eps=1e-10, maxiter=1000, maxhalf=20, maxlogstep=2)@\\
\mbox{}\verb@{@\\
\mbox{}\verb@    f0Control <- as.list(environment())@\\
\mbox{}\verb@    class(f0Control) <- "f0Control"@\\
\mbox{}\verb@    f0Control@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@\\
\mbox{}\verb@#' f0 optimization routine@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#'@{\tt @}\verb@param y Vector of response values.@\\
\mbox{}\verb@#'@{\tt @}\verb@param spt Vector of unique observed support points in the response.@\\
\mbox{}\verb@#'@{\tt @}\verb@param ySptIndex Index of each \code{y} value within \code{spt}.@\\
\mbox{}\verb@#'@{\tt @}\verb@param sptFreq Vector containing frequency of each \code{spt} value.@\\
\mbox{}\verb@#'@{\tt @}\verb@param sampprobs Optional matrix of sampling probabilities.@\\
\mbox{}\verb@#'@{\tt @}\verb@param mu Fitted mean for each observation. Only used if \code{sampprobs=NULL}.@\\
\mbox{}\verb@#'@{\tt @}\verb@param mu0 Mean constraing for f0.@\\
\mbox{}\verb@#'@{\tt @}\verb@param f0Start Starting f0 values. (Typically the estimate from the previous@\\
\mbox{}\verb@#' iteration.)@\\
\mbox{}\verb@#'@{\tt @}\verb@param thStart Starting theta values. Needs to be a list of values matching@\\
\mbox{}\verb@#' the output of the \code{getTheta} function.@\\
\mbox{}\verb@#'@{\tt @}\verb@param thetaControl A "thetaControl" object returned from the \code{theta.control}@\\
\mbox{}\verb@#' function.@\\
\mbox{}\verb@#'@{\tt @}\verb@param f0Control An "f0Control" object returned from the \code{f0.control}@\\
\mbox{}\verb@#' function.@\\
\mbox{}\verb@#' trace Logical. If TRUE, then progress is printed to terminal at each iteration.@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#' @{\tt @}\verb@return A list containing the following:@\\
\mbox{}\verb@#' \itemize{@\\
\mbox{}\verb@#' \item \code{f0} Updated values.@\\
\mbox{}\verb@#' \item \code{llik} Updated log-likelihood.@\\
\mbox{}\verb@#' \item \code{th} Updated list returned from the \code{getTheta} function.@\\
\mbox{}\verb@#' \item \code{conv} Convergence indicator.@\\
\mbox{}\verb@#' \item \code{iter} Number of iterations until convergence.@\\
\mbox{}\verb@#' \item \code{nhalf} The number of half steps taken on the last iteration if the@\\
\mbox{}\verb@#' initial BFGS update did not improve the log-likelihood.@\\
\mbox{}\verb@#' \item \code{score.log} Score function with respect to log(f0) at convergence.@\\
\mbox{}\verb@#' \item \code{info.log} Information matrix with respect to log(f0) at convergence.@\\
\mbox{}\verb@#' }@\\
\mbox{}\verb@#'@\\
\mbox{}\verb@#' @{\tt @}\verb@keywords internal@\\
\mbox{}\verb@getf0 <- function(y, spt, ySptIndex, sptFreq, weights, sptFreq.weighted, mu, mu0, f0Start, thStart,@\\
\mbox{}\verb@                  thetaControl=theta.control(), f0Control=f0.control(), trace=FALSE)@\\
\mbox{}\verb@{@\\
\mbox{}\verb@    ## Extract theta control arguments@\\
\mbox{}\verb@    if (class(f0Control) != "f0Control")@\\
\mbox{}\verb@        stop("f0Control must be an object of class f0Control returned by f0Control() function.")@\\
\mbox{}\verb@    eps <- f0Control$eps@\\
\mbox{}\verb@    maxiter <- f0Control$maxiter@\\
\mbox{}\verb@    maxhalf <- f0Control$maxhalf@\\
\mbox{}\verb@    maxlogstep <- f0Control$maxlogstep@\\
\mbox{}\verb@@\\
\mbox{}\verb@    f0 <- f0Start  # assumes sum(f0Start) = 1 and sum(f0Start * spt) = mu0@\\
\mbox{}\verb@    th <- thStart@\\
\mbox{}\verb@    llik <- th$llik@\\
\mbox{}\verb@    score.log <- NULL@\\
\mbox{}\verb@    smm <- outer(spt, mu, "-")@\\
\mbox{}\verb@    ymm <- y - mu@\\
\mbox{}\verb@    yeqmu <- which(abs(ymm) < 1e-15)@\\
\mbox{}\verb@    @\\
\mbox{}\verb@@\\
\mbox{}\verb@    conv <- FALSE@\\
\mbox{}\verb@    iter <- 0@\\
\mbox{}\verb@    while (!conv && iter<maxiter) {@\\
\mbox{}\verb@        iter <- iter + 1@\\
\mbox{}\verb@@\\
\mbox{}\verb@        # Score calculation@\\
\mbox{}\verb@        score.logOld <- score.log@\\
\mbox{}\verb@        @\\
\mbox{}\verb@        fTiltSWSums <- rowSums(th$fTiltSW)@\\
\mbox{}\verb@        fTiltSWSumsWeighted <- apply(th$fTiltSW, MARGIN=1, function(x) sum(weights * x)) @\\
\mbox{}\verb@        smmfTiltSW <- smm * th$fTiltSW@\\
\mbox{}\verb@        ystd <- ymm / th$bPrime2SW@\\
\mbox{}\verb@        ystdWeighted <- weights * ystd@\\
\mbox{}\verb@        ystd[yeqmu] <- 0  # prevent 0/0@\\
\mbox{}\verb@        ystdWeighted[yeqmu] <- 0  # prevent 0/0@\\
\mbox{}\verb@    @\\
\mbox{}\verb@@\\
\mbox{}\verb@        score.logT1 <- sptFreq.weighted@\\
\mbox{}\verb@        score.logT2 <- fTiltSWSumsWeighted@\\
\mbox{}\verb@        score.logT3 <- c(smmfTiltSW %*% ystdWeighted)@\\
\mbox{}\verb@        score.log <- score.logT1 - score.logT2 - score.logT3@\\
\mbox{}\verb@@\\
\mbox{}\verb@        # Inverse info, score step, and f0 step are on the log scale (score is not)@\\
\mbox{}\verb@        if (iter == 1) {@\\
\mbox{}\verb@            d1 <- min(fTiltSWSumsWeighted)  # max inverse diagonal of first information term, on log scale@\\
\mbox{}\verb@            d2 <- max(abs(score.log)) / maxlogstep@\\
\mbox{}\verb@            d <- max(d1, d2)@\\
\mbox{}\verb@            infoinvBFGS.log <- diag(1/d, nrow=length(f0))@\\
\mbox{}\verb@        } else {@\\
\mbox{}\verb@            scorestep.log <- score.log - score.logOld@\\
\mbox{}\verb@           # f0step.log <- log(f0) - log(f0old)@\\
\mbox{}\verb@            # to prevent the 0/0 situation @\\
\mbox{}\verb@            ratiof0f0old <- f0 / f0old@\\
\mbox{}\verb@            ratiof0f0old[is.na(ratiof0f0old)] <- 1@\\
\mbox{}\verb@            f0step.log <- log(ratiof0f0old)@\\
\mbox{}\verb@            @\\
\mbox{}\verb@            sy <- sum(f0step.log * scorestep.log)@\\
\mbox{}\verb@            yiy <- c(crossprod(scorestep.log, infoinvBFGS.log %*% scorestep.log))@\\
\mbox{}\verb@            iys <- tcrossprod(infoinvBFGS.log %*% scorestep.log, f0step.log)@\\
\mbox{}\verb@            infoinvBFGS.log <- infoinvBFGS.log + ((yiy - sy) / sy^2) * tcrossprod(f0step.log) - (1 / sy) * (iys + t(iys))@\\
\mbox{}\verb@        }@\\
\mbox{}\verb@        logstep <- c(infoinvBFGS.log %*% score.log)@\\
\mbox{}\verb@@\\
\mbox{}\verb@        # Cap log(f0) step size@\\
\mbox{}\verb@        logstep.max <- max(abs(logstep))@\\
\mbox{}\verb@        if (logstep.max > maxlogstep)@\\
\mbox{}\verb@            logstep <- logstep * (maxlogstep / logstep.max)@\\
\mbox{}\verb@@\\
\mbox{}\verb@        # Save values from previous iteration@\\
\mbox{}\verb@        f0old <- f0@\\
\mbox{}\verb@        thold <- th@\\
\mbox{}\verb@        llikold <- llik@\\
\mbox{}\verb@@\\
\mbox{}\verb@        # Take update step@\\
\mbox{}\verb@        f0 <- exp(log(f0) + logstep)@\\
\mbox{}\verb@        # Scale and tilt f0@\\
\mbox{}\verb@        f0 <- f0 / sum(f0)@\\
\mbox{}\verb@        f0 <- getTheta(spt=spt, f0=f0, mu=mu0, weights=weights, ySptIndex=1,@\\
\mbox{}\verb@                       thetaStart=0, thetaControl=thetaControl)$fTilt[, 1]@\\
\mbox{}\verb@        # Update theta and likelihood@\\
\mbox{}\verb@        thold <- th@\\
\mbox{}\verb@        llikold <- llik@\\
\mbox{}\verb@        th <- getTheta(spt=spt, f0=f0, mu=mu, weights=weights, ySptIndex=ySptIndex,@\\
\mbox{}\verb@                       thetaStart=th$theta, thetaControl=thetaControl)@\\
\mbox{}\verb@        llik <- th$llik@\\
\mbox{}\verb@        conv <- abs((llik - llikold) / (llik + 1e-100)) < eps@\\
\mbox{}\verb@@\\
\mbox{}\verb@        # If log-likelihood does not improve, change step direction to be along gradient@\\
\mbox{}\verb@        # Take half steps until likelihood improves@\\
\mbox{}\verb@        # Continue taking half steps until log likelihood no longer improves@\\
\mbox{}\verb@        nhalf <- 0@\\
\mbox{}\verb@        if (llik<llikold) {@\\
\mbox{}\verb@            llikprev <- -Inf@\\
\mbox{}\verb@            while ((llik<llikold || llik>llikprev) && nhalf<maxhalf) {@\\
\mbox{}\verb@                nhalf <- nhalf + 1@\\
\mbox{}\verb@@\\
\mbox{}\verb@                # Set previous values@\\
\mbox{}\verb@                llikprev <- llik@\\
\mbox{}\verb@                thprev <- th@\\
\mbox{}\verb@                f0prev <- f0@\\
\mbox{}\verb@                infoinvBFGS.logprev <- infoinvBFGS.log@\\
\mbox{}\verb@@\\
\mbox{}\verb@                f0 <- exp((log(f0) + log(f0old)) / 2)@\\
\mbox{}\verb@                f0 <- f0 / sum(f0)@\\
\mbox{}\verb@                f0 <- getTheta(spt=spt, f0=f0, mu=mu0, weights=weights, ySptIndex=1,@\\
\mbox{}\verb@                               thetaStart=0, thetaControl=thetaControl)$fTilt[, 1]@\\
\mbox{}\verb@                th <- getTheta(spt=spt, f0=f0, mu=mu, weights=weights, ySptIndex=ySptIndex,@\\
\mbox{}\verb@                               thetaStart=th$theta, thetaControl=thetaControl)@\\
\mbox{}\verb@                llik <- th$llik@\\
\mbox{}\verb@                infoinvBFGS.log <- infoinvBFGS.log / 2@\\
\mbox{}\verb@            }@\\
\mbox{}\verb@@\\
\mbox{}\verb@            if (llik < llikprev) {@\\
\mbox{}\verb@                nhalf <- nhalf - 1@\\
\mbox{}\verb@                llik <- llikprev@\\
\mbox{}\verb@                th <- thprev@\\
\mbox{}\verb@                f0 <- f0prev@\\
\mbox{}\verb@                infoinvBFGS.log <- infoinvBFGS.logprev@\\
\mbox{}\verb@            }@\\
\mbox{}\verb@@\\
\mbox{}\verb@            conv <- abs((llik - llikold) / (llik + 1e-100)) < eps@\\
\mbox{}\verb@        }@\\
\mbox{}\verb@@\\
\mbox{}\verb@        if (llik < llikold) {@\\
\mbox{}\verb@            f0 <- f0old@\\
\mbox{}\verb@            th <- thold@\\
\mbox{}\verb@            llik <- llikold@\\
\mbox{}\verb@            conv <- TRUE@\\
\mbox{}\verb@        }@\\
\mbox{}\verb@@\\
\mbox{}\verb@        if (trace) {@\\
\mbox{}\verb@            printout <- paste0("iter ", iter, ": llik=", llik)@\\
\mbox{}\verb@            if (nhalf > 0)@\\
\mbox{}\verb@                printout <- paste0(printout, "; ", nhalf, " half steps")@\\
\mbox{}\verb@            cat(printout, "\n")@\\
\mbox{}\verb@        }@\\
\mbox{}\verb@    }@\\
\mbox{}\verb@@\\
\mbox{}\verb@    # Final score calculation@\\
\mbox{}\verb@        smm <- outer(spt, th$bPrimeSW, "-")@\\
\mbox{}\verb@        ymm <- y - th$bPrimeSW@\\
\mbox{}\verb@        yeqmu <- which(abs(ymm) < 1e-15)@\\
\mbox{}\verb@        @\\
\mbox{}\verb@          fTiltSWSums <- rowSums(th$fTiltSW)@\\
\mbox{}\verb@  fTiltSWSumsWeighted <- apply(th$fTiltSW, MARGIN=1, function(x) sum(weights * x)) @\\
\mbox{}\verb@  smmfTiltSW <- smm * th$fTiltSW@\\
\mbox{}\verb@  ystd <- ymm / th$bPrime2SW@\\
\mbox{}\verb@  ystdWeighted <- weights * ystd@\\
\mbox{}\verb@  ystd[yeqmu] <- 0  # prevent 0/0@\\
\mbox{}\verb@  ystdWeighted[yeqmu] <- 0  # prevent 0/0@\\
\mbox{}\verb@  ystd[yeqmu] <- 0  # prevent 0/0@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  score.logT1 <- sptFreq.weighted@\\
\mbox{}\verb@  score.logT2 <- fTiltSWSumsWeighted@\\
\mbox{}\verb@  score.logT3 <- c(smmfTiltSW %*% ystdWeighted)@\\
\mbox{}\verb@  score.log <- score.logT1 - score.logT2 - score.logT3@\\
\mbox{}\verb@  @\\
\mbox{}\verb@  # Final info calculation@\\
\mbox{}\verb@  info.logT1 <- diag(fTiltSWSumsWeighted)@\\
\mbox{}\verb@  info.logT2 <- tcrossprod(th$fTiltSW)@\\
\mbox{}\verb@  info.logT3 <- tcrossprod(smmfTiltSW, smmfTiltSW * rep(ystdWeighted, each=nrow(smmfTiltSW)))@\\
\mbox{}\verb@  info.log <- info.logT1 - info.logT2 - info.logT3@\\
\mbox{}\verb@  @\\
\mbox{}\verb@@\\
\mbox{}\verb@    list(f0=f0, llik=llik, th=th, conv=conv, iter=iter, nhalf=nhalf,@\\
\mbox{}\verb@         score.log=score.log, info.log=info.log)@\\
\mbox{}\verb@}@\\
\mbox{}\verb@@{\NWsep}
\end{list}
\vspace{-1.5ex}
\footnotesize
\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb4}{, 4}\NWlink{nuweb6}{, 6}\NWlink{nuweb10}{, 10}.

\item{}
\end{list}
\vspace{4ex}
\end{flushleft}
\end{document}

### 2021-03-04 R/SPGLM.R

@@ -1,5 +1,6 @@
 
-
+#' Fit semi-parametric GLM
+#'
 #'@rdname spglm
 #'@param formula a one-sided formula of the form \code{cbind(r, s) ~ predictors} where \code{r} and \code{s} give the number of responses and non-responses within each cluster, respectively (so the cluster size is \code{r+s}), and \code{predictors} describes the covariates.
 #'@param data  an optional matrix or data frame containing the variables in the formula \code{formula}. By default the variables are taken from \code{environment(formula).}
@@ -12,7 +13,6 @@
 #'@export
 #' @importFrom stats terms model.matrix
 
-#' Semi-parametric generalized linear model
 spglm <- function(formula, data, subset, weights, link="logit", start=NULL, control=list(eps=0.001, maxit=100), ...){
     fam <- binomial(link=link)
     
@@ -44,7 +44,7 @@ spglm <- function(formula, data, subset, weights, link="logit", start=NULL, cont
             stop("negative weights not allowed")
     
     
-     while (iter<itermax & difference>tol) {
+     while (iter < control$maxit & difference > control$eps) {
         iter <- iter + 1
         referencef0Pre <- referencef0
         tiltingWsPre <- tiltingWs
@@ -69,10 +69,7 @@ spglm <- function(formula, data, subset, weights, link="logit", start=NULL, cont
         betas <- res$regressionEst
         # monitor parameter estimates for baseline distribution, tilting parameters, regression coefficients
         difference <- sum(abs(referencef0Pre - referencef0)) + sum(abs(betasPre - betas))
-        # difference <- sum(abs(referencef0Pre - referencef0)) + sum(abs(tiltingWsPre - tiltingWs)) + sum(abs(betasPre - betas))
-        cat("In iteration", iter, ", the corresponding log-likelihood is", observedLikelihood(CBData, referencef0, tiltingWs), ". \n")
       }
-      # plot(referencef0, type="b")
       
       list(referencef0 = referencef0,
            tiltingWs = tiltingWs,

### 2021-03-04 R/SPreg.R

@@ -29,6 +29,8 @@ lambda_from_p <- function(p.vec, n=length(p.vec)-1){
   c(H %*% p.vec)
 }
 
+#' Fit semi-parametric relative risk model
+#'
 #'@rdname sprr
 #'@param formula a one-sided formula of the form \code{cbind(r, s) ~ predictors} where \code{r} and \code{s} give the number of responses and non-responses within each cluster, respectively (so the cluster size is \code{r+s}), and \code{predictors} describes the covariates.
 #'@param data  an optional matrix or data frame containing the variables in the formula \code{formula}. By default the variables are taken from \code{environment(formula).}
@@ -42,7 +44,6 @@ lambda_from_p <- function(p.vec, n=length(p.vec)-1){
 #'@export
 #' @importFrom stats terms model.matrix
 
-#' Semi-parametric relative risk model
 sprr <- function(formula, data, subset, weights, link="cloglog", mu1=NULL, start=NULL, control=list(eps=0.001, maxit=100), ...){
     fam <- binomial(link=link)
     model_fun <- fam$linkinv

### 2021-03-04 R/wgldrm.R

@@ -356,7 +356,7 @@ g <- function(mu, m, M) log(mu-m) - log(M-mu)
 #'
 #' @return Object of S3 class "thetaControl", which is a list of control arguments.
 #'
-#' @export
+#' @keywords internal
 theta.control <- function(eps=1e-10, maxiter=100, maxhalf=20, maxtheta=500,
                           logit=TRUE, logsumexp=FALSE)
 {
@@ -573,7 +573,7 @@ getTheta <- function(spt, f0, mu, weights, ySptIndex, thetaStart=NULL, thetaCont
 #'
 #' @return Object of S3 class "f0Control", which is a list of control arguments.
 #'
-#' @export
+#' @keywords internal
 f0.control <- function(eps=1e-10, maxiter=1000, maxhalf=20, maxlogstep=2)
 {
     f0Control <- as.list(environment())

### 2021-03-04 nuweb/SPreg_develop.R

@@ -1,19 +1,23 @@
 library(devtools)
-source('z:/RForge/Nuweb.R')
-#source('/home/aszabo/RForge/Nuweb_linux.R')
+source('nuweb/Nuweb.R')
+
+#  setting up the package and its infrastructure
+create("../exchreg")
+use_package("CorrBin")
+use_build_ignore("nuweb")
 
 
-run_nuweb(file="SPregress.w", path=getwd())
-run_nuweb(file="SPGLM.w", path=getwd())
-run_nuweb(file="wgldrm.w", path=getwd())
+ex <- as.package("../exchreg")
+nuweb(ex)
 
-source('R/SPreg.R', echo=FALSE)
-source('R/SPGLM.R', echo=FALSE)
-source('R/wgldrm.R', echo=FALSE)
+shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet --run-viewer  SPregress.tex")
+shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet --run-viewer  SPGLM.tex")
+shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet --run-viewer  wgldrm.tex")
 
-shell("cd c:/exchreg/ && texify --pdf --quiet --run-viewer --clean SPGLM.tex")
 
-data(shelltox)
+document(ex)
+load_all(ex)
+
 ######### Testing wGLDRM
 run_gldrm <- 
 function (formula, data = NULL, link = "identity", mu0 = NULL,  weights=NULL,
@@ -71,13 +75,21 @@ function (formula, data = NULL, link = "identity", mu0 = NULL,  weights=NULL,
   modZ
 }
 
+data(shelltox)
 sh2 <- shelltox[rep(1:nrow(shelltox), shelltox$Freq), ]
+
+# original code
+res0 <- gldrm::gldrm(I(NResp/ClusterSize) ~ Trt, data=sh2, link="logit")
 # full dataset without frequency weights
 res1 <- run_gldrm(I(NResp/ClusterSize) ~ Trt, data=sh2, link="logit")
 # reduced dataset with frequency weights
 res2 <- run_gldrm(I(NResp/ClusterSize) ~ Trt, data=shelltox, link="logit", weights=shelltox$Freq)
 # rescale weights to a sum of 1
 res3 <- run_gldrm(I(NResp/ClusterSize) ~ Trt, data=shelltox, link="logit", weights=shelltox$Freq/sum(shelltox$Freq))
+
[…]

### 2021-03-04 nuweb/SPGLM.tex
Set up package infrastructure

@@
 \begin{flushleft} \small\label{scrap1}\raggedright\small
-\NWtarget{nuweb1}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {1}}$\equiv$
+\NWtarget{nuweb?}{} \verb@"R/SPGLM.R"@\nobreak\ {\footnotesize {?}}$\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@#' Fit semi-parametric GLM@\\
-\mbox{}\verb@#'@\\
+\mbox{}\verb@@\\
 \mbox{}\verb@#'@{\tt @}\verb@rdname spglm@\\
@@
 \mbox{}\verb@@\\
+\mbox{}\verb@#' Semi-parametric generalized linear model@\\
 \mbox{}\verb@spglm <- function(formula, data, subset, weights, link="logit", start=NULL, control=list(eps=0.001, maxit=100), ...){@\\
@@
 \mbox{}\verb@    @\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize \NWlink{nuweb2}{2}}$\,\rangle$}\verb@@\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape Fit model}\nobreak\ {\footnotesize \NWlink{nuweb4}{4}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize \NWlink{nuweb?}{?}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape Fit model}\nobreak\ {\footnotesize \NWlink{nuweb?}{?}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@@\\
@@
 \begin{flushleft} \small\label{scrap2}\raggedright\small
-\NWtarget{nuweb2}{} $\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize {2}}$\,\rangle\equiv$
+\NWtarget{nuweb?}{} $\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize {?}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
-\item \NWtxtMacroRefIn\ \NWlink{nuweb1}{1}.
+\item \NWtxtMacroRefIn\ \NWlink{nuweb?}{?}.
 
@@
 \begin{flushleft} \small\label{scrap3}\raggedright\small
-\NWtarget{nuweb4}{} $\langle\,${\itshape Fit model}\nobreak\ {\footnotesize {4}}$\,\rangle\equiv$
+\NWtarget{nuweb?}{} $\langle\,${\itshape Fit model}\nobreak\ {\footnotesize {?}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@ while (iter < control$maxit & difference > control$eps) {@\\
+\mbox{}\verb@ while (iter<itermax & difference>tol) {@\\
 \mbox{}\verb@    iter <- iter + 1@\\
@@
 \mbox{}\verb@    difference <- sum(abs(referencef0Pre - referencef0)) + sum(abs(betasPre - betas))@\\
+\mbox{}\verb@    # difference <- sum(abs(referencef0Pre - referencef0)) + sum(abs(tiltingWsPre - tiltingWs)) + sum(abs(betasPre - betas))@\\
+\mbox{}\verb@    cat("In iteration", iter, ", the corresponding log-likelihood is", observedLikelihood(CBData, referencef0, tiltingWs), ". \n")@\\
 \mbox{}\verb@  }@\\
+\mbox{}\verb@  # plot(referencef0, type="b")@\\
 \mbox{}\verb@  @\\

### 2021-03-04 SPreg_develop.R

@@ -1,4 +1,5 @@
 library(devtools)
+# create("../exchreg")
 source('z:/RForge/Nuweb.R')
 #source('/home/aszabo/RForge/Nuweb_linux.R')
 
@@ -11,6 +12,7 @@ source('R/SPreg.R', echo=FALSE)
 source('R/SPGLM.R', echo=FALSE)
 source('R/wgldrm.R', echo=FALSE)
 
+shell("cd c:/exchreg/ && texify --pdf --quiet --run-viewer --clean SPregress.tex")
 shell("cd c:/exchreg/ && texify --pdf --quiet --run-viewer --clean SPGLM.tex")
 
 data(shelltox)

### 2021-03-04 nuweb/SPreg_develop.R
Resolve merge conflicts

@@ -1,11 +1,6 @@
 library(devtools)
 <<<<<<< HEAD:nuweb/SPreg_develop.R
 source('nuweb/Nuweb.R')
-=======
-# create("../exchreg")
-source('z:/RForge/Nuweb.R')
-#source('/home/aszabo/RForge/Nuweb_linux.R')
->>>>>>> e91f19d9246eeff8aac80ff853ad9aaf61539ae9:SPreg_develop.R
 
 #  setting up the package and its infrastructure
 create("../exchreg")
@@ -16,7 +11,6 @@ use_build_ignore("nuweb")
 ex <- as.package("../exchreg")
 nuweb(ex)
 
-<<<<<<< HEAD:nuweb/SPreg_develop.R
 shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet --run-viewer  SPregress.tex")
 shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet --run-viewer  SPGLM.tex")
 shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet --run-viewer  wgldrm.tex")
@@ -24,10 +18,6 @@ shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet --run-viewer  wgldrm.tex")
 
 document(ex)
 load_all(ex)
-=======
-shell("cd c:/exchreg/ && texify --pdf --quiet --run-viewer --clean SPregress.tex")
-shell("cd c:/exchreg/ && texify --pdf --quiet --run-viewer --clean SPGLM.tex")
->>>>>>> e91f19d9246eeff8aac80ff853ad9aaf61539ae9:SPreg_develop.R
 
 ######### Testing wGLDRM
 run_gldrm <- 

### 2021-03-04 nuweb/SPreg_develop.R
Edit package DESCRIPTION

@@ -1,12 +1,11 @@
 library(devtools)
-<<<<<<< HEAD:nuweb/SPreg_develop.R
 source('nuweb/Nuweb.R')
 
 #  setting up the package and its infrastructure
 create("../exchreg")
 use_package("CorrBin")
 use_build_ignore("nuweb")
-
+use_gpl3_license()
 
 ex <- as.package("../exchreg")
 nuweb(ex)

### 2021-03-04 R/boric_acid.R
Add boric acid data

@@ -1,0 +1,16 @@
+#' Developmental toxicity of boric acid in mice
+#'
+#' A dataset containing the outcomes of a developmental toxicity experiment of boric acid by the National Toxicology
+#' Program. Pregnant mice were provided feed containing boric acid. At the end of the study, each mouse's uterus was 
+#' examined to evaluate the fetal outcomes.
+#'
+#' @format A data frame with 67 rows and 4 variables:
+#' \describe{
+#'   \item{Trt}{Exposure level to boric acid, %}
+#'   \item{ClusterSize}{Number of fetuses in the litter}
+#'   \item{NResp}{Number of malformed, dead, or resorped fetuses}
+#'   \item{Freq}{Number of litters with the same exposure/response pattern}
+#' }
+#' @source Heindel, J J and Price, C J and Schwetz, B A. The developmental toxicity of boric acid in mice, rats, and rabbits.
+#'	Environmental Health Perspectives, 102 Suppl 7,	1994, 107--12,	.
+"boric_acid"

### 2021-03-04 nuweb/SPreg_develop.R

@@ -6,6 +6,13 @@ create("../exchreg")
 use_package("CorrBin")
 use_build_ignore("nuweb")
 use_gpl3_license()
+desc::desc_add_author(given = "Xinran", family = "Qi", email = "xinqi@mcw.edu", role = "aut",
+               comment = NULL, file = ".", normalize = FALSE)
+
+# add data
+boric_acid <- CorrBin::read.CBData("z:/EOGeorge/Data/Binary/BoricAcidMousedata_processed.csv",
+                  sep=",", skip=1)
+use_data(boric_acid)
 
 ex <- as.package("../exchreg")
 nuweb(ex)

### 2021-03-07 nuweb/SPGLM.tex
Implementing E-step of spglm, calculating log-likelihood

--- 2021-03-04 nuweb/SPGLM.tex
@@
 \mbox{}\verb@#'@{\tt @}\verb@param link      a link function for the mean.@\\
-\mbox{}\verb@#'@{\tt @}\verb@param start     an optional list with elements named \code{beta}, \code{q}, and/or \code{mu1} giving starting values for estimation.
-If none or only some are specified, starting values are selected automatically.@\\
+\mbox{}\verb@#'@{\tt @}\verb@param mu0       an optional numeric value constraining the mean of the baseline distribution@\\
 \mbox{}\verb@#'@{\tt @}\verb@param control a list with parameters controlling the algorithm.@\\
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@spglm <- function(formula, data, subset, weights, link="logit", start=NULL, control=list(eps=0.001, maxit=100), ...){@\\
-\mbox{}\verb@    fam <- binomial(link=link)@\\
-\mbox{}\verb@    @\\
+\mbox{}\verb@spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL, @\\
+\mbox{}\verb@                  control=list(eps=0.001, maxit=100), ...){@\\
+\mbox{}\verb@@\\
 \mbox{}\verb@    @\hbox{$\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize \NWlink{nuweb2}{2}}$\,\rangle$}\verb@@\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape Fit model}\nobreak\ {\footnotesize \NWlink{nuweb4}{4}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape Fit model}\nobreak\ {\footnotesize \NWlink{nuweb3b}{3b}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@@\\
 \mbox{}\verb@    mt <- attr(mf, "terms")@\\
-\mbox{}\verb@    res <- list(coefficients = beta_new, q=q_new, niter = iter, loglik=logl,@\\
+\mbox{}\verb@    res <- list(coefficients = betas, f0=referencef0, mu0=mu0, niter = iter, loglik=mod$llik,@\\
 \mbox{}\verb@                link = link, call = mc, terms = mt,@\\
 \mbox{}\verb@                xlevels = .getXlevels(mt, mf),@\\
-\mbox{}\verb@                data_object=list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights),@\\
-\mbox{}\verb@                model_fun=model_fun)@\\
+\mbox{}\verb@                data_object=list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset))@\\
 \mbox{}\verb@    class(res) <- "spglm"@\\
@@
 \begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
+\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3a}{, 3a}.
 
@@
 \mbox{}\verb@    mc <- match.call(expand.dots = FALSE)@\\
-\mbox{}\verb@    m <- match(c("formula", "data", "subset", "weights"), names(mc), 0L)@\\
+\mbox{}\verb@    m <- match(c("formula", "data", "subset", "weights", "offset"), names(mc), 0L)@\\
 \mbox{}\verb@    m <- mc[c(1L,m)]@\\
@@
 \mbox{}\verb@    mm <- model.matrix(formula, data=mf)@\\
+\mbox{}\verb@    @\\
+\mbox{}\verb@    if (is.character(link)) {@\\
+\mbox{}\verb@      link <- stats::make.link(link)@\\
+\mbox{}\verb@    }@\\
+\mbox{}\verb@    else if (!is.list(link) || !(all(c("linkfun", "linkinv", @\\
+\mbox{}\verb@                                     "mu.eta") %in% names(link)))) {@\\
+\mbox{}\verb@      stop(paste0("link should be a character string or a list containing ", @\\
+\mbox{}\verb@                "functions named linkfun, linkinv, and mu.eta"))@\\
+\mbox{}\verb@    }@\\
 \mbox{}\verb@@\\
+\mbox{}\verb@    # extract offset@\\
+\mbox{}\verb@    offset <- as.vector(model.offset(mf))@\\
+\mbox{}\verb@    if (is.null(offset)) @\\
+\mbox{}\verb@      offset <- rep(0, nrow(mm))@\\
+\mbox{}\verb@    @\\
 \mbox{}\verb@    # extract weights@\\
@@
 \mbox{}\verb@        stop("negative weights not allowed")@\\
+\mbox{}\verb@     if (is.null(weights))@\\
+\mbox{}\verb@        weights <- rep(1, nrow(mm))@\\
 \mbox{}\verb@@{\NWsep}
@@
 
-\subsubsection{Observed data log-likelihood}
+\subsection{Observed data log-likelihood}
 Under the marginal compatibility assumption, the observed log-likelihood function is:
@@
     \ell (q_{y,N}, \boldsymbol{\beta}) &= \sum_{i=1}^I \log \{ q_{y_i,N_i} (\boldsymbol{Z}_i, \boldsymbol{\beta}) \} \\
-    &= \sum_{i=1}^I \log \bigg[ \binom{N_i}{y_i} \sum_{t=y}^{N-N_i+y_i} \dfrac{\binom{N-N_i}{t-y_i}}{\binom{N}{Y_i}} \times \dfrac{q_{t,N}^{(0)} \times \exp \{ \omega(\boldsymbol{Z}_i; \boldsymbol{\beta}) t \}} {\sum_{t'=0}^{N} q_{t',N}^{(0)} \times \exp \{ \omega(\boldsymbol{Z}_i; \boldsymbol{\beta}) t' \}}  \bigg].
+    &= \sum_{i=1}^I \log \bigg[ \binom{N_i}{y_i} \sum_{t=y}^{N-N_i+y_i} \dfrac{\binom{N-N_i}{t-y_i}}{\binom{N}{t}} \times \dfrac{q_{t,N}^{(0)} \times \exp \{ \omega(\boldsymbol{Z}_i; \boldsymbol{\beta}) t \}} {\sum_{t'=0}^{N} q_{t',N}^{(0)} \times \exp \{ \omega(\boldsymbol{Z}_i; \boldsymbol{\beta}) t' \}}  \bigg].
 \end{split}
@@
 
+\subsection{Model prediction and likelihood}
+We define internal function to calculate predicted values and the log-likelihood.
+
+\begin{flushleft} \small\label{scrap3}\raggedright\small
+\NWtarget{nuweb3a}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {3a}}$\equiv$
+\vspace{-1ex}
+\begin{list}{}{} \item
+\mbox{}\verb@@\\
+\mbox{}\verb@spglm_pred_mean <- function(beta, data_object, link){@\\
+\mbox{}\verb@  eta <- c(data_object$model_matrix %*% beta + data_object$offset)@\\
+\mbox{}\verb@  mu <- link$linkinv(eta)@\\
+\mbox{}\verb@  mu@\\
+\mbox{}\verb@}@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@spglm_loglik <- function(beta, f0, data_object, link){@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@    mu <- spglm_pred_mean(beta=beta, data_object=data_object, link=link)@\\
+\mbox{}\verb@    N <- max(data_object$n)@\\
+\mbox{}\verb@    nobs <- nrow(data_object$model_matrix)@\\
+\mbox{}\verb@    @\\
+\mbox{}\verb@    # ySptIndex is only used to calculate the log-likelihood, which we will not be using@\\
+\mbox{}\verb@    th <- getTheta(spt=(0:N)/N, f0=f0, mu=mu, weights=data_object$weights, ySptIndex=rep(1, nobs),@\\
+\mbox{}\verb@                   thetaStart=NULL, thetaControl=theta.control())@\\
+\mbox{}\verb@    @\\
+\mbox{}\verb@    hp <- sapply(0:N, function(t)dhyper(x=data_object$resp[,1], m=data_object$n, n=N-data_object$n, k=t))   @\\
+\mbox{}\verb@    llik_term <- rowSums(t(th$fTilt) * hp)@\\
+\mbox{}\verb@    llik <- llik_term %*% data_object$weights@\\
+\mbox{}\verb@    c(llik)@\\
+\mbox{}\verb@}@\\
+\mbox{}\verb@@{\NWsep}
+\end{list}
+\vspace{-1.5ex}
+\footnotesize
+\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
+\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3a}{, 3a}.
+
+\item{}
+\end{list}
+\vspace{4ex}
+\end{flushleft}
 It is difficult to utilize the observed log-likelihood directly for parameter estimation.
@@
 
-\subsubsection{Missing data setup}
+\begin{flushleft} \small\label{scrap4}\raggedright\small
+\NWtarget{nuweb3b}{} $\langle\,${\itshape Fit model}\nobreak\ {\footnotesize {3b}}$\,\rangle\equiv$
+\vspace{-1ex}
+\begin{list}{}{} \item
+\mbox{}\verb@@\\
+\mbox{}\verb@ @\hbox{$\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize \NWlink{nuweb6}{6}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@ @\hbox{$\langle\,${\itshape Set up for E-step}\nobreak\ {\footnotesize \NWlink{nuweb4}{4}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@ @\\
+\mbox{}\verb@ iter <- 0@\\
+\mbox{}\verb@ difference <- 100@\\
+\mbox{}\verb@ while (iter < control$maxit & difference > control$eps) {@\\
+\mbox{}\verb@    iter <- iter + 1@\\
+\mbox{}\verb@    referencef0Pre <- referencef0@\\
+\mbox{}\verb@    betasPre <- betas@\\
+\mbox{}\verb@    @\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape E-step}\nobreak\ {\footnotesize \NWlink{nuweb5a}{5a}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape M-step}\nobreak\ {\footnotesize \NWlink{nuweb5b}{5b}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\\
+\mbox{}\verb@    difference <- sum(abs(referencef0Pre - referencef0)) + sum(abs(betasPre - betas))@\\
+\mbox{}\verb@  }@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@@{\NWsep}
+\end{list}
+\vspace{-1.5ex}
+\footnotesize
+\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
+\item \NWtxtMacroRefIn\ \NWlink{nuweb1}{1}.
+
+\item{}
+\end{list}
+\vspace{4ex}
+\end{flushleft}
+\subsection{Missing data setup}
 Stefanescu et al (2003) have proved that the marginal compatibility assumption is equivalent to assuming that each cluster of size $N'$ arises from a cluster of maximal cluster size $N$ with some binary observations missing completely missing at random (MCAR).
@@
 
-\subsubsection{Expectation step}
-%For complete data in which all clusters have $N$ observations, the marginal probability of achieving $y$ events out of $N$ observations within a cluster is denoted with parameter $q_{y,N}$. 
+Equation \ref{E:logLikelihoodCompleteMapFromObs} can be interpreted as the log-likelihood of a model with fixed cluster size $N$ for an expanded data set: each observation with cluster size $N_i$ and number of responses $y_i$ is expanded to $(N+1)$ replicates with $0, 1, \ldots, N$, responses, respectively.
+
+\begin{flushleft} \small\label{scrap5}\raggedright\small
+\NWtarget{nuweb4}{} $\langle\,${\itshape Set up for E-step}\nobreak\ {\footnotesize {4}}$\,\rangle\equiv$
+\vspace{-1ex}
+\begin{list}{}{} \item
+\mbox{}\verb@@\\
+\mbox{}\verb@  # replace each cluster with N+1 clusters of size N@\\
+\mbox{}\verb@  new <- cbind(y=0:N)@\\
+\mbox{}\verb@  rep_idx <- rep(1:nrow(Y), each=nrow(new))@\\
+\mbox{}\verb@  Y2 <- cbind(1:nrow(Y), Y)[rep_idx,]@\\
+\mbox{}\verb@  colnames(Y2) <- c("i", "resp","nonresp")@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@  rep_idx2 <- rep(1:nrow(new), times=nrow(Y))@\\
+\mbox{}\verb@  Ycomb <- cbind(Y2, new[rep_idx2, ,drop=FALSE])@\\
+\mbox{}\verb@  # select possible combinations@\\
+\mbox{}\verb@  possible <- (Ycomb[,"y"] >= Ycomb[,"resp"]) & (N-Ycomb[,"y"] >= Ycomb[,"nonresp"])@\\
+\mbox{}\verb@  Ycomb <- Ycomb[possible,]@\\
+\mbox{}\verb@ @\\
+\mbox{}\verb@  mm2 <- mm[Ycomb[,"i"], ,drop=FALSE]@\\
+\mbox{}\verb@  weights2 <- weights[Ycomb[,"i"]]@\\
+\mbox{}\verb@  offset2 <- offset[Ycomb[,"i"]]@\\
+\mbox{}\verb@@{\NWsep}
+\end{list}
+\vspace{-1.5ex}
+\footnotesize
+\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
+\item \NWtxtMacroRefIn\ \NWlink{nuweb3b}{3b}.
+
+\item{}
+\end{list}
+\vspace{4ex}
+\end{flushleft}
+\subsection{Expectation step}
+
 We take the expectation of complete data log-likelihood \eqref{E:logLikelihoodCompleteMapFromObs} given parameters from the previous $k^{th}$ iteration, which is shown as follows:
@@
 
-\subsubsection{Maximization step}
+The value of $q_{y,N}^{(k)}((\boldsymbol{Z}_i; \boldsymbol{\beta}^{(k)})$ can be obtained from ${q_{y,N}^{(0)}}^{(k)}$ using \ref{M:densitySPGLM}.
+But it is also returned by \texttt{gldrmFit} as \texttt{fTiltMatrix}.
+
+\begin{flushleft} \small\label{scrap6}\raggedright\small
+\NWtarget{nuweb5a}{} $\langle\,${\itshape E-step}\nobreak\ {\footnotesize {5a}}$\,\rangle\equiv$
+\vspace{-1ex}
+\begin{list}{}{} \item
+\mbox{}\verb@@\\
+\mbox{}\verb@  # convert fTiltMatrix to long vector, watching out for potentially different support@\\
+\mbox{}\verb@  y_idx <- match(Ycomb[,"y"], spt)@\\
+\mbox{}\verb@  fTilt2 <- ifelse(!is.na(y_idx), fTiltMatrix[cbind(Ycomb[,"i"], y_idx)], 0)@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  # numerator of weights@\\
+\mbox{}\verb@  pp_num <- choose(n=Ycomb[,"y"], k=Ycomb[,"resp"]) * @\\
+\mbox{}\verb@           choose(n=N-Ycomb[,"y"], k=Ycomb[,"nonresp"]) * @\\
+\mbox{}\verb@           fTilt2@\\
+\mbox{}\verb@  # denominator @\\
+\mbox{}\verb@  pp_denom <- tapply(pp_num, list(i=Ycomb[,"i"]), sum, simplify=TRUE)@\\
+\mbox{}\verb@  pp <- c(pp_num / pp_denom[Ycomb[,"i"]])@\\
+\mbox{}\verb@  if (!is.null(weights)) pp <- weights2 * pp@\\
+\mbox{}\verb@@{\NWsep}
+\end{list}
+\vspace{-1.5ex}
+\footnotesize
+\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
+\item \NWtxtMacroRefIn\ \NWlink{nuweb3b}{3b}.
+
+\item{}
+\end{list}
+\vspace{4ex}
+\end{flushleft}
+\subsection{Maximization step}
 
@@
 
-\begin{flushleft} \small\label{scrap3}\raggedright\small
-\NWtarget{nuweb4}{} $\langle\,${\itshape Fit model}\nobreak\ {\footnotesize {4}}$\,\rangle\equiv$
+Default settings are used for the algorithm, except the starting values is updated based on the previous iteration
+
+\begin{flushleft} \small\label{scrap7}\raggedright\small
+\NWtarget{nuweb5b}{} $\langle\,${\itshape M-step}\nobreak\ {\footnotesize {5b}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@ while (iter < control$maxit & difference > control$eps) {@\\
-\mbox{}\verb@    iter <- iter + 1@\\
-\mbox{}\verb@    referencef0Pre <- referencef0@\\
-\mbox{}\verb@    tiltingWsPre <- tiltingWs@\\
-\mbox{}\verb@    betasPre <- betas@\\
-\mbox{}\verb@    # passing parameter updates to the next iteration@\\
-\mbox{}\verb@    if (iter == 1) {@\\
-\mbox{}\verb@      startUpdates <- list(iter = iter,@\\
-\mbox{}\verb@                           f0StartValue = NULL,@\\
-\mbox{}\verb@                           thetaStartValue = NULL,@\\
-\mbox{}\verb@                           betaStartValue = NULL)@\\
-\mbox{}\verb@    } else {@\\
-\mbox{}\verb@      startUpdates <- list(iter = iter,@\\
-\mbox{}\verb@                           f0StartValue = referencef0Pre,@\\
-\mbox{}\verb@                           thetaStartValue = tiltingWsPre,@\\
-\mbox{}\verb@                           betaStartValue = betasPre)@\\
-\mbox{}\verb@    }@\\
-\mbox{}\verb@    # EM NR algorithm@\\
-\mbox{}\verb@    res <- gldrMaximize(DesignMatrix.CONST, CBData, referencef0Pre, tiltingWsPre, startUpdates)@\\
-\mbox{}\verb@    referencef0 <- res$density.ref @\\
-\mbox{}\verb@    # notice that this tilting parameter estimates are w.r.t the intermediate dataset, original dataset extended by max(cluster size)+1 times@\\
-\mbox{}\verb@    tiltingWs <- res$tiltParam[(0:(nrow(CBData)-1))*length(referencef0)+1]@\\
-\mbox{}\verb@    betas <- res$regressionEst@\\
-\mbox{}\verb@    # monitor parameter estimates for baseline distribution, tilting parameters, regression coefficients@\\
-\mbox{}\verb@    difference <- sum(abs(referencef0Pre - referencef0)) + sum(abs(betasPre - betas))@\\
-\mbox{}\verb@  }@\\
+\mbox{}\verb@  gldrmControl0 <- gldrm.control(returnfTiltMatrix = TRUE, returnf0ScoreInfo = FALSE, print=FALSE,@\\
+\mbox{}\verb@                                betaStart = betasPre, f0Start = referencef0Pre)@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@  mod <- gldrmFit(x = mm2, y=Ycomb[,"y"]/N, linkfun=link$linkfun, linkinv = link$linkinv,@\\
+\mbox{}\verb@                  mu.eta = link$mu.eta, mu0 = mu0, offset = offset2, weights = pp, @\\
+\mbox{}\verb@                  gldrmControl = gldrmControl0,  thetaControl=theta.control(),@\\
+\mbox{}\verb@                  betaControl=beta.control(), f0Control=f0.control())@\\
 \mbox{}\verb@                  @\\
-\mbox{}\verb@  list(referencef0 = referencef0,@\\
-\mbox{}\verb@       tiltingWs = tiltingWs,@\\
-\mbox{}\verb@       betas = betas,@\\
-\mbox{}\verb@       EMgldrmFitRes = res)@\\
+\mbox{}\verb@  fTiltMatrix <- mod$fTiltMatrix@\\
+\mbox{}\verb@  betas <- mod$beta@\\
+\mbox{}\verb@  referencef0 <- mod$f0@\\
+\mbox{}\verb@  spt <- round(mod$spt * N)@\\
 \mbox{}\verb@@{\NWsep}
@@
 \begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
-\item \NWtxtMacroRefIn\ \NWlink{nuweb1}{1}.
+\item \NWtxtMacroRefIn\ \NWlink{nuweb3b}{3b}.
+
+\item{}
+\end{list}
+\vspace{4ex}
+\end{flushleft}
+\subsection{Starting values}
+We start the regression coefficients are set to 0, so $q^{(0)}_N$ would be the overall estimate under marginally compatibility. The default value of \texttt{mu0} is set to the mean of $y_i/n_i$.
+
+\begin{flushleft} \small\label{scrap8}\raggedright\small
+\NWtarget{nuweb6}{} $\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize {6}}$\,\rangle\equiv$
+\vspace{-1ex}
+\begin{list}{}{} \item
+\mbox{}\verb@@\\
+\mbox{}\verb@  N <- max(rowSums(Y))@\\
+\mbox{}\verb@  nobs <- nrow(mm)@\\
+\mbox{}\verb@  p <- ncol(mm)@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  betas <- rep(0, times= p)@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@  pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), @\\
+\mbox{}\verb@                    trt="Trt", clustersize="ClusterSize", nresp="NResp")@\\
+\mbox{}\verb@  est <- mc.est(pooled)@\\
+\mbox{}\verb@  referencef0 <- est$Prob[est$ClusterSize == N]@\\
+\mbox{}\verb@  # ensure all positive values@\\
+\mbox{}\verb@  referencef0 <- (referencef0 + 1e-6)/(1+(N+1)*1e-6)@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  fTiltMatrix <- matrix(rep(referencef0, times=nobs), byrow=TRUE, nrow=nobs, ncol=N+1)@\\
+\mbox{}\verb@  spt <- 0:N@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@  if (is.null(mu0))@\\
+\mbox{}\verb@    mu0 <- weighted.mean(Y[,1]/rowSums(Y), weights)@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@@{\NWsep}
+\end{list}
+\vspace{-1.5ex}
+\footnotesize
+\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
+\item \NWtxtMacroRefIn\ \NWlink{nuweb3b}{3b}.
 

### 2021-03-07 nuweb/wgldrm.tex
Set up package infrastructure
Implementing E-step of spglm, calculating log-likelihood

@@
 \mbox{}\verb@    maxiter <- gldrmControl$maxiter@\\
-\mbox{}\verb@    returnHess <- gldrmControl$returnHess@\\
 \mbox{}\verb@    returnfTiltMatrix <- gldrmControl$returnfTiltMatrix@\\
@@
 \mbox{}\verb@    # create a weight matrix for score.logT1 calculation@\\
-\mbox{}\verb@  weightsMatrix <- matrix(0, nrow=length(y), ncol=length(unique(y)), @\\
-\mbox{}\verb@        dimnames=list(paste0("obs", 1:length(y)), paste0("response", sort(unique(y)))))@\\
-\mbox{}\verb@  # fulfillment of weightMatrix@\\
-\mbox{}\verb@  for (obs in 1:length(y)) {@\\
-\mbox{}\verb@    weightsMatrix[obs, y[obs]==sort(unique(y))] <- weights[obs]@\\
-\mbox{}\verb@  }@\\
+\mbox{}\verb@    weightsMatrix <- matrix(0, nrow=n, ncol=length(spt))@\\
+\mbox{}\verb@    weightsMatrix[cbind(1:n, ySptIndex)] <- weights@\\
+\mbox{}\verb@    @\\
 \mbox{}\verb@    # weighted version of "sptFreq" @\\
 \mbox{}\verb@    sptFreq.weighted <- colSums(weightsMatrix)@\\
-\mbox{}\verb@@\\
-\mbox{}\verb@@\\
 \mbox{}\verb@@\\
@@
 \mbox{}\verb@        ## update f0 and theta, with fixed beta (mu)@\\
-\mbox{}\verb@        ff <- getf0(y=y, spt=spt, ySptIndex=ySptIndex, sptFreq=sptFreq,@\\
+\mbox{}\verb@        ff <- getf0(y=y, spt=spt, ySptIndex=ySptIndex, @\\
 \mbox{}\verb@                    weights=weights, sptFreq.weighted=sptFreq.weighted, mu=mu, mu0=mu0, f0Start=f0, thStart=th,@\\
@@
 \mbox{}\verb@#'@{\tt @}\verb@param ySptIndex Index of each \code{y} value within \code{spt}.@\\
-\mbox{}\verb@#'@{\tt @}\verb@param sptFreq Vector containing frequency of each \code{spt} value.@\\
-\mbox{}\verb@#'@{\tt @}\verb@param sampprobs Optional matrix of sampling probabilities.@\\
+\mbox{}\verb@#'@{\tt @}\verb@param sptFreq.weighted Vector containing weighted frequency of each \code{spt} value.@\\
 \mbox{}\verb@#'@{\tt @}\verb@param mu Fitted mean for each observation. Only used if \code{sampprobs=NULL}.@\\
@@
 \mbox{}\verb@#' @{\tt @}\verb@keywords internal@\\
-\mbox{}\verb@getf0 <- function(y, spt, ySptIndex, sptFreq, weights, sptFreq.weighted, mu, mu0, f0Start, thStart,@\\
+\mbox{}\verb@getf0 <- function(y, spt, ySptIndex, weights, sptFreq.weighted, mu, mu0, f0Start, thStart,@\\
 \mbox{}\verb@                  thetaControl=theta.control(), f0Control=f0.control(), trace=FALSE)@\\

### 2021-03-07 R/SPGLM.R

@@ -7,22 +7,22 @@
 #'@param subset an optional vector specifying a subset of observations to be used.
 #'@param weight an optional vector specifying observation weights.
 #'@param link     a link function for the mean.
-#'@param start    an optional list with elements named \code{beta}, \code{q}, and/or \code{mu1} giving starting values for estimation. If none or only some are specified, starting values are selected automatically.
+#'@param mu0      an optional numeric value constraining the mean of the baseline distribution
 #'@param control        a list with parameters controlling the algorithm.
 #'@return an object of class \code{spglm} with the fitted model.
 #'@export
 #' @importFrom stats terms model.matrix
 
-spglm <- function(formula, data, subset, weights, link="logit", start=NULL, control=list(eps=0.001, maxit=100), ...){
-    fam <- binomial(link=link)
-    
+spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL, 
+                  control=list(eps=0.001, maxit=100), ...){
+
     
        if (missing(formula) || (length(formula) != 3L))
             stop("'formula' missing or incorrect")
        if (missing(data))
             data <- environment(formula)
         mc <- match.call(expand.dots = FALSE)
-        m <- match(c("formula", "data", "subset", "weights"), names(mc), 0L)
+        m <- match(c("formula", "data", "subset", "weights", "offset"), names(mc), 0L)
         m <- mc[c(1L,m)]
         if (is.matrix(eval(m$data, parent.frame())))
             m$data <- as.data.frame(data)
@@ -35,55 +35,139 @@ spglm <- function(formula, data, subset, weights, link="logit", start=NULL, cont
 
         # create model matrix
         mm <- model.matrix(formula, data=mf)
+        
+        if (is.character(link)) {
+          link <- stats::make.link(link)
+        }
+        else if (!is.list(link) || !(all(c("linkfun", "linkinv", 
+                                         "mu.eta") %in% names(link)))) {
+          stop(paste0("link should be a character string or a list containing ", 
+                    "functions named linkfun, linkinv, and mu.eta"))
+        }
 
+        # extract offset
[…]

### 2021-03-07 R/wgldrm.R

@@ -23,7 +23,6 @@ gldrmFit <- function(x, y, linkfun, linkinv, mu.eta, mu0=NULL, offset=NULL, weig
               gldrmControl() function.")
     eps <- gldrmControl$eps
     maxiter <- gldrmControl$maxiter
-    returnHess <- gldrmControl$returnHess
     returnfTiltMatrix <- gldrmControl$returnfTiltMatrix
     returnf0ScoreInfo <- gldrmControl$returnf0ScoreInfo
     print <- gldrmControl$print
@@ -37,17 +36,12 @@ gldrmFit <- function(x, y, linkfun, linkinv, mu.eta, mu0=NULL, offset=NULL, weig
     # sptFreq <- table(ySptIndex)
     # attributes(sptFreq) <- NULL
 
-  # create a weight matrix for score.logT1 calculation
-  weightsMatrix <- matrix(0, nrow=length(y), ncol=length(unique(y)), 
-        dimnames=list(paste0("obs", 1:length(y)), paste0("response", sort(unique(y)))))
-  # fulfillment of weightMatrix
-  for (obs in 1:length(y)) {
-    weightsMatrix[obs, y[obs]==sort(unique(y))] <- weights[obs]
-  }
-  # weighted version of "sptFreq" 
-  sptFreq.weighted <- colSums(weightsMatrix)
-
-
+    # create a weight matrix for score.logT1 calculation
+    weightsMatrix <- matrix(0, nrow=n, ncol=length(spt))
+    weightsMatrix[cbind(1:n, ySptIndex)] <- weights
+    
+    # weighted version of "sptFreq" 
+    sptFreq.weighted <- colSums(weightsMatrix)
 
     ## Initialize offset
     if (is.null(offset))
@@ -127,7 +121,7 @@ gldrmFit <- function(x, y, linkfun, linkinv, mu.eta, mu0=NULL, offset=NULL, weig
         beta <- bb$beta
 
         ## update f0 and theta, with fixed beta (mu)
-        ff <- getf0(y=y, spt=spt, ySptIndex=ySptIndex, sptFreq=sptFreq,
+        ff <- getf0(y=y, spt=spt, ySptIndex=ySptIndex, 
                     weights=weights, sptFreq.weighted=sptFreq.weighted, mu=mu, mu0=mu0, f0Start=f0, thStart=th,
                     thetaControl=thetaControl, f0Control=f0Control, trace=FALSE)
         th <- ff$th
[…]

### 2021-03-07 nuweb/SPreg_develop.R

@@ -25,6 +25,19 @@ shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet --run-viewer  wgldrm.tex")
 document(ex)
 load_all(ex)
 
+## test SPGLM
+data(boric_acid)
+m <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=boric_acid, link="logit",
+           weights=boric_acid$Freq)
+
+ba2 <- boric_acid[rep(1:nrow(boric_acid), boric_acid$Freq),]
+m2 <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=ba2, link="logit")
+
+all.equal(coef(m), coef(m2))
+
+a <- spglm_pred_mean(m$coefficients, m$data_object, m$link)
+b <- spglm_loglik(beta=m$coefficients, f0=m$f0, m$data_object, m$link)
+
 ######### Testing wGLDRM
 run_gldrm <- 
 function (formula, data = NULL, link = "identity", mu0 = NULL,  weights=NULL,
@@ -139,15 +152,14 @@ b <- sprr(cbind(NResp, ClusterSize-NResp) ~ Trt+log(ClusterSize),
                 control=list(eps=0.01,maxit=100))
 
 #
-ba <- read.CBData("z:/EOGeorge/Data/Binary/BoricAcidMousedata_processed.csv",
-                  sep=",", skip=1)
+data("boric_acid")
 ba.mod <- sprr(cbind(NResp, ClusterSize-NResp) ~ Trt,
-          data=ba, weights=ba$Freq,link="log",
+          data=boric_acid, weights=boric_acid$Freq,link="log",
           control=list(eps=0.01,maxit=100))
 
 
 ba.mod1 <- sprr(cbind(NResp, ClusterSize-NResp) ~ Trt + log(ClusterSize),
-               data=ba, weights=ba$Freq, link="log",
+               data=boric_acid, weights=boric_acid$Freq, link="log",
                control=list(eps=0.01,maxit=100),
                start=list(q=ba.mod$q, beta=c(coef(ba.mod), 0),
                           mu1=lambda_from_p(ba.mod$q)[2]))

### 2021-03-07 nuweb/SPGLM.tex
Monitor log-likelihood during convergence

@@
 \mbox{}\verb@    mt <- attr(mf, "terms")@\\
-\mbox{}\verb@    res <- list(coefficients = betas, f0=referencef0, mu0=mu0, niter = iter, loglik=mod$llik,@\\
+\mbox{}\verb@    res <- list(coefficients = betas, f0=referencef0, mu0=mu0, niter = iter, loglik=llik,@\\
 \mbox{}\verb@                link = link, call = mc, terms = mt,@\\
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@ @\hbox{$\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize \NWlink{nuweb6}{6}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@ @\hbox{$\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize \NWlink{nuweb6b}{6b}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@ @\hbox{$\langle\,${\itshape Set up for E-step}\nobreak\ {\footnotesize \NWlink{nuweb4}{4}}$\,\rangle$}\verb@@\\
@@
 \mbox{}\verb@    betasPre <- betas@\\
+\mbox{}\verb@    llikPre <- llik@\\
 \mbox{}\verb@    @\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape E-step}\nobreak\ {\footnotesize \NWlink{nuweb5a}{5a}}$\,\rangle$}\verb@@\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape M-step}\nobreak\ {\footnotesize \NWlink{nuweb5b}{5b}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape E-step}\nobreak\ {\footnotesize \NWlink{nuweb5}{5}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape M-step}\nobreak\ {\footnotesize \NWlink{nuweb6a}{6a}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@    @\\
-\mbox{}\verb@    difference <- sum(abs(referencef0Pre - referencef0)) + sum(abs(betasPre - betas))@\\
+\mbox{}\verb@    difference <- abs(llik - llikPre)@\\
 \mbox{}\verb@  }@\\
@@
 \mbox{}\verb@  @\\
+\mbox{}\verb@  obs_start <- match(1:nobs, Ycomb[,"i"])@\\
+\mbox{}\verb@ @\\
 \mbox{}\verb@  mm2 <- mm[Ycomb[,"i"], ,drop=FALSE]@\\
@@
 \begin{flushleft} \small\label{scrap6}\raggedright\small
-\NWtarget{nuweb5a}{} $\langle\,${\itshape E-step}\nobreak\ {\footnotesize {5a}}$\,\rangle\equiv$
+\NWtarget{nuweb5}{} $\langle\,${\itshape E-step}\nobreak\ {\footnotesize {5}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap7}\raggedright\small
-\NWtarget{nuweb5b}{} $\langle\,${\itshape M-step}\nobreak\ {\footnotesize {5b}}$\,\rangle\equiv$
+\NWtarget{nuweb6a}{} $\langle\,${\itshape M-step}\nobreak\ {\footnotesize {6a}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@                  @\\
-\mbox{}\verb@  fTiltMatrix <- mod$fTiltMatrix@\\
+\mbox{}\verb@  fTiltMatrix <- mod$fTiltMatrix[obs_start,]@\\
 \mbox{}\verb@  betas <- mod$beta@\\
@@
 \mbox{}\verb@  spt <- round(mod$spt * N)@\\
+\mbox{}\verb@  llik <- c(rowSums(fTiltMatrix * hp) %*% weights)@\\
 \mbox{}\verb@@{\NWsep}
@@
 \begin{flushleft} \small\label{scrap8}\raggedright\small
-\NWtarget{nuweb6}{} $\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize {6}}$\,\rangle\equiv$
+\NWtarget{nuweb6b}{} $\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize {6b}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@  @\\
-\mbox{}\verb@  betas <- rep(0, times= p)@\\
-\mbox{}\verb@@\\
+\mbox{}\verb@  betas <- NULL@\\
 \mbox{}\verb@  pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), @\\
@@
 \mbox{}\verb@  @\\
+\mbox{}\verb@  # hypergeometric terms for log-likelihood calculation@\\
+\mbox{}\verb@  hp <- sapply(0:N, function(t)dhyper(x=Y[,1], m=rowSums(Y), n=N-rowSums(Y), k=t))   @\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  # initial log-likelihood@\\
+\mbox{}\verb@  llik <- rowSums(fTiltMatrix * hp) %*% weights@\\
+\mbox{}\verb@@\\
 \mbox{}\verb@@{\NWsep}

### 2021-03-07 R/SPGLM.R

@@ -65,8 +65,7 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
        nobs <- nrow(mm)
        p <- ncol(mm)
        
-       betas <- rep(0, times= p)
-
+       betas <- NULL
        pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), 
                          trt="Trt", clustersize="ClusterSize", nresp="NResp")
        est <- mc.est(pooled)
@@ -80,6 +79,12 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
        if (is.null(mu0))
          mu0 <- weighted.mean(Y[,1]/rowSums(Y), weights)
        
+       # hypergeometric terms for log-likelihood calculation
+       hp <- sapply(0:N, function(t)dhyper(x=Y[,1], m=rowSums(Y), n=N-rowSums(Y), k=t))   
+       
+       # initial log-likelihood
+       llik <- rowSums(fTiltMatrix * hp) %*% weights
+
      
      
        # replace each cluster with N+1 clusters of size N
@@ -93,6 +98,8 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
        # select possible combinations
        possible <- (Ycomb[,"y"] >= Ycomb[,"resp"]) & (N-Ycomb[,"y"] >= Ycomb[,"nonresp"])
        Ycomb <- Ycomb[possible,]
+       
+       obs_start <- match(1:nobs, Ycomb[,"i"])
       
        mm2 <- mm[Ycomb[,"i"], ,drop=FALSE]
        weights2 <- weights[Ycomb[,"i"]]
@@ -105,6 +112,7 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
         iter <- iter + 1
         referencef0Pre <- referencef0
         betasPre <- betas
+        llikPre <- llik
         
         
           # convert fTiltMatrix to long vector, watching out for potentially different support
@@ -129,19 +137,20 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
                           gldrmControl = gldrmControl0,  thetaControl=theta.control(),
                           betaControl=beta.control(), f0Control=f0.control())
                           
-          fTiltMatrix <- mod$fTiltMatrix
+          fTiltMatrix <- mod$fTiltMatrix[obs_start,]
[…]

### 2021-03-07 nuweb/SPreg_develop.R

@@ -28,15 +28,28 @@ load_all(ex)
 ## test SPGLM
 data(boric_acid)
 m <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=boric_acid, link="logit",
-           weights=boric_acid$Freq)
+           weights=boric_acid$Freq, control = list(eps=1e-6, maxit=1000))
 
 ba2 <- boric_acid[rep(1:nrow(boric_acid), boric_acid$Freq),]
-m2 <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=ba2, link="logit")
+m2 <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=ba2, link="logit", control = list(eps=1e-6, maxit=1000))
 
 all.equal(coef(m), coef(m2))
+all.equal(m$loglik, m2$loglik)
 
 a <- spglm_pred_mean(m$coefficients, m$data_object, m$link)
 b <- spglm_loglik(beta=m$coefficients, f0=m$f0, m$data_object, m$link)
+b2 <- spglm_loglik(beta=m2$coefficients, f0=m2$f0, m$data_object, m$link)
+
+# Hessian matrix calculation
+library(numDeriv)
+my_ll <- function(x, data_object, link){
+  p <- ncol(data_object$model_matrix)
+  spglm_loglik(beta=x[1:p], f0 = exp(x[-(1:p)]), data_object, link)
+}
+hess <- hessian(func= my_ll,  x=c(coef(m), log(m$f0)), data_object=m$data_object, link=m$link)
+# inverse of Hessian matrix == Fisher information (used for SE estimation) for regression coefficients
+fish <- solve(-hess)
+
 
 ######### Testing wGLDRM
 run_gldrm <- 

### 2021-03-08 nuweb/wgldrm.tex
Remove sampling-weighted terms from gldrm

@@
 \mbox{}\verb@#' distribution for each observation.@\\
-\mbox{}\verb@#' \item \code{fTiltSW} Matrix containing the exponentially tilted distribution for each@\\
-\mbox{}\verb@#' observation, conditional on that observation being sampled, i.e. f(y|X=x, S=1).@\\
-\mbox{}\verb@#' If \code{sampprobs=NULL}, then \code{fTiltSW} matches \code{fTilt}.@\\
-\mbox{}\verb@#' \item \code{bPrimeSW} Vector containing the mean for each observation, conditional@\\
-\mbox{}\verb@#' on that observation being sampled. If \code{sampprobs=NULL}, then \code{bPrimeSW}@\\
-\mbox{}\verb@#' matches \code{bPrime}.@\\
-\mbox{}\verb@#' \item \code{bPrime2SW} Vector containing the variance for each observation, conditional@\\
-\mbox{}\verb@#' on that observation being sampled. If \code{sampprobs=NULL}, then \code{bPrime2SW}@\\
 \mbox{}\verb@#' matches \code{bPrime2}.@\\
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@    fTiltSW <- fTilt@\\
-\mbox{}\verb@    bPrimeSW <- bPrime@\\
-\mbox{}\verb@    bPrime2SW <- bPrime2@\\
-\mbox{}\verb@  @\\
-\mbox{}\verb@@\\
 \mbox{}\verb@    ## Calculate log-likelihood@\\
-\mbox{}\verb@   # llik <- sum(log(fTiltSW[cbind(ySptIndex, seq_along(ySptIndex))]))@\\
-\mbox{}\verb@    fTiltSW.extracted <- fTiltSW[cbind(ySptIndex, seq_along(ySptIndex))]@\\
-\mbox{}\verb@    llik <- sum(weights[fTiltSW.extracted>0] * log(fTiltSW.extracted[fTiltSW.extracted>0]))@\\
+\mbox{}\verb@    fTilt.extracted <- fTilt[cbind(ySptIndex, seq_along(ySptIndex))]@\\
+\mbox{}\verb@    llik <- sum(weights[fTilt.extracted>0] * log(fTilt.extracted[fTilt.extracted>0]))@\\
 \mbox{}\verb@ @\\
 \mbox{}\verb@    list(theta=theta, fTilt=fTilt, bPrime=bPrime, bPrime2=bPrime2,@\\
-\mbox{}\verb@         fTiltSW=fTiltSW, bPrimeSW=bPrimeSW, bPrime2SW=bPrime2SW,@\\
 \mbox{}\verb@         llik=llik, conv=conv, iter=iter)@\\
@@
 \mbox{}\verb@        @\\
-\mbox{}\verb@        fTiltSWSums <- rowSums(th$fTiltSW)@\\
-\mbox{}\verb@        fTiltSWSumsWeighted <- apply(th$fTiltSW, MARGIN=1, function(x) sum(weights * x)) @\\
-\mbox{}\verb@        smmfTiltSW <- smm * th$fTiltSW@\\
-\mbox{}\verb@        ystd <- ymm / th$bPrime2SW@\\
+\mbox{}\verb@        fTiltSums <- rowSums(th$fTilt)@\\
+\mbox{}\verb@        fTiltSumsWeighted <- apply(th$fTilt, MARGIN=1, function(x) sum(weights * x)) @\\
+\mbox{}\verb@        smmfTilt <- smm * th$fTilt@\\
+\mbox{}\verb@        ystd <- ymm / th$bPrime2@\\
 \mbox{}\verb@        ystdWeighted <- weights * ystd@\\
@@
 \mbox{}\verb@        score.logT1 <- sptFreq.weighted@\\
-\mbox{}\verb@        score.logT2 <- fTiltSWSumsWeighted@\\
-\mbox{}\verb@        score.logT3 <- c(smmfTiltSW %*% ystdWeighted)@\\
+\mbox{}\verb@        score.logT2 <- fTiltSumsWeighted@\\
+\mbox{}\verb@        score.logT3 <- c(smmfTilt %*% ystdWeighted)@\\
 \mbox{}\verb@        score.log <- score.logT1 - score.logT2 - score.logT3@\\
@@
 \mbox{}\verb@        if (iter == 1) {@\\
-\mbox{}\verb@            d1 <- min(fTiltSWSumsWeighted)  # max inverse diagonal of first information term, on log scale@\\
+\mbox{}\verb@            d1 <- min(fTiltSumsWeighted)  # max inverse diagonal of first information term, on log scale@\\
 \mbox{}\verb@            d2 <- max(abs(score.log)) / maxlogstep@\\
@@
 \mbox{}\verb@    # Final score calculation@\\
-\mbox{}\verb@        smm <- outer(spt, th$bPrimeSW, "-")@\\
-\mbox{}\verb@        ymm <- y - th$bPrimeSW@\\
+\mbox{}\verb@        smm <- outer(spt, th$bPrime, "-")@\\
+\mbox{}\verb@        ymm <- y - th$bPrime@\\
 \mbox{}\verb@        yeqmu <- which(abs(ymm) < 1e-15)@\\
 \mbox{}\verb@        @\\
-\mbox{}\verb@          fTiltSWSums <- rowSums(th$fTiltSW)@\\
-\mbox{}\verb@  fTiltSWSumsWeighted <- apply(th$fTiltSW, MARGIN=1, function(x) sum(weights * x)) @\\
-\mbox{}\verb@  smmfTiltSW <- smm * th$fTiltSW@\\
-\mbox{}\verb@  ystd <- ymm / th$bPrime2SW@\\
+\mbox{}\verb@          fTiltSums <- rowSums(th$fTilt)@\\
+\mbox{}\verb@  fTiltSumsWeighted <- apply(th$fTilt, MARGIN=1, function(x) sum(weights * x)) @\\
+\mbox{}\verb@  smmfTilt <- smm * th$fTilt@\\
+\mbox{}\verb@  ystd <- ymm / th$bPrime2@\\
 \mbox{}\verb@  ystdWeighted <- weights * ystd@\\
@@
 \mbox{}\verb@  score.logT1 <- sptFreq.weighted@\\
-\mbox{}\verb@  score.logT2 <- fTiltSWSumsWeighted@\\
-\mbox{}\verb@  score.logT3 <- c(smmfTiltSW %*% ystdWeighted)@\\
+\mbox{}\verb@  score.logT2 <- fTiltSumsWeighted@\\
+\mbox{}\verb@  score.logT3 <- c(smmfTilt %*% ystdWeighted)@\\
 \mbox{}\verb@  score.log <- score.logT1 - score.logT2 - score.logT3@\\
@@
 \mbox{}\verb@  # Final info calculation@\\
-\mbox{}\verb@  info.logT1 <- diag(fTiltSWSumsWeighted)@\\
-\mbox{}\verb@  info.logT2 <- tcrossprod(th$fTiltSW)@\\
-\mbox{}\verb@  info.logT3 <- tcrossprod(smmfTiltSW, smmfTiltSW * rep(ystdWeighted, each=nrow(smmfTiltSW)))@\\
+\mbox{}\verb@  info.logT1 <- diag(fTiltSumsWeighted)@\\
+\mbox{}\verb@  info.logT2 <- tcrossprod(th$fTilt)@\\
+\mbox{}\verb@  info.logT3 <- tcrossprod(smmfTilt, smmfTilt * rep(ystdWeighted, each=nrow(smmfTilt)))@\\
 \mbox{}\verb@  info.log <- info.logT1 - info.logT2 - info.logT3@\\

### 2021-03-08 R/SPGLM.R

@@ -83,7 +83,7 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
        hp <- sapply(0:N, function(t)dhyper(x=Y[,1], m=rowSums(Y), n=N-rowSums(Y), k=t))   
        
        # initial log-likelihood
-       llik <- rowSums(fTiltMatrix * hp) %*% weights
+       llik <- log(rowSums(fTiltMatrix * hp)) %*% weights
 
      
      
@@ -141,7 +141,7 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
           betas <- mod$beta
           referencef0 <- mod$f0
           spt <- round(mod$spt * N)
-          llik <- c(rowSums(fTiltMatrix * hp) %*% weights)
+          llik <- c(log(rowSums(fTiltMatrix * hp)) %*% weights)
         
         
         difference <- abs(llik - llikPre)
@@ -176,7 +176,7 @@ spglm_loglik <- function(beta, f0, data_object, link){
                    thetaStart=NULL, thetaControl=theta.control())
     
     hp <- sapply(0:N, function(t)dhyper(x=data_object$resp[,1], m=data_object$n, n=N-data_object$n, k=t))   
-    llik_term <- rowSums(t(th$fTilt) * hp)
+    llik_term <- log(rowSums(t(th$fTilt) * hp))
     llik <- llik_term %*% data_object$weights
     c(llik)
 }

### 2021-03-08 R/wgldrm.R

@@ -387,14 +387,6 @@ theta.control <- function(eps=1e-10, maxiter=100, maxhalf=20, maxtheta=500,
 #' for each observation. Should match \code{mu} argument very closely.
 #' \item \code{bPrime2} Vector containing the variance of the exponentially tilted
 #' distribution for each observation.
-#' \item \code{fTiltSW} Matrix containing the exponentially tilted distribution for each
-#' observation, conditional on that observation being sampled, i.e. f(y|X=x, S=1).
-#' If \code{sampprobs=NULL}, then \code{fTiltSW} matches \code{fTilt}.
-#' \item \code{bPrimeSW} Vector containing the mean for each observation, conditional
-#' on that observation being sampled. If \code{sampprobs=NULL}, then \code{bPrimeSW}
-#' matches \code{bPrime}.
-#' \item \code{bPrime2SW} Vector containing the variance for each observation, conditional
-#' on that observation being sampled. If \code{sampprobs=NULL}, then \code{bPrime2SW}
 #' matches \code{bPrime2}.
 #' \item \code{llik} Semiparametric log-likelihood, evaluated at the current beta
 #' and f0 values. If sampling weights are used, then the log-likelihood is conditional
@@ -534,18 +526,11 @@ getTheta <- function(spt, f0, mu, weights, ySptIndex, thetaStart=NULL, thetaCont
     }
 
 
-    fTiltSW <- fTilt
-    bPrimeSW <- bPrime
-    bPrime2SW <- bPrime2
-  
-
     ## Calculate log-likelihood
-   # llik <- sum(log(fTiltSW[cbind(ySptIndex, seq_along(ySptIndex))]))
-    fTiltSW.extracted <- fTiltSW[cbind(ySptIndex, seq_along(ySptIndex))]
-    llik <- sum(weights[fTiltSW.extracted>0] * log(fTiltSW.extracted[fTiltSW.extracted>0]))
+    fTilt.extracted <- fTilt[cbind(ySptIndex, seq_along(ySptIndex))]
+    llik <- sum(weights[fTilt.extracted>0] * log(fTilt.extracted[fTilt.extracted>0]))
  
     list(theta=theta, fTilt=fTilt, bPrime=bPrime, bPrime2=bPrime2,
[…]

### 2021-03-08 nuweb/SPreg_develop.R

@@ -42,14 +42,25 @@ b2 <- spglm_loglik(beta=m2$coefficients, f0=m2$f0, m$data_object, m$link)
 
 # Hessian matrix calculation
 library(numDeriv)
-my_ll <- function(x, data_object, link){
+my_ll <- function(x, data_object, link, mu0){
   p <- ncol(data_object$model_matrix)
-  spglm_loglik(beta=x[1:p], f0 = exp(x[-(1:p)]), data_object, link)
+  f <- exp(x[-(1:p)])
+  spglm_loglik(beta=x[1:p], f0 = f, data_object, link)
 }
-hess <- hessian(func= my_ll,  x=c(coef(m), log(m$f0)), data_object=m$data_object, link=m$link)
-# inverse of Hessian matrix == Fisher information (used for SE estimation) for regression coefficients
-fish <- solve(-hess)
 
+hess <- hessian(func= my_ll,  x=c(coef(m), log(m$f0)), data_object=m$data_object, link=m$link,
+                mu0 = m$mu0)
+grad <- c(rep(1,length(coef(m))), 1/m$f0[-zf_idx])
+hess1 <- diag(grad) %*% hess %*% diag(grad)
+# inverse of Hessian matrix == variance-covariance matri (used for SE estimation) for regression coefficients
+vc <- solve(-hess1)
+sqrt(diag(vc))[1:4]
+
+# fix f0
+hess0 <- hessian(func= spglm_loglik,  x=coef(m),f0=m$f0, data_object=m$data_object, link=m$link)
+# inverse of Hessian matrix == Fisher information (used for SE estimation) for regression coefficients
+fish0 <- solve(-hess0)
+sqrt(diag(fish0))
 
 ######### Testing wGLDRM
 run_gldrm <- 

### 2021-03-08 nuweb/SPGLM.tex
Remove sampling-weighted terms from gldrm
Fix bug incorporating frequency weights into initial estimate

@@
 \mbox{}\verb@    hp <- sapply(0:N, function(t)dhyper(x=data_object$resp[,1], m=data_object$n, n=N-data_object$n, k=t))   @\\
-\mbox{}\verb@    llik_term <- rowSums(t(th$fTilt) * hp)@\\
+\mbox{}\verb@    llik_term <- log(rowSums(t(th$fTilt) * hp))@\\
 \mbox{}\verb@    llik <- llik_term %*% data_object$weights@\\
@@
 \mbox{}\verb@  spt <- round(mod$spt * N)@\\
-\mbox{}\verb@  llik <- c(rowSums(fTiltMatrix * hp) %*% weights)@\\
+\mbox{}\verb@  llik <- c(log(rowSums(fTiltMatrix * hp)) %*% weights)@\\
 \mbox{}\verb@@{\NWsep}
@@
 \mbox{}\verb@  pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), @\\
-\mbox{}\verb@                    trt="Trt", clustersize="ClusterSize", nresp="NResp")@\\
+\mbox{}\verb@                    trt="Trt", clustersize="ClusterSize", nresp="NResp", freq="Freq")@\\
 \mbox{}\verb@  est <- mc.est(pooled)@\\
@@
 \mbox{}\verb@  # initial log-likelihood@\\
-\mbox{}\verb@  llik <- rowSums(fTiltMatrix * hp) %*% weights@\\
+\mbox{}\verb@  llik <- log(rowSums(fTiltMatrix * hp)) %*% weights@\\
 \mbox{}\verb@@\\

### 2021-03-08 nuweb/SPregress.tex
Rearrange folder structure
Set up package infrastructure
Implementing E-step of spglm, calculating log-likelihood
Monitor log-likelihood during convergence
Fix bug incorporating frequency weights into initial estimate

--- 2019-09-25 SPregress.tex
@@
 \begin{flushleft} \small\label{scrap1}\raggedright\small
-\NWtarget{nuweb1a}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {1a}}$\equiv$
+\NWtarget{nuweb1a}{} \verb@"../R/SPreg.R"@\nobreak\ {\footnotesize {1a}}$\equiv$
 \vspace{-1ex}
@@
 \begin{equation}\label{E:1to1}
-	p_{r,n}= \binom{n}{r} \sum_{j=0}^{n-r} {(-1)^j} \binom{n-r}{j} \lambda_{r+j} = \binom{n}{r} \sum_{s=r}^{n} {(-1)^(s-r)} \binom{n-r}{s-r} \lambda_{s},
+	p_{r,n}= \binom{n}{r} \sum_{j=0}^{n-r} {(-1)^j} \binom{n-r}{j} \lambda_{r+j} = \binom{n}{r} \sum_{s=r}^{n} {(-1)^{(s-r)}} \binom{n-r}{s-r} \lambda_{s},
 \end{equation}
@@
 \begin{flushleft} \small\label{scrap2}\raggedright\small
-\NWtarget{nuweb1b}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {1b}}$\equiv$
+\NWtarget{nuweb1b}{} \verb@"../R/SPreg.R"@\nobreak\ {\footnotesize {1b}}$\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap3}\raggedright\small
-\NWtarget{nuweb2}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {2}}$\equiv$
+\NWtarget{nuweb2}{} \verb@"../R/SPreg.R"@\nobreak\ {\footnotesize {2}}$\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@@\\
+\mbox{}\verb@#' Fit semi-parametric relative risk model@\\
+\mbox{}\verb@#'@\\
 \mbox{}\verb@#'@{\tt @}\verb@rdname sprr@\\
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@#' Semi-parametric relative risk model@\\
 \mbox{}\verb@sprr <- function(formula, data, subset, weights, link="cloglog", mu1=NULL, start=NULL, control=list(eps=0.001, maxit=100), ...){@\\
@@
 \begin{flushleft} \small\label{scrap5}\raggedright\small
-\NWtarget{nuweb3b}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {3b}}$\equiv$
+\NWtarget{nuweb3b}{} \verb@"../R/SPreg.R"@\nobreak\ {\footnotesize {3b}}$\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap6}\raggedright\small
-\NWtarget{nuweb4}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {4}}$\equiv$
+\NWtarget{nuweb4}{} \verb@"../R/SPreg.R"@\nobreak\ {\footnotesize {4}}$\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap7}\raggedright\small
-\NWtarget{nuweb5}{} \verb@"R/SPreg.R"@\nobreak\ {\footnotesize {5}}$\equiv$
+\NWtarget{nuweb5}{} \verb@"../R/SPreg.R"@\nobreak\ {\footnotesize {5}}$\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@    mod <- glm(cbind(Ycomb[,"s"], Ycomb[,"y"]-Ycomb[,"s"]) ~ mm2+0, family=fam, weights=ew)@\\
+\mbox{}\verb@    mod <- glm(cbind(Ycomb[,"s"], Ycomb[,"y"]-Ycomb[,"s"]) ~ mm2+0, family=fam, weights=ew, start=beta_old)@\\
 \mbox{}\verb@    beta_new <- coef(mod)@\\
@@
 \mbox{}\verb@ if (is.null(start$q)){@\\
+\mbox{}\verb@    if (is.null(weights))       @\\
 \mbox{}\verb@      pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y)), trt="Trt",@\\
 \mbox{}\verb@                       clustersize="ClusterSize", nresp="NResp")@\\
+\mbox{}\verb@    else                       @\\
+\mbox{}\verb@       pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), @\\
+\mbox{}\verb@                        trt="Trt", clustersize="ClusterSize", nresp="NResp", freq="Freq")@\\
 \mbox{}\verb@   est <- mc.est(pooled)@\\

### 2021-03-08 R/SPGLM.R

@@ -67,7 +67,7 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
        
        betas <- NULL
        pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), 
-                         trt="Trt", clustersize="ClusterSize", nresp="NResp")
+                         trt="Trt", clustersize="ClusterSize", nresp="NResp", freq="Freq")
        est <- mc.est(pooled)
        referencef0 <- est$Prob[est$ClusterSize == N]
        # ensure all positive values

### 2021-03-08 R/SPreg.R

@@ -109,9 +109,13 @@ sprr <- function(formula, data, subset, weights, link="cloglog", mu1=NULL, start
          }
          
          if (is.null(start$q)){
-            pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y)), trt="Trt",
-                             clustersize="ClusterSize", nresp="NResp")
-            est <- mc.est(pooled)
+            if (is.null(weights))       
+              pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y)), trt="Trt",
+                               clustersize="ClusterSize", nresp="NResp")
+            else                       
+               pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), 
+                                trt="Trt", clustersize="ClusterSize", nresp="NResp", freq="Freq")
+           est <- mc.est(pooled)
             q0 <- est$Prob[est$ClusterSize == N]
             if (is.null(mu1) && is.null(start$mu1))
                 q_new <- mean_constrained_probs(pmax(q0, 0.01))

### 2021-03-08 nuweb/SPreg_develop.R

@@ -14,6 +14,7 @@ boric_acid <- CorrBin::read.CBData("z:/EOGeorge/Data/Binary/BoricAcidMousedata_p
                   sep=",", skip=1)
 use_data(boric_acid)
 
+#############
 ex <- as.package("../exchreg")
 nuweb(ex)
 

### 2021-03-08 nuweb/SPGLM.tex
Add standard error calculation

@@
 \mbox{}\verb@    @\hbox{$\langle\,${\itshape Create model matrix from formula and data}\nobreak\ {\footnotesize \NWlink{nuweb2}{2}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\\
+\mbox{}\verb@    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset)@\\
+\mbox{}\verb@    @\\
 \mbox{}\verb@    @\hbox{$\langle\,${\itshape Fit model}\nobreak\ {\footnotesize \NWlink{nuweb3b}{3b}}$\,\rangle$}\verb@@\\
@@
 \mbox{}\verb@    mt <- attr(mf, "terms")@\\
-\mbox{}\verb@    res <- list(coefficients = betas, f0=referencef0, mu0=mu0, niter = iter, loglik=llik,@\\
+\mbox{}\verb@    res <- list(coefficients = betas, SE = SEbeta, f0=referencef0, mu0=mu0, niter = iter, loglik=llik,@\\
 \mbox{}\verb@                link = link, call = mc, terms = mt,@\\
 \mbox{}\verb@                xlevels = .getXlevels(mt, mf),@\\
-\mbox{}\verb@                data_object=list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset))@\\
+\mbox{}\verb@                data_object=data_object)@\\
 \mbox{}\verb@    class(res) <- "spglm"@\\
@@
 \mbox{}\verb@  }@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  @\hbox{$\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize \NWlink{nuweb?}{?}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@  @\\
@@
 \end{flushleft}
+\subsection{Standard errors}
+
+We will use numeric derivation to obtain the hessian of the observed data likelihood.
+The variance-covariance matrix can be estimated as the inverse of the negative hessian.
+The reference distribution is included in the hessian on the log-scale.
+
+\begin{flushleft} \small\label{scrap9}\raggedright\small
+\NWtarget{nuweb?}{} $\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize {?}}$\,\rangle\equiv$
+\vspace{-1ex}
+\begin{list}{}{} \item
+\mbox{}\verb@@\\
+\mbox{}\verb@  ll <- function(x){@\\
+\mbox{}\verb@    spglm_loglik(beta=x[1:p], f0 = exp(x[-(1:p)]), data_object=data_object, link=link)@\\
+\mbox{}\verb@  }@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@  hess <- numDeriv::hessian(func=ll,  x=c(betas, log(referencef0)))@\\
+\mbox{}\verb@  vc <- solve(-hess1)@\\
+\mbox{}\verb@  SEbeta <- sqrt(diag(vc)[1:p])@\\
+\mbox{}\verb@@{\NWsep}
+\end{list}
+\vspace{-1.5ex}
+\footnotesize
+\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
+\item \NWtxtMacroRefIn\ \NWlink{nuweb3b}{3b}.
+
+\item{}
+\end{list}
+\vspace{4ex}
+\end{flushleft}
 \end{document}

### 2021-03-08 R/SPGLM.R

@@ -60,6 +60,9 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
             weights <- rep(1, nrow(mm))
     
     
+    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset)
+    
+    
      
        N <- max(rowSums(Y))
        nobs <- nrow(mm)
@@ -147,13 +150,23 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
         difference <- abs(llik - llikPre)
       }
       
+      
+        ll <- function(x){
+          spglm_loglik(beta=x[1:p], f0 = exp(x[-(1:p)]), data_object=data_object, link=link)
+        }
+
+        hess <- numDeriv::hessian(func=ll,  x=c(betas, log(referencef0)))
+        vc <- solve(-hess1)
+        SEbeta <- sqrt(diag(vc)[1:p])
+      
+      
     
 
     mt <- attr(mf, "terms")
-    res <- list(coefficients = betas, f0=referencef0, mu0=mu0, niter = iter, loglik=llik,
+    res <- list(coefficients = betas, SE = SEbeta, f0=referencef0, mu0=mu0, niter = iter, loglik=llik,
                 link = link, call = mc, terms = mt,
                 xlevels = .getXlevels(mt, mf),
-                data_object=list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset))
+                data_object=data_object)
     class(res) <- "spglm"
     res
 

### 2021-03-08 nuweb/SPreg_develop.R

@@ -4,6 +4,7 @@ source('nuweb/Nuweb.R')
 #  setting up the package and its infrastructure
 create("../exchreg")
 use_package("CorrBin")
+use_package("numDeriv")
 use_build_ignore("nuweb")
 use_gpl3_license()
 desc::desc_add_author(given = "Xinran", family = "Qi", email = "xinqi@mcw.edu", role = "aut",
@@ -29,10 +30,10 @@ load_all(ex)
 ## test SPGLM
 data(boric_acid)
 m <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=boric_acid, link="logit",
-           weights=boric_acid$Freq, control = list(eps=1e-6, maxit=1000))
+           weights=boric_acid$Freq, control = list(eps=1e-5, maxit=1000))
 
 ba2 <- boric_acid[rep(1:nrow(boric_acid), boric_acid$Freq),]
-m2 <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=ba2, link="logit", control = list(eps=1e-6, maxit=1000))
+m2 <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=ba2, link="logit", control = list(eps=1e-5, maxit=1000))
 
 all.equal(coef(m), coef(m2))
 all.equal(m$loglik, m2$loglik)

### 2021-03-19 nuweb/SPGLM.tex
Adapt variance calculation to incorporate constraints

@@
 \mbox{}\verb@    mt <- attr(mf, "terms")@\\
-\mbox{}\verb@    res <- list(coefficients = betas, SE = SEbeta, f0=referencef0, mu0=mu0, niter = iter, loglik=llik,@\\
+\mbox{}\verb@    res <- list(coefficients = betas, SE = SEbeta, f0=referencef0, SEf0 = SEf0, mu0=mu0, niter = iter, loglik=llik,@\\
 \mbox{}\verb@                link = link, call = mc, terms = mt,@\\
@@
 The variance-covariance matrix can be estimated as the inverse of the negative hessian.
-The reference distribution is included in the hessian on the log-scale.
+The sum-to-one and fixed-mean constraints of the reference distribution are included by expanding the hessian to a bordered hessian before inversion by including the derivatives of the constraints (this can be derived based on the Lagrange multiplier method).
+The reference distribution is included in the hessian on the log-scale, and is multiplied by the gradient before inversion to revert to the original scale.
 
@@
 \mbox{}\verb@  hess <- numDeriv::hessian(func=ll,  x=c(betas, log(referencef0)))@\\
-\mbox{}\verb@  vc <- solve(-hess1)@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  # revert to unlogged f0@\\
+\mbox{}\verb@  grad <- c(rep(1, p), 1/referencef0)@\\
+\mbox{}\verb@  hess1 <- diag(grad) %*% hess %*% diag(grad)@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  # create bordered hessian@\\
+\mbox{}\verb@  border1 <- c(rep(0, p), rep(1, N+1))  # gradient of sum-to-one constraint@\\
+\mbox{}\verb@  border2 <- c(rep(0, p), 0:N)          # gradient of fixed-mean constraint@\\
+\mbox{}\verb@  bhess <- rbind(cbind(hess1, border1, border2), c(border1,0,0), c(border2,0,0))@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  # calculate variance-covariance matrix@\\
+\mbox{}\verb@  vc <- solve(-bhess)@\\
 \mbox{}\verb@  SEbeta <- sqrt(diag(vc)[1:p])@\\
+\mbox{}\verb@  SEf0 <- sqrt(diag(vc)[p+1+(0:N)])@\\
 \mbox{}\verb@@{\NWsep}

### 2021-03-19 R/SPGLM.R

@@ -156,14 +156,26 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
         }
 
         hess <- numDeriv::hessian(func=ll,  x=c(betas, log(referencef0)))
-        vc <- solve(-hess1)
+        
+        # revert to unlogged f0
+        grad <- c(rep(1, p), 1/referencef0)
+        hess1 <- diag(grad) %*% hess %*% diag(grad)
+        
+        # create bordered hessian
+        border1 <- c(rep(0, p), rep(1, N+1))  # gradient of sum-to-one constraint
+        border2 <- c(rep(0, p), 0:N)          # gradient of fixed-mean constraint
+        bhess <- rbind(cbind(hess1, border1, border2), c(border1,0,0), c(border2,0,0))
+        
+        # calculate variance-covariance matrix
+        vc <- solve(-bhess)
         SEbeta <- sqrt(diag(vc)[1:p])
+        SEf0 <- sqrt(diag(vc)[p+1+(0:N)])
       
       
     
 
     mt <- attr(mf, "terms")
-    res <- list(coefficients = betas, SE = SEbeta, f0=referencef0, mu0=mu0, niter = iter, loglik=llik,
+    res <- list(coefficients = betas, SE = SEbeta, f0=referencef0, SEf0 = SEf0, mu0=mu0, niter = iter, loglik=llik,
                 link = link, call = mc, terms = mt,
                 xlevels = .getXlevels(mt, mf),
                 data_object=data_object)

### 2021-03-19 nuweb/SPreg_develop.R

@@ -42,27 +42,20 @@ a <- spglm_pred_mean(m$coefficients, m$data_object, m$link)
 b <- spglm_loglik(beta=m$coefficients, f0=m$f0, m$data_object, m$link)
 b2 <- spglm_loglik(beta=m2$coefficients, f0=m2$f0, m$data_object, m$link)
 
-# Hessian matrix calculation
-library(numDeriv)
-my_ll <- function(x, data_object, link, mu0){
-  p <- ncol(data_object$model_matrix)
-  f <- exp(x[-(1:p)])
-  spglm_loglik(beta=x[1:p], f0 = f, data_object, link)
-}
+mn <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=boric_acid, link="log",
+              weights=boric_acid$Freq, control = list(eps=1e-5, maxit=1000))
 
-hess <- hessian(func= my_ll,  x=c(coef(m), log(m$f0)), data_object=m$data_object, link=m$link,
-                mu0 = m$mu0)
-grad <- c(rep(1,length(coef(m))), 1/m$f0[-zf_idx])
-hess1 <- diag(grad) %*% hess %*% diag(grad)
-# inverse of Hessian matrix == variance-covariance matri (used for SE estimation) for regression coefficients
-vc <- solve(-hess1)
-sqrt(diag(vc))[1:4]
 
-# fix f0
-hess0 <- hessian(func= spglm_loglik,  x=coef(m),f0=m$f0, data_object=m$data_object, link=m$link)
-# inverse of Hessian matrix == Fisher information (used for SE estimation) for regression coefficients
-fish0 <- solve(-hess0)
-sqrt(diag(fish0))
+# laplace link
+laplace.link <- function(mu) ifelse(mu>0.5, -log(2-2*mu), log(2*mu))
+laplace.invlink <- function(eta) 0.5 + 0.5*sign(eta)*(1-exp(-abs(eta)))
+laplace.mu.eta <- function(eta) pmax(0.5 * exp(-abs(eta)), .Machine$double.eps)
+
+mll <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=boric_acid, 
+             link=list(linkfun=laplace.link, linkinv=laplace.invlink, mu.eta=laplace.mu.eta),
+            weights=boric_acid$Freq, control = list(eps=1e-5, maxit=1000))
+
+
 
 ######### Testing wGLDRM
 run_gldrm <- 

### 2021-06-30 nuweb/SPGLM.tex
Add predict and print methods for `spglm` objects

@@
 \mbox{}\verb@    @\\
-\mbox{}\verb@    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset)@\\
+\mbox{}\verb@    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset, maxN=N)@\\
 \mbox{}\verb@    @\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape Fit model}\nobreak\ {\footnotesize \NWlink{nuweb3b}{3b}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape Fit model}\nobreak\ {\footnotesize \NWlink{nuweb6}{6}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@@\\
 \mbox{}\verb@    mt <- attr(mf, "terms")@\\
-\mbox{}\verb@    res <- list(coefficients = betas, SE = SEbeta, f0=referencef0, SEf0 = SEf0, mu0=mu0, niter = iter, loglik=llik,@\\
-\mbox{}\verb@                link = link, call = mc, terms = mt,@\\
+\mbox{}\verb@    names(betas) <- colnames(mm)@\\
+\mbox{}\verb@    names(referencef0) <- 0:N@\\
+\mbox{}\verb@    res <- list(coefficients = betas, SE = SEbeta, f0=referencef0, SEf0 = SEf0, mu0=mu0, niter = iter, @\\
+\mbox{}\verb@                loglik=llik, link = link, call = mc, terms = mt,@\\
 \mbox{}\verb@                xlevels = .getXlevels(mt, mf),@\\
@@
 \begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
-\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3a}{, 3a}.
+\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3}{, 3}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}.
 
@@
 \mbox{}\verb@        weights <- rep(1, nrow(mm))@\\
+\mbox{}\verb@   @\\
+\mbox{}\verb@   # define dimensions     @\\
+\mbox{}\verb@    N <- max(rowSums(Y))@\\
+\mbox{}\verb@    nobs <- nrow(mm)@\\
+\mbox{}\verb@    p <- ncol(mm)@\\
+\mbox{}\verb@  @\\
 \mbox{}\verb@@{\NWsep}
@@
 \item \NWtxtMacroRefIn\ \NWlink{nuweb1}{1}.
+
+\item{}
+\end{list}
+\vspace{4ex}
+\end{flushleft}
+\subsection{Methods for the \texttt{spglm} class}
+
+First, define a printing method which does not show the saved data and model matrix
+
+\begin{flushleft} \small\label{scrap3}\raggedright\small
+\NWtarget{nuweb3}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {3}}$\equiv$
+\vspace{-1ex}
+\begin{list}{}{} \item
+\mbox{}\verb@@\\
+\mbox{}\verb@#' @{\tt @}\verb@rdname spglm@\\
+\mbox{}\verb@#' @{\tt @}\verb@export@\\
+\mbox{}\verb@print.spglm <- function(x, digits = max(3L, getOption("digits") - 3L),...){@\\
+\mbox{}\verb@  # based on print.lm, summary.lm, and print.summary.lm@\\
+\mbox{}\verb@  cat("\nA semi-parametric generalized linear regression model fit\n")@\\
+\mbox{}\verb@  cat("\nCall:\n", paste(deparse(x$call), sep = "\n", collapse = "\n"),@\\
+\mbox{}\verb@        "\n\n", sep = "")@\\
+\mbox{}\verb@  if (length(coef(x))) {@\\
+\mbox{}\verb@    beta <- x$coefficients@\\
+\mbox{}\verb@    SEbeta <- x$SE@\\
+\mbox{}\verb@    z <- beta / SEbeta@\\
+\mbox{}\verb@    pval <- 2 * stats::pnorm(abs(z), lower.tail=FALSE)@\\
+\mbox{}\verb@    coefmat <-  cbind(Estimate = beta, `Std. Error` = SEbeta, @\\
+\mbox{}\verb@                      `z value` = z, `Pr(>|z|)` = pval)@\\
+\mbox{}\verb@    cat("Coefficients:\n")@\\
+\mbox{}\verb@    printCoefmat(coefmat, digits = digits, na.print = "NA", ...)@\\
+\mbox{}\verb@    }@\\
+\mbox{}\verb@    else cat("No coefficients\n")@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@    if (length(x$f0)){@\\
+\mbox{}\verb@        cat("\n Baseline probabilities (q0):\n")        @\\
+\mbox{}\verb@        q0 <- x$f0@\\
+\mbox{}\verb@         print.default(format(q0, digits = digits),@\\
+\mbox{}\verb@            print.gap = 2, quote = FALSE)@\\
+\mbox{}\verb@    }@\\
+\mbox{}\verb@    else cat("No baseline joint probabilities\n")@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@   cat("\n Log-likelihood: ", format(x$loglik, digits=digits), "\n")@\\
+\mbox{}\verb@   invisible(x)@\\
+\mbox{}\verb@}@\\
+\mbox{}\verb@@{\NWsep}
+\end{list}
+\vspace{-1.5ex}
+\footnotesize
+\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
+\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3}{, 3}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}.
+
+\item{}
+\end{list}
+\vspace{4ex}
+\end{flushleft}
+The prediction method will predict for a variety of scenarios:
+\begin{itemize}
+\item Input data set:
+  \begin{itemize}
+    \item the data used in the fitting;
+    \item new data.
+  \end{itemize}
+\item Results:
+  \begin{itemize}
+    \item $\mu(z,\beta)$: the mean event probability, given $z$;
+    \item $p_{r,n}(z)$: the probability of observing the given $r$ responses with cluster size $n$, given $z$;
+    \item $\{p_{\cdot,n}(z)\}$: the entire vector of response probabilities for cluster size $n$, given $z$ (will be a list due to varying lengths);
+    \item $\beta'z$: the linear predictor value $z$;
+    \item $\omega(\beta'z)$: the tilting parameter at predictor value $z$.
+  \end{itemize}
+\end{itemize}
+
+\begin{flushleft} \small\label{scrap4}\raggedright\small
+\NWtarget{nuweb4}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {4}}$\equiv$
+\vspace{-1ex}
+\begin{list}{}{} \item
+\mbox{}\verb@@\\
+\mbox{}\verb@#' Predict methods for SPGLM fits@\\
+\mbox{}\verb@#' Obtains predictions from a fitted semi-parametric generalized linear model. Note that \code{offset}@\\
+\mbox{}\verb@#' and \code{weight} terms are not implemented for predicting from new data.@\\
+\mbox{}\verb@#' @{\tt @}\verb@param object fitted model of class \code{spglm}@\\
+\mbox{}\verb@#' @{\tt @}\verb@param newdata optionally, a data frame in which to look for covariate values for prediction. @\\
+\mbox{}\verb@#'  If NULL, the original data set is used@\\
+\mbox{}\verb@#' @{\tt @}\verb@param type the type of prediction requested. The default is "mean", the mean event probability;@\\
+\mbox{}\verb@#' "prob" requests the probability of the observation (given number of responses with given cluster size), @\\
+\mbox{}\verb@#' "tilt" returns the tilting parameter which achieves the modeled mean from the baseline distribution;@\\
+\mbox{}\verb@#' and "lp" requests the linear predictor.@\\
+\mbox{}\verb@#' @{\tt @}\verb@param newn if \code{newdata} is provided and \code{type="prob"}, an integer or integer vector specifying the @\\
+\mbox{}\verb@#' @{\tt @}\verb@param newevents if \code{newdata} is provided and \code{type="prob"}, an integer or integer vector specifying the @\\
+\mbox{}\verb@#'  number of events for the predictions@\\
+\mbox{}\verb@#' @{\tt @}\verb@export@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@predict.spglm <- function(object, newdata=NULL,@\\
+\mbox{}\verb@                              type=c("mean", "prob", "tilt", "lp"),@\\
+\mbox{}\verb@                              newn=NULL, newevents=NULL, ...){@\\
+\mbox{}\verb@  type <- match.arg(type)@\\
+\mbox{}\verb@  tt <- terms(object)@\\
+\mbox{}\verb@  if (!missing(newdata)){    @\\
+\mbox{}\verb@    Terms <- delete.response(tt)@\\
+\mbox{}\verb@    m <- model.frame(Terms, newdata, xlev = object$xlevels)@\\
+\mbox{}\verb@    if (!is.null(cl <- attr(Terms, "dataClasses")))@\\
+\mbox{}\verb@            .checkMFClasses(cl, m)@\\
+\mbox{}\verb@    mm <- model.matrix(Terms, m)@\\
+\mbox{}\verb@    data_object <- list(model_matrix=mm, offset = rep(0, nrow(mm)), weights=rep(1, nrow(mm)),@\\
+\mbox{}\verb@                        maxN = object$data_object$maxN)@\\
+\mbox{}\verb@    if (!missing(newn)){@\\
+\mbox{}\verb@       if (!(length(newn) == 1L || length(newn) == nrow(newdata)))@\\
+\mbox{}\verb@          stop("'newn' should have length 1 or equal to the number of rows of 'newdata'")@\\
+\mbox{}\verb@        data_object$n <- rep(newn, length=nrow(newdata))@\\
+\mbox{}\verb@        if (!all(data_object$n %in% 0:data_object$maxN))@\\
+\mbox{}\verb@          stop("Values in 'newn' should be integers between 0 and the maximum cluster size in the original data")@\\
+\mbox{}\verb@       }@\\
+\mbox{}\verb@    if (!missing(newevents)){@\\
+\mbox{}\verb@       if (!(length(newevents) == 1L || length(newevents) == nrow(newdata)))@\\
+\mbox{}\verb@          stop("'newevents' should have length 1 or equal to the number of rows of 'newdata'")@\\
+\mbox{}\verb@        data_object$resp <- cbind(rep(newevents, length=nrow(newdata)))@\\
+\mbox{}\verb@       }@\\
+\mbox{}\verb@  } else {@\\
+\mbox{}\verb@    data_object <- object$data_object@\\
+\mbox{}\verb@  }@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  if (type=="mean"){@\\
+\mbox{}\verb@    pred <- spglm_pred_mean(beta=object$coefficients, data_object, link=object$link)@\\
+\mbox{}\verb@  } else@\\
+\mbox{}\verb@  if (type=="prob"){@\\
+\mbox{}\verb@    if (!missing(newdata) && (missing(newn) || missing(newevents)))@\\
+\mbox{}\verb@       stop("For prediction of probability vectors with new data, cluster sizes should be specified in 'newn' and number of events in 'newevents'.")     @\\
+\mbox{}\verb@     pred <- spglm_probs(beta=object$coefficients, f0=object$f0, @\\
+\mbox{}\verb@                              data_object=data_object, link=object$link)@\\
+\mbox{}\verb@  } else@\\
+\mbox{}\verb@  if (type == "lp"){@\\
+\mbox{}\verb@     pred <- spglm_lp(beta=object$coefficients, data_object=data_object)@\\
+\mbox{}\verb@  } else@\\
+\mbox{}\verb@  if (type == "tilt"){@\\
+\mbox{}\verb@     pred <- spglm_tilt(beta=object$coefficients, f0=object$f0, @\\
+\mbox{}\verb@                       data_object=data_object, link=object$link)@\\
+\mbox{}\verb@  }@\\
+\mbox{}\verb@  return(pred)    @\\
+\mbox{}\verb@}@\\
+\mbox{}\verb@@{\NWsep}
+\end{list}
+\vspace{-1.5ex}
+\footnotesize
+\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
+\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3}{, 3}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}.
 
@@
 
-\begin{flushleft} \small\label{scrap3}\raggedright\small
-\NWtarget{nuweb3a}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {3a}}$\equiv$
+\begin{flushleft} \small\label{scrap5}\raggedright\small
+\NWtarget{nuweb5}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {5}}$\equiv$
 \vspace{-1ex}
 \begin{list}{}{} \item
+\mbox{}\verb@@\\
+\mbox{}\verb@#' @{\tt @}\verb@keywords internal@\\
+\mbox{}\verb@spglm_lp <- function(beta, data_object){@\\
+\mbox{}\verb@  eta <- c(data_object$model_matrix %*% beta + data_object$offset)@\\
+\mbox{}\verb@  eta@\\
+\mbox{}\verb@}@\\
 \mbox{}\verb@@\\
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@spglm_loglik <- function(beta, f0, data_object, link){@\\
+\mbox{}\verb@spglm_tilt <- function(beta, f0, data_object, link){@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@    mu <- spglm_pred_mean(beta=beta, data_object=data_object, link=link)@\\
+\mbox{}\verb@    N <- data_object$maxN@\\
+\mbox{}\verb@    nobs <- nrow(data_object$model_matrix)@\\
 \mbox{}\verb@    @\\
+\mbox{}\verb@    # ySptIndex is only used to calculate the log-likelihood, which we will not be using@\\
+\mbox{}\verb@    th <- getTheta(spt=(0:N)/N, f0=f0, mu=mu, weights=data_object$weights, ySptIndex=rep(1, nobs),@\\
+\mbox{}\verb@                   thetaStart=NULL, thetaControl=theta.control())@\\
+\mbox{}\verb@    th$theta@\\
+\mbox{}\verb@}@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@spglm_probs <- function(beta, f0, data_object, link){@\\
+\mbox{}\verb@@\\
 \mbox{}\verb@    mu <- spglm_pred_mean(beta=beta, data_object=data_object, link=link)@\\
-\mbox{}\verb@    N <- max(data_object$n)@\\
+\mbox{}\verb@    N <- data_object$maxN@\\
 \mbox{}\verb@    nobs <- nrow(data_object$model_matrix)@\\
@@
 \mbox{}\verb@    hp <- sapply(0:N, function(t)dhyper(x=data_object$resp[,1], m=data_object$n, n=N-data_object$n, k=t))   @\\
-\mbox{}\verb@    llik_term <- log(rowSums(t(th$fTilt) * hp))@\\
-\mbox{}\verb@    llik <- llik_term %*% data_object$weights@\\
+\mbox{}\verb@    probs <- rowSums(t(th$fTilt) * hp)@\\
+\mbox{}\verb@    probs@\\
+\mbox{}\verb@}@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@spglm_loglik <- function(beta, f0, data_object, link){@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@    probs <- spglm_probs(beta, f0, data_object, link)@\\
+\mbox{}\verb@    llik <- log(probs) %*% data_object$weights@\\
 \mbox{}\verb@    c(llik)@\\
@@
 
-\begin{flushleft} \small\label{scrap4}\raggedright\small
-\NWtarget{nuweb3b}{} $\langle\,${\itshape Fit model}\nobreak\ {\footnotesize {3b}}$\,\rangle\equiv$
+\begin{flushleft} \small\label{scrap6}\raggedright\small
+\NWtarget{nuweb6}{} $\langle\,${\itshape Fit model}\nobreak\ {\footnotesize {6}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@ @\hbox{$\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize \NWlink{nuweb6b}{6b}}$\,\rangle$}\verb@@\\
-\mbox{}\verb@ @\hbox{$\langle\,${\itshape Set up for E-step}\nobreak\ {\footnotesize \NWlink{nuweb4}{4}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@ @\hbox{$\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize \NWlink{nuweb9a}{9a}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@ @\hbox{$\langle\,${\itshape Set up for E-step}\nobreak\ {\footnotesize \NWlink{nuweb7}{7}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@ @\\
@@
 \mbox{}\verb@    @\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape E-step}\nobreak\ {\footnotesize \NWlink{nuweb5}{5}}$\,\rangle$}\verb@@\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape M-step}\nobreak\ {\footnotesize \NWlink{nuweb6a}{6a}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape E-step}\nobreak\ {\footnotesize \NWlink{nuweb8a}{8a}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape M-step}\nobreak\ {\footnotesize \NWlink{nuweb8b}{8b}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@    @\\
@@
 \mbox{}\verb@  @\\
-\mbox{}\verb@  @\hbox{$\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize \NWlink{nuweb?}{?}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@  @\hbox{$\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize \NWlink{nuweb9b}{9b}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@  @\\
@@
 
-\begin{flushleft} \small\label{scrap5}\raggedright\small
-\NWtarget{nuweb4}{} $\langle\,${\itshape Set up for E-step}\nobreak\ {\footnotesize {4}}$\,\rangle\equiv$
+\begin{flushleft} \small\label{scrap7}\raggedright\small
+\NWtarget{nuweb7}{} $\langle\,${\itshape Set up for E-step}\nobreak\ {\footnotesize {7}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
-\item \NWtxtMacroRefIn\ \NWlink{nuweb3b}{3b}.
+\item \NWtxtMacroRefIn\ \NWlink{nuweb6}{6}.
 
@@
 
-\begin{flushleft} \small\label{scrap6}\raggedright\small
-\NWtarget{nuweb5}{} $\langle\,${\itshape E-step}\nobreak\ {\footnotesize {5}}$\,\rangle\equiv$
+\begin{flushleft} \small\label{scrap8}\raggedright\small
+\NWtarget{nuweb8a}{} $\langle\,${\itshape E-step}\nobreak\ {\footnotesize {8a}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 
-\begin{flushleft} \small\label{scrap7}\raggedright\small
-\NWtarget{nuweb6a}{} $\langle\,${\itshape M-step}\nobreak\ {\footnotesize {6a}}$\,\rangle\equiv$
+\begin{flushleft} \small\label{scrap9}\raggedright\small
+\NWtarget{nuweb8b}{} $\langle\,${\itshape M-step}\nobreak\ {\footnotesize {8b}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 
-\begin{flushleft} \small\label{scrap8}\raggedright\small
-\NWtarget{nuweb6b}{} $\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize {6b}}$\,\rangle\equiv$
+\begin{flushleft} \small\label{scrap10}\raggedright\small
+\NWtarget{nuweb9a}{} $\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize {9a}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@  N <- max(rowSums(Y))@\\
-\mbox{}\verb@  nobs <- nrow(mm)@\\
-\mbox{}\verb@  p <- ncol(mm)@\\
-\mbox{}\verb@  @\\
 \mbox{}\verb@  betas <- NULL@\\
@@
 
-\begin{flushleft} \small\label{scrap9}\raggedright\small
-\NWtarget{nuweb?}{} $\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize {?}}$\,\rangle\equiv$
+\begin{flushleft} \small\label{scrap11}\raggedright\small
+\NWtarget{nuweb9b}{} $\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize {9b}}$\,\rangle\equiv$
 \vspace{-1ex}

### 2021-06-30 R/SPGLM.R

@@ -56,18 +56,20 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
             stop("'weights' must be a numeric vector")
         if (!is.null(weights) && any(weights < 0))
             stop("negative weights not allowed")
-         if (is.null(weights))
+        if (is.null(weights))
             weights <- rep(1, nrow(mm))
+       
+       # define dimensions     
+        N <- max(rowSums(Y))
+        nobs <- nrow(mm)
+        p <- ncol(mm)
+      
     
     
-    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset)
+    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset, maxN=N)
     
     
      
-       N <- max(rowSums(Y))
-       nobs <- nrow(mm)
-       p <- ncol(mm)
-       
        betas <- NULL
        pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), 
                          trt="Trt", clustersize="ClusterSize", nresp="NResp", freq="Freq")
@@ -175,13 +177,115 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
     
 
     mt <- attr(mf, "terms")
-    res <- list(coefficients = betas, SE = SEbeta, f0=referencef0, SEf0 = SEf0, mu0=mu0, niter = iter, loglik=llik,
-                link = link, call = mc, terms = mt,
+    names(betas) <- colnames(mm)
+    names(referencef0) <- 0:N
+    res <- list(coefficients = betas, SE = SEbeta, f0=referencef0, SEf0 = SEf0, mu0=mu0, niter = iter, 
+                loglik=llik, link = link, call = mc, terms = mt,
                 xlevels = .getXlevels(mt, mf),
                 data_object=data_object)
     class(res) <- "spglm"
     res
 
+}
+
+#' @rdname spglm
+#' @export
+print.spglm <- function(x, digits = max(3L, getOption("digits") - 3L),...){
+  # based on print.lm, summary.lm, and print.summary.lm
+  cat("\nA semi-parametric generalized linear regression model fit\n")
+  cat("\nCall:\n", paste(deparse(x$call), sep = "\n", collapse = "\n"),
+        "\n\n", sep = "")
[…]

### 2021-06-30 nuweb/SPreg_develop.R

@@ -38,13 +38,31 @@ m2 <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=ba2, link="logit", con
 all.equal(coef(m), coef(m2))
 all.equal(m$loglik, m2$loglik)
 
+# offset
+moff <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, offset=Freq, data=boric_acid)
+# prediction
 a <- spglm_pred_mean(m$coefficients, m$data_object, m$link)
 b <- spglm_loglik(beta=m$coefficients, f0=m$f0, m$data_object, m$link)
 b2 <- spglm_loglik(beta=m2$coefficients, f0=m2$f0, m$data_object, m$link)
 
+nd <- data.frame(Trt = unique(boric_acid$Trt))
+nd$p <- predict(m, newdata=nd, type="mean")
+nd$p2 <- predict(moff, newdata=nd, type="mean")
+nd
+
+b3 <- predict(m, type="prob")
+predict(m, newdata=nd, newn=5, newevents=0, type="prob")
+nd_long <- merge(nd, data.frame(Resp=0:5))
+nd_long$p3 <- predict(m, newdata=nd_long, newevents = nd_long$Resp, newn=5, type="prob")
+p3mat <- matrix(nd_long$p3, nrow=nrow(nd), 
+                dimnames = list(Trt = nd$Trt, Resp=0:5))
+matplot(0:5, t(p3mat), type="l")
+
 mn <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=boric_acid, link="log",
               weights=boric_acid$Freq, control = list(eps=1e-5, maxit=1000))
 
+theta <- predict(m, newdata=nd, type="tilt")
+predict(moff, newdata=nd, type="lp")
 
 # laplace link
 laplace.link <- function(mu) ifelse(mu>0.5, -log(2-2*mu), log(2*mu))

### 2021-06-30 README.Rmd
Add simple readme

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Regression models for clustered exchangeable binary data

<!-- badges: start -->
<!-- badges: end -->

The goal of the `exchreg` is to fit semi-parametric relative-risk and generalized linear regression models for exchangeable clustered binary data with between-cluster covariates. You can browse its [source code](https://github.com/anikoszabo/exchreg).

## Installation

In the future, this package will be released on CRAN, but currently the development version is available on [GitHub](https://github.com/) with:

``` {r}
# install.packages("devtools")
devtools::install_github("anikoszabo/exchreg")
```
## Example: developmental toxicology study

The `boric_acid` dataset contains the outcomes of a developmental toxicity experiment of boric acid by the National Toxicology Program. Pregnant mice were provided feed containing boric acid at different concentrations. At the end of the study, each mouse's uterus was examined to evaluate the fetal outcomes. The data contains the size of each litter, the number of fetuses with an adverse event (resorption, intra-uterine death, or malformation), as well as the dose group. Since the same combination of clustersize/response count can occur multiple times, a frequency count column records the number of such replicates.

```{r DataExtract}
library(exchreg)
library(CorrBin)
head(boric_acid)
```

The goal is to model the effect of the covariates (here, treatment dose) on the probability of adverse event, while providing predictions for the entire distribution of the outcomes, ie the probability of observing $x$ events in a cluster (litter) of size $n$.

### Semi-parametric GLM

We can fit a semi-parametric GLM with a logit-link

```{r SPGLMexample}
mod1 <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=boric_acid, weights=Freq)
mod1
```

The prediction method can be used to obtain the predicted probabilities for any given outcome. For example, we estimate the probability of having 0 - 5 affected fetuses in a litter of size 5 at dose level 0.4:

```{r PredictModel}
pred <- predict(mod1, newdata = data.frame(Trt = rep("0.4", 6)), newn=5, newevents = 0:5, type="prob")
plot(0:5, pred, type="h", xlab="Number affected", ylab="Probability", ylim=c(0,max(pred)))

```

### 2021-06-30 nuweb/SPreg_develop.R

@@ -9,6 +9,7 @@ use_build_ignore("nuweb")
 use_gpl3_license()
 desc::desc_add_author(given = "Xinran", family = "Qi", email = "xinqi@mcw.edu", role = "aut",
                comment = NULL, file = ".", normalize = FALSE)
+use_readme_rmd()
 
 # add data
 boric_acid <- CorrBin::read.CBData("z:/EOGeorge/Data/Binary/BoricAcidMousedata_processed.csv",
@@ -26,6 +27,11 @@ shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet --run-viewer  wgldrm.tex")
 
 document(ex)
 load_all(ex)
+
+build_readme()
+
+check(ex, check_dir = "c:/Temp", cran = TRUE, manual=TRUE)
+
 
 ## test SPGLM
 data(boric_acid)

### 2021-06-30 nuweb/SPGLM.tex
Add multiple importFrom to pass checks

@@
 \mbox{}\verb@#'@{\tt @}\verb@export@\\
-\mbox{}\verb@#' @{\tt @}\verb@importFrom stats terms model.matrix@\\
+\mbox{}\verb@#' @{\tt @}\verb@importFrom stats terms model.matrix model.frame model.offset model.response model.weights@\\
 \mbox{}\verb@@\\
@@
 \mbox{}\verb@#' @{\tt @}\verb@export@\\
+\mbox{}\verb@#' @{\tt @}\verb@importFrom stats printCoefmat@\\
 \mbox{}\verb@print.spglm <- function(x, digits = max(3L, getOption("digits") - 3L),...){@\\
@@
 \mbox{}\verb@        "\n\n", sep = "")@\\
-\mbox{}\verb@  if (length(coef(x))) {@\\
+\mbox{}\verb@  if (length(x$coefficients)) {@\\
 \mbox{}\verb@    beta <- x$coefficients@\\
@@
 \mbox{}\verb@#' @{\tt @}\verb@export@\\
+\mbox{}\verb@#' @{\tt @}\verb@importFrom stats  .checkMFClasses .getXlevels delete.response@\\
 \mbox{}\verb@@\\
@@
 \mbox{}\verb@#' @{\tt @}\verb@keywords internal@\\
+\mbox{}\verb@#' @{\tt @}\verb@importFrom stats dhyper@\\
 \mbox{}\verb@spglm_lp <- function(beta, data_object){@\\
@@
 \mbox{}\verb@  betas <- NULL@\\
-\mbox{}\verb@  pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), @\\
+\mbox{}\verb@  pooled <- CorrBin::CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), @\\
 \mbox{}\verb@                    trt="Trt", clustersize="ClusterSize", nresp="NResp", freq="Freq")@\\
-\mbox{}\verb@  est <- mc.est(pooled)@\\
+\mbox{}\verb@  est <- CorrBin::mc.est(pooled)@\\
 \mbox{}\verb@  referencef0 <- est$Prob[est$ClusterSize == N]@\\

### 2021-06-30 R/SPGLM.R
Add multiple importFrom to pass checks

@@ -11,7 +11,7 @@
 #'@param control        a list with parameters controlling the algorithm.
 #'@return an object of class \code{spglm} with the fitted model.
 #'@export
-#' @importFrom stats terms model.matrix
+#' @importFrom stats terms model.matrix model.frame model.offset model.response model.weights
 
 spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL, 
                   control=list(eps=0.001, maxit=100), ...){
@@ -71,9 +71,9 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
     
      
        betas <- NULL
-       pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), 
+       pooled <- CorrBin::CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), 
                          trt="Trt", clustersize="ClusterSize", nresp="NResp", freq="Freq")
-       est <- mc.est(pooled)
+       est <- CorrBin::mc.est(pooled)
        referencef0 <- est$Prob[est$ClusterSize == N]
        # ensure all positive values
        referencef0 <- (referencef0 + 1e-6)/(1+(N+1)*1e-6)
@@ -190,12 +190,13 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
 
 #' @rdname spglm
 #' @export
+#' @importFrom stats printCoefmat
 print.spglm <- function(x, digits = max(3L, getOption("digits") - 3L),...){
   # based on print.lm, summary.lm, and print.summary.lm
   cat("\nA semi-parametric generalized linear regression model fit\n")
   cat("\nCall:\n", paste(deparse(x$call), sep = "\n", collapse = "\n"),
         "\n\n", sep = "")
-  if (length(coef(x))) {
+  if (length(x$coefficients)) {
     beta <- x$coefficients
     SEbeta <- x$SE
     z <- beta / SEbeta
@@ -233,6 +234,7 @@ print.spglm <- function(x, digits = max(3L, getOption("digits") - 3L),...){
 #' @param newevents if \code{newdata} is provided and \code{type="prob"}, an integer or integer vector specifying the 
 #'  number of events for the predictions
 #' @export
+#' @importFrom stats  .checkMFClasses .getXlevels delete.response
 
[…]

### 2021-06-30 R/wgldrm.R

@@ -4,6 +4,7 @@
 #' This function is called by the main \code{gldrm} function.
 #'
 #' @keywords internal
+#' @importFrom stats lm.fit lm.wfit weighted.mean
 gldrm.control <- function(eps=1e-10, maxiter=100, returnfTiltMatrix=TRUE,
                           returnf0ScoreInfo=FALSE, print=FALSE,
                           betaStart=NULL, f0Start=NULL)

### 2021-08-15 ManuscriptCode/Manuscript.Rmd
Reproducing manuscript results

---
title: Code to reproduce results in _Semi-parametric generalized linear model for
  binomial data with varying cluster sizes_
author: "Aniko Szabo"
output:
  html_document:
    df_print: paged
---

```{r setup, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, autodep=TRUE)

library(exchreg, quietly=TRUE)
library(knitr, quietly=TRUE)
library(dplyr, quietly=TRUE)
library(tidyr, quietly=TRUE)
library(ggplot2, quietly=TRUE)
theme_set(theme_classic())

library(CorrBin, quietly=TRUE) ## utilities for working with correlated binomial data
library(VGAM, quietly=TRUE) ## for beta-binomial regression
library(geepack, quietly=TRUE)  ## for GEE

set.seed(20210814)
```

# Application to boric acid developmental toxicity data

The boric acid data is included in the package as an example dataset `boric_acid`. We fit a GEE, beta-binomial, and semi-parameteric GLM model with a logit-link.

```{r BA_gee}
boric_acid$Trt <- factor(boric_acid$Trt, levels=c("0","0.1", "0.2","0.4"),
                         labels = c("0%", "0.1%", "0.2%", "0.4%"))
# convert data into one-row per cluster-member format needed for GEE
ba_long <- unwrap(boric_acid)

# run GEE
ba_gee <- geeglm(Resp ~ Trt, data=ba_long, id = ID, family=binomial,
                 corstr = "exch")
ba_gee
```

```{r BA_betabin}
# fit beta-binomial model
ba_bb <- vgam(cbind(NResp, ClusterSize - NResp) ~ Trt, weights = Freq,
                 data=boric_acid, family =  betabinomial(lmu="logitlink"))
summaryvglm(ba_bb)
```

```{r BA_spglm}
ba_spglm <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=boric_acid,
                  weights  = Freq)

ba_spglm
```

```{r Table1}
table1 <- data.frame(GEE = cbind(Est = coef(ba_gee), SE=sqrt(diag(vcov(ba_gee)))),
                BB = cbind(Est = coef(ba_bb)[-2],      # second value is logit(rho)   
                           SE=sqrt(diag(vcov(ba_bb)))[-2]),  
                SPGLM = cbind(Est = coef(ba_spglm), SE=ba_spglm$SE))
kable(table1, digits=3)
```

Fitted values for the outcome probabilities can be obtained for the SPGLM and Beta-binomial models, as well as using non-parametric estimates within each group.

```{r PredictBB}
newd <- data.frame(Trt=unique(boric_acid$Trt))

# beta-binomial
mns_rho <- predictvglm(ba_bb, type="link", newdata=newd, untransform=TRUE)

fit_bb <- tibble(newd, as_tibble(mns_rho)) %>%
  group_by(Trt) %>%
  mutate(ClusterSize = 10,
         NResp = list(0:10),
         Prob = list(dbetabinom(0:10, size=10, prob=mu, rho=rho))) %>%
  unnest(cols=c(NResp, Prob)) %>%
  mutate(Model="Beta-binomial")

```

```{r PredictSPGLM}
fit_spglm <- crossing(newd, NResp=0:10) %>%
  mutate(ClusterSize = 10,
         Prob = predict(ba_spglm, newdata=., type="prob", 
                          newn=ClusterSize, newevents = NResp),
         Model = "SPGLM")
```

```{r PredictNonparam}
fit_nonpar <- mc.est(boric_acid) %>%
  filter(ClusterSize == 10) %>%
  mutate(Model="Non-parametric")

```

```{r PredictPlot}
fits <- bind_rows(fit_bb, fit_spglm, fit_nonpar)

ggplot(fits, aes(x=NResp, y=Prob, colour=Model)) +
  facet_wrap(~Trt) +
  geom_point() +
  geom_line() +
  scale_x_continuous("# of affected animals", breaks=seq(0,10,by=2)) +
  ylab("Probability") +
  theme_bw()

```

# Simulation study

Goal:

 * simulate data with either a beta-binomial or a q-power backbone distribution that follows a semi-parametric logistic regression model
 * fit the data using our semi-parametric regression model, beta-binomial regression (potentially misspecified distribution), and GEE (marginal only)

## Simulating data

The dose levels will have mean 0.2, 0.4, 0.6, and 0.8, and intra-cluster correlation 0.4.

```{r SimOne}
generate_data <- function(pdf.fun, maxClusterSize = 10, aveClusterSize = 9,
                          doseLevels = 4, nPerDose = 25){
  # generate binomially distributed frequencies for each cluster-size
  ss <- expand.grid(Trt = 1:doseLevels, ClusterSize = 0:maxClusterSize) %>%
    group_by(Trt) %>%
    mutate(Freq = rmultinom(1, size=nPerDose, 
                            prob=dbinom(ClusterSize, size=maxClusterSize, 
                                        prob=aveClusterSize / maxClusterSize))) %>%
    filter(Freq > 0)

  res <- ran.CBData(sample.sizes = ss, 
                p.gen.fun = function(g) 0.2*g,
                rho.gen.fun = function(g) 0.4,
                pdf.fun = pdf.fun)
  res$Z <- rnorm(nrow(res), mean=0, sd=1)  
  res
}
```

## Analyzing a simulated dataset

```{r AnalyzeOne}
run_analysis <- function(dat){
  #GEE
  dat_long <- unwrap(dat)
  res_gee <- geeglm(Resp ~ Trt + Z, data=dat_long, id = ID, family=binomial,
                 corstr = "exch")
  #Beta-binomial
  res_bb <- vgam(cbind(NResp, ClusterSize - NResp) ~ Trt + Z, weights = Freq,
                 data=dat, family =  betabinomial(lmu="logitlink"))
  #SPGLM
  op <- options(warn=-1)
  res_spglm <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt + Z, data=dat,
                   weights  = Freq)
  options(op)

  nms <- names(coef(res_spglm))
  res <- bind_rows(
    GEE = tibble(Term = nms, 
                 Est=coef(res_gee), 
                 SE = sqrt(diag(vcov(res_gee)))),
    BB = tibble(Term = nms,
                Est = coef(res_bb)[-2],      # second value is logit(rho)   
                SE=sqrt(diag(vcov(res_bb)))[-2]),
    SPGLM = tibble(Term = nms,
                   Est = coef(res_spglm), 
                   SE=res_spglm$SE),
    .id = "Model")
  res
}
```

## Run simulations

```{r ActualSim}
R <- 2

# beta-binomial PDF
simBB.list <- 
  replicate(R,
            {simd <- generate_data(betabin.pdf);
             run_analysis(simd)}, 
            simplify = FALSE)

# q-power PDF
simQP.list <- 
  replicate(R,
            {simd <- generate_data(qpower.pdf);
             run_analysis(simd)}, 
            simplify = FALSE)

simBB <- bind_rows(simBB.list, .id="ID")
simQP <- bind_rows(simQP.list, .id="ID")

```

## Compare results

```{r Compare}
comb <- bind_rows(`Beta-Binomial` = simBB, `q-Power`= simQP, .id="Generator")

# get correct values of parameters on a logit-model scale
mns <- c(0.2, 0.4, 0.6, 0.8)
logit.mns <- log(mns/(1-mns))
correct <- tibble(Term = c("(Intercept)","Trt2","Trt3", "Trt4","Z"),
                  Correct = c(logit.mns[1], logit.mns[-1] - logit.mns[1], 0))
comb <- left_join(comb, correct, by="Term")

scomb <- comb %>%
  group_by(Generator, Term, Model) %>%
  summarize(bias = mean(Est-Correct, na.rm=TRUE),
            mc.se = sd(Est, na.rm=TRUE),
            se = mean(SE, na.rm=TRUE),
            .groups="drop_last") %>%
  mutate(rel.eff = mc.se / mc.se[Model=="GEE"]) %>%
  ungroup() %>%
  arrange(Generator, Model, Term)
kable(scomb, digits=3)
```

```{r ComparePlot}
ggplot(comb, aes(x=Model, y=Est, colour=Model)) +
  facet_grid(Generator~Term) +
  geom_hline(aes(yintercept=Correct)) +
  geom_jitter(alpha=0.3, height=0)
```

### 2021-08-15 nuweb/SPGLM.tex
Add extraction methods; clean up some documentation bugs to pass check

@@
 \mbox{}\verb@#'@{\tt @}\verb@param subset  an optional vector specifying a subset of observations to be used.@\\
-\mbox{}\verb@#'@{\tt @}\verb@param weight  an optional vector specifying observation weights.@\\
+\mbox{}\verb@#'@{\tt @}\verb@param weights an optional vector specifying observation weights.@\\
+\mbox{}\verb@#'@{\tt @}\verb@param offset  an optional vector specifying an offset term (predictor with a fixed coefficient of 1).@\\
 \mbox{}\verb@#'@{\tt @}\verb@param link      a link function for the mean.@\\
@@
 \mbox{}\verb@spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL, @\\
-\mbox{}\verb@                  control=list(eps=0.001, maxit=100), ...){@\\
+\mbox{}\verb@                  control=list(eps=0.001, maxit=100)){@\\
 \mbox{}\verb@@\\
@@
 \mbox{}\verb@    @\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape Fit model}\nobreak\ {\footnotesize \NWlink{nuweb6}{6}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape Fit model}\nobreak\ {\footnotesize \NWlink{nuweb7}{7}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@@\\
@@
 \mbox{}\verb@    names(referencef0) <- 0:N@\\
-\mbox{}\verb@    res <- list(coefficients = betas, SE = SEbeta, f0=referencef0, SEf0 = SEf0, mu0=mu0, niter = iter, @\\
+\mbox{}\verb@    rownames(vc) <- colnames(vc) <- c(names(betas), names(referencef0))@\\
+\mbox{}\verb@    res <- list(coefficients = betas, f0=referencef0, vcov = vc, mu0=mu0, niter = iter, @\\
 \mbox{}\verb@                loglik=llik, link = link, call = mc, terms = mt,@\\
@@
 \begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
-\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3}{, 3}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}.
+\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3}{, 3}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}\NWlink{nuweb6}{, 6}.
 
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@#' @{\tt @}\verb@rdname spglm@\\
 \mbox{}\verb@#' @{\tt @}\verb@export@\\
@@
 \mbox{}\verb@    beta <- x$coefficients@\\
-\mbox{}\verb@    SEbeta <- x$SE@\\
+\mbox{}\verb@    p <- length(beta)@\\
+\mbox{}\verb@    SEbeta <- sqrt(diag(x$vcov[1:p, 1:p]))@\\
 \mbox{}\verb@    z <- beta / SEbeta@\\
@@
 \begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
-\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3}{, 3}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}.
+\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3}{, 3}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}\NWlink{nuweb6}{, 6}.
+
+\item{}
+\end{list}
+\vspace{4ex}
+\end{flushleft}
+Custom extractors for the coefficients and variance-covariance matrix, with option to get only beta's (default), only the backbone pdf, or both.
+
+\begin{flushleft} \small\label{scrap4}\raggedright\small
+\NWtarget{nuweb4}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {4}}$\equiv$
+\vspace{-1ex}
+\begin{list}{}{} \item
+\mbox{}\verb@@\\
+\mbox{}\verb@#' Extracting components of fitted `spglm` object@\\
+\mbox{}\verb@#'@\\
+\mbox{}\verb@#' @{\tt @}\verb@param object fitted model of class \code{spglm}@\\
+\mbox{}\verb@#' @{\tt @}\verb@param beta logical, whether the regression parameter estimates / variances should be returned@\\
+\mbox{}\verb@#' @{\tt @}\verb@param f0 logical, whether the backbone density estimates / variances should be returned@\\
+\mbox{}\verb@#' @{\tt @}\verb@param ... not used, present fo consistency with generics@\\
+\mbox{}\verb@#' @{\tt @}\verb@importFrom stats coef@\\
+\mbox{}\verb@#' @{\tt @}\verb@export@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@coef.spglm <- function(object, beta=TRUE, f0=FALSE, ...){@\\
+\mbox{}\verb@  res <- NULL@\\
+\mbox{}\verb@  if (beta) res <- c(res, object$coefficients)@\\
+\mbox{}\verb@  if (f0) res <- c(res, object$f0)@\\
+\mbox{}\verb@  res@\\
+\mbox{}\verb@}@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@#' @{\tt @}\verb@rdname coef.spglm@\\
+\mbox{}\verb@#'@\\
+\mbox{}\verb@#' @{\tt @}\verb@importFrom stats vcov@\\
+\mbox{}\verb@#' @{\tt @}\verb@export@\\
+\mbox{}\verb@vcov.spglm <- function(object, beta=TRUE, f0=FALSE, ...){@\\
+\mbox{}\verb@  p <- length(object$coefficients)@\\
+\mbox{}\verb@  m <- nrow(object$vcov)@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  idx <- NULL@\\
+\mbox{}\verb@  if (beta) idx <- c(idx, 1:p)@\\
+\mbox{}\verb@  if (f0) idx <- c(idx, (p+1):m)@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  vc <- object$vcov[idx,idx]@\\
+\mbox{}\verb@  vc@\\
+\mbox{}\verb@}@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@#' @{\tt @}\verb@rdname coef.spglm@\\
+\mbox{}\verb@#'@\\
+\mbox{}\verb@#' @{\tt @}\verb@importFrom stats logLik@\\
+\mbox{}\verb@#' @{\tt @}\verb@export@\\
+\mbox{}\verb@logLik.spglm <- function(object, ...){@\\
+\mbox{}\verb@  object$loglik@\\
+\mbox{}\verb@}@\\
+\mbox{}\verb@                              @\\
+\mbox{}\verb@@{\NWsep}
+\end{list}
+\vspace{-1.5ex}
+\footnotesize
+\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
+\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3}{, 3}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}\NWlink{nuweb6}{, 6}.
 
@@
 
-\begin{flushleft} \small\label{scrap4}\raggedright\small
-\NWtarget{nuweb4}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {4}}$\equiv$
+\begin{flushleft} \small\label{scrap5}\raggedright\small
+\NWtarget{nuweb5}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {5}}$\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@#' Predict methods for SPGLM fits@\\
+\mbox{}\verb@#'@\\
 \mbox{}\verb@#' Obtains predictions from a fitted semi-parametric generalized linear model. Note that \code{offset}@\\
@@
 \mbox{}\verb@#' and "lp" requests the linear predictor.@\\
-\mbox{}\verb@#' @{\tt @}\verb@param newn if \code{newdata} is provided and \code{type="prob"}, an integer or integer vector specifying the @\\
+\mbox{}\verb@#' @{\tt @}\verb@param newn if \code{newdata} is provided and \code{type="prob"}, an integer or integer vector specifying the clustersize for the predictors@\\
 \mbox{}\verb@#' @{\tt @}\verb@param newevents if \code{newdata} is provided and \code{type="prob"}, an integer or integer vector specifying the @\\
 \mbox{}\verb@#'  number of events for the predictions@\\
+\mbox{}\verb@#' @{\tt @}\verb@param ... not used, present for consistency with generic@\\
 \mbox{}\verb@#' @{\tt @}\verb@export@\\
@@
 
-\begin{flushleft} \small\label{scrap5}\raggedright\small
-\NWtarget{nuweb5}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {5}}$\equiv$
+\begin{flushleft} \small\label{scrap6}\raggedright\small
+\NWtarget{nuweb6}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {6}}$\equiv$
 \vspace{-1ex}
@@
 
-\begin{flushleft} \small\label{scrap6}\raggedright\small
-\NWtarget{nuweb6}{} $\langle\,${\itshape Fit model}\nobreak\ {\footnotesize {6}}$\,\rangle\equiv$
+\begin{flushleft} \small\label{scrap7}\raggedright\small
+\NWtarget{nuweb7}{} $\langle\,${\itshape Fit model}\nobreak\ {\footnotesize {7}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@ @\hbox{$\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize \NWlink{nuweb9a}{9a}}$\,\rangle$}\verb@@\\
-\mbox{}\verb@ @\hbox{$\langle\,${\itshape Set up for E-step}\nobreak\ {\footnotesize \NWlink{nuweb7}{7}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@ @\hbox{$\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize \NWlink{nuweb10a}{10a}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@ @\hbox{$\langle\,${\itshape Set up for E-step}\nobreak\ {\footnotesize \NWlink{nuweb8}{8}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@ @\\
@@
 \mbox{}\verb@    @\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape E-step}\nobreak\ {\footnotesize \NWlink{nuweb8a}{8a}}$\,\rangle$}\verb@@\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape M-step}\nobreak\ {\footnotesize \NWlink{nuweb8b}{8b}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape E-step}\nobreak\ {\footnotesize \NWlink{nuweb9a}{9a}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape M-step}\nobreak\ {\footnotesize \NWlink{nuweb9b}{9b}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@    @\\
@@
 \mbox{}\verb@  @\\
-\mbox{}\verb@  @\hbox{$\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize \NWlink{nuweb9b}{9b}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@  @\hbox{$\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize \NWlink{nuweb10b}{10b}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@  @\\
@@
 
-\begin{flushleft} \small\label{scrap7}\raggedright\small
-\NWtarget{nuweb7}{} $\langle\,${\itshape Set up for E-step}\nobreak\ {\footnotesize {7}}$\,\rangle\equiv$
+\begin{flushleft} \small\label{scrap8}\raggedright\small
+\NWtarget{nuweb8}{} $\langle\,${\itshape Set up for E-step}\nobreak\ {\footnotesize {8}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
-\item \NWtxtMacroRefIn\ \NWlink{nuweb6}{6}.
+\item \NWtxtMacroRefIn\ \NWlink{nuweb7}{7}.
 
@@
 
-The value of $q_{y,N}^{(k)}((\boldsymbol{Z}_i; \boldsymbol{\beta}^{(k)})$ can be obtained from ${q_{y,N}^{(0)}}^{(k)}$ using \ref{M:densitySPGLM}.
+The value of $q_{y,N}^{(k)}(\boldsymbol{Z}_i; \boldsymbol{\beta}^{(k)})$ can be obtained from ${q_{y,N}^{(0)}}^{(k)}$ using \ref{M:densitySPGLM}.
 But it is also returned by \texttt{gldrmFit} as \texttt{fTiltMatrix}.
 
-\begin{flushleft} \small\label{scrap8}\raggedright\small
-\NWtarget{nuweb8a}{} $\langle\,${\itshape E-step}\nobreak\ {\footnotesize {8a}}$\,\rangle\equiv$
+\begin{flushleft} \small\label{scrap9}\raggedright\small
+\NWtarget{nuweb9a}{} $\langle\,${\itshape E-step}\nobreak\ {\footnotesize {9a}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 
-\begin{flushleft} \small\label{scrap9}\raggedright\small
-\NWtarget{nuweb8b}{} $\langle\,${\itshape M-step}\nobreak\ {\footnotesize {8b}}$\,\rangle\equiv$
+\begin{flushleft} \small\label{scrap10}\raggedright\small
+\NWtarget{nuweb9b}{} $\langle\,${\itshape M-step}\nobreak\ {\footnotesize {9b}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 
-\begin{flushleft} \small\label{scrap10}\raggedright\small
-\NWtarget{nuweb9a}{} $\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize {9a}}$\,\rangle\equiv$
+\begin{flushleft} \small\label{scrap11}\raggedright\small
+\NWtarget{nuweb10a}{} $\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize {10a}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 
-\begin{flushleft} \small\label{scrap11}\raggedright\small
-\NWtarget{nuweb9b}{} $\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize {9b}}$\,\rangle\equiv$
+\begin{flushleft} \small\label{scrap12}\raggedright\small
+\NWtarget{nuweb10b}{} $\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize {10b}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@  # calculate variance-covariance matrix@\\
-\mbox{}\verb@  vc <- solve(-bhess)@\\
-\mbox{}\verb@  SEbeta <- sqrt(diag(vc)[1:p])@\\
-\mbox{}\verb@  SEf0 <- sqrt(diag(vc)[p+1+(0:N)])@\\
+\mbox{}\verb@  bvc <- solve(-bhess)@\\
+\mbox{}\verb@  vc <- bvc[1:(p+N+1), 1:(p+N+1)]      # remove border@\\
 \mbox{}\verb@@{\NWsep}

### 2021-08-15 nuweb/SPregress.tex
Add extraction methods; clean up some documentation bugs to pass check

@@
 \mbox{}\verb@#'@{\tt @}\verb@param subset  an optional vector specifying a subset of observations to be used.@\\
-\mbox{}\verb@#'@{\tt @}\verb@param weight  an optional vector specifying observation weights.@\\
+\mbox{}\verb@#'@{\tt @}\verb@param weights an optional vector specifying observation weights.@\\
 \mbox{}\verb@#'@{\tt @}\verb@param link      a link function for the binomial distribution, the inverse of which models the covariate effects.@\\
@@
 \mbox{}\verb@#'@{\tt @}\verb@export@\\
-\mbox{}\verb@#' @{\tt @}\verb@importFrom stats terms model.matrix@\\
+\mbox{}\verb@#' @{\tt @}\verb@importFrom stats terms model.matrix binomial dbinom glm uniroot@\\
 \mbox{}\verb@@\\
-\mbox{}\verb@sprr <- function(formula, data, subset, weights, link="cloglog", mu1=NULL, start=NULL, control=list(eps=0.001, maxit=100), ...){@\\
+\mbox{}\verb@sprr <- function(formula, data, subset, weights, link="cloglog", mu1=NULL, start=NULL, control=list(eps=0.001, maxit=100)){@\\
 \mbox{}\verb@    fam <- binomial(link=link)@\\
@@
 \mbox{}\verb@    if (is.null(weights))       @\\
-\mbox{}\verb@      pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y)), trt="Trt",@\\
+\mbox{}\verb@      pooled <- CorrBin::CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y)), trt="Trt",@\\
 \mbox{}\verb@                       clustersize="ClusterSize", nresp="NResp")@\\
 \mbox{}\verb@    else                       @\\
-\mbox{}\verb@       pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), @\\
+\mbox{}\verb@       pooled <- CorrBin::CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), @\\
 \mbox{}\verb@                        trt="Trt", clustersize="ClusterSize", nresp="NResp", freq="Freq")@\\
-\mbox{}\verb@   est <- mc.est(pooled)@\\
+\mbox{}\verb@   est <- CorrBin::mc.est(pooled)@\\
 \mbox{}\verb@    q0 <- est$Prob[est$ClusterSize == N]@\\

### 2021-08-15 R/SPGLM.R

@@ -5,7 +5,8 @@
 #'@param formula a one-sided formula of the form \code{cbind(r, s) ~ predictors} where \code{r} and \code{s} give the number of responses and non-responses within each cluster, respectively (so the cluster size is \code{r+s}), and \code{predictors} describes the covariates.
 #'@param data  an optional matrix or data frame containing the variables in the formula \code{formula}. By default the variables are taken from \code{environment(formula).}
 #'@param subset an optional vector specifying a subset of observations to be used.
-#'@param weight an optional vector specifying observation weights.
+#'@param weights        an optional vector specifying observation weights.
+#'@param offset an optional vector specifying an offset term (predictor with a fixed coefficient of 1).
 #'@param link     a link function for the mean.
 #'@param mu0      an optional numeric value constraining the mean of the baseline distribution
 #'@param control        a list with parameters controlling the algorithm.
@@ -14,7 +15,7 @@
 #' @importFrom stats terms model.matrix model.frame model.offset model.response model.weights
 
 spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL, 
-                  control=list(eps=0.001, maxit=100), ...){
+                  control=list(eps=0.001, maxit=100)){
 
     
        if (missing(formula) || (length(formula) != 3L))
@@ -169,9 +170,8 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
         bhess <- rbind(cbind(hess1, border1, border2), c(border1,0,0), c(border2,0,0))
         
         # calculate variance-covariance matrix
-        vc <- solve(-bhess)
-        SEbeta <- sqrt(diag(vc)[1:p])
-        SEf0 <- sqrt(diag(vc)[p+1+(0:N)])
+        bvc <- solve(-bhess)
+        vc <- bvc[1:(p+N+1), 1:(p+N+1)]      # remove border
       
       
     
@@ -179,7 +179,8 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
     mt <- attr(mf, "terms")
     names(betas) <- colnames(mm)
     names(referencef0) <- 0:N
-    res <- list(coefficients = betas, SE = SEbeta, f0=referencef0, SEf0 = SEf0, mu0=mu0, niter = iter, 
[…]

### 2021-08-15 R/SPreg.R

@@ -35,16 +35,16 @@ lambda_from_p <- function(p.vec, n=length(p.vec)-1){
 #'@param formula a one-sided formula of the form \code{cbind(r, s) ~ predictors} where \code{r} and \code{s} give the number of responses and non-responses within each cluster, respectively (so the cluster size is \code{r+s}), and \code{predictors} describes the covariates.
 #'@param data  an optional matrix or data frame containing the variables in the formula \code{formula}. By default the variables are taken from \code{environment(formula).}
 #'@param subset an optional vector specifying a subset of observations to be used.
-#'@param weight an optional vector specifying observation weights.
+#'@param weights        an optional vector specifying observation weights.
 #'@param link     a link function for the binomial distribution, the inverse of which models the covariate effects.
 #'@param mu1      an optional value between 0 and 1 giving the maximal predicted marginal probability. If set to NULL (default), the algorithm will try to estimate it from the data.
 #'@param start    an optional list with elements named \code{beta}, \code{q}, and/or \code{mu1} giving starting values for estimation. If none or only some are specified, starting values are selected automatically.
 #'@param control        a list with parameters controlling the algorithm.
 #'@return an object of class \code{sprr} with the fitted model.
 #'@export
-#' @importFrom stats terms model.matrix
+#' @importFrom stats terms model.matrix binomial dbinom glm uniroot
 
-sprr <- function(formula, data, subset, weights, link="cloglog", mu1=NULL, start=NULL, control=list(eps=0.001, maxit=100), ...){
+sprr <- function(formula, data, subset, weights, link="cloglog", mu1=NULL, start=NULL, control=list(eps=0.001, maxit=100)){
     fam <- binomial(link=link)
     model_fun <- fam$linkinv
 
@@ -110,12 +110,12 @@ sprr <- function(formula, data, subset, weights, link="cloglog", mu1=NULL, start
          
          if (is.null(start$q)){
             if (is.null(weights))       
-              pooled <- CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y)), trt="Trt",
+              pooled <- CorrBin::CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y)), trt="Trt",
                                clustersize="ClusterSize", nresp="NResp")
             else                       
[…]

### 2021-08-15 nuweb/SPreg_develop.R

@@ -36,7 +36,7 @@ check(ex, check_dir = "c:/Temp", cran = TRUE, manual=TRUE)
 ## test SPGLM
 data(boric_acid)
 m <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=boric_acid, link="logit",
-           weights=boric_acid$Freq, control = list(eps=1e-5, maxit=1000))
+           weights=boric_acid$Freq, control = list(eps=1e-5, maxit=100))
 
 ba2 <- boric_acid[rep(1:nrow(boric_acid), boric_acid$Freq),]
 m2 <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=ba2, link="logit", control = list(eps=1e-5, maxit=1000))

### 2021-08-15 nuweb/SPGLM.tex
Accomodate zero values in backbone pdf

@@
 \mbox{}\verb@    @\\
-\mbox{}\verb@    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset, maxN=N)@\\
+\mbox{}\verb@    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset, maxN=N, spt=0:N)@\\
 \mbox{}\verb@    @\\
@@
 \mbox{}\verb@    nobs <- nrow(data_object$model_matrix)@\\
+\mbox{}\verb@    spt <- data_object$spt@\\
 \mbox{}\verb@    @\\
 \mbox{}\verb@    # ySptIndex is only used to calculate the log-likelihood, which we will not be using@\\
-\mbox{}\verb@    th <- getTheta(spt=(0:N)/N, f0=f0, mu=mu, weights=data_object$weights, ySptIndex=rep(1, nobs),@\\
+\mbox{}\verb@    th <- getTheta(spt=spt/N, f0=f0[spt+1], mu=mu, weights=data_object$weights, ySptIndex=rep(1, nobs),@\\
 \mbox{}\verb@                   thetaStart=NULL, thetaControl=theta.control())@\\
@@
 \mbox{}\verb@    nobs <- nrow(data_object$model_matrix)@\\
+\mbox{}\verb@    spt <- data_object$spt@\\
 \mbox{}\verb@   @\\
 \mbox{}\verb@    # ySptIndex is only used to calculate the log-likelihood, which we will not be using@\\
-\mbox{}\verb@    th <- getTheta(spt=(0:N)/N, f0=f0, mu=mu, weights=data_object$weights, ySptIndex=rep(1, nobs),@\\
+\mbox{}\verb@    th <- getTheta(spt=spt/N, f0=f0[spt+1], mu=mu, weights=data_object$weights, ySptIndex=rep(1, nobs),@\\
 \mbox{}\verb@                   thetaStart=NULL, thetaControl=theta.control())@\\
@@
 \mbox{}\verb@    hp <- sapply(0:N, function(t)dhyper(x=data_object$resp[,1], m=data_object$n, n=N-data_object$n, k=t))  @\\
-\mbox{}\verb@    probs <- rowSums(t(th$fTilt) * hp)@\\
+\mbox{}\verb@    @\\
+\mbox{}\verb@    # fill in 0's for values with no support@\\
+\mbox{}\verb@    fTilt <- matrix(0, ncol = nobs, nrow = N+1)@\\
+\mbox{}\verb@    fTilt[spt+1,] <- th$fTilt@\\
+\mbox{}\verb@    #probs <- rowSums(t(th$fTilt) * hp)@\\
+\mbox{}\verb@    probs <- rowSums(t(fTilt) * hp)@\\
 \mbox{}\verb@    probs@\\
@@
 \mbox{}\verb@ @\hbox{$\langle\,${\itshape Set up for E-step}\nobreak\ {\footnotesize \NWlink{nuweb8}{8}}$\,\rangle$}\verb@@\\
-\mbox{}\verb@ @\\
+\mbox{}\verb@ @\hbox{$\langle\,${\itshape Identify zero support}\nobreak\ {\footnotesize \NWlink{nuweb10b}{10b}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@@\\
 \mbox{}\verb@ iter <- 0@\\
@@
 \mbox{}\verb@  @\\
-\mbox{}\verb@  @\hbox{$\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize \NWlink{nuweb10b}{10b}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@  @\hbox{$\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize \NWlink{nuweb?}{?}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@  @\\
@@
 \mbox{}\verb@  offset2 <- offset[Ycomb[,"i"]]@\\
+\mbox{}\verb@@\\
 \mbox{}\verb@@{\NWsep}
@@
 \mbox{}\verb@  # convert fTiltMatrix to long vector, watching out for potentially different support@\\
-\mbox{}\verb@  y_idx <- match(Ycomb[,"y"], spt)@\\
+\mbox{}\verb@  y_idx <- match(Ycomb[,"y"], 0:N)@\\
 \mbox{}\verb@  fTilt2 <- ifelse(!is.na(y_idx), fTiltMatrix[cbind(Ycomb[,"i"], y_idx)], 0)@\\
@@
 \mbox{}\verb@  gldrmControl0 <- gldrm.control(returnfTiltMatrix = TRUE, returnf0ScoreInfo = FALSE, print=FALSE,@\\
-\mbox{}\verb@                                betaStart = betasPre, f0Start = referencef0Pre)@\\
+\mbox{}\verb@                                betaStart = betasPre, f0Start = referencef0Pre[spt+1])@\\
 \mbox{}\verb@@\\
@@
 \mbox{}\verb@                  @\\
-\mbox{}\verb@  fTiltMatrix <- mod$fTiltMatrix[obs_start,]@\\
+\mbox{}\verb@  fTiltMatrix[,spt+1] <- mod$fTiltMatrix[obs_start,]@\\
 \mbox{}\verb@  betas <- mod$beta@\\
-\mbox{}\verb@  referencef0 <- mod$f0@\\
-\mbox{}\verb@  spt <- round(mod$spt * N)@\\
+\mbox{}\verb@  referencef0[spt+1] <- mod$f0@\\
+\mbox{}\verb@#  spt <- round(mod$spt * N)@\\
 \mbox{}\verb@  llik <- c(log(rowSums(fTiltMatrix * hp)) %*% weights)@\\
@@
 \mbox{}\verb@  referencef0 <- est$Prob[est$ClusterSize == N]@\\
-\mbox{}\verb@  # ensure all positive values@\\
-\mbox{}\verb@  referencef0 <- (referencef0 + 1e-6)/(1+(N+1)*1e-6)@\\
 \mbox{}\verb@  @\\
-\mbox{}\verb@  fTiltMatrix <- matrix(rep(referencef0, times=nobs), byrow=TRUE, nrow=nobs, ncol=N+1)@\\
-\mbox{}\verb@  spt <- 0:N@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@@{\NWsep}
+\end{list}
+\vspace{-1.5ex}
+\footnotesize
+\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
+\item \NWtxtMacroRefIn\ \NWlink{nuweb7}{7}.
+
+\item{}
+\end{list}
+\vspace{4ex}
+\end{flushleft}
+Values of $y$ that cannot be a possible number of responses in a cluster of size $N$ corresponding to an observed cluster will always have $q_{y,N}=0$, so they need to be removed from the support during the `gldrm` estimation process (which will use $\log q_{y,N}$)
+
+\begin{flushleft} \small\label{scrap12}\raggedright\small
+\NWtarget{nuweb10b}{} $\langle\,${\itshape Identify zero support}\nobreak\ {\footnotesize {10b}}$\,\rangle\equiv$
+\vspace{-1ex}
+\begin{list}{}{} \item
+\mbox{}\verb@@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  # identify possible y-values@\\
+\mbox{}\verb@  spt <- sort(unique(Ycomb[,"y"]))@\\
+\mbox{}\verb@  data_object$spt <- spt@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  # ensure all positive values within support@\\
+\mbox{}\verb@  referencef0[spt+1] <- referencef0[spt+1] + 1e-6@\\
+\mbox{}\verb@  referencef0 <- referencef0 / sum(referencef0)@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  fTiltMatrix <- matrix(rep(referencef0, times=nobs), byrow=TRUE, @\\
+\mbox{}\verb@      nrow=nobs, ncol=N+1)@\\
 \mbox{}\verb@@\\
@@
 
-\begin{flushleft} \small\label{scrap12}\raggedright\small
-\NWtarget{nuweb10b}{} $\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize {10b}}$\,\rangle\equiv$
+\begin{flushleft} \small\label{scrap13}\raggedright\small
+\NWtarget{nuweb?}{} $\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize {?}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@  hess <- numDeriv::hessian(func=ll,  x=c(betas, log(referencef0)))@\\
+\mbox{}\verb@  hess_idx <- c(1:p, p+spt+1) # betas and f0 elements with non-zero support@\\
 \mbox{}\verb@  @\\
 \mbox{}\verb@  # revert to unlogged f0@\\
-\mbox{}\verb@  grad <- c(rep(1, p), 1/referencef0)@\\
-\mbox{}\verb@  hess1 <- diag(grad) %*% hess %*% diag(grad)@\\
+\mbox{}\verb@  grad <- c(rep(1, p), 1/referencef0[spt+1])@\\
+\mbox{}\verb@  hess1 <- diag(grad) %*% hess[hess_idx, hess_idx] %*% diag(grad)@\\
 \mbox{}\verb@  @\\
 \mbox{}\verb@  # create bordered hessian@\\
-\mbox{}\verb@  border1 <- c(rep(0, p), rep(1, N+1))  # gradient of sum-to-one constraint@\\
-\mbox{}\verb@  border2 <- c(rep(0, p), 0:N)          # gradient of fixed-mean constraint@\\
+\mbox{}\verb@  border1 <- c(rep(0, p), rep(1, length(spt)))  # gradient of sum-to-one constraint@\\
+\mbox{}\verb@  border2 <- c(rep(0, p), spt)          # gradient of fixed-mean constraint@\\
 \mbox{}\verb@  bhess <- rbind(cbind(hess1, border1, border2), c(border1,0,0), c(border2,0,0))@\\
@@
 \mbox{}\verb@  bvc <- solve(-bhess)@\\
-\mbox{}\verb@  vc <- bvc[1:(p+N+1), 1:(p+N+1)]      # remove border@\\
+\mbox{}\verb@  # remove border and set to 0 for zero-support values@\\
+\mbox{}\verb@  vc <- matrix(0, nrow=p+N+1, ncol=p+N+1)@\\
+\mbox{}\verb@  vc[hess_idx, hess_idx] <- bvc[1:(p+length(spt)), 1:(p+length(spt))]     @\\
 \mbox{}\verb@@{\NWsep}

### 2021-08-15 R/SPGLM.R

@@ -67,7 +67,7 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
       
     
     
-    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset, maxN=N)
+    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset, maxN=N, spt=0:N)
     
     
      
@@ -76,21 +76,8 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
                          trt="Trt", clustersize="ClusterSize", nresp="NResp", freq="Freq")
        est <- CorrBin::mc.est(pooled)
        referencef0 <- est$Prob[est$ClusterSize == N]
-       # ensure all positive values
-       referencef0 <- (referencef0 + 1e-6)/(1+(N+1)*1e-6)
        
-       fTiltMatrix <- matrix(rep(referencef0, times=nobs), byrow=TRUE, nrow=nobs, ncol=N+1)
-       spt <- 0:N
-
-       if (is.null(mu0))
-         mu0 <- weighted.mean(Y[,1]/rowSums(Y), weights)
        
-       # hypergeometric terms for log-likelihood calculation
-       hp <- sapply(0:N, function(t)dhyper(x=Y[,1], m=rowSums(Y), n=N-rowSums(Y), k=t))   
-       
-       # initial log-likelihood
-       llik <- log(rowSums(fTiltMatrix * hp)) %*% weights
-
      
      
        # replace each cluster with N+1 clusters of size N
@@ -110,8 +97,33 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
        mm2 <- mm[Ycomb[,"i"], ,drop=FALSE]
        weights2 <- weights[Ycomb[,"i"]]
        offset2 <- offset[Ycomb[,"i"]]
+
      
      
+       
+       # identify possible y-values
+       spt <- sort(unique(Ycomb[,"y"]))
+       data_object$spt <- spt
+       
+       # ensure all positive values within support
+       referencef0[spt+1] <- referencef0[spt+1] + 1e-6
+       referencef0 <- referencef0 / sum(referencef0)
+
+       
+       fTiltMatrix <- matrix(rep(referencef0, times=nobs), byrow=TRUE, 
+           nrow=nobs, ncol=N+1)
+
[…]

### 2021-08-15 nuweb/SPreg_develop.R

@@ -44,6 +44,11 @@ m2 <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=ba2, link="logit", con
 all.equal(coef(m), coef(m2))
 all.equal(m$loglik, m2$loglik)
 
+# values with 0 support
+m12 <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, 
+             data=subset(boric_acid, ClusterSize < 13),
+             link="logit", weights=Freq, control = list(eps=1e-5, maxit=100))
+
 # offset
 moff <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, offset=Freq, data=boric_acid)
 # prediction

### 2021-08-15 R/SPGLM.R
Make 'predict' an S3 method

@@ -293,7 +293,7 @@ logLik.spglm <- function(object, ...){
 #'  number of events for the predictions
 #' @param ... not used, present for consistency with generic
 #' @export
-#' @importFrom stats  .checkMFClasses .getXlevels delete.response
+#' @importFrom stats predict .checkMFClasses .getXlevels delete.response
 
 predict.spglm <- function(object, newdata=NULL,
                               type=c("mean", "prob", "tilt", "lp"),
@@ -307,7 +307,7 @@ predict.spglm <- function(object, newdata=NULL,
             .checkMFClasses(cl, m)
     mm <- model.matrix(Terms, m)
     data_object <- list(model_matrix=mm, offset = rep(0, nrow(mm)), weights=rep(1, nrow(mm)),
-                        maxN = object$data_object$maxN)
+                        maxN = object$data_object$maxN, spt = object$data_object$spt)
     if (!missing(newn)){
        if (!(length(newn) == 1L || length(newn) == nrow(newdata)))
           stop("'newn' should have length 1 or equal to the number of rows of 'newdata'")

### 2021-08-15 nuweb/SPreg_develop.R

@@ -20,9 +20,10 @@ use_data(boric_acid)
 ex <- as.package("../exchreg")
 nuweb(ex)
 
-shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet --run-viewer  SPregress.tex")
-shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet --run-viewer  SPGLM.tex")
-shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet --run-viewer  wgldrm.tex")
+#shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet --run-viewer  SPregress.tex")
+shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet  SPregress.tex")
+shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet SPGLM.tex")
+shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet wgldrm.tex")
 
 
 document(ex)

### 2021-08-16 ManuscriptCode/Manuscript.Rmd
Cleaner output for manuscript

@@
 
-```{r setup, warning=FALSE}
+```{r setup, warning=FALSE, message=FALSE}
 knitr::opts_chunk$set(echo = TRUE, cache=TRUE, autodep=TRUE)
 
-library(exchreg, quietly=TRUE)
-library(knitr, quietly=TRUE)
-library(dplyr, quietly=TRUE)
-library(tidyr, quietly=TRUE)
-library(ggplot2, quietly=TRUE)
+library(exchreg)
+library(knitr)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
 theme_set(theme_classic())
 
-library(CorrBin, quietly=TRUE) ## utilities for working with correlated binomial data
-library(VGAM, quietly=TRUE) ## for beta-binomial regression
-library(geepack, quietly=TRUE)  ## for GEE
+library(CorrBin) ## utilities for working with correlated binomial data
+library(VGAM) ## for beta-binomial regression
+library(geepack)  ## for GEE
 
 set.seed(20210814)
+data(boric_acid)
 ```
@@
 
-```{r Table1}
+```{r Table1a}
+table1 <- list(GEE = cbind(Est = coef(ba_gee), SE=sqrt(diag(vcov(ba_gee)))),
+                BB = cbind(Est = coef(ba_bb)[-2],      # second value is logit(rho)   
+                           SE=sqrt(diag(vcov(ba_bb)))[-2]),  
+                SPGLM = cbind(Est = coef(ba_spglm), SE=sqrt(diag(vcov(ba_spglm)))))
+table1
+```
+
+```{r Table1, eval=FALSE}
 table1 <- data.frame(GEE = cbind(Est = coef(ba_gee), SE=sqrt(diag(vcov(ba_gee)))),
@@
                            SE=sqrt(diag(vcov(ba_bb)))[-2]),  
-                SPGLM = cbind(Est = coef(ba_spglm), SE=ba_spglm$SE))
+                SPGLM = cbind(Est = coef(ba_spglm), SE=sqrt(diag(vcov(ba_spglm)))))
 kable(table1, digits=3)
@@
                    Est = coef(res_spglm), 
-                   SE=res_spglm$SE),
+                   SE= sqrt(diag(vcov(res_spglm)))),
     .id = "Model")
@@
 ```{r ActualSim}
-R <- 2
+R <- 50
 
@@
   geom_hline(aes(yintercept=Correct)) +
+  geom_boxplot(outlier.shape = NA) +
   geom_jitter(alpha=0.3, height=0)
 ```
+
+# Session information
+```{r Info}
+sessionInfo()
+```

### 2021-08-29 nuweb/SPGLM.tex
Make 'predict' an S3 method
Generation of random data following SPGLM

@@
 \begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
-\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3}{, 3}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}\NWlink{nuweb6}{, 6}.
+\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3}{, 3}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}\NWlink{nuweb6}{, 6}\NWlink{nuweb?}{, ?}.
 
@@
 \mbox{}\verb@#' @{\tt @}\verb@export@\\
-\mbox{}\verb@#' @{\tt @}\verb@importFrom stats  .checkMFClasses .getXlevels delete.response@\\
+\mbox{}\verb@#' @{\tt @}\verb@importFrom stats predict .checkMFClasses .getXlevels delete.response@\\
 \mbox{}\verb@@\\
@@
 \mbox{}\verb@    data_object <- list(model_matrix=mm, offset = rep(0, nrow(mm)), weights=rep(1, nrow(mm)),@\\
-\mbox{}\verb@                        maxN = object$data_object$maxN)@\\
+\mbox{}\verb@                        maxN = object$data_object$maxN, spt = object$data_object$spt)@\\
 \mbox{}\verb@    if (!missing(newn)){@\\
@@
 \mbox{}\verb@  @\\
-\mbox{}\verb@  @\hbox{$\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize \NWlink{nuweb?}{?}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@  @\hbox{$\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize \NWlink{nuweb11}{11}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@  @\\
@@
 \begin{flushleft} \small\label{scrap13}\raggedright\small
-\NWtarget{nuweb?}{} $\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize {?}}$\,\rangle\equiv$
+\NWtarget{nuweb11}{} $\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize {11}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \end{flushleft}
+\section{Simulate data from semi-parametric model}
+
+\begin{flushleft} \small\label{scrap14}\raggedright\small
+\NWtarget{nuweb?}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {?}}$\equiv$
+\vspace{-1ex}
+\begin{list}{}{} \item
+\mbox{}\verb@@\\
+\mbox{}\verb@#' Simulate data that follows an SPGLM @\\
+\mbox{}\verb@#'@\\
+\mbox{}\verb@#' @\\
+\mbox{}\verb@#' @{\tt @}\verb@param n vector of cluster-sizes for the simulated data.@\\
+\mbox{}\verb@#' @{\tt @}\verb@param means vector of mean event probability for each cluster. Values should be between 0 and 1. @\\
+\mbox{}\verb@#' @{\tt @}\verb@param q0 the backbone probability distribution of number of events in a cluster of maximal size. @\\
+\mbox{}\verb@#' @{\tt @}\verb@return a data frame with columns `Mean`, `ClusterSize`, `NResp`@\\
+\mbox{}\verb@#' @{\tt @}\verb@export@\\
+\mbox{}\verb@#' @{\tt @}\verb@importFrom stats rhyper@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@ran.spglm <- function(n, means, q0){@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@  if (length(n) != length(means)) stop("`n` and `means` should have the same length")@\\
+\mbox{}\verb@  if (any(means < 0) || any(means > 1)) stop("All values of `means` should be between 0 and 1")@\\
+\mbox{}\verb@  if (any(q0 < 0)) stop("All values of `q0` should be non-negative")@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  q0 <- q0 / sum(q0)@\\
+\mbox{}\verb@  N <- length(q0) - 1@\\
+\mbox{}\verb@  spt <- which(q0 > 0) - 1@\\
+\mbox{}\verb@  nobs <- length(n)@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  # get event probabilities at max clustersize for each cluster@\\
+\mbox{}\verb@  th <- getTheta(spt = spt/N, f0 = q0[spt+1], mu = means, @\\
+\mbox{}\verb@                 weights = rep(1, nobs), ySptIndex = rep(1, nobs))@\\
+\mbox{}\verb@                 @\\
+\mbox{}\verb@  probmat <- t(th$fTilt)@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  # random number of events at max clustersize@\\
+\mbox{}\verb@  nevents <- apply(probmat, 1, function(pvec)sample(x=spt, size=1, replace=FALSE, prob=pvec))@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  # random number of events in actual clusters using marginal compatibility@\\
+\mbox{}\verb@  nresp <- rhyper(nobs, m=nevents, n=N-nevents, k=n)@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  res <- data.frame(Mean = means, ClusterSize = n, NResp = nresp)@\\
+\mbox{}\verb@  res@\\
+\mbox{}\verb@}@\\
+\mbox{}\verb@@{\NWsep}
+\end{list}
+\vspace{-1.5ex}
+\footnotesize
+\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
+\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3}{, 3}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}\NWlink{nuweb6}{, 6}\NWlink{nuweb?}{, ?}.
+
+\item{}
+\end{list}
+\vspace{4ex}
+\end{flushleft}
 \end{document}

### 2021-08-29 R/SPGLM.R

@@ -396,3 +396,40 @@ spglm_loglik <- function(beta, f0, data_object, link){
     llik <- log(probs) %*% data_object$weights
     c(llik)
 }
+
+#' Simulate data that follows an SPGLM 
+#'
+#' 
+#' @param n vector of cluster-sizes for the simulated data.
+#' @param means vector of mean event probability for each cluster. Values should be between 0 and 1. 
+#' @param q0 the backbone probability distribution of number of events in a cluster of maximal size. 
+#' @return a data frame with columns `Mean`, `ClusterSize`, `NResp`
+#' @export
+#' @importFrom stats rhyper
+
+ran.spglm <- function(n, means, q0){
+
+  if (length(n) != length(means)) stop("`n` and `means` should have the same length")
+  if (any(means < 0) || any(means > 1)) stop("All values of `means` should be between 0 and 1")
+  if (any(q0 < 0)) stop("All values of `q0` should be non-negative")
+  
+  q0 <- q0 / sum(q0)
+  N <- length(q0) - 1
+  spt <- which(q0 > 0) - 1
+  nobs <- length(n)
+  
+  # get event probabilities at max clustersize for each cluster
+  th <- getTheta(spt = spt/N, f0 = q0[spt+1], mu = means, 
+                 weights = rep(1, nobs), ySptIndex = rep(1, nobs))
+                 
+  probmat <- t(th$fTilt)
+  
+  # random number of events at max clustersize
+  nevents <- apply(probmat, 1, function(pvec)sample(x=spt, size=1, replace=FALSE, prob=pvec))
+  
+  # random number of events in actual clusters using marginal compatibility
+  nresp <- rhyper(nobs, m=nevents, n=N-nevents, k=n)
+  
+  res <- data.frame(Mean = means, ClusterSize = n, NResp = nresp)
+  res
+}

### 2021-08-29 nuweb/SPreg_develop.R

@@ -85,7 +85,9 @@ mll <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=boric_acid,
              link=list(linkfun=laplace.link, linkinv=laplace.invlink, mu.eta=laplace.mu.eta),
             weights=boric_acid$Freq, control = list(eps=1e-5, maxit=1000))
 
-
+# random data
+dd <- ran.spglm(1:5, means=seq(0.2, 0.6, length.out = 5), q0 = dbinom(0:5, size=5, prob=0.3))
+dd
 
 ######### Testing wGLDRM
 run_gldrm <- 

### 2021-08-29 nuweb/SPGLM.tex
Insist on full range of probabilities

@@
 \mbox{}\verb@    @\\
-\mbox{}\verb@    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset, maxN=N, spt=0:N)@\\
+\mbox{}\verb@    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset, @\\
+\mbox{}\verb@                        maxN=N, spt=0:N)@\\
 \mbox{}\verb@    @\\
@@
 
+To ensure that the range of possible response probabilities covers 0 -- 1, we need to ensure that clusters with 0 and $N$ responses both are possible, buy adding small-weight fake clusters if needed.
+
 \begin{flushleft} \small\label{scrap8}\raggedright\small
@@
 \mbox{}\verb@@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  # identify if fake observations are needed @\\
+\mbox{}\verb@  # we need at least one observation with 0 events and one with 0 non-events@\\
+\mbox{}\verb@  fakeID <- numeric()@\\
+\mbox{}\verb@  # save original values@\\
+\mbox{}\verb@  Y0 <- Y@\\
+\mbox{}\verb@  mm0 <- mm@\\
+\mbox{}\verb@  offset0 <- offset@\\
+\mbox{}\verb@  weights0 <- weights@\\
+\mbox{}\verb@  nobs0 <- nobs@\\
+\mbox{}\verb@ @\\
+\mbox{}\verb@  if (!(0 %in% Y[,1])){  # 0 events@\\
+\mbox{}\verb@    Y <- rbind(Y, c(0, N))@\\
+\mbox{}\verb@    mm <- rbind(mm, colMeans(mm0))@\\
+\mbox{}\verb@    offset <- c(offset, mean(offset0))@\\
+\mbox{}\verb@    weights <- c(weights, 1e-7)@\\
+\mbox{}\verb@    nobs <- nobs + 1@\\
+\mbox{}\verb@    fakeID <- c(fakeID, nobs)@\\
+\mbox{}\verb@  }@\\
+\mbox{}\verb@  if (!(0 %in% Y[,2])){ # 0 non-events@\\
+\mbox{}\verb@    Y <- rbind(Y, c(N, 0))@\\
+\mbox{}\verb@    mm <- rbind(mm, colMeans(mm0))@\\
+\mbox{}\verb@    offset <- c(offset, mean(offset0))@\\
+\mbox{}\verb@    weights <- c(weights, 1e-7)@\\
+\mbox{}\verb@    nobs <- nobs + 1@\\
+\mbox{}\verb@    fakeID <- c(fakeID, nobs)@\\
+\mbox{}\verb@  }@\\
+\mbox{}\verb@  @\\
 \mbox{}\verb@  # replace each cluster with N+1 clusters of size N@\\
@@
 \mbox{}\verb@  @\\
+\mbox{}\verb@  #real_spt <- sort(unique(Ycomb[!(Ycomb[,"i"] %in% fakeID),"y"]))@\\
+\mbox{}\verb@  @\\
 \mbox{}\verb@  # ensure all positive values within support@\\
@@
 \mbox{}\verb@@\\
+\mbox{}\verb@  data_object_mod <- list(model_matrix = mm, resp=Y, n=rowSums(Y), @\\
+\mbox{}\verb@                          weights = weights, offset = offset, maxN = N, @\\
+\mbox{}\verb@                          spt = spt)@\\
 \mbox{}\verb@  ll <- function(x){@\\
-\mbox{}\verb@    spglm_loglik(beta=x[1:p], f0 = exp(x[-(1:p)]), data_object=data_object, link=link)@\\
+\mbox{}\verb@    spglm_loglik(beta=x[1:p], f0 = exp(x[-(1:p)]), data_object=data_object_mod, link=link)@\\
 \mbox{}\verb@  }@\\
@@
 \mbox{}\verb@  hess <- numDeriv::hessian(func=ll,  x=c(betas, log(referencef0)))@\\
-\mbox{}\verb@  hess_idx <- c(1:p, p+spt+1) # betas and f0 elements with non-zero support@\\
+\mbox{}\verb@  hess_idx <- c(1:p, p+spt+1) # betas and f0 elements with non-zero real support@\\
 \mbox{}\verb@  @\\

### 2021-08-29 R/SPGLM.R

@@ -67,7 +67,8 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
       
     
     
-    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset, maxN=N, spt=0:N)
+    data_object <- list(model_matrix=mm, resp=Y, n=rowSums(Y), weights=weights, offset=offset, 
+                        maxN=N, spt=0:N)
     
     
      
@@ -80,6 +81,34 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
        
      
      
+       
+       # identify if fake observations are needed 
+       # we need at least one observation with 0 events and one with 0 non-events
+       fakeID <- numeric()
+       # save original values
+       Y0 <- Y
+       mm0 <- mm
+       offset0 <- offset
+       weights0 <- weights
+       nobs0 <- nobs
+      
+       if (!(0 %in% Y[,1])){  # 0 events
+         Y <- rbind(Y, c(0, N))
+         mm <- rbind(mm, colMeans(mm0))
+         offset <- c(offset, mean(offset0))
+         weights <- c(weights, 1e-7)
+         nobs <- nobs + 1
+         fakeID <- c(fakeID, nobs)
+       }
+       if (!(0 %in% Y[,2])){ # 0 non-events
+         Y <- rbind(Y, c(N, 0))
+         mm <- rbind(mm, colMeans(mm0))
+         offset <- c(offset, mean(offset0))
+         weights <- c(weights, 1e-7)
+         nobs <- nobs + 1
+         fakeID <- c(fakeID, nobs)
+       }
+       
        # replace each cluster with N+1 clusters of size N
        new <- cbind(y=0:N)
        rep_idx <- rep(1:nrow(Y), each=nrow(new))
@@ -104,6 +133,8 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
        # identify possible y-values
        spt <- sort(unique(Ycomb[,"y"]))
        data_object$spt <- spt
+       
+       #real_spt <- sort(unique(Ycomb[!(Ycomb[,"i"] %in% fakeID),"y"]))
        
        # ensure all positive values within support
        referencef0[spt+1] <- referencef0[spt+1] + 1e-6
@@ -166,12 +197,15 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
       }
       
       
[…]

### 2021-08-29 nuweb/SPreg_develop.R

@@ -45,10 +45,21 @@ m2 <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=ba2, link="logit", con
 all.equal(coef(m), coef(m2))
 all.equal(m$loglik, m2$loglik)
 
-# values with 0 support
+# values with 0 support for in-between values
 m12 <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, 
              data=subset(boric_acid, ClusterSize < 13),
              link="logit", weights=Freq, control = list(eps=1e-5, maxit=100))
+
+# values with 0 support for 0 responses
+mzero <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, 
+             data=subset(boric_acid, NResp > 0),
+             link="logit", weights=Freq, control = list(eps=1e-5, maxit=100))
+
+# values with 0 support for 0 & N responses 
+mzero2 <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, 
+               data=subset(boric_acid, (NResp > 0) & (NResp < ClusterSize)),
+               link="logit", weights=Freq, control = list(eps=1e-5, maxit=100))
+
 
 # offset
 moff <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, offset=Freq, data=boric_acid)

### 2021-08-29 ManuscriptCode/Manuscript.Rmd
Update simulation study

@@
 
- * simulate data with either a beta-binomial or a q-power backbone distribution that follows a semi-parametric logistic regression model
+ * simulate data with four settings:
+
+  1. Both the beta-binomial and semi-parameteric model are correct - a binomial backbone, which belongs to the exponential family
+  2. Only the beta-binomial model is correct - generated from a beta-binomial distribution
+  3. Only the semi-parametric model is correct - generated from the spglm, with a zero-inflated beta-binomial backbone
+  4. Neither model is correct - generated from a zero-inflated beta-binomial distribution
  * fit the data using our semi-parametric regression model, beta-binomial regression (potentially misspecified distribution), and GEE (marginal only)
@@
 
-The dose levels will have mean 0.2, 0.4, 0.6, and 0.8, and intra-cluster correlation 0.4.
+We will have one binary predictor $X_1 \sim Bern(0.5)$ with $beta_1=0.5$ and one continuous predictor $X_2 \sim N(0,1)$ with $\beta_2 = 1$. The cluster size distribution will be binomial.
 
 ```{r SimOne}
-generate_data <- function(pdf.fun, maxClusterSize = 10, aveClusterSize = 9,
-                          doseLevels = 4, nPerDose = 25){
-  # generate binomially distributed frequencies for each cluster-size
-  ss <- expand.grid(Trt = 1:doseLevels, ClusterSize = 0:maxClusterSize) %>%
-    group_by(Trt) %>%
-    mutate(Freq = rmultinom(1, size=nPerDose, 
-                            prob=dbinom(ClusterSize, size=maxClusterSize, 
-                                        prob=aveClusterSize / maxClusterSize))) %>%
-    filter(Freq > 0)
+beta_correct <- c(0, 0.5, 1)
+sim_covariates <- function(n, maxClusterSize = 10, aveClusterSize = 7, beta=beta_correct){
+  x1 <- rbinom(n, size=1, prob=0.5)
+  x2 <- rnorm(n)
+  lp <- cbind(1, x1, x2) %*% beta
+  mns <- 1/(1+exp(-lp))
 
-  res <- ran.CBData(sample.sizes = ss, 
-                p.gen.fun = function(g) 0.2*g,
-                rho.gen.fun = function(g) 0.4,
-                pdf.fun = pdf.fun)
-  res$Z <- rnorm(nrow(res), mean=0, sd=1)  
+  cs <- rbinom(n, size=maxClusterSize, prob=aveClusterSize/maxClusterSize)
+
+  data.frame(ClusterSize = cs, X1=x1, X2=x2, p_resp = mns)
+}
+
+# add random response counts using SPGLM model
+add_response_spglm <- function(cov_data, q0){
+  # cov_data should come from sim_covariates
+
+  simres <- ran.spglm(n=cov_data$ClusterSize, means = cov_data$p_resp, q0 = q0)
+  res <- cov_data
+  res$NResp <- simres$NResp
+  res$Trt <- "All"
+  res$Freq <- 1
+  class(res) <- c("CBData", class(res))
+  res
+}
+
+# add random response counts from zero-inflated beta-binomial 
+add_response_zoibb <- function(cov_data, pstr0, rho){
+  # cov_data should come from sim_covariates
+
+  res <- cov_data
+  res$NResp <- rzoibetabinom(n = nrow(cov_data),
+                             size = cov_data$ClusterSize,
+                             prob = cov_data$p_resp,
+                             rho = rho,
+                             pstr0 = pstr0)
+  res$Trt <- "All"
+  res$Freq <- 1
+  class(res) <- c("CBData", class(res))
   res
@@
   dat_long <- unwrap(dat)
-  res_gee <- geeglm(Resp ~ Trt + Z, data=dat_long, id = ID, family=binomial,
+  res_gee <- geeglm(Resp ~ X1 + X2, data=dat_long, id = ID, family=binomial,
                  corstr = "exch")
   #Beta-binomial
-  res_bb <- vgam(cbind(NResp, ClusterSize - NResp) ~ Trt + Z, weights = Freq,
+  res_bb <- vgam(cbind(NResp, ClusterSize - NResp) ~  X1 + X2, weights = Freq,
                  data=dat, family =  betabinomial(lmu="logitlink"))
@@
   op <- options(warn=-1)
-  res_spglm <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt + Z, data=dat,
+  res_spglm <- spglm(cbind(NResp, ClusterSize - NResp) ~ X1 + X2, data=dat,
                    weights  = Freq)
@@
 
-```{r ActualSim}
+```{r ActualSim1, warning=FALSE}
 R <- 50
 
-# beta-binomial PDF
-simBB.list <- 
+# Scenario 1: binomial PDF
+sim1 <- 
   replicate(R,
-            {simd <- generate_data(betabin.pdf);
+            {simd <- sim_covariates(100) %>% add_response_zoibb(pstr=0, rho=0);
              run_analysis(simd)}, 
-            simplify = FALSE)
+            simplify = FALSE)  %>%
+   bind_rows(.id="ID")
+```
 
-# q-power PDF
-simQP.list <- 
+```{r ActualSim2, warning=FALSE}
+# Scenario 2: binomial PDF
+sim2 <- 
   replicate(R,
-            {simd <- generate_data(qpower.pdf);
+            {simd <- sim_covariates(100) %>% add_response_zoibb(pstr=0, rho=0.3);
              run_analysis(simd)}, 
-            simplify = FALSE)
+            simplify = FALSE) %>%
+   bind_rows(.id="ID")
 
-simBB <- bind_rows(simBB.list, .id="ID")
-simQP <- bind_rows(simQP.list, .id="ID")
+```
+
+```{r ActualSim3, warning=FALSE}
+# Scenario 3: SPGLM with zero-inflated betabinomial backbone
+sim3 <- 
+  replicate(R,
+            {simd <- sim_covariates(100) %>% 
+              add_response_spglm(q0 = dzoibetabinom(0:10, size=10, prob=0.4, rho = 0.3, pstr0 = 0.1));
+             run_analysis(simd)}, 
+            simplify = FALSE)  %>%
+   bind_rows(.id="ID")
+
+```
+
+```{r ActualSim4, warning=FALSE}
+# Scenario 4: zero-inflated betabinomial 
+sim4 <- 
+  replicate(R,
+            {simd <- sim_covariates(100) %>% add_response_zoibb(pstr=0.1, rho=0.3);
+             run_analysis(simd)}, 
+            simplify = FALSE)  %>%
+   bind_rows(.id="ID")
 
@@
 ```{r Compare}
-comb <- bind_rows(`Beta-Binomial` = simBB, `q-Power`= simQP, .id="Generator")
+comb <- bind_rows(`Both correct` = sim1, 
+                  `BB correct`= sim2,
+                  `SPGLM correct` = sim3, 
+                  `Neither correct` = sim4,
+                  .id="Generator")
 
 # get correct values of parameters on a logit-model scale
-mns <- c(0.2, 0.4, 0.6, 0.8)
-logit.mns <- log(mns/(1-mns))
-correct <- tibble(Term = c("(Intercept)","Trt2","Trt3", "Trt4","Z"),
-                  Correct = c(logit.mns[1], logit.mns[-1] - logit.mns[1], 0))
+correct <- tibble(Term = c("(Intercept)","X1","X2"),
+                  Correct = beta_correct)
 comb <- left_join(comb, correct, by="Term")

### 2021-09-03 R/SPGLM.R
Protect against singular hessian

@@ -217,10 +217,15 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
         bhess <- rbind(cbind(hess1, border1, border2), c(border1,0,0), c(border2,0,0))
         
         # calculate variance-covariance matrix
-        bvc <- solve(-bhess)
-        # remove border and set to 0 for zero-support values
-        vc <- matrix(0, nrow=p+N+1, ncol=p+N+1)
-        vc[hess_idx, hess_idx] <- bvc[1:(p+length(spt)), 1:(p+length(spt))]     
+        bvc <- tryCatch(solve(-bhess), error = function(e)NA)
+        # bvc <- solve(-bhess)
+        if (!is.na(bvc)){
+          # remove border and set to 0 for zero-support values
+          vc <- matrix(0, nrow=p+N+1, ncol=p+N+1)
+          vc[hess_idx, hess_idx] <- bvc[1:(p+length(spt)), 1:(p+length(spt))]     
+        } else {
+          vc <- matrix(NA, nrow=p+N+1, ncol=p+N+1)
+        }
       
       
     

### 2021-11-19 ManuscriptCode/Manuscript.Rmd
Switch to q-power backbone for misspecified model

@@
   2. Only the beta-binomial model is correct - generated from a beta-binomial distribution
-  3. Only the semi-parametric model is correct - generated from the spglm, with a zero-inflated beta-binomial backbone
-  4. Neither model is correct - generated from a zero-inflated beta-binomial distribution
+  3. Only the semi-parametric model is correct - generated from the spglm, with a q-power backbone
+  4. Neither model is correct - generated from a q-power distribution
  * fit the data using our semi-parametric regression model, beta-binomial regression (potentially misspecified distribution), and GEE (marginal only)
@@
 
-# add random response counts from zero-inflated beta-binomial 
-add_response_zoibb <- function(cov_data, pstr0, rho){
+# add random response counts from binomial, beta-binomial, or q-power
+add_response_param <- function(cov_data, rfun=function(n, cs, p)rbinom(n, size=cs, prob=p)){
   # cov_data should come from sim_covariates
@@
   res <- cov_data
-  res$NResp <- rzoibetabinom(n = nrow(cov_data),
-                             size = cov_data$ClusterSize,
-                             prob = cov_data$p_resp,
-                             rho = rho,
-                             pstr0 = pstr0)
+  res$NResp <- rfun(n = nrow(cov_data),
+                    cs = cov_data$ClusterSize,
+                    p = cov_data$p_resp)
   res$Trt <- "All"
@@
   class(res) <- c("CBData", class(res))
+  res
+}
+
+# random number generation from the q-power pdf
+rqpower <- function(n, size, prob, rho){
+  size <- rep(size, length.out = n)
+  prob <- rep(prob, length.out = n)
+  rho <- rep(rho, length.out = n)
+
+  res <- mapply(
+    function(s,p,r){
+      pdf <- qpower.pdf(p=p, rho=r, n=s)
+      sample(0:s, size=1, prob=pdf)},
+    size, prob, rho)
   res
@@
 ```{r AnalyzeOne}
-run_analysis <- function(dat){
+run_analysis <- function(dat, cs_predict=5){
   #GEE
@@
 
+  # predict PDF at cluster-size cs_predict & all covariates = 0
+  newd <- data.frame(X1 = 0, X2 = 0)
+  # beta-binomial
+  mns_rho <- predictvglm(res_bb, type="link", newdata=newd, untransform=TRUE)
+
+  fit_bb <- newd %>% 
+    crossing(NResp=0:cs_predict) %>%
+    mutate(ClusterSize = cs_predict,
+           Prob = dbetabinom(NResp, size=cs_predict, prob=mns_rho[1,"mu"], rho=mns_rho[1,"rho"])) %>%
+  mutate(Model="BB")
+
+  fit_spglm <- newd %>% 
+    crossing(NResp=0:cs_predict) %>%
+    mutate(ClusterSize = cs_predict,
+           Prob = predict(res_spglm, newdata=., type="prob", 
+                          newn=ClusterSize, newevents = NResp),
+           Model = "SPGLM")
+
   nms <- names(coef(res_spglm))
@@
     .id = "Model")
-  res
+
+  fit <- bind_rows(fit_bb, fit_spglm) %>%
+    mutate(Type = "PDF")
+
+  out <- res %>%
+    mutate(Type = "Beta") %>%
+    bind_rows(fit %>% 
+                mutate(Term = as.character(NResp)) %>%
+                select(Type, Model, Term, Est=Prob))
+
+  out
 }
@@
 ```{r ActualSim1, warning=FALSE}
-R <- 50
+R <- 100
+cs_predict <- 5
+p_predict <- 1/(1+exp(-c(beta_correct %*% c(0,0,0))))
 
@@
   replicate(R,
-            {simd <- sim_covariates(100) %>% add_response_zoibb(pstr=0, rho=0);
-             run_analysis(simd)}, 
+            {simd <- sim_covariates(100) %>% 
+                add_response_param(rfun=function(n, cs, p)rbinom(n, size=cs, prob=p));
+             run_analysis(simd, cs_predict = cs_predict)}, 
             simplify = FALSE)  %>%
    bind_rows(.id="ID")
+
+pdf1 <- tibble(NResp = 0:cs_predict) %>%
+  mutate(Correct = dbinom(NResp, size=cs_predict, prob=p_predict))
 ```
@@
 ```{r ActualSim2, warning=FALSE}
-# Scenario 2: binomial PDF
+# Scenario 2: beta-binomial PDF
 sim2 <- 
   replicate(R,
-            {simd <- sim_covariates(100) %>% add_response_zoibb(pstr=0, rho=0.3);
-             run_analysis(simd)}, 
+            {simd <- sim_covariates(100) %>% 
+                add_response_param(rfun=function(n, cs, p)rbetabinom(n, size=cs, prob=p, rho=0.3));
+             run_analysis(simd, cs_predict = cs_predict)}, 
             simplify = FALSE) %>%
@@
 
+pdf2 <- tibble(NResp = 0:cs_predict) %>%
+  mutate(Correct = dbetabinom(NResp, size=cs_predict, prob=p_predict, rho=0.3))
 ```
@@
 ```{r ActualSim3, warning=FALSE}
-# Scenario 3: SPGLM with zero-inflated betabinomial backbone
+# Scenario 3: SPGLM with q-power backbone
 sim3 <- 
@@
             {simd <- sim_covariates(100) %>% 
-              add_response_spglm(q0 = dzoibetabinom(0:10, size=10, prob=0.4, rho = 0.3, pstr0 = 0.1));
-             run_analysis(simd)}, 
+              add_response_spglm(q0 = qpower.pdf(n=10, p=0.4, rho = 0.3));
+             run_analysis(simd, cs_predict = cs_predict)}, 
             simplify = FALSE)  %>%
    bind_rows(.id="ID")
+
+# simulation-based pdf
+pdf3 <- ran.spglm(n=rep(cs_predict, 1e6), means = rep(p_predict, 1e6), 
+                  q0 = qpower.pdf(n=10, p=0.4, rho = 0.3)) %>%
+  count(NResp) %>%
+  mutate(Correct = n/sum(n)) %>%
+  select(-n)
 
@@
 ```{r ActualSim4, warning=FALSE}
-# Scenario 4: zero-inflated betabinomial 
+# Scenario 4: q-power 
 sim4 <- 
   replicate(R,
-            {simd <- sim_covariates(100) %>% add_response_zoibb(pstr=0.1, rho=0.3);
-             run_analysis(simd)}, 
+            {simd <- sim_covariates(100) %>% 
+                add_response_param(rfun=function(n, cs, p)rqpower(n, size=cs, prob=p, rho=0.3));
+             run_analysis(simd, cs_predict = cs_predict)}, 
             simplify = FALSE)  %>%
    bind_rows(.id="ID")
+
+pdf4 <- tibble(NResp = 0:cs_predict) %>%
+  mutate(Correct = qpower.pdf(n=cs_predict, p=p_predict, rho=0.3))
 
@@
 
-# get correct values of parameters on a logit-model scale
-correct <- tibble(Term = c("(Intercept)","X1","X2"),
+# get correct values of parameters 
+correct_beta <- tibble(Term = c("(Intercept)","X1","X2"),
                   Correct = beta_correct)
-comb <- left_join(comb, correct, by="Term")
+correct_pdf <- bind_rows(`Both correct` = pdf1, 
+                         `BB correct`= pdf2,
+                         `SPGLM correct` = pdf3, 
+                         `Neither correct` = pdf4,
+                         .id="Generator") %>%
+  mutate(Term = as.character(NResp))
 
-scomb <- comb %>%
+comb_beta <- left_join(comb %>% filter(Type == "Beta"), correct_beta, by="Term")
+
+comb_pdf <- left_join(comb %>% filter(Type=="PDF"), 
+                      correct_pdf, by=c("Term","Generator"))
+
+scomb <- comb_beta %>%
   group_by(Generator, Term, Model) %>%
@@
 ```{r ComparePlot}
-ggplot(comb, aes(x=Model, y=Est, colour=Model)) +
-  facet_grid(Generator~Term) +
-  geom_hline(aes(yintercept=Correct)) +
-  geom_boxplot(outlier.shape = NA) +
-  geom_jitter(alpha=0.3, height=0)
+ggplot(comb_beta, aes(x=Term, y=Est-Correct, colour=Model, shape=Model)) +
+  facet_wrap(~Generator) +
+  geom_hline(aes(yintercept=0)) +
+  stat_summary(fun.data =  mean_sdl, geom="pointrange", position=position_dodge(width=0.4))
+
+```
+
+```{r PDFplot}
+ggplot(comb_pdf, aes(x=NResp, y=Est, colour=Model)) +
+  facet_wrap(~Generator) +
+  geom_line(aes(y=Correct), col="black") +
+  stat_summary(fun.data =  mean_sdl, geom="pointrange", position=position_dodge(width=0.1))
+
 ```

### 2021-11-20 R/RcppExports.R
Convert setup part of getTheta to C++

@@ -1,0 +1,7 @@
+# Generated by using Rcpp::compileAttributes() -> do not edit by hand
+# Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
+
+getThetaC <- function(spt, f0, mu, weights, thetaStart, thetaControl) {
+    .Call(`_exchreg_getThetaC`, spt, f0, mu, weights, thetaStart, thetaControl)
+}
+

### 2021-11-20 R/wgldrm.R

@@ -319,6 +319,11 @@ getBeta <- function(x, y, spt, ySptIndex, f0, linkinv, mu.eta, offset, weights,
     return(list(beta=beta, mu=mu, th=th, llik=llik, iter=iter, conv=conv))
 }
 
+## usethis namespace: start
+#' @useDynLib exchreg, .registration = TRUE
+#' @importFrom Rcpp sourceCpp
+## usethis namespace: end
+
 ## Computes log(sum(exp(x))) with better precision
 logSumExp <- function(x)
 {

### 2021-11-20 nuweb/SPreg_develop.R

@@ -10,6 +10,7 @@ use_gpl3_license()
 desc::desc_add_author(given = "Xinran", family = "Qi", email = "xinqi@mcw.edu", role = "aut",
                comment = NULL, file = ".", normalize = FALSE)
 use_readme_rmd()
+use_rcpp_armadillo()
 
 # add data
 boric_acid <- CorrBin::read.CBData("z:/EOGeorge/Data/Binary/BoricAcidMousedata_processed.csv",
@@ -38,6 +39,7 @@ check(ex, check_dir = "c:/Temp", cran = TRUE, manual=TRUE)
 data(boric_acid)
 m <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=boric_acid, link="logit",
            weights=boric_acid$Freq, control = list(eps=1e-5, maxit=100))
+
 
 ba2 <- boric_acid[rep(1:nrow(boric_acid), boric_acid$Freq),]
 m2 <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=ba2, link="logit", control = list(eps=1e-5, maxit=1000))
@@ -99,6 +101,10 @@ mll <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=boric_acid,
 # random data
 dd <- ran.spglm(1:5, means=seq(0.2, 0.6, length.out = 5), q0 = dbinom(0:5, size=5, prob=0.3))
 dd
+
+# profiling -> getTheta takes most of the time
+profvis::profvis({m <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=boric_acid, link="logit",
+                   weights=boric_acid$Freq, control = list(eps=1e-5, maxit=100))})
 
 ######### Testing wGLDRM
 run_gldrm <- 
@@ -178,6 +184,7 @@ all.equal(res1[c("conv", "iter", "beta", "f0", "mu0", "llik")],
 
 all.equal(res1[c("conv", "iter", "beta", "f0", "mu0", "llik")],
           c(res3[c("conv", "iter", "beta", "f0", "mu0")], llik = res3$llik * sum(shelltox$Freq)))
+
 
 ######### Testing SPreg
 

### 2021-11-20 src/RcppExports.cpp

@@ -1,0 +1,39 @@
+// Generated by using Rcpp::compileAttributes() -> do not edit by hand
+// Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
+
+#include <RcppArmadillo.h>
+#include <Rcpp.h>
+
+using namespace Rcpp;
+
+#ifdef RCPP_USE_GLOBAL_ROSTREAM
+Rcpp::Rostream<true>&  Rcpp::Rcout = Rcpp::Rcpp_cout_get();
+Rcpp::Rostream<false>& Rcpp::Rcerr = Rcpp::Rcpp_cerr_get();
+#endif
+
+// getThetaC
+List getThetaC(arma::vec spt, arma::vec f0, arma::vec mu, arma::vec weights, arma::vec thetaStart, List thetaControl);
+RcppExport SEXP _exchreg_getThetaC(SEXP sptSEXP, SEXP f0SEXP, SEXP muSEXP, SEXP weightsSEXP, SEXP thetaStartSEXP, SEXP thetaControlSEXP) {
+BEGIN_RCPP
+    Rcpp::RObject rcpp_result_gen;
+    Rcpp::RNGScope rcpp_rngScope_gen;
+    Rcpp::traits::input_parameter< arma::vec >::type spt(sptSEXP);
+    Rcpp::traits::input_parameter< arma::vec >::type f0(f0SEXP);
+    Rcpp::traits::input_parameter< arma::vec >::type mu(muSEXP);
+    Rcpp::traits::input_parameter< arma::vec >::type weights(weightsSEXP);
+    Rcpp::traits::input_parameter< arma::vec >::type thetaStart(thetaStartSEXP);
+    Rcpp::traits::input_parameter< List >::type thetaControl(thetaControlSEXP);
+    rcpp_result_gen = Rcpp::wrap(getThetaC(spt, f0, mu, weights, thetaStart, thetaControl));
+    return rcpp_result_gen;
+END_RCPP
+}
+
+static const R_CallMethodDef CallEntries[] = {
+    {"_exchreg_getThetaC", (DL_FUNC) &_exchreg_getThetaC, 6},
+    {NULL, NULL, 0}
+};
+
+RcppExport void R_init_exchreg(DllInfo *dll) {
+    R_registerRoutines(dll, NULL, CallEntries, NULL, NULL);
+    R_useDynamicSymbols(dll, FALSE);
+}

### 2021-11-20 src/getTheta.cpp

@@ -1,0 +1,72 @@
+#include <RcppArmadillo.h>
+using namespace Rcpp;
+using namespace arma;
+// [[Rcpp::depends(RcppArmadillo)]]
+
+// Computes log(sum(exp(x))) with better precision
+double logSumExp(arma::vec x){
+  NumericVector xa = wrap(x);
+  int i = which_max(xa);
+  double m = xa[i];
+  xa.erase(i);
+  double lse = log1p(sum(exp(xa-m))) + m;
+  return lse;
+}
+
+// g function (logit transformation from appendix)
+arma::vec g(arma::vec mu, double m, double M){
+  arma::vec res = log(mu-m) - log(M-mu);
+  return res;
+}  
+  
+
+// [[Rcpp::export]]
+List getThetaC(arma::vec spt, arma::vec f0, arma::vec mu, arma::vec weights, 
+               arma::vec thetaStart, List thetaControl){
+ // bool logit  = thetaControl["logit"];
+//  double eps = thetaControl["eps"];
+//  int maxiter = thetaControl["maxiter"];
+//  double maxhalf = thetaControl["maxhalf"];
+//  double maxtheta = thetaControl["maxtheta"];
+ // bool logsumexp = thetaControl["logsumexp"];
+  
+  int sptN = spt.size();
+//  double m = min(spt);
+//  double M = max(spt);
+  int n = mu.size();
+
+  //initialize values
+  arma::vec theta = thetaStart;
+  arma::vec thetaOld(n);
+  arma::vec bPrimeErrOld(n);
+  LogicalVector conv(n);
+  conv.fill(FALSE);
+  LogicalVector maxedOut(n);
+  maxedOut.fill(FALSE);
+  
+  arma::rowvec oo = ones(sptN);
+    
+  mat fUnstd = f0 * exp(spt * theta.t()); // |spt| x n matrix of tilted f0 values
+  rowvec b = oo * fUnstd;  //column sums;
+  mat fTilt = fUnstd.each_row() / b;  // normalized
+  
+  colvec bPrime = fTilt.t() * spt;  // mean as a function of theta
+  colvec bPrime2(sptN); // variance as a function of theta
+  for (int j=0; j<sptN; j++){
+    bPrime2(j) = 0;
+    for (int i=0; i<n; i++){
+      bPrime2(j) += pow(spt(j) - bPrime(i), 2.0) * fTilt(i,j);
+    }
+  }
[…]

### 2021-11-21 nuweb/wgldrm.tex
Add multiple importFrom to pass checks
Convert setup part of getTheta to C++
Complete C++ re-implementation of getTheta

@@
 \mbox{}\verb@#' @{\tt @}\verb@keywords internal@\\
+\mbox{}\verb@#' @{\tt @}\verb@importFrom stats lm.fit lm.wfit weighted.mean@\\
 \mbox{}\verb@gldrm.control <- function(eps=1e-10, maxiter=100, returnfTiltMatrix=TRUE,@\\
@@
 \begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
-\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb4}{, 4}\NWlink{nuweb6}{, 6}\NWlink{nuweb10}{, 10}.
+\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb4}{, 4}\NWlink{nuweb6}{, 6}\NWlink{nuweb?}{, ?}.
 
@@
 \begin{list}{}{} \item
+\mbox{}\verb@@\\
+\mbox{}\verb@## usethis namespace: start@\\
+\mbox{}\verb@#' @{\tt @}\verb@useDynLib exchreg, .registration = TRUE@\\
+\mbox{}\verb@#' @{\tt @}\verb@importFrom Rcpp sourceCpp@\\
+\mbox{}\verb@## usethis namespace: end@\\
 \mbox{}\verb@@\\
@@
 \mbox{}\verb@             thetaControl() function.")@\\
-\mbox{}\verb@    logit <- thetaControl$logit@\\
-\mbox{}\verb@    eps <- thetaControl$eps@\\
-\mbox{}\verb@    maxiter <- thetaControl$maxiter@\\
-\mbox{}\verb@    maxhalf <- thetaControl$maxhalf@\\
-\mbox{}\verb@    maxtheta <- thetaControl$maxtheta@\\
-\mbox{}\verb@    logsumexp <- thetaControl$logsumexp@\\
-\mbox{}\verb@@\\
+\mbox{}\verb@             @\\
 \mbox{}\verb@    ## Define value from inputs@\\
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@    ## Value does not change@\\
-\mbox{}\verb@    gMu <- g(mu, m, M)@\\
-\mbox{}\verb@@\\
-\mbox{}\verb@    ## Initialize values@\\
-\mbox{}\verb@    theta <- thetaStart  # initial values required@\\
-\mbox{}\verb@    thetaOld <- bPrimeErrOld <- rep(NA, n)@\\
-\mbox{}\verb@    conv <- rep(FALSE, n)@\\
-\mbox{}\verb@    maxedOut <- rep(FALSE, n)@\\
-\mbox{}\verb@@\\
-\mbox{}\verb@    if (logsumexp) {@\\
-\mbox{}\verb@        logf0 <- log(f0)@\\
-\mbox{}\verb@        logfUnstd <- logf0 + tcrossprod(spt, theta)@\\
-\mbox{}\verb@        logb <- apply(logfUnstd, 2, logSumExp)@\\
-\mbox{}\verb@        fTilt <- exp(logfUnstd - rep(logb, each=sptN))@\\
-\mbox{}\verb@        normfact <- colSums(fTilt)@\\
-\mbox{}\verb@        normss <- which(normfact != 1)@\\
-\mbox{}\verb@        fTilt[, normss] <- fTilt[, normss, drop=FALSE] / rep(normfact[normss], each=sptN)@\\
-\mbox{}\verb@    } else {@\\
-\mbox{}\verb@        fUnstd <- f0 * exp(tcrossprod(spt, theta))  # |spt| x n matrix of tilted f0 values@\\
-\mbox{}\verb@        b <- colSums(fUnstd)@\\
-\mbox{}\verb@        fTilt <- fUnstd / rep(b, each=sptN)  # normalized@\\
-\mbox{}\verb@    }@\\
-\mbox{}\verb@    bPrime <- colSums(spt*fTilt)  # mean as a function of theta@\\
-\mbox{}\verb@    bPrime2 <- colSums(outer(spt, bPrime, "-")^2 * fTilt)  # variance as a function of theta@\\
-\mbox{}\verb@    bPrimeErr <- bPrime - mu  # used to assess convergence@\\
-\mbox{}\verb@@\\
-\mbox{}\verb@    ## Update theta until convergence@\\
-\mbox{}\verb@    conv <- (abs(bPrimeErr) < eps) |@\\
-\mbox{}\verb@        (theta==maxtheta & bPrimeErr<0) |@\\
-\mbox{}\verb@        (theta==-maxtheta & bPrimeErr>0)@\\
-\mbox{}\verb@    s <- which(!conv)@\\
-\mbox{}\verb@    iter <- 0@\\
-\mbox{}\verb@    while(length(s)>0 && iter<maxiter) {@\\
-\mbox{}\verb@        iter <- iter + 1@\\
-\mbox{}\verb@        bPrimeErrOld[s] <- bPrimeErr[s]  # used to assess convergence@\\
-\mbox{}\verb@@\\
-\mbox{}\verb@        ## 1) Update theta@\\
-\mbox{}\verb@        thetaOld[s] <- theta[s]@\\
-\mbox{}\verb@        if (logit) {@\\
-\mbox{}\verb@            tPrimeS <- (M-m) / ((bPrime[s]-m) * (M-bPrime[s])) * bPrime2[s]@\\
-\mbox{}\verb@                # t'(theta) from paper: temporary variable@\\
-\mbox{}\verb@                # only needed for the subset of observations that have not converged@\\
-\mbox{}\verb@            tPrimeS[is.na(tPrimeS) | tPrimeS==Inf] <- 0@\\
-\mbox{}\verb@                # If bPrime is on the boundary, then bPrime2 should be zero.@\\
-\mbox{}\verb@                # Exceptions are due to rounding error.@\\
-\mbox{}\verb@            thetaS <- theta[s] - (g(bPrime[s], m, M) - gMu[s]) / tPrimeS@\\
-\mbox{}\verb@        } else {@\\
-\mbox{}\verb@            thetaS <- theta[s] - bPrimeErr[s] / bPrime2[s]@\\
-\mbox{}\verb@        }@\\
-\mbox{}\verb@        thetaS[thetaS > maxtheta] <- maxtheta@\\
-\mbox{}\verb@        thetaS[thetaS < -maxtheta] <- -maxtheta@\\
-\mbox{}\verb@        theta[s] <- thetaS@\\
-\mbox{}\verb@@\\
-\mbox{}\verb@        ## 2) Update fTilt, bPrime, and bPrime2 and take half steps if bPrimeErr not improved@\\
-\mbox{}\verb@        ss <- s@\\
-\mbox{}\verb@        nhalf <- 0@\\
-\mbox{}\verb@        while(length(ss)>0 && nhalf<maxhalf) {@\\
-\mbox{}\verb@            ## 2a) Update fTilt, bPrime, and bPrime2@\\
-\mbox{}\verb@            if (logsumexp) {@\\
-\mbox{}\verb@                logfUnstd[, ss] <- logf0 + tcrossprod(spt, theta[ss])@\\
-\mbox{}\verb@                logb[ss] <- apply(logfUnstd[, ss, drop=FALSE], 2, logSumExp)@\\
-\mbox{}\verb@                fTilt[, ss] <- exp(logfUnstd[, ss, drop=FALSE] - rep(logb[ss], each=sptN))@\\
-\mbox{}\verb@                normfact <- colSums(fTilt[, ss, drop=FALSE])@\\
-\mbox{}\verb@                normss <- which(normfact != 1)@\\
-\mbox{}\verb@                fTilt[, ss[normss]] <- fTilt[, ss[normss], drop=FALSE] / rep(normfact[normss], each=sptN)@\\
-\mbox{}\verb@            } else {@\\
-\mbox{}\verb@                fUnstd[, ss] <- f0*exp(tcrossprod(spt, theta[ss]))  # |spt| x n matrix of tilted f0 values@\\
-\mbox{}\verb@                b[ss] <- colSums(fUnstd[, ss, drop=FALSE])@\\
-\mbox{}\verb@                fTilt[, ss] <- fUnstd[, ss, drop=FALSE] / rep(b[ss], each=sptN)  # normalized@\\
-\mbox{}\verb@            }@\\
-\mbox{}\verb@            bPrime[ss] <- colSums(spt*fTilt[, ss, drop=FALSE])  # mean as a function of theta@\\
-\mbox{}\verb@            bPrime2[ss] <- colSums(outer(spt, bPrime[ss], "-")^2 * fTilt[, ss, drop=FALSE])  # variance as a function of theta@\\
-\mbox{}\verb@            bPrimeErr[ss] <- bPrime[ss] - mu[ss]  # used to assess convergence@\\
-\mbox{}\verb@@\\
-\mbox{}\verb@            ## 2b) Take half steps if necessary@\\
-\mbox{}\verb@            ss <- ss[abs(bPrimeErr[ss]) > abs(bPrimeErrOld[ss])]@\\
-\mbox{}\verb@            if (length(ss) > 0) nhalf <- nhalf + 1@\\
-\mbox{}\verb@            theta[ss] <- (theta[ss] + thetaOld[ss]) / 2@\\
-\mbox{}\verb@        }@\\
-\mbox{}\verb@        ## If maximum half steps are exceeded, set theta to previous value@\\
-\mbox{}\verb@        maxedOut[ss] <- TRUE@\\
-\mbox{}\verb@        theta[ss] <- thetaOld[ss]@\\
-\mbox{}\verb@@\\
-\mbox{}\verb@        ## 3) Check convergence@\\
-\mbox{}\verb@        conv[s] <- (abs(bPrimeErr[s]) < eps) |@\\
-\mbox{}\verb@            (theta[s]==maxtheta & bPrimeErr[s]<0) |@\\
-\mbox{}\verb@            (theta[s]==-maxtheta & bPrimeErr[s]>0)@\\
-\mbox{}\verb@        s <- s[!conv[s] & !maxedOut[s]]@\\
-\mbox{}\verb@    }@\\
-\mbox{}\verb@@\\
+\mbox{}\verb@    res <- getThetaC(spt, f0, mu, thetaStart, thetaControl)@\\
 \mbox{}\verb@@\\
 \mbox{}\verb@    ## Calculate log-likelihood@\\
-\mbox{}\verb@    fTilt.extracted <- fTilt[cbind(ySptIndex, seq_along(ySptIndex))]@\\
+\mbox{}\verb@    fTilt.extracted <- res$fTilt[cbind(ySptIndex, seq_along(ySptIndex))]@\\
 \mbox{}\verb@    llik <- sum(weights[fTilt.extracted>0] * log(fTilt.extracted[fTilt.extracted>0]))@\\
 \mbox{}\verb@ @\\
-\mbox{}\verb@    list(theta=theta, fTilt=fTilt, bPrime=bPrime, bPrime2=bPrime2,@\\
-\mbox{}\verb@         llik=llik, conv=conv, iter=iter)@\\
+\mbox{}\verb@    list(theta=c(res$theta), fTilt=res$fTilt, bPrime=c(res$bPrime), bPrime2=c(res$bPrime2),@\\
+\mbox{}\verb@         llik=llik, conv=c(res$conv), iter=res$iter)@\\
 \mbox{}\verb@}@\\
@@
 \end{flushleft}
+The computationally intensive part of \texttt{getTheta} is implemented in C++. For now, only the \texttt{logsumexp = FALSE} and \texttt{logit = FALSE} settings are implemented.
+
 \begin{flushleft} \small\label{scrap4}\raggedright\small
-\NWtarget{nuweb10}{} \verb@"../R/wgldrm.R"@\nobreak\ {\footnotesize {10}}$\equiv$
+\NWtarget{nuweb10}{} \verb@"../src/getTheta.cpp"@\nobreak\ {\footnotesize {10}}$\equiv$
+\vspace{-1ex}
+\begin{list}{}{} \item
+\mbox{}\verb@@\\
+\mbox{}\verb@#include <RcppArmadillo.h>@\\
+\mbox{}\verb@using namespace Rcpp;@\\
+\mbox{}\verb@using namespace arma;@\\
+\mbox{}\verb@// [[Rcpp::depends(RcppArmadillo)]]@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@// Computes log(sum(exp(x))) with better precision@\\
+\mbox{}\verb@double logSumExp(arma::vec x){@\\
+\mbox{}\verb@  uword i = x.index_max();@\\
+\mbox{}\verb@  double m = x(i);@\\
+\mbox{}\verb@  x.shed_row(i);@\\
+\mbox{}\verb@  double lse = log1p(sum(exp(x-m))) + m;@\\
+\mbox{}\verb@  return lse;@\\
+\mbox{}\verb@}@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@// g function (logit transformation from appendix)@\\
+\mbox{}\verb@arma::vec g(arma::vec mu, double m, double M){@\\
+\mbox{}\verb@  arma::vec res = log(mu-m) - log(M-mu);@\\
+\mbox{}\verb@  return res;@\\
+\mbox{}\verb@}  @\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@@\\
+\mbox{}\verb@// [[Rcpp::export]]@\\
+\mbox{}\verb@List getThetaC(arma::vec spt, arma::vec f0, arma::vec mu,  @\\
+\mbox{}\verb@               arma::vec thetaStart, List thetaControl){@\\
+\mbox{}\verb@ // bool logit  = thetaControl["logit"];@\\
+\mbox{}\verb@  double eps = thetaControl["eps"];@\\
+\mbox{}\verb@  int maxiter = thetaControl["maxiter"];@\\
+\mbox{}\verb@  double maxhalf = thetaControl["maxhalf"];@\\
+\mbox{}\verb@  double maxtheta = thetaControl["maxtheta"];@\\
+\mbox{}\verb@ // bool logsumexp = thetaControl["logsumexp"];@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  int sptN = spt.size();@\\
+\mbox{}\verb@//  double m = min(spt);@\\
+\mbox{}\verb@//  double M = max(spt);@\\
+\mbox{}\verb@  int n = mu.size();@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@  //initialize values@\\
+\mbox{}\verb@  arma::vec theta = thetaStart;@\\
+\mbox{}\verb@  arma::vec thetaOld(n);@\\
+\mbox{}\verb@  arma::vec bPrimeErrOld(n);@\\
+\mbox{}\verb@  uvec conv(n, fill::zeros);@\\
+\mbox{}\verb@  uvec maxedOut(n, fill::zeros);@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  arma::rowvec oo(sptN, fill::ones);@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@  mat fUnstd = exp(spt * theta.t());@\\
+\mbox{}\verb@  fUnstd.each_col() %= f0;  // |spt| x n matrix of tilted f0 values@\\
+\mbox{}\verb@  rowvec b = oo * fUnstd;  //column sums;@\\
+\mbox{}\verb@  mat fTilt = fUnstd.each_row() / b;  // normalized@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@  colvec bPrime = fTilt.t() * spt;  // mean as a function of theta@\\
+\mbox{}\verb@  colvec bPrime2(n); // variance as a function of theta@\\
+\mbox{}\verb@  for (int j=0; j<n; j++){       //iterate over the columns of fTilt@\\
+\mbox{}\verb@    bPrime2(j) = 0;@\\
+\mbox{}\verb@    for (int i=0; i<sptN; i++){  //iterate over the rows of fTilt@\\
+\mbox{}\verb@      bPrime2(j) += pow(spt(i) - bPrime(j), 2.0) * fTilt(i,j);@\\
+\mbox{}\verb@    }@\\
+\mbox{}\verb@  }@\\
+\mbox{}\verb@  colvec bPrimeErr = bPrime - mu;  // used to assess convergence@\\
+\mbox{}\verb@    @\\
+\mbox{}\verb@  conv = (abs(bPrimeErr) < eps) || (theta==maxtheta && bPrimeErr<0) ||@\\
+\mbox{}\verb@      (theta==-maxtheta && bPrimeErr>0);@\\
+\mbox{}\verb@  uvec s = find(conv == 0);@\\
+\mbox{}\verb@  int iter = 0;@\\
+\mbox{}\verb@    @\\
+\mbox{}\verb@@\\
+\mbox{}\verb@ while(s.size() > 0 && iter < maxiter) {@\\
+\mbox{}\verb@    iter++;@\\
+\mbox{}\verb@    bPrimeErrOld(s) = bPrimeErr(s);  // used to assess convergence@\\
+\mbox{}\verb@    @\\
+\mbox{}\verb@// 1) Update theta@\\
+\mbox{}\verb@    thetaOld(s) = theta(s);@\\
+\mbox{}\verb@    colvec thetaS = theta(s) - bPrimeErr(s) / bPrime2(s);@\\
+\mbox{}\verb@    thetaS(find(thetaS > maxtheta)).fill(maxtheta);@\\
+\mbox{}\verb@    thetaS(find(thetaS < -maxtheta)).fill(-maxtheta);@\\
+\mbox{}\verb@    theta(s)= thetaS;@\\
+\mbox{}\verb@    @\\
+\mbox{}\verb@// 2) Update fTilt, bPrime, and bPrime2 and take half steps if bPrimeErr not improved@\\
+\mbox{}\verb@    uvec ss = s;@\\
+\mbox{}\verb@    int nhalf = 0;@\\
+\mbox{}\verb@    while(ss.size() > 0 && nhalf < maxhalf) {@\\
+\mbox{}\verb@// 2a) Update fTilt, bPrime, and bPrime2@\\
+\mbox{}\verb@      fUnstd.cols(ss) = exp(spt * theta(ss).t());@\\
+\mbox{}\verb@      fUnstd.each_col(ss) %= f0;  // |spt| x n matrix of tilted f0 values@\\
+\mbox{}\verb@      b = oo * fUnstd.cols(ss);  //column sums;@\\
+\mbox{}\verb@      mat tmp = fUnstd.cols(ss);@\\
+\mbox{}\verb@      fTilt.cols(ss) = tmp.each_row() / b;  // normalized@\\
+\mbox{}\verb@      @\\
+\mbox{}\verb@      bPrime(ss) = fTilt.cols(ss).t() * spt;  // mean as a function of theta@\\
+\mbox{}\verb@      for (uword j=0; j<ss.size(); j++){       //iterate over the columns of fTilt[,ss]@\\
+\mbox{}\verb@        bPrime2(ss(j)) = 0;@\\
+\mbox{}\verb@        for (int i=0; i<sptN; i++){  //iterate over the rows of fTilt[,ss]@\\
+\mbox{}\verb@          bPrime2(ss(j)) += pow(spt(i) - bPrime(ss(j)), 2.0) * fTilt(i,ss(j));@\\
+\mbox{}\verb@        }@\\
+\mbox{}\verb@      }@\\
+\mbox{}\verb@      bPrimeErr(ss) = bPrime(ss) - mu(ss); @\\
+\mbox{}\verb@      @\\
+\mbox{}\verb@// 2b) Take half steps if necessary@\\
+\mbox{}\verb@      ss = ss(find(abs(bPrimeErr(ss)) > abs(bPrimeErrOld(ss))));@\\
+\mbox{}\verb@      if (ss.size() > 0){@\\
+\mbox{}\verb@        nhalf++;@\\
+\mbox{}\verb@      } @\\
+\mbox{}\verb@      theta(ss) = (theta(ss) + thetaOld(ss)) / 2;@\\
+\mbox{}\verb@    }@\\
+\mbox{}\verb@// If maximum half steps are exceeded, set theta to previous value@\\
+\mbox{}\verb@    maxedOut(ss).fill(1);@\\
+\mbox{}\verb@    theta(ss) = thetaOld(ss);@\\
+\mbox{}\verb@    @\\
+\mbox{}\verb@// 3) Check convergence@\\
+\mbox{}\verb@    conv(s) = (abs(bPrimeErr(s)) < eps) ||  (theta(s)==maxtheta && bPrimeErr(s) < 0) ||@\\
+\mbox{}\verb@      (theta(s)==-maxtheta && bPrimeErr(s) > 0);@\\
+\mbox{}\verb@    s = s(find(conv(s) == 0 && maxedOut(s) == 0));@\\
+\mbox{}\verb@ }@\\
+\mbox{}\verb@      @\\
+\mbox{}\verb@  List res;@\\
+\mbox{}\verb@  res["theta"] = theta;@\\
+\mbox{}\verb@  res["fTilt"] = fTilt;@\\
+\mbox{}\verb@  res["bPrime"] = bPrime;@\\
+\mbox{}\verb@  res["bPrime2"] = bPrime2;@\\
+\mbox{}\verb@  res["conv"] = conv;@\\
+\mbox{}\verb@  res["iter"] = iter;@\\
+\mbox{}\verb@  return res;@\\
+\mbox{}\verb@}@\\
+\mbox{}\verb@@\\
+\mbox{}\verb@@{\NWsep}
+\end{list}
+\vspace{-1.5ex}
+\footnotesize
+\begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
+
+\item{}
+\end{list}
+\vspace{4ex}
+\end{flushleft}
+\begin{flushleft} \small\label{scrap5}\raggedright\small
+\NWtarget{nuweb?}{} \verb@"../R/wgldrm.R"@\nobreak\ {\footnotesize {?}}$\equiv$
 \vspace{-1ex}

### 2021-11-21 R/RcppExports.R

@@ -1,7 +1,7 @@
 # Generated by using Rcpp::compileAttributes() -> do not edit by hand
 # Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
 
-getThetaC <- function(spt, f0, mu, weights, thetaStart, thetaControl) {
-    .Call(`_exchreg_getThetaC`, spt, f0, mu, weights, thetaStart, thetaControl)
+getThetaC <- function(spt, f0, mu, thetaStart, thetaControl) {
+    .Call(`_exchreg_getThetaC`, spt, f0, mu, thetaStart, thetaControl)
 }
 

### 2021-11-21 R/wgldrm.R

@@ -408,13 +408,7 @@ getTheta <- function(spt, f0, mu, weights, ySptIndex, thetaStart=NULL, thetaCont
     if (class(thetaControl) != "thetaControl")
         stop("thetaControl must be an object of class \'thetaControl\' returned by
              thetaControl() function.")
-    logit <- thetaControl$logit
-    eps <- thetaControl$eps
-    maxiter <- thetaControl$maxiter
-    maxhalf <- thetaControl$maxhalf
-    maxtheta <- thetaControl$maxtheta
-    logsumexp <- thetaControl$logsumexp
-
+             
     ## Define value from inputs
     sptN <- length(spt)
     m <- min(spt)
@@ -441,103 +435,14 @@ getTheta <- function(spt, f0, mu, weights, ySptIndex, thetaStart=NULL, thetaCont
     if (length(thetaStart) != n)
         stop("thetaStart must be a vector with length equal length(mu)")
 
-    ## Value does not change
-    gMu <- g(mu, m, M)
-
-    ## Initialize values
-    theta <- thetaStart  # initial values required
-    thetaOld <- bPrimeErrOld <- rep(NA, n)
-    conv <- rep(FALSE, n)
-    maxedOut <- rep(FALSE, n)
-
-    if (logsumexp) {
-        logf0 <- log(f0)
-        logfUnstd <- logf0 + tcrossprod(spt, theta)
-        logb <- apply(logfUnstd, 2, logSumExp)
-        fTilt <- exp(logfUnstd - rep(logb, each=sptN))
-        normfact <- colSums(fTilt)
-        normss <- which(normfact != 1)
-        fTilt[, normss] <- fTilt[, normss, drop=FALSE] / rep(normfact[normss], each=sptN)
-    } else {
-        fUnstd <- f0 * exp(tcrossprod(spt, theta))  # |spt| x n matrix of tilted f0 values
-        b <- colSums(fUnstd)
-        fTilt <- fUnstd / rep(b, each=sptN)  # normalized
-    }
-    bPrime <- colSums(spt*fTilt)  # mean as a function of theta
-    bPrime2 <- colSums(outer(spt, bPrime, "-")^2 * fTilt)  # variance as a function of theta
-    bPrimeErr <- bPrime - mu  # used to assess convergence
-
-    ## Update theta until convergence
-    conv <- (abs(bPrimeErr) < eps) |
[…]

### 2021-11-21 nuweb/SPreg_develop.R

@@ -102,7 +102,8 @@ mll <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=boric_acid,
 dd <- ran.spglm(1:5, means=seq(0.2, 0.6, length.out = 5), q0 = dbinom(0:5, size=5, prob=0.3))
 dd
 
-# profiling -> getTheta takes most of the time
+# profiling -> getTheta takes most of the time originally
+# after rewriting it in C, the dhyper calls within spglm_probs stands out
 profvis::profvis({m <- spglm(cbind(NResp, ClusterSize - NResp) ~ Trt, data=boric_acid, link="logit",
                    weights=boric_acid$Freq, control = list(eps=1e-5, maxit=100))})
 
@@ -186,6 +187,114 @@ all.equal(res1[c("conv", "iter", "beta", "f0", "mu0", "llik")],
           c(res3[c("conv", "iter", "beta", "f0", "mu0")], llik = res3$llik * sum(shelltox$Freq)))
 
 
+## Testing getTheta C-version
+getThetaR <- function(spt, f0, mu, thetaStart=NULL, thetaControl=theta.control())
+{
+  ## Extract control arguments
+  if (class(thetaControl) != "thetaControl")
+    stop("thetaControl must be an object of class \'thetaControl\' returned by
+             thetaControl() function.")
+  logit <- thetaControl$logit
+  eps <- thetaControl$eps
+  maxiter <- thetaControl$maxiter
+  maxhalf <- thetaControl$maxhalf
+  maxtheta <- thetaControl$maxtheta
+
+  ## Define value from inputs
+  sptN <- length(spt)
+  m <- min(spt)
+  M <- max(spt)
+  n <- length(mu)
+  
+  ## Format arguments
+  spt <- as.vector(spt)
+  f0 <- as.vector(f0)
+  mu <- as.vector(mu)
+  thetaStart <- as.vector(thetaStart)
+  
+  ## Value does not change
+  gMu <- exchreg:::g(mu, m, M)
+  
+  ## Initialize values
+  theta <- thetaStart  # initial values required
+  thetaOld <- bPrimeErrOld <- rep(NA, n)
+  conv <- rep(FALSE, n)
+  maxedOut <- rep(FALSE, n)
+  
+
[…]

### 2021-11-21 src/RcppExports.cpp

@@ -12,24 +12,23 @@ Rcpp::Rostream<false>& Rcpp::Rcerr = Rcpp::Rcpp_cerr_get();
 #endif
 
 // getThetaC
-List getThetaC(arma::vec spt, arma::vec f0, arma::vec mu, arma::vec weights, arma::vec thetaStart, List thetaControl);
-RcppExport SEXP _exchreg_getThetaC(SEXP sptSEXP, SEXP f0SEXP, SEXP muSEXP, SEXP weightsSEXP, SEXP thetaStartSEXP, SEXP thetaControlSEXP) {
+List getThetaC(arma::vec spt, arma::vec f0, arma::vec mu, arma::vec thetaStart, List thetaControl);
+RcppExport SEXP _exchreg_getThetaC(SEXP sptSEXP, SEXP f0SEXP, SEXP muSEXP, SEXP thetaStartSEXP, SEXP thetaControlSEXP) {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
     Rcpp::traits::input_parameter< arma::vec >::type spt(sptSEXP);
     Rcpp::traits::input_parameter< arma::vec >::type f0(f0SEXP);
     Rcpp::traits::input_parameter< arma::vec >::type mu(muSEXP);
-    Rcpp::traits::input_parameter< arma::vec >::type weights(weightsSEXP);
     Rcpp::traits::input_parameter< arma::vec >::type thetaStart(thetaStartSEXP);
     Rcpp::traits::input_parameter< List >::type thetaControl(thetaControlSEXP);
-    rcpp_result_gen = Rcpp::wrap(getThetaC(spt, f0, mu, weights, thetaStart, thetaControl));
+    rcpp_result_gen = Rcpp::wrap(getThetaC(spt, f0, mu, thetaStart, thetaControl));
     return rcpp_result_gen;
 END_RCPP
 }
 
 static const R_CallMethodDef CallEntries[] = {
-    {"_exchreg_getThetaC", (DL_FUNC) &_exchreg_getThetaC, 6},
+    {"_exchreg_getThetaC", (DL_FUNC) &_exchreg_getThetaC, 5},
     {NULL, NULL, 0}
 };
 

### 2021-11-21 src/getTheta.cpp

@@ -1,3 +1,4 @@
+
 #include <RcppArmadillo.h>
 using namespace Rcpp;
 using namespace arma;
@@ -5,11 +6,10 @@ using namespace arma;
 
 // Computes log(sum(exp(x))) with better precision
 double logSumExp(arma::vec x){
-  NumericVector xa = wrap(x);
-  int i = which_max(xa);
-  double m = xa[i];
-  xa.erase(i);
-  double lse = log1p(sum(exp(xa-m))) + m;
+  uword i = x.index_max();
+  double m = x(i);
+  x.shed_row(i);
+  double lse = log1p(sum(exp(x-m))) + m;
   return lse;
 }
 
@@ -21,13 +21,13 @@ arma::vec g(arma::vec mu, double m, double M){
   
 
 // [[Rcpp::export]]
-List getThetaC(arma::vec spt, arma::vec f0, arma::vec mu, arma::vec weights, 
+List getThetaC(arma::vec spt, arma::vec f0, arma::vec mu,  
                arma::vec thetaStart, List thetaControl){
  // bool logit  = thetaControl["logit"];
-//  double eps = thetaControl["eps"];
-//  int maxiter = thetaControl["maxiter"];
-//  double maxhalf = thetaControl["maxhalf"];
-//  double maxtheta = thetaControl["maxtheta"];
+  double eps = thetaControl["eps"];
+  int maxiter = thetaControl["maxiter"];
+  double maxhalf = thetaControl["maxhalf"];
+  double maxtheta = thetaControl["maxtheta"];
  // bool logsumexp = thetaControl["logsumexp"];
   
   int sptN = spt.size();
@@ -39,28 +39,80 @@ List getThetaC(arma::vec spt, arma::vec f0, arma::vec mu, arma::vec weights,
   arma::vec theta = thetaStart;
   arma::vec thetaOld(n);
   arma::vec bPrimeErrOld(n);
-  LogicalVector conv(n);
-  conv.fill(FALSE);
-  LogicalVector maxedOut(n);
-  maxedOut.fill(FALSE);
+  uvec conv(n, fill::zeros);
+  uvec maxedOut(n, fill::zeros);
   
-  arma::rowvec oo = ones(sptN);
-    
-  mat fUnstd = f0 * exp(spt * theta.t()); // |spt| x n matrix of tilted f0 values
+  arma::rowvec oo(sptN, fill::ones);
+
[…]

### 2021-11-21 SPsims/SPsim.Rmd
Split off running SAS from loading results

@@
 }
+```
+
+```{r LoadResults}
 gee.res <- read.delim("GEEest.txt") %>%

### 2021-11-23 nuweb/SPGLM.tex
Protect against singular hessian
Complete C++ re-implementation of getTheta
Reuse precomputed hypergeometric probabilities in Hessian calculation

@@
 \begin{list}{}{\setlength{\itemsep}{-\parsep}\setlength{\itemindent}{-\leftmargin}}
-\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3}{, 3}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}\NWlink{nuweb6}{, 6}\NWlink{nuweb?}{, ?}.
+\item \NWtxtFileDefBy\ \NWlink{nuweb1}{1}\NWlink{nuweb3}{, 3}\NWlink{nuweb4}{, 4}\NWlink{nuweb5}{, 5}\NWlink{nuweb6}{, 6}\NWlink{nuweb12}{, 12}.
 
@@
 \mbox{}\verb@                   @\\
-\mbox{}\verb@    hp <- sapply(0:N, function(t)dhyper(x=data_object$resp[,1], m=data_object$n, n=N-data_object$n, k=t))  @\\
+\mbox{}\verb@    if (is.null(data_object$hyper_probs)){                   @\\
+\mbox{}\verb@      hp <- sapply(0:N, function(t)dhyper(x=data_object$resp[,1], m=data_object$n, @\\
+\mbox{}\verb@                   n=N-data_object$n, k=t))@\\
+\mbox{}\verb@    } else {@\\
+\mbox{}\verb@      hp <- data_object$hyper_probs@\\
+\mbox{}\verb@    }@\\
 \mbox{}\verb@    @\\
@@
 \mbox{}\verb@@\\
-\mbox{}\verb@ @\hbox{$\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize \NWlink{nuweb10a}{10a}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@ @\hbox{$\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize \NWlink{nuweb10b}{10b}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@ @\hbox{$\langle\,${\itshape Set up for E-step}\nobreak\ {\footnotesize \NWlink{nuweb8}{8}}$\,\rangle$}\verb@@\\
-\mbox{}\verb@ @\hbox{$\langle\,${\itshape Identify zero support}\nobreak\ {\footnotesize \NWlink{nuweb10b}{10b}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@ @\hbox{$\langle\,${\itshape Identify zero support}\nobreak\ {\footnotesize \NWlink{nuweb11a}{11a}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@@\\
@@
 \mbox{}\verb@    @\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape E-step}\nobreak\ {\footnotesize \NWlink{nuweb9a}{9a}}$\,\rangle$}\verb@@\\
-\mbox{}\verb@    @\hbox{$\langle\,${\itshape M-step}\nobreak\ {\footnotesize \NWlink{nuweb9b}{9b}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape E-step}\nobreak\ {\footnotesize \NWlink{nuweb9}{9}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@    @\hbox{$\langle\,${\itshape M-step}\nobreak\ {\footnotesize \NWlink{nuweb10a}{10a}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@    @\\
@@
 \mbox{}\verb@  @\\
-\mbox{}\verb@  @\hbox{$\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize \NWlink{nuweb11}{11}}$\,\rangle$}\verb@@\\
+\mbox{}\verb@  @\hbox{$\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize \NWlink{nuweb11b}{11b}}$\,\rangle$}\verb@@\\
 \mbox{}\verb@  @\\
@@
 \begin{flushleft} \small\label{scrap9}\raggedright\small
-\NWtarget{nuweb9a}{} $\langle\,${\itshape E-step}\nobreak\ {\footnotesize {9a}}$\,\rangle\equiv$
+\NWtarget{nuweb9}{} $\langle\,${\itshape E-step}\nobreak\ {\footnotesize {9}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap10}\raggedright\small
-\NWtarget{nuweb9b}{} $\langle\,${\itshape M-step}\nobreak\ {\footnotesize {9b}}$\,\rangle\equiv$
+\NWtarget{nuweb10a}{} $\langle\,${\itshape M-step}\nobreak\ {\footnotesize {10a}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap11}\raggedright\small
-\NWtarget{nuweb10a}{} $\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize {10a}}$\,\rangle\equiv$
+\NWtarget{nuweb10b}{} $\langle\,${\itshape Set starting values}\nobreak\ {\footnotesize {10b}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap12}\raggedright\small
-\NWtarget{nuweb10b}{} $\langle\,${\itshape Identify zero support}\nobreak\ {\footnotesize {10b}}$\,\rangle\equiv$
+\NWtarget{nuweb11a}{} $\langle\,${\itshape Identify zero support}\nobreak\ {\footnotesize {11a}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \begin{flushleft} \small\label{scrap13}\raggedright\small
-\NWtarget{nuweb11}{} $\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize {11}}$\,\rangle\equiv$
+\NWtarget{nuweb11b}{} $\langle\,${\itshape Calculate standard errors}\nobreak\ {\footnotesize {11b}}$\,\rangle\equiv$
 \vspace{-1ex}
@@
 \mbox{}\verb@                          weights = weights, offset = offset, maxN = N, @\\
-\mbox{}\verb@                          spt = spt)@\\
+\mbox{}\verb@                          spt = spt, hyper_probs = hp)@\\
 \mbox{}\verb@  ll <- function(x){@\\
@@
 \mbox{}\verb@  # calculate variance-covariance matrix@\\
-\mbox{}\verb@  bvc <- solve(-bhess)@\\
+\mbox{}\verb@  bvc <- tryCatch(solve(-bhess), error = function(e)NULL)@\\
+\mbox{}\verb@  # bvc <- solve(-bhess)@\\
+\mbox{}\verb@  if (!is.null(bvc)){@\\
 \mbox{}\verb@    # remove border and set to 0 for zero-support values@\\
@@
 \mbox{}\verb@    vc[hess_idx, hess_idx] <- bvc[1:(p+length(spt)), 1:(p+length(spt))]     @\\
+\mbox{}\verb@  } else {@\\
+\mbox{}\verb@    vc <- matrix(NA, nrow=p+N+1, ncol=p+N+1)@\\
+\mbox{}\verb@  }@\\
 \mbox{}\verb@@{\NWsep}
@@
 \begin{flushleft} \small\label{scrap14}\raggedright\small
-\NWtarget{nuweb?}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {?}}$\equiv$
+\NWtarget{nuweb12}{} \verb@"../R/SPGLM.R"@\nobreak\ {\footnotesize {12}}$\equiv$
 \vspace{-1ex}

### 2021-11-23 R/SPGLM.R

@@ -199,7 +199,7 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
       
         data_object_mod <- list(model_matrix = mm, resp=Y, n=rowSums(Y), 
                                 weights = weights, offset = offset, maxN = N, 
-                                spt = spt)
+                                spt = spt, hyper_probs = hp)
         ll <- function(x){
           spglm_loglik(beta=x[1:p], f0 = exp(x[-(1:p)]), data_object=data_object_mod, link=link)
         }
@@ -419,7 +419,12 @@ spglm_probs <- function(beta, f0, data_object, link){
     th <- getTheta(spt=spt/N, f0=f0[spt+1], mu=mu, weights=data_object$weights, ySptIndex=rep(1, nobs),
                    thetaStart=NULL, thetaControl=theta.control())
                    
-    hp <- sapply(0:N, function(t)dhyper(x=data_object$resp[,1], m=data_object$n, n=N-data_object$n, k=t))  
+    if (is.null(data_object$hyper_probs)){                   
+      hp <- sapply(0:N, function(t)dhyper(x=data_object$resp[,1], m=data_object$n, 
+                   n=N-data_object$n, k=t))
+    } else {
+      hp <- data_object$hyper_probs
+    }
     
     # fill in 0's for values with no support
     fTilt <- matrix(0, ncol = nobs, nrow = N+1)

### 2021-12-23 R/SPreg.R
Remove relative risk regression (sprr)

@@ -1,314 +1,0 @@
-
-require(CorrBin)
-
-# lambda-to-p weight matrix, rows:r, cols:s
-# (-1)^(s-r)*choose(n,r)*choose(n-r,s-r)
-
-weight_mat <- function(n){
-  s <- r <- 0:n
-  res <- outer(r, s, function(x,y)(-1)^(y-x)*choose(n,x)*choose(n-x,y-x))
-  res
-}
-
-p_from_lambda <- function(lambda.vec, n){
-  H <- weight_mat(n)
-  c(H %*% lambda.vec[1:(n+1)])
-}
-
-# p-to-lambda weight matrix, rows:k, cols:r
-# (choose(n-k, n-r)/choose(n,r)
-
-weight_mat2 <- function(n){
-  k <- r <- 0:n
-  res <- outer(k, r, function(x,y)choose(n-x, n-y)/choose(n,y))
-  res
-}
-
-lambda_from_p <- function(p.vec, n=length(p.vec)-1){
-  H <- weight_mat2(n)
-  c(H %*% p.vec)
-}
-
-#' Fit semi-parametric relative risk model
-#'
-#'@rdname sprr
-#'@param formula a one-sided formula of the form \code{cbind(r, s) ~ predictors} where \code{r} and \code{s} give the number of responses and non-responses within each cluster, respectively (so the cluster size is \code{r+s}), and \code{predictors} describes the covariates.
-#'@param data  an optional matrix or data frame containing the variables in the formula \code{formula}. By default the variables are taken from \code{environment(formula).}
-#'@param subset an optional vector specifying a subset of observations to be used.
-#'@param weights        an optional vector specifying observation weights.
-#'@param link     a link function for the binomial distribution, the inverse of which models the covariate effects.
-#'@param mu1      an optional value between 0 and 1 giving the maximal predicted marginal probability. If set to NULL (default), the algorithm will try to estimate it from the data.
-#'@param start    an optional list with elements named \code{beta}, \code{q}, and/or \code{mu1} giving starting values for estimation. If none or only some are specified, starting values are selected automatically.
-#'@param control        a list with parameters controlling the algorithm.
-#'@return an object of class \code{sprr} with the fitted model.
-#'@export
-#' @importFrom stats terms model.matrix binomial dbinom glm uniroot
-
[…]

### 2021-12-23 nuweb/SPreg_develop.R

@@ -21,8 +21,7 @@ use_data(boric_acid)
 ex <- as.package("../exchreg")
 nuweb(ex)
 
-#shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet --run-viewer  SPregress.tex")
-shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet  SPregress.tex")
+#shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet --run-viewer  SPGLM.tex")
 shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet SPGLM.tex")
 shell("cd c:/exchreg/nuweb/ && texify --pdf --quiet wgldrm.tex")
 
@@ -295,21 +294,7 @@ microbenchmark::microbenchmark(
   getThetaC(.spt, .f0, .mu, .thetaStart, exchreg:::theta.control()),
   getThetaR(.spt, .f0, .mu, .thetaStart, exchreg:::theta.control())
 )
-######### Testing SPreg
 
-
-a <- sprr(cbind(NResp, ClusterSize-NResp) ~ Trt, data=shelltox, weights=shelltox$Freq,
-                control=list(eps=0.001, maxit=1000), 
-          start = list(mu1=0.4))
-
-nd <- data.frame(Trt = unique(shelltox$Trt))
-nd$p <- predict(a, newdata=nd)
-nd
-
-a2 <- sprr(cbind(NResp, ClusterSize-NResp) ~ Trt + log(ClusterSize), 
-           data=shelltox, weights=shelltox$Freq,
-           control=list(eps=0.001,maxit=100),
-           start = list(beta=c(coef(a),0), q=a$q))
 
 
 ab <- sprr(cbind(NResp, ClusterSize-NResp) ~ Trt, data=shelltox, weights=shelltox$Freq,

### 2021-12-23 ManuscriptCode/Manuscript.Rmd
Prettify manuscript plots

@@
 
-set.seed(20210814)
+set.seed(20211223)
 data(boric_acid)
@@
 
-We will have one binary predictor $X_1 \sim Bern(0.5)$ with $beta_1=0.5$ and one continuous predictor $X_2 \sim N(0,1)$ with $\beta_2 = 1$. The cluster size distribution will be binomial.
+We will have one binary predictor $X_1 \sim Bern(0.5)$ with $\beta_1=0.5$ and one continuous predictor $X_2 \sim N(0,1)$ with $\beta_2 = 1$. The cluster size distribution will be binomial.
 
@@
   cs <- rbinom(n, size=maxClusterSize, prob=aveClusterSize/maxClusterSize)
+  cs <- pmax(1, cs) # don't want clusters of size 0
 
@@
                   `Neither correct` = sim4,
-                  .id="Generator")
+                  .id="Generator") %>%
+  mutate(Generator= factor(Generator, levels = c("Both correct",
+                                                 "BB correct",
+                                                 "SPGLM correct",
+                                                 "Neither correct")),
+         Model = factor(Model, levels = c("BB","SPGLM","GEE")))
 
@@
                          .id="Generator") %>%
-  mutate(Term = as.character(NResp))
+  mutate(Term = as.character(NResp),
+         Generator= factor(Generator, levels = c("Both correct",
+                                                 "BB correct",
+                                                 "SPGLM correct",
+                                                 "Neither correct")))
 
@@
 ```{r ComparePlot}
-ggplot(comb_beta, aes(x=Term, y=Est-Correct, colour=Model, shape=Model)) +
+cols <- RColorBrewer::brewer.pal(3, "Set2")
+names(cols) <- levels(comb$Model)
+ggplot(comb_beta, aes(x=Term, y=Est-Correct, fill=Model, outlier.fill=Model)) +
   facet_wrap(~Generator) +
   geom_hline(aes(yintercept=0)) +
-  stat_summary(fun.data =  mean_sdl, geom="pointrange", position=position_dodge(width=0.4))
+  geom_boxplot(position=position_dodge(width=0.4), width=0.3, 
+               outlier.size =1, outlier.alpha=0.5) +
+  scale_y_continuous("Bias = Estimate - correct value") +
+  scale_fill_manual(breaks=names(cols), values=cols)
 
@@
 ```{r PDFplot}
-ggplot(comb_pdf, aes(x=NResp, y=Est, colour=Model)) +
+ggplot(comb_pdf, aes(x=NResp, y=Est, fill=Model, outlier.fill=Model)) +
   facet_wrap(~Generator) +
-  geom_line(aes(y=Correct), col="black") +
-  stat_summary(fun.data =  mean_sdl, geom="pointrange", position=position_dodge(width=0.1))
+  geom_line(aes(y=Correct), colour="black", size=1) +
+  geom_boxplot(aes(group=interaction(Model, NResp)), 
+               position=position_dodge(width=0.6), width=0.5, 
+               outlier.size =1, outlier.alpha=0.5) +
+  scale_y_continuous("Probability") +
+  scale_x_continuous("# of responses", breaks = 0:cs_predict) +
+  scale_fill_manual(breaks=names(cols), values=cols)
 

### 2022-03-08 ManuscriptCode/Manuscript.Rmd
Add binomial fit

@@
                  data=dat, family =  betabinomial(lmu="logitlink"))
+  #Binomial
+  res_bin <- glm(cbind(NResp, ClusterSize - NResp) ~  X1 + X2, weights = Freq,
+                 data=dat, family =  binomial)
+
   #SPGLM
@@
 
+   # binomial
+  mns <- predict(res_bin, type="response", newdata=newd)
+
+  fit_bin <- newd %>% 
+    crossing(NResp=0:cs_predict) %>%
+    mutate(ClusterSize = cs_predict,
+           Prob = dbinom(NResp, size=cs_predict, prob=mns)) %>%
+  mutate(Model="Bin")
+
   fit_spglm <- newd %>% 
@@
                 SE=sqrt(diag(vcov(res_bb)))[-2]),
+    Bin = tibble(Term = nms,
+                 Est = coef(res_bin),
+                 SE = sqrt(diag(vcov(res_bin)))),
     SPGLM = tibble(Term = nms,
@@
 
-  fit <- bind_rows(fit_bb, fit_spglm) %>%
+  fit <- bind_rows(fit_bb, fit_bin, fit_spglm) %>%
     mutate(Type = "PDF")

### 2022-03-11 nuweb/SPGLM.tex
Improve robustness of variance calculation

@@
 \mbox{}\verb@  hess <- numDeriv::hessian(func=ll,  x=c(betas, log(referencef0)))@\\
-\mbox{}\verb@  hess_idx <- c(1:p, p+spt+1) # betas and f0 elements with non-zero real support@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  # determine non-degenerate support@\\
+\mbox{}\verb@  hess_spt <- which(referencef0 > sqrt(.Machine$double.eps))@\\
+\mbox{}\verb@  @\\
+\mbox{}\verb@  hess_idx <- c(1:p, p+hess_spt) # betas and f0 elements with non-zero real support@\\
 \mbox{}\verb@  @\\
 \mbox{}\verb@  # revert to unlogged f0@\\
-\mbox{}\verb@  grad <- c(rep(1, p), 1/referencef0[spt+1])@\\
+\mbox{}\verb@  grad <- c(rep(1, p), 1/referencef0[hess_spt])@\\
 \mbox{}\verb@  hess1 <- diag(grad) %*% hess[hess_idx, hess_idx] %*% diag(grad)@\\
@@
 \mbox{}\verb@  # create bordered hessian@\\
-\mbox{}\verb@  border1 <- c(rep(0, p), rep(1, length(spt)))  # gradient of sum-to-one constraint@\\
-\mbox{}\verb@  border2 <- c(rep(0, p), spt)          # gradient of fixed-mean constraint@\\
+\mbox{}\verb@  border1 <- c(rep(0, p), rep(1, length(hess_spt)))  # gradient of sum-to-one constraint@\\
+\mbox{}\verb@  border2 <- c(rep(0, p), hess_spt-1)          # gradient of fixed-mean constraint@\\
 \mbox{}\verb@  bhess <- rbind(cbind(hess1, border1, border2), c(border1,0,0), c(border2,0,0))@\\
@@
 \mbox{}\verb@    vc <- matrix(0, nrow=p+N+1, ncol=p+N+1)@\\
-\mbox{}\verb@    vc[hess_idx, hess_idx] <- bvc[1:(p+length(spt)), 1:(p+length(spt))]     @\\
+\mbox{}\verb@    vc[hess_idx, hess_idx] <- bvc[1:(p+length(hess_spt)), 1:(p+length(hess_spt))]     @\\
 \mbox{}\verb@  } else {@\\

### 2022-03-11 R/SPGLM.R

@@ -205,15 +205,19 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
         }
 
         hess <- numDeriv::hessian(func=ll,  x=c(betas, log(referencef0)))
-        hess_idx <- c(1:p, p+spt+1) # betas and f0 elements with non-zero real support
+        
+        # determine non-degenerate support
+        hess_spt <- which(referencef0 > sqrt(.Machine$double.eps))
+        
+        hess_idx <- c(1:p, p+hess_spt) # betas and f0 elements with non-zero real support
         
         # revert to unlogged f0
-        grad <- c(rep(1, p), 1/referencef0[spt+1])
+        grad <- c(rep(1, p), 1/referencef0[hess_spt])
         hess1 <- diag(grad) %*% hess[hess_idx, hess_idx] %*% diag(grad)
         
         # create bordered hessian
-        border1 <- c(rep(0, p), rep(1, length(spt)))  # gradient of sum-to-one constraint
-        border2 <- c(rep(0, p), spt)          # gradient of fixed-mean constraint
+        border1 <- c(rep(0, p), rep(1, length(hess_spt)))  # gradient of sum-to-one constraint
+        border2 <- c(rep(0, p), hess_spt-1)          # gradient of fixed-mean constraint
         bhess <- rbind(cbind(hess1, border1, border2), c(border1,0,0), c(border2,0,0))
         
         # calculate variance-covariance matrix
@@ -222,7 +226,7 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
         if (!is.null(bvc)){
           # remove border and set to 0 for zero-support values
           vc <- matrix(0, nrow=p+N+1, ncol=p+N+1)
-          vc[hess_idx, hess_idx] <- bvc[1:(p+length(spt)), 1:(p+length(spt))]     
+          vc[hess_idx, hess_idx] <- bvc[1:(p+length(hess_spt)), 1:(p+length(hess_spt))]     
         } else {
           vc <- matrix(NA, nrow=p+N+1, ncol=p+N+1)
         }

### 2022-03-11 ManuscriptCode/Manuscript.Rmd
Change BB to binomial for "both correct" setting

--- 2021-12-23 ManuscriptCode/Manuscript.Rmd
@@
 
-  1. Both the beta-binomial and semi-parameteric model are correct - a binomial backbone, which belongs to the exponential family
-  2. Only the beta-binomial model is correct - generated from a beta-binomial distribution
+  1. Both the parametric (binomial) and semi-parameteric model are correct - a binomial backbone, which belongs to the exponential family
+  2. Only the parametric (beta-binomial) model is correct - generated from a beta-binomial distribution
   3. Only the semi-parametric model is correct - generated from the spglm, with a q-power backbone
@@
 ```{r AnalyzeOne}
-run_analysis <- function(dat, cs_predict=5){
+run_analysis <- function(dat, cs_predict=5, Binom = FALSE){
   #GEE
@@
                  corstr = "exch")
+
+  if (Binom){
+    #Binomial
+    res_bin <- glm(cbind(NResp, ClusterSize - NResp) ~  X1 + X2, weights = Freq,
+                  data=dat, family =  binomial)
+  } else {
     #Beta-binomial
@@
                  data=dat, family =  betabinomial(lmu="logitlink"))
+  }
+
   #SPGLM
@@
   newd <- data.frame(X1 = 0, X2 = 0)
+  nms <- names(coef(res_spglm))
+
+  if (Binom){   
+    # binomial
+    mns <- predict(res_bin, type="response", newdata=newd)
+
+   fit_binbb <- newd %>% 
+      crossing(NResp=0:cs_predict) %>%
+      mutate(ClusterSize = cs_predict,
+            Prob = dbinom(NResp, size=cs_predict, prob=mns)) %>%
+    mutate(Model="Binom/BB")
+
+   ests_binbb <- tibble(Term = nms,
+                 Est = coef(res_bin),
+                 SE = sqrt(diag(vcov(res_bin))))
+  } else {
     # beta-binomial
@@
 
-  fit_bb <- newd %>% 
+    fit_binbb <- newd %>% 
       crossing(NResp=0:cs_predict) %>%
       mutate(ClusterSize = cs_predict,
-           Prob = dbetabinom(NResp, size=cs_predict, prob=mns_rho[1,"mu"], rho=mns_rho[1,"rho"])) %>%
-  mutate(Model="BB")
+            Prob = dbetabinom(NResp, size=cs_predict, prob=mns_rho[1,"mu"],
+                              rho=mns_rho[1,"rho"])) %>%
+    mutate(Model="Binom/BB")
+
+    ests_binbb <- tibble(Term = nms,
+                Est = coef(res_bb)[-2],      # second value is logit(rho)   
+                SE=sqrt(diag(vcov(res_bb)))[-2])
+  }
 
@@
 
-  nms <- names(coef(res_spglm))
   res <- bind_rows(
@@
                  SE = sqrt(diag(vcov(res_gee)))),
-    BB = tibble(Term = nms,
-                Est = coef(res_bb)[-2],      # second value is logit(rho)   
-                SE=sqrt(diag(vcov(res_bb)))[-2]),
+    `Binom/BB` = ests_binbb,
     SPGLM = tibble(Term = nms,
@@
 
-  fit <- bind_rows(fit_bb, fit_spglm) %>%
+  fit <- bind_rows(fit_binbb, fit_spglm) %>%
     mutate(Type = "PDF")
@@
 ```{r ActualSim1, warning=FALSE}
-R <- 100
+R <- 500
 cs_predict <- 5
@@
                 add_response_param(rfun=function(n, cs, p)rbinom(n, size=cs, prob=p));
-             run_analysis(simd, cs_predict = cs_predict)}, 
+             run_analysis(simd, cs_predict = cs_predict, Binom = TRUE)}, 
             simplify = FALSE)  %>%
@@
 comb <- bind_rows(`Both correct` = sim1, 
-                  `BB correct`= sim2,
+                  `Parametric correct`= sim2,
                   `SPGLM correct` = sim3, 
@@
   mutate(Generator= factor(Generator, levels = c("Both correct",
-                                                 "BB correct",
+                                                 "Parametric correct",
                                                  "SPGLM correct",
                                                  "Neither correct")),
-         Model = factor(Model, levels = c("BB","SPGLM","GEE")))
+         Model = factor(Model, levels = c("Binom/BB","SPGLM","GEE")))
 
@@
 correct_pdf <- bind_rows(`Both correct` = pdf1, 
-                         `BB correct`= pdf2,
+                         `Parametric correct`= pdf2,
                          `SPGLM correct` = pdf3, 
@@
          Generator= factor(Generator, levels = c("Both correct",
-                                                 "BB correct",
+                                                 "Parametric correct",
                                                  "SPGLM correct",
@@
             .groups="drop_last") %>%
-  mutate(rel.eff = mc.se / mc.se[Model=="GEE"]) %>%
+  mutate(rel.eff = mc.se / mc.se[Model=="GEE"],
+         se.rel.bias = mc.se/se) %>%
   ungroup() %>%
@@
 
+ggplot(comb_pdf, aes(x=NResp, y=Est-Correct, fill=Model, outlier.fill=Model)) +
+  facet_wrap(~Generator) +
+  geom_hline(yintercept = 0, colour="black", size=1) +
+  stat_summary(aes(color=Model), fun=mean, geom="line") +
+  geom_boxplot(aes(group=interaction(Model, NResp)), 
+               position=position_dodge(width=0.6), width=0.5, 
+               outlier.size =1, outlier.alpha=0.5) +
+  scale_y_continuous("Bias of probability") +
+  scale_x_continuous("# of responses", breaks = 0:cs_predict) +
+  scale_fill_manual(breaks=names(cols), values=cols)
+
 ```

### 2022-03-14 nuweb/SPGLM.tex
Change starting value for f0

@@
 \mbox{}\verb@  betas <- NULL@\\
-\mbox{}\verb@  pooled <- CorrBin::CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), @\\
-\mbox{}\verb@                    trt="Trt", clustersize="ClusterSize", nresp="NResp", freq="Freq")@\\
-\mbox{}\verb@  est <- CorrBin::mc.est(pooled)@\\
-\mbox{}\verb@  referencef0 <- est$Prob[est$ClusterSize == N]@\\
+\mbox{}\verb@  #pooled <- CorrBin::CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), @\\
+\mbox{}\verb@  #                  trt="Trt", clustersize="ClusterSize", nresp="NResp", freq="Freq")@\\
+\mbox{}\verb@  #est <- CorrBin::mc.est(pooled)@\\
+\mbox{}\verb@  #referencef0 <- est$Prob[est$ClusterSize == N]@\\
+\mbox{}\verb@  referencef0 <- rep(1/(N+1), N+1)@\\
 \mbox{}\verb@  @\\

### 2022-03-14 R/SPGLM.R

@@ -73,10 +73,11 @@ spglm <- function(formula, data, subset, weights, offset, link="logit", mu0=NULL
     
      
        betas <- NULL
-       pooled <- CorrBin::CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), 
-                         trt="Trt", clustersize="ClusterSize", nresp="NResp", freq="Freq")
-       est <- CorrBin::mc.est(pooled)
-       referencef0 <- est$Prob[est$ClusterSize == N]
+       #pooled <- CorrBin::CBData(data.frame(Trt = "All", NResp = Y[,1], ClusterSize = rowSums(Y), Freq=ceiling(weights)), 
+       #                  trt="Trt", clustersize="ClusterSize", nresp="NResp", freq="Freq")
+       #est <- CorrBin::mc.est(pooled)
+       #referencef0 <- est$Prob[est$ClusterSize == N]
+       referencef0 <- rep(1/(N+1), N+1)
        
        
      

### 2022-03-19 ManuscriptCode/Manuscript.Rmd
Tweak manuscript plots

@@
 output:
+  pdf_document: 
+    keep_tex: true
   html_document:
@@
 
-ggplot(fits, aes(x=NResp, y=Prob, colour=Model)) +
+linecols <- colorspace::darken(RColorBrewer::brewer.pal(3, "Set3"), 0.2)
+names(linecols) <- unique(fits$Model)
+
+ggplot(fits, aes(x=NResp, y=Prob, colour=Model, shape=Model, linetype=Model)) +
   facet_wrap(~Trt) +
@@
   scale_x_continuous("# of affected animals", breaks=seq(0,10,by=2)) +
+  scale_colour_manual(values=linecols) +
   ylab("Probability") +
@@
 
+```{r SimParams}
+R <- 500
+beta_correct <- c(0, 0.5, 1)
+cs_predict <- 5
+p_predict <- 1/(1+exp(-c(beta_correct %*% c(0,0,0))))
+```
+
 ```{r SimOne}
-beta_correct <- c(0, 0.5, 1)
 sim_covariates <- function(n, maxClusterSize = 10, aveClusterSize = 7, beta=beta_correct){
@@
 ```{r ActualSim1, warning=FALSE}
-R <- 500
-cs_predict <- 5
-p_predict <- 1/(1+exp(-c(beta_correct %*% c(0,0,0))))
-
+set.seed(20211223)
 # Scenario 1: binomial PDF
@@
 ```{r ActualSim2, warning=FALSE}
+set.seed(20211223)
 # Scenario 2: beta-binomial PDF
@@
 ```{r ActualSim3, warning=FALSE}
+set.seed(20211223)
 # Scenario 3: SPGLM with q-power backbone
@@
 ```{r ActualSim4, warning=FALSE}
+set.seed(20211223)
 # Scenario 4: q-power 
@@
 
-comb_beta <- left_join(comb %>% filter(Type == "Beta"), correct_beta, by="Term")
+comb_beta <- left_join(comb %>% filter(Type == "Beta"), correct_beta, by="Term") %>%
+  mutate(Coefficient = recode(Term, "(Intercept)" = "beta[0]", 
+                              "X1"="beta[1]", 
+                              X2="beta[2]"),
+         Term = recode(Term,  "(Intercept)" = "$\\beta_0$", 
+                       "X1"="$\\beta_1$", 
+                       X2="$\\beta_2$"))
 
@@
             .groups="drop_last") %>%
-  mutate(rel.eff = mc.se / mc.se[Model=="GEE"],
-         se.rel.bias = mc.se/se) %>%
+  mutate(rel.eff = mc.se / mc.se[Model=="GEE"]) %>%
   ungroup() %>%
@@
 ```{r ComparePlot}
-cols <- RColorBrewer::brewer.pal(3, "Set2")
+cols <- RColorBrewer::brewer.pal(3, "Set3")
 names(cols) <- levels(comb$Model)
-ggplot(comb_beta, aes(x=Term, y=Est-Correct, fill=Model, outlier.fill=Model)) +
+ggplot(comb_beta, aes(x=Coefficient, y=Est-Correct, fill=Model, outlier.fill=Model)) +
   facet_wrap(~Generator) +
   geom_hline(aes(yintercept=0)) +
-  geom_boxplot(position=position_dodge(width=0.4), width=0.3, 
+  geom_boxplot(position=position_dodge(width=0.8), width=0.7, 
                outlier.size =1, outlier.alpha=0.5) +
-  scale_y_continuous("Bias = Estimate - correct value") +
-  scale_fill_manual(breaks=names(cols), values=cols)
+  scale_x_discrete(labels = scales::label_parse()) +
+  scale_y_continuous("Bias of parameter estimate") +
+  scale_fill_manual(breaks=names(cols), values=cols) +
+  theme(legend.position = "top")
 
@@
   scale_x_continuous("# of responses", breaks = 0:cs_predict) +
-  scale_fill_manual(breaks=names(cols), values=cols)
+  scale_fill_manual(breaks=names(cols), values=cols)  +
+  theme(legend.position = "top")
 
@@
   facet_wrap(~Generator) +
-  geom_hline(yintercept = 0, colour="black", size=1) +
-  stat_summary(aes(color=Model), fun=mean, geom="line") +
+  geom_hline(yintercept = 0, colour="black") +
+ # stat_summary(aes(color=Model), fun=mean, geom="line") +
   geom_boxplot(aes(group=interaction(Model, NResp)), 
-               position=position_dodge(width=0.6), width=0.5, 
+               position=position_dodge(width=0.8), width=0.7, 
                outlier.size =1, outlier.alpha=0.5) +
-  scale_y_continuous("Bias of probability") +
+  scale_y_continuous("Bias of event probability estimate") +
   scale_x_continuous("# of responses", breaks = 0:cs_predict) +
-  scale_fill_manual(breaks=names(cols), values=cols)
+  scale_fill_manual(breaks=names(cols), values=cols) +
+  theme(legend.position = "top")
 

### 2022-03-20 ManuscriptCode/Manuscript.Rmd
Save plot data for Zenodo upload

@@
   theme_bw()
+
+# save data
+write.csv(fits %>% select(Model, Trt, ClusterSize, NResp, Prob),
+          file="PredictPlot.csv")
 
@@
             .groups="drop_last") %>%
-  mutate(rel.eff = mc.se / mc.se[Model=="GEE"]) %>%
+  mutate(rel.eff = mc.se^2 / mc.se[Model=="GEE"]^2) %>%
   ungroup() %>%
-  arrange(Generator, Model, Term)
+  arrange(Generator, Term, Model)
 kable(scomb, digits=3)
@@
 
+# save data
+write.csv(comb_beta %>% select(Generator, ID, Model, Coefficient, Est, SE, Correct),
+          file="ComparePlot.csv")
 ```
@@
 
+# save data
+write.csv(comb_pdf %>% select(Generator, ID, Model, NResp, Est, Correct),
+          file="PDFPlot.csv")
 ```
